---
title: "Add_feature_IDs"
output: html_document
date: "2025-01-29"
params: 
  metabolomics_pos: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231120_UDA/231120_Patient_plasma_BB23-00001_1-IPS22-00063_2_2024-03-13_ESIpos.tsv
  metabolomics_neg: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231128_UDA/231128_Patient_plasma_BB23-00001_1-IPS22-00063_2_2024-03-13_ESIneg.tsv
  outdir: Z:/data/iPSC_data/Results/MixedModels/Metabolomics_anno
---

## Corrected or Uncorrected data
```{r}
input_dir <- 'Z:/data/iPSC_data/Results/MixedModels'
#input_dir <- 'Z:/data/iPSC_data/Results/MixedModels_CORRECTED/'
```

## Params
```{r}
print(params)
```


```{r}
library(openxlsx)
library(dplyr)
library(tidyverse)
```


## Set Assay to metabolomics data
```{r}
assay_name <- 'Metabolomics, TML'
```


## Read mixed models results
```{r}
split_string <- "_MixedModel"

xlsx_files <- list.files(input_dir, pattern = "\\.xlsx$", full.names=T)

data_list <- list()

for (file in xlsx_files) {
  data_list[[file]] <- openxlsx::read.xlsx(file, sheet = assay_name)
}
```


Feature metadata
```{r}
metabolomics_IDs_pos <- read.delim(params$metabolomics_pos, sep = '\t')
metabolomics_IDs_pos$Feature_ID <- paste0('TML_pos', metabolomics_IDs_pos$Feature_ID)

metabolomics_IDs_neg <- read.delim(params$metabolomics_neg, sep = '\t')
metabolomics_IDs_neg$Feature_ID <- paste0('TML_neg', metabolomics_IDs_neg$Feature_ID)
```


## Add retention time and mass
```{r}
# Use first df as mapping table
metabolomics <- data_list[[1]]

features_RT <- lapply(metabolomics$feature, function(x) {
    if (grepl('_pos', x)) {
      rt <- metabolomics_IDs_pos[metabolomics_IDs_pos$Feature_ID == x, 'Feature_RT']
    }
    else if (grepl('_neg', x)) {
      rt <- metabolomics_IDs_neg[metabolomics_IDs_neg$Feature_ID == x, 'Feature_RT']
    }
    unique(rt)
})

features_Mass <- lapply(metabolomics$feature, function(x) {
    if (grepl('_pos', x)) {
      rt <- metabolomics_IDs_pos[metabolomics_IDs_pos$Feature_ID == x, 'Feature_Mass']
    }
    else if (grepl('_neg', x)) {
      rt <- metabolomics_IDs_neg[metabolomics_IDs_neg$Feature_ID == x, 'Feature_Mass']
    }
    unique(rt)
})

df_ids <- data.frame(Feature_Mass = as.character(features_Mass), 
                     Feature_RT = as.character(features_RT))
```


## Find HMDB ids per feature 
```{r}
hmdb_mappings <- lapply(metabolomics$feature, function(x) {
    if (grepl('_pos', x)) {
      HMDB_ids <- metabolomics_IDs_pos[metabolomics_IDs_pos$Feature_ID == x, 'HMP']
    }
    else if (grepl('_neg', x)) {
      HMDB_ids <- metabolomics_IDs_neg[metabolomics_IDs_neg$Feature_ID == x, 'HMP']
    }
    HMDB_ids
})
```


## Add to new dataframe, add this to MixedModel results table
```{r}
df <- tibble(original = hmdb_mappings) %>%
  mutate(row_id = row_number()) %>%
  unnest_wider(original, names_sep ="_")%>%
  select(-row_id)
colnames(df) <- paste0("Put_annotation", seq_along(df))
```


## Create mapping table of feature metadata
```{r}
df_ids_temp1 <- cbind(df_ids, df)
## Mapping table
mapping_table <- cbind(metabolomics, df_ids_temp1)

#dir.create(params$outdir)
write.csv(mapping_table, file.path(params$outdir, 'HMDBMetabolomicsMappings.csv'))
```


## Filter HDMBs putative annoation for metabolic panel
```{r}
hmdb_mappings_MP <- lapply(metabolomics$feature, function(x) {
    if (grepl('_pos', x)) {
      HMDB_ids <- metabolomics_IDs_pos[metabolomics_IDs_pos$Feature_ID == x & metabolomics_IDs_pos$MP == 'YES', 'HMP']
    }
    else if (grepl('_neg', x)) {
      HMDB_ids <- metabolomics_IDs_neg[metabolomics_IDs_neg$Feature_ID == x & metabolomics_IDs_neg$MP == 'YES', 'HMP']
    }
    HMDB_ids
})


df_panel <- tibble(original = hmdb_mappings_MP) %>%
  mutate(row_id = row_number()) %>%
  unnest_wider(original, names_sep ="_")%>%
  select(-row_id)
colnames(df_panel) <- paste0("Put_annotation", seq_along(df_panel))


mapping_table$Feature_Mass.1 <- NULL
df_ids <- cbind(df_ids, df_panel)
df_ids$Feature_Mass <- NULL
df_ids$Feature_RT <- NULL
##Create mapping table
mapping_table_panel <- cbind(metabolomics, df_ids)


mapping_table_panel$Feature_Mass <- mapping_table$Feature_Mass
mapping_table_panel$Feature_RT <- mapping_table$Feature_RT

write.csv(mapping_table_panel , file.path(params$outdir, 'HMDBMetabolomicsMappings_PANEL.csv'))
```


## Add feature metadata (only panel annotations) to Mixed Model results
```{r}
updated_list <- lapply(data_list, function(df) {
  df <- df %>%
    left_join(mapping_table_panel %>% select(all_of(c("feature",
                                                           "Feature_Mass",
                                                           "Feature_RT")),
                                                    starts_with("Put_annotation")),
              by = "feature")
  return(df)
})
```



## Save to the Excel file
```{r}
lapply(names(updated_list), function(name_df) {
  df <- updated_list[[name_df]]
  wb <- loadWorkbook(name_df)
  writeData(wb, sheet = assay_name, x = df, withFilter = F)
  saveWorkbook(wb, name_df, overwrite = T)
})
```


## Save HMDB ids and more in separate table
```{r}
# hmdb_extra_table <- rbind(metabolomics_IDs_pos, metabolomics_IDs_neg)
# hmdb_extra_table <- hmdb_extra_table[hmdb_extra_table$Feature_ID %in% metabolomics$feature,]
# hmdb_extra_table <- hmdb_extra_table[ , c('HMP', 'Formula', 'Adduct', 'Isotope')]
# write.csv(hmdb_extra_table, 'Z:/data/iPSC_data/Results/MixedModels/Metabolomics_HMDB_annotations.csv')
```







