---
title: "QC_WES"
author: "Cenna Doornbos"
output: html_document
params:
  cnv_wes_ed_data_tab: Z:/data/iPSC_data/iPSC_CNV-WES/Data_integration/cnv_ed.csv
---

## Additional QC analysis for WES data

#### Script parameters
```{r, params}
print(Sys.Date())
print(params)
```
### Libraries
```{r, read data}
library(readxl)
library(tidyverse)
library(ggrepel)
library(plotly)
library(ggplot2)
```

# Read data
```{r, read data}
data_frame_ed <- read.csv(params$cnv_wes_ed_data_tab)
```





# CHD2 P3 clones
```{r}
CHD2_P3_names <- c('CHD2_p3c1', 'CHD2_p3c2')
data_frame_ed_CHD2_P3 <- data_frame_ed[data_frame_ed$sample_name %in% CHD2_P3_names,]
```


Number of mutations found per clone per chromosome
```{r}
table(data_frame_ed_CHD2_P3$sample_name, data_frame_ed_CHD2_P3$chromosome )
```

Type of mutations found per clone 
```{r}
table(data_frame_ed_CHD2_P3$sample_name, data_frame_ed_CHD2_P3$type )
```

## How many mutations are not consistent across clones?
```{r}
## Table with id frequency
tab <- table(data_frame_ed_CHD2_P3$id)

message("IDs found in both 3 clones: ")
print(length(tab[tab>1]))
```

## Filter deletions/duplications not mapped to exons
```{r}
data_frame_ed_CHD2_P3_exons <- data_frame_ed_CHD2_P3[data_frame_ed_CHD2_P3$nexons >= 1,]
data_frame_ed_CHD2_P3_exons
```




# DM1 P2 clones
```{r}
DM1_P2_names <- c('DM1_p2c1', 'DM1_p2c2', 'DM1_p2c3')
data_frame_ed_DM_P2 <- data_frame_ed[data_frame_ed$sample_name %in% DM1_P2_names,]
```


Number of mutations found per clone per chromosome
```{r}
table(data_frame_ed_DM_P2$sample_name, data_frame_ed_DM_P2$chromosome )
```

Type of mutations found per clone 
```{r}
table(data_frame_ed_DM_P2$sample_name, data_frame_ed_DM_P2$type )
```

## How many mutations are not consistent across clones?
```{r}
## Table with id frequency
tab <- table(data_frame_ed_DM_P2$id)

message("IDs found in at least two of 3 clones: ")
print(length(tab[tab>1]))

message("IDs found in all 3 clones: ")
print(length(tab[tab>2]))
```



## Filter deletions/duplications not mapped to exons
```{r}
data_frame_ed_DM_P2_exons <- data_frame_ed_DM_P2[data_frame_ed_DM_P2$nexons >= 1,]
data_frame_ed_DM_P2_exons
```


## DM P2 clones on chromsome 12
```{r}
data_frame_ed_DM_P2_Chr12 <- data_frame_ed_DM_P2[data_frame_ed_DM_P2$chromosome == 12,]
data_frame_ed_DM_P2_Chr12
```

NANOG_2
```{r}
data_frame_ed_DM_P2_Chr12[!is.na(data_frame_ed_DM_P2_Chr12$exons.hg19) & startsWith(data_frame_ed_DM_P2_Chr12$exons.hg19, 'NANO'),]
```

## Clone 2 on only chromsome 12
```{r}
data_frame_ed_DM_P2_Chr12_Clone2 <- data_frame_ed_DM_P2_Chr12[data_frame_ed_DM_P2_Chr12$sample_name == 'DM1_p2c2',]
```



## Compare with DMRs found between clone 2 and other clones
```{r}
DMRs_c2_df <- read.csv('Z:/data/iPSC_data/Results/DMR/DMRS_DM1_c2.csv', row.names=1)
DMRs_c2_df_Chr12 <- DMRs_c2_df[DMRs_c2_df$chr == 'chr12',]
```


## Visualize overlap
```{r}
df_Meth <- DMRs_c2_df_Chr12[, c('start', 'end')]
df_WES <- data_frame_ed_DM_P2_Chr12_Clone2[, c('start', 'end')]
```


```{r, fig.width=14, fig.height=3}
df_Meth$group <- 'DMR'
df_WES$group <- 'WES'

df_combined <- rbind(df_Meth, df_WES)
df_combined$id <- seq_len(nrow(df_combined))

plot_WES_Meth_overlap <- function(df, linetype = 'butt') {
    df$y_jitter <- as.numeric(factor(df$group)) + runif(nrow(df), -0.3, 0.3)
    
    ggp <- ggplot(df) +
      geom_segment(aes(x=start, xend = end,
                   y = y_jitter, yend = y_jitter,
                   group = id, color = group),
                   size = 2, alpha = 1,
                   lineend = linetype) +
      #scale_color_viridis_d(option= "cividis", name = "group") +
      theme_minimal()
  return(ggp)
}

ggp <- plot_WES_Meth_overlap(df_combined, linetype = 'round')
ggp
ggsave('Chr12_WES_EMseq.pdf', ggp, width=20, height=3, dpi=300)
```


## Compare All samples on this Chromosome 12 locus
```{r}
data_frame_ed_Chr12 <- data_frame_ed[data_frame_ed$chromosome == 12,]
df_WES <- data_frame_ed_Chr12[, c('start', 'end')]


df_Meth <- DMRs_c2_df_Chr12[, c('start', 'end')]


df_Meth$group <- 'DMR'
df_WES$group <- 'WES'

df_combined <- rbind(df_Meth, df_WES)
df_combined$id <- seq_len(nrow(df_combined))

plot_WES_Meth_overlap <- function(df, linetype = 'round') {
    
    df$y_jitter <- as.numeric(factor(df$group)) + runif(nrow(df), -0.3, 0.3)
    
    ggp <- ggplot(df) +
      geom_segment(aes(x=start, xend = end,
                   y = y_jitter, yend = y_jitter,
                   group = id, color = group),
                   size = 2, alpha = 1,
                   lineend = linetype) +
      #scale_color_viridis_d(option= "cividis", name = "group") +
      theme_minimal()
        
  return(ggp)
}

ggp <- plot_WES_Meth_overlap(df_combined)
ggp
ggsave('Chr12_WES_EMseq_ALL_SAMPLES.pdf', ggp, width=20, height=3, dpi=300)
```





Zoom in in region with many mutations
```{r}
ggp <- plot_WES_Meth_overlap(df_combined[df_combined$start > 125000000, ]) + 
  xlim(c(125000000, 132000000)) +
  xlab('Genomic coordinate') +
  ylab(element_blank()) + 
  ggtitle('WES and Methylation data on Chr 12')
ggp
```

