---
title: "Metabolomics  to tabular format"
output: html_document
params:
  
  metabomics_feature_neg_metadata: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231128_XCMS/XCMSR.annotated.Report_231128_ESIneg.tsv
  metabomics_feature_pos_metadata: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231120_XCMS/XCMSR.annotated.Report_231120_ESIpos.tsv
  metabolomics_pos_data: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231120_UDA/231120_ControlsvsPatients_2024-03-13_ESIpos.tsv
  metabolomics_neg_data: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231128_UDA/231128_ControlsvsPatients_2024-03-13_ESIneg.tsv
  
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  
  outdir: Z:/data/iPSC_data/iPSC_MAE
---

## Read processed metabolomics data and save to tabular formats for the MultiAssayExperiment object

### Libraries
```{r}
library(dplyr)
```

#### Script parameters
```{r}
print(Sys.Date())
print(params)
```


#### Functions to read data
```{r}

## Function to read metabolomics data from TML
read_mtblcs_data_tml <- function(filepath, id_string) {
  
  ### Read data
  mtblmcs_data <- read.delim(filepath, header = FALSE)
  mtblmcs_data$V105 <- NULL
  
  ### Remove columns
  colnames(mtblmcs_data) <- mtblmcs_data[3,]
  rows <- nrow(mtblmcs_data)
  mtblmcs_data <- mtblmcs_data[4:rows,]
  
  mtblmcs_data$Mass <- NULL
  mtblmcs_data$R <- NULL
  
  ## Rename metabolomics features
  feature_names <- lapply(mtblmcs_data$F, function(x) {
    y <- paste0(id_string, x)
    y
  })
  rownames(mtblmcs_data) <- feature_names
  mtblmcs_data$F <- NULL
  
  return(mtblmcs_data)
}

### Save sample/filename mapping dataframe
create_sample_metadata_df <- function(filepath) {
  sample_mapping <- read.delim(filepath, header = FALSE)
  sample_mapping <- sample_mapping[1:3,4:104]
  sample_mapping <- data.frame(t(sample_mapping))
  sample_mapping$X2 <- NULL
  colnames(sample_mapping) <- c('File_name', 'Sample_name')
  return(sample_mapping)
}


### Store Mass and Retention time in other dataframe
create_feature_metadata_df  <- function(filepath) {
  data <- read.delim(filepath, header = FALSE)
  colnames(data) <- data[3,]
  rows <- nrow(data)
  data <- data[4:rows,]
  feature_metadata_df <- data[, 2:3]
  return(feature_metadata_df)
}  
```

## Read data for positive ionization mode
```{r}
## measurement data
metabolomics_pos <- read_mtblcs_data_tml(params$metabolomics_pos_data, 'TML_pos')

## sample metadata
sample_mapping <- create_sample_metadata_df(params$metabolomics_pos_data)

## feature metadata
metabolomics_features_pos_ids <- create_feature_metadata_df(params$metabolomics_pos_data)
```


## Read data for negative ionization mode
```{r}
## measurement data
metabolomics_neg <- read_mtblcs_data_tml(params$metabolomics_neg_data, 'TML_neg')

## feature metadata
metabolomics_features_neg_ids <- create_feature_metadata_df(params$metabolomics_neg_data)
```


#### Feature metadata (positive)
```{r }
metabolomics_features_pos <- read.delim(params$metabomics_feature_pos_metadata, row.names = 1)
columns_keep <- c("mzmed", "mzmin", "mzmax", "rtmed", 
                  "rtmin", "rtmax", "npeaks", "X1", "ms_level",
                  "isotopes", "adduct", "pcgroup")
metabolomics_features_pos <- metabolomics_features_pos[ , names(metabolomics_features_pos) %in% columns_keep]

## Add mass and retention time from UDA file
metabolomics_features_pos <- cbind(metabolomics_features_pos_ids, metabolomics_features_pos)

## Add feature IDs (temporarily)
rownames(metabolomics_features_pos) <- rownames(metabolomics_pos)
head(metabolomics_features_pos)
print(dim(metabolomics_features_pos))
```


### Filter features with too much missing values
```{r}
library(dplyr)
filter_na_columns <- function(data, threshold = 0.2) {
  
  # calculate NA values per column
  na_proportion <- colSums((data==0) / nrow(data))
  # Filter out columns about threshold
  data_out <- data %>% select(which(na_proportion <= threshold))
  return(data_out)
}

# Remove features
metabolomics_pos <- data.frame(t(metabolomics_pos))
metabolomics_filtered_pos <- filter_na_columns(metabolomics_pos)
dim(metabolomics_filtered_pos)


# Remove these features from feature metadata table
metabolomics_features_pos <- metabolomics_features_pos[colnames(metabolomics_filtered_pos),]
```
### Replace zero's by half of the minimum value
```{r}
### Make columns numericals
metabolomics_filtered_pos[] <- lapply(metabolomics_filtered_pos, as.numeric)

## Replace zero's by half of minimum
metabolomics_pos_noZero <- metabolomics_filtered_pos %>% mutate(across(everything(), ~if_else(. == 0, min(.[. != 0])/2, .)))
```

### Log-transform data
```{r}
metabolomics_pos_noZero_log <- log(metabolomics_pos_noZero + 1)
```

### Remove QC samples
```{r}
metabolomics_pos_noZero_noQC <- metabolomics_pos_noZero[!grepl('^QC', rownames(metabolomics_pos_noZero)),]
metabolomics_pos_noZero_log_noQC <- metabolomics_pos_noZero_log[!grepl('^QC', rownames(metabolomics_pos_noZero_log)),]
```

### Sample names matching
```{r}
sample_mapping_noQC <- sample_mapping[!grepl('^QC', sample_mapping$Sample_name),]

sample_ids_temp <- sub(".*?-.*?-(.*)", "\\1", sample_mapping_noQC$Sample_name)
sample_ids <- sub("^([^_]*)_", "\\1", sample_ids_temp)

sample_mapping_noQC$Sample_name <- sample_ids


# New column for ISA names
sample_identifiers_ISA <- lapply(sample_mapping_noQC$Sample_name, function(x){
    name <- substr(x, start = 1, stop = (nchar(x) - 1))
    name
})
sample_mapping_noQC$Sample_name_ISA <- as.character(sample_identifiers_ISA)
```


### Merge technical replicates by taking mean
```{r}
metabolomics_pos_noZero_log_noQC_merged <- metabolomics_pos_noZero_log_noQC
metabolomics_pos_noZero_noQC_merged <- metabolomics_pos_noZero_noQC

## Function to rename sample IDs to isa names
rename_samples <- function(sample_id) {
  
  ## Remove ".1", indicating technical replicate
  if (substring(sample_id, nchar(sample_id)-1, nchar(sample_id)) == ".1") {
    temp_name <- substring(sample_id, 1, nchar(sample_id)-2)
  } else {
    temp_name <- sample_id
  }
  
  ## Split based on - char
  temp_name <- substr(temp_name, 21, nchar(temp_name))
  temp_name <- gsub("\\.", "-", temp_name)
  parts <- strsplit(temp_name, "-")

  ## Select parts that are relevent for naming  
  if (length(parts[[1]]) > 2) {
    new_sample_id <- paste0(parts[[1]][[2]], '-', parts[[1]][[3]])
  }
  else {
    new_sample_id <- parts[[1]][[2]]
  }
  return(new_sample_id)
}

# Rename Sample IDs without .1
new_names <- lapply(rownames(metabolomics_pos_noZero_log_noQC_merged), rename_samples)
metabolomics_pos_noZero_log_noQC_merged$Sample_ID <- as.character(new_names)
#metabolomics_pos_noZero_log_noQC_merged$SampleName <- NULL

# Aggregate technical replicates with mean
metabolomics_pos_noZero_log_noQC_merged <- metabolomics_pos_noZero_log_noQC_merged %>%
  group_by(Sample_ID)  %>%
  summarize(across(everything(), mean))


## Same for non-log transformed data
# Rename Sample IDs without .1
new_names <- lapply(rownames(metabolomics_pos_noZero_noQC_merged), rename_samples)
metabolomics_pos_noZero_noQC_merged$Sample_ID <- as.character(new_names)
#metabolomics_pos_noZero_log_noQC_merged$SampleName <- NULL

# Aggregate technical replicates with mean
metabolomics_pos_noZero_noQC_merged <- metabolomics_pos_noZero_noQC_merged %>%
  group_by(Sample_ID)  %>%
  summarize(across(everything(), mean))
  
```

## ISA files with phenotype data
```{r}
isa_assay_prot <- readxl::read_xlsx('C:/Users/Casper.deVisser/Documents/ipsc-data-analysis/20240822_Xomics_iPSC_Neurons_new_READ-ONLY.xlsx', sheet = 'Assay - proteomics')
isa_assay_prot$SampleName <- paste0(isa_assay_prot$`sample identifier`, '_', isa_assay_prot$`biological replicate`)
# Match sample names
metabolomics_pos_noZero_log_noQC_merged$Sample_ID[metabolomics_pos_noZero_log_noQC_merged$Sample_ID == 'IPS21-00151_2'] <- 'IPS22-00151_2'
```


## Phenotype data
```{r}
library(MultiAssayExperiment)
library(HDF5Array)

mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')

phenotype_data <- as.data.frame(colData(mae))

## Add phenotype col
add_pheno_col <- function(metabolomicds_matching_isa) {
  phenotypes <- lapply(metabolomicds_matching_isa$Sample_ID, function(x) {
  pheno <- phenotype_data[phenotype_data$sample.identifier == x, "phenotype"]
  pheno
})
  metabolomicds_matching_isa$phenotype <- as.character(phenotypes)
  metabolomicds_matching_isa <- metabolomicds_matching_isa[metabolomicds_matching_isa$phenotype != 'character(0)',]
  return(metabolomicds_matching_isa)
  
}
metabolomics_pos_noZero_log_noQC_merged <- add_pheno_col(metabolomics_pos_noZero_log_noQC_merged)

metabolomics_pos_noZero_noQC_merged <- add_pheno_col(metabolomics_pos_noZero_noQC_merged)
```


## Plot raw measurments per metabolite
```{r}
plot(metabolomics_pos$TML_pos2019)
plot(metabolomics_pos_noZero$TML_pos2019)
plot(metabolomics_pos_noZero_log$TML_pos2019)
plot(metabolomics_pos_noZero_log_noQC_merged$TML_pos2019)
plot(metabolomics_pos_noZero_noQC_merged$TML_pos2019)
```



### Boxplots per sample
```{r, fig.width=9}
## Function to make boxplot of top associated -omics feature
boxplot_per_group <- function(df_input, feature, disease_group = 'Disease', y_lab, phenotype='phenotype') {
  #print(feature)
  df2plot <- df_input[, c(feature, phenotype)]
  colnames(df2plot) <- c("y_feature", 'group')
  
  ggp <- ggplot(df2plot, 
             aes(x = group, y = y_feature, fill = group)) +
        geom_boxplot(position = position_dodge(width = 0.9))  +
        #scale_fill_okabe_ito() + 
         ggtitle(stringr::str_wrap(paste0(
          feature), 
          width = 60)) +
        theme_bw() +
        labs(y = paste0(y_lab)) +
        theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1),
              axis.text.y = element_text(angle =90))
  #full_path <- file.path(outdir, disease_group, assay_name)
  # full_path <- file.path(outdir, disease_group) #TODO separate function/parameter for this
  # if (!dir.exists(full_path)) {
  #   dir.create(full_path, recursive = T)
  # }
  # file_path <- file.path(full_path, paste0(feature, '.png'))
  # ggsave(file_path, ggp,
  #     width = 25.04, height = 17.46, units = "cm", dpi = 300)
  return(ggp)
}
boxplot_per_group(metabolomics_pos_noZero_log_noQC_merged, 'TML_pos2019', y_lab = 'Log-transformed values')
boxplot_per_group(metabolomics_pos_noZero_noQC_merged, 'TML_pos2019', y_lab = 'Non-transormed values')

boxplot_per_group(metabolomics_pos_noZero_log_noQC_merged, 'TML_pos6127', y_lab = 'Log-transformed values')
boxplot_per_group(metabolomics_pos_noZero_noQC_merged, 'TML_pos6127', y_lab = 'Non-transormed values')

```


## Save to CSV
```{r}
write.csv(metabolomics_final, file =  file.path(params$outdir, 'metabolomics_tml_data.csv'), row.names = F)
write.csv(feature_metadata, file =  file.path(params$outdir, 'metabolomics_tml_feature_metadata.csv'))
write.csv(sample_mapping_noQC, file =  file.path(params$outdir, 'metabolomics_tml_sample_metadata.csv'))
```


```{r}
sessionInfo()   
```

