---
title: "CNV_WES2Tabular"
output: html_document
params:
  cnv_wes_ed_data: Z:/data/iPSC_data/iPSC_CNV-WES/CNV_processed/
  cnv_wes_conifer_data: Z:/data/iPSC_data/iPSC_CNV-WES/DNA24-01729_calls.conifer.txt

---

## Read processed CNV-WES data and save to tabular formats for the MultiAssayExperiment object

#### Script parameters
```{r}
print(Sys.Date())
print(params)
```

### Function
```{r}
read_ed_file <- function(dir) {
  
  # ED file
  ed_files <- dir(dir, pattern = "ED.txt")
  ed_file <- ed_files[[1]]
  
  # Extract sample name from file name
  sample_name <- gsub(params$cnv_wes_ed_data, "", dir)
  sample_name <- gsub("_Genomics", "", sample_name) ##TODO
  
  file <- read.delim(file.path(dir, ed_file))
  
  file$sample_name <- rep(sample_name, nrow(file))
  return(file)
}

```

### Read all ED files
```{r}
parent.folder <- params$cnv_wes_ed_data
sub.folders <- list.dirs(parent.folder, recursive = TRUE)[-1]

data_frame_ed <-  read_ed_file(sub.folders[[1]])

for (sub in sub.folders[2:16]) {
  temp <- read_ed_file(sub)
  data_frame_ed <- rbind(data_frame_ed, temp)
}

write.csv(data_frame_ed, 'Z:/data/iPSC_data/iPSC_CNV-WES/Data_integration/cnv_ed.csv', row.names = FALSE)
```

## Number of CNV calls per sample
```{r}
table(data_frame_ed$sample_name)
```

### Retrieve unique features and sample names
```{r}
unique_features <- unique(data_frame_ed$id)
samples <- unique(data_frame_ed$sample_name)
```

### Combine data into classical sample/feature data frame
```{r}
library(plyr)

df_per_sample_list <-lapply(samples, function(sample) {
  df_temp <- data_frame_ed[ data_frame_ed$sample_name == sample,] 
  df_temp <- df_temp[, c( 'id', 'reads.ratio')]
  rownames(df_temp) <- df_temp$id
  df_temp$id <- NULL
  df_temp <- data.frame(t(df_temp))
})
names(df_per_sample_list) <- samples

df_combined <- plyr::rbind.fill(df_per_sample_list)
rownames(df_combined) <- samples

df_combined[,1:5]
```


### Visualize missingness in data
```{r}
source("support_functions/heatmap_missingness.R")
heatmap_missing(df_combined, 'CNV-WES')
```

### CNV calls present in most samples
```{r}
s <- sort(apply(is.na(df_combined), 1, sum))
df_combined_ordered <- df_combined[,order(colSums(is.na(df_combined)))]
df_combined_ordered[, 1:10]
```




## Principal Component analysis
```{r  fig.width=14}

# df_pca_input <- data.frame(t(df_combined_ordered[, 1:100])) 
# performPCA(df_pca_input, "X409B", "WTC") 

# TODO: Modify imputation method for this plot, or in separate function
```

```{r}
source("support_functions/performPCA.R")
## Manual PCA code (impute with 1)

data_table <- data.frame(t(df_combined_ordered[, 1:100])) 
start_col_name <-  "X409B"
end_col_name <- "WTC"

start = as.numeric(which(colnames(data_table) == start_col_name))
end = as.numeric(which(colnames(data_table) == end_col_name))
  
data_table_sub <- data_table[,start:end]
  
#transpose matrix
data_table_sub <- t(data_table_sub)
# Convert NA to 0
data_table_sub[is.na(data_table_sub)] <- 1
  
# Add a new column to indicate sample categories
# The categories have been extracted from the rownames, as the first word before row names
# NANS_p3c1_Proteomics_2_S2_A11_1 is NANS
sampleType <- sub("^(.*?)_.*","\\1", rownames(data_table_sub))
colnames(data_table_sub)[ncol(data_table_sub)] <- "sampleType"
  
data.pca <- prcomp(data_table_sub, center = TRUE, scale.= TRUE)
g <- ggbiplot(
    data.pca,
    groups = sampleType,
    var.axes = F,
    labels = rownames(data_table_sub)
  )
ggsave("Z:/data/iPSC_data/Results/QC/WES_PCA.pdf",
       plot = g, width = 7, height = 7)
```