---
title: "Create MultiAssayExperiment object"
output:
  html_document:
    df_print: paged
  pdf_document: default
params:

  isa_files: Z:/data/iPSC_data/iPSC_MAE/XomicsiPSCneurons_march_7.xlsx
  
  metabolomics_tml_data_Log: Z:/data/iPSC_data/iPSC_MAE/metabolomics_tml_data.csv
  metabolomics_tml_feature_metadata: Z:/data/iPSC_data/iPSC_MAE/metabolomics_tml_feature_metadata.csv
  metabolomics_tml_sample_metadata: Z:/data/iPSC_data/iPSC_MAE/metabolomics_tml_sample_metadata.csv
  
  proteomics_data: Z:/data/iPSC_data/iPSC_MAE/proteomics_data.csv
  proteomics_feature_metadata: Z:/data/iPSC_data/iPSC_MAE/proteomics_feature_metadata.csv
  
  lumc_sample_matching: 'Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Sample list.xlsx'
  
  metabolomics_lumc_data_total: Z:/data/iPSC_data/iPSC_MAE/metabolomics_lumc_data_total.csv
  metabolomics_lumc_data_protein: Z:/data/iPSC_data/iPSC_MAE/metabolomics_lumc_data_protein.csv
  metabolomics_lumc_sample_metadata: Z:/data/iPSC_data/iPSC_MAE/metabolomics_lumc_sample_metadata.csv
  metabolomics_lumc_feature_metadata: Z:/data/iPSC_data/iPSC_MAE/metabolomics_lumc_feature_metadata.csv
  
  metabolomics_lumc_data_total_ann: Z:/data/iPSC_data/iPSC_MAE/metabolomics_lumc_data_total_Ann.csv
  metabolomics_lumc_feature_metadata_ann: Z:/data/iPSC_data/iPSC_MAE/metabolomics_lumc_feature_metadata_Ann.csv
  
  metabolomics_lumc_data_unfiltered: Z:/data/iPSC_data/iPSC_MAE/metabolomics_lumc_data_raw_log.csv
  metabolomics_lumc_feature_metadata_unfiltered: Z:/data/iPSC_data/iPSC_MAE/metabolomics_lumc_feature_metadata_raw.csv
  
  lipidomics_T_log_data: Z:/data/iPSC_data/iPSC_MAE/lipidomics_T_imputed_log_data.csv
  lipidomics_T_TA_log_data: Z:/data/iPSC_data/iPSC_MAE/lipidomics_T_TA_imputed_data.csv
  lipidomics_T_sample_metadata: Z:/data/iPSC_data/iPSC_MAE/lipidomics_T_sample_metadata.csv
  lipidomics_T_feature_metadata: Z:/data/iPSC_data/iPSC_MAE/lipidomics_T_feature_metadata.csv
  
  lipidomics_UT_data: Z:/data/iPSC_data/iPSC_MAE/lipidomics_UT_data.csv
  lipidomics_UT_feature_metadata: Z:/data/iPSC_data/iPSC_MAE/lipidomics_UT_feature_metadata.csv
  lipidomics_UT_sample_metadata: Z:/data/iPSC_data/iPSC_MAE/lipidomics_UT_sample_metadata.csv
  lipidomics_UT_data_filtered: Z:/data/iPSC_data/iPSC_MAE/lipidomics_UT_data_filtered.csv
  lipidomics_UT_feature_metadata_filtered: Z:/data/iPSC_data/iPSC_MAE/lipidomics_UT_feature_metadata_filtered.csv
  lipidomics_UT_sample_metadata_filtered: Z:/data/iPSC_data/iPSC_MAE/lipidomics_UT_sample_metadata_filtered.csv
  
  mRNAseq_data_cpm: Z:/data/iPSC_data/iPSC_MAE/mrna_cpm_log.csv
  mRNAseq_data_deseq: Z:/data/iPSC_data/iPSC_MAE/mrna_deseq_log.csv
  mRNAseq_feature_metadata: Z:/data/iPSC_data/iPSC_MAE/mrna_feature_metadata.csv
  mRNAseq_feature_metadata_rt: Z:/data/iPSC_data/iPSC_MAE/mrna_feature_metadata_rt.csv
  
  mRNAseq_data_rtc_vst: Z:/data/iPSC_data/iPSC_MAE/mrna_deseq_rt_vst.csv
  
  EMseq_filt_sd: Z:/data/iPSC_data/iPSC_MAE/meth_filt_sd.RDS
  EMseq_filt_rand: Z:/data/iPSC_data/iPSC_MAE/meth_filt_rand.RDS
  EMseq_mapping_table: Z:/data/iPSC_data/iPSC_EMseq/MappingCPG2Gene.csv
  
  h5_outdir: Z:/data/iPSC_data/iPSC_MAE/h5MAE
---


This R notebook creates a MultiAssayExperiment using the *CSV* files containing preprocessed omics measurements.


## Libraries
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
suppressMessages(library(SummarizedExperiment))
suppressMessages(library(MultiAssayExperiment))
suppressMessages(library(HDF5Array))
```


```{r parameters}
print(Sys.Date())
print(params)
```


## Phenotype data

Read phenotype information from CSV file.

```{r read_phenotype_data}
pheno_df <- readxl::read_xlsx(params$isa_files, sheet = 4)
# pheno_df <- read.csv(params$phenotype_data, stringsAsFactors = TRUE)
cat("Column names of pheno_df:\n")
print(names(pheno_df))
cat("Dimensions of pheno_df:\n")
dim(pheno_df)
```

## Add Sex column from observation unit tab
```{r}
observationUnit_df <- readxl::read_excel(params$isa_files, sheet =3)

# Add new column for gender, to match observation unit df
gender_col <- lapply(pheno_df$`observation unit identifier`, function(x) {
    gender <- observationUnit_df[observationUnit_df$`observation unit identifier` == x, 7]
    gender <- gender[[1]]
})

pheno_df$sex <- as.character(gender_col)
```


### Tripling number of rows (adding biological replicates)
```{r}
library(stringr)
library(purrr)

source('./createMAE_functions/tripling_rows')

pheno_df_replicates <- tripling_rows(pheno_df)

## Correct passage number columns
NewPassageNumbers <- sapply(seq_len(nrow(pheno_df_replicates)), function(i) {
  idx <- as.integer(sub(".*(\\d+)$", "\\1", pheno_df_replicates$`sample identifier`[i]))
  vals <- strsplit(pheno_df_replicates$`passage number`[i], ";")[[1]]
  vals[idx]
})

pheno_df_replicates$PassageNumber <- as.numeric(NewPassageNumbers)
pheno_df_replicates$`passage number` <- NULL
## TODO: Remove replicates that do not exist for all -omics layers?
```


### Sample IDs will be altered by R when added as colnames (in assays), add similar names here
```{r}
rownames(pheno_df_replicates) <- pheno_df_replicates$`sample identifier`
pheno_df_temp <- data.frame(t(pheno_df_replicates))
sample_colnames <- colnames(pheno_df_temp)
pheno_df_replicates <- data.frame(t(pheno_df_temp))
# Replace "." by underscore
rownames(pheno_df_replicates) <- gsub("\\.", "_", rownames(pheno_df_replicates))
```

## Numerical columns to numerical
```{r}
pheno_df_replicates$age.at.sampling.from.host <- as.numeric(pheno_df_replicates$age.at.sampling.from.host)

pheno_df_replicates$biosafety.level <- as.numeric(pheno_df_replicates$biosafety.level)

pheno_df_replicates$days.in.vitro <- as.numeric(pheno_df_replicates$days.in.vitro)
```


## Add second observation unit column with other values for the CRISPR lines
```{r}
pheno_df_replicates$observation.unit.identifier.crispr <- pheno_df_replicates$observation.unit.identifier
pheno_df_replicates[pheno_df_replicates$sample.name == 'DM1 patient 2 clone 1 repeat excised', 'observation.unit.identifier.crispr'] <- 'DM1_Rescue'
pheno_df_replicates[pheno_df_replicates$sample.name == 'control 2 (originally CHD2 het in WTC control)', 'observation.unit.identifier.crispr'] <- 'CHD2_CRISPR'
```


## Create SummarizedExperiment per omics

### Metabolomics TML
```{r}
feature_metadata <- read.csv(params$metabolomics_tml_feature_metadata, row.names = 1)
feature_names <- rownames(feature_metadata)

mtblmcs_data <- data.frame(t(read.csv(params$metabolomics_tml_data_Log, row.names = 1)))
# Replace "." by underscore
colnames(mtblmcs_data) <- gsub("\\.", "_", colnames(mtblmcs_data))
# Ensure features are all in feature metadata
mtblmcs_data <- mtblmcs_data[feature_names,]
# Make columns numericals
mtblmcs_data <- data.frame(sapply(mtblmcs_data, as.numeric))
rownames(mtblmcs_data) <- feature_names
# Fix typo in one sample name
sample_names <- colnames(mtblmcs_data)
sample_names[sample_names == 'IPS21_00151_2'] <- 'IPS22_00151_2'
colnames(mtblmcs_data) <- sample_names 

my_assay_list = list(
  `Metabolomics, TML` = mtblmcs_data
  )


se <- SummarizedExperiment(
    assays = my_assay_list,
    # # rows: features
    rowData = feature_metadata, 
    # # columns: samples
    #colData = sample_metadata, 
    # # metadata about data processing
    # metadata = list(),
    checkDimnames = TRUE)

experiment_list <- ExperimentList(list(`Metabolomics, TML` = se))

```


### Proteomics data and metadata
```{r}
## Feature metadata
feature_metadata <- read.csv(params$proteomics_feature_metadata, row.names = 2)

## Read proteomics data
proteomics_data <- read.csv(params$proteomics_data, row.names = 1)

# CDH2 --> CHD2
rownames(proteomics_data) <- gsub("CDH2", "CHD2", rownames(proteomics_data))

## Map to unified IDs in ISA Proteomics file
isa_assay_prot <- readxl::read_excel(params$isa_files, sheet = 'Assay - Proteomics')
### Add assay identifier column without tech rep number
isa_assay_prot$`assay identifier2` <- substr(isa_assay_prot$`assay identifier`, 1, nchar(isa_assay_prot$`assay identifier`)-2)


## Change sample order to ISA order
proteomics_data <- proteomics_data[unique(match(isa_assay_prot$`assay identifier2`, rownames(proteomics_data))),]
unique_ids <- lapply(1:nrow(proteomics_data), function(x) {
    name1 <- isa_assay_prot[x, "sample identifier"]
    s <- isa_assay_prot[x, "assay identifier2"]
    rep <- substring(s, nchar(s), nchar(s))
    paste0(name1, "_", rep)
})
rownames(proteomics_data) <- as.character(unique_ids)
proteomics_data <- data.frame(t(proteomics_data))

## Ensure the feature names match
proteomics_data <- proteomics_data[rownames(feature_metadata),]

## Replace . by underscore
colnames(proteomics_data) <- gsub("\\.", "_", colnames(proteomics_data))

## Remove columns with missing values
proteomics_data_with_409B <- proteomics_data
proteomics_data <- proteomics_data[ , colSums(is.na(proteomics_data))==0]


my_assay_list = list(
  `Proteomics` = proteomics_data
)


#colnames(my_assay_list[[1]]) <- rownames(sample_metadata)
#print(identical(rownames(sample_metadata), colnames(my_assay)))


se <- SummarizedExperiment(
    assays = my_assay_list,
    # # rows: features
    rowData = feature_metadata,  
    # # columns: samples
    #colData = sample_metadata, 
    # # metadata about data processing
    # metadata = list(),
    checkDimnames = TRUE)

experiment_list <- append(experiment_list, (list(`Proteomics` = se)))
```


## Match sample names across all -omics data
```{r}
sample_names_overview <- data.frame(colnames(proteomics_data_with_409B), colnames(mtblmcs_data))
colnames(sample_names_overview) <- c('Proteomics', 'Mtblmcs_TML')
sample_names_overview
```


### Add Metabolomics LUMC sample names to overview
```{r}
# Read and transpose metabolomics data
mtblmcs_data_lumc <- data.frame(t(read.csv(params$metabolomics_lumc_data_total, row.names = 1)))

# Replace "." with "_" in column names to standardize sample names
lumc_names_with_underscore <- gsub("\\.", "_", colnames(mtblmcs_data_lumc))

# Match LUMC metabolomics sample names from overview to actual column names
sample_names_overview$Mtblmcs_LUMC <- sapply(sample_names_overview$Mtblmcs_TML, function(x) {
  if (x %in% lumc_names_with_underscore) {
    x
  } else if (x == "IPS21_00151_2") {
    # Manual fix for known mismatch
    "IPS22_00151_2"
  } else {
    NA
  }
})
sample_names_overview
```


### Metabolomics data and metadata (LUMC)

Prepare several `SummarizedExperiment` objects for the LUMC metabolomics data. 

```{r prepare_metabolomics_lumc_experiments}

lumc_sample_metadata <- read.csv(params$metabolomics_lumc_sample_metadata, row.names = 1)
lumc_sample_metadata$Sample_Overview_names <- as.character(lumc_names_with_underscore)

metabolomics_lumc_list <-  lapply(list(
     `Metabolomics, LUMC | total area normalized` = list(
      assay_name =  "Metabolomics, LUMC | total area normalized",
      assay_data = params$metabolomics_lumc_data_total,
      assay_sample_metadata = params$metabolomics_lumc_sample_metadata,
      assay_feature_metadata = params$metabolomics_lumc_feature_metadata
      ),
     `Metabolomics, LUMC | total area normalized | annotated` = list(
      assay_name =  "Metabolomics, LUMC | total area normalized | annotated",
      assay_data = params$metabolomics_lumc_data_total_ann,
      assay_sample_metadata = params$metabolomics_lumc_sample_metadata,
      assay_feature_metadata = params$metabolomics_lumc_feature_metadata_ann
      ),
     # `Metabolomics, LUMC | protein content normalized` = list(
     #  assay_name = "Metabolomics, LUMC | protein content normalized",
     #  assay_data = params$metabolomics_lumc_data_protein,
     #  assay_sample_metadata = params$metabolomics_lumc_sample_metadata,
     #  assay_feature_metadata = params$metabolomics_lumc_feature_metadata
     #  ),
     `Metabolomics, LUMC | unfiltered` = list(
      assay_name = "Metabolomics, LUMC | unfiltered",
      assay_data = params$metabolomics_lumc_data_unfiltered,
      assay_sample_metadata = params$metabolomics_lumc_sample_metadata,
      assay_feature_metadata = params$metabolomics_lumc_feature_metadata_unfiltered
      )
  ), function(a) {
  ## Measurement data
  my_assay <- data.frame(t(read.csv(a$assay_data, row.names = 1)))
  ## Sample metadata
  sample_metadata <- read.csv(a$assay_sample_metadata, row.names = 1)
  sample_metadata$Sample_Overview_names <- as.character(lumc_names_with_underscore)
  rownames(sample_metadata) <- sample_metadata$Sample_Overview_names
  colnames(my_assay) <- rownames(sample_metadata)
  ## Feature metadata
  feature_metadata <- read.csv(a$assay_feature_metadata, row.names = 1)
  rownames(feature_metadata) <- feature_metadata$FeatureID
  
  my_assay_list <- list(my_assay)
  names(my_assay_list) <- c(a$assay_name)

  se <- SummarizedExperiment(
      assays = my_assay_list,
      # # rows: features
      rowData = feature_metadata,
      # # columns: samples
      colData = sample_metadata,
      # # metadata about data processing
      # metadata = list(),
      checkDimnames = TRUE)
})

experiment_list <- append(experiment_list, metabolomics_lumc_list)
```


### Lipidomics data (Targeted)
```{r prepare_lipidomics}
lipidomics_targeted_experiments_list <-  lapply(list(
    `Lipidomics, Targeted | imputed | log` = list(
      assay_name = "Lipidomics, Targeted | imputed | log",
      assay_data = params$lipidomics_T_log_data,
      assay_sample_metadata = params$lipidomics_T_sample_metadata,
      assay_feature_metadata = params$lipidomics_T_feature_metadata
      ),
    `Lipidomics, Targeted | imputed | log  | TA` = list(
      assay_name = "Lipidomics, Targeted | imputed | log  | TA",
      assay_data = params$lipidomics_T_TA_log_data,
      assay_sample_metadata = params$lipidomics_T_sample_metadata,
      assay_feature_metadata = params$lipidomics_T_feature_metadata
      )
  ), function(a) {
    # sample metadata
    lip_T_sample_metadata <- read.csv(params$lipidomics_T_sample_metadata, row.names = 1)
    lip_T_sample_metadata$Sample_Overview_names <- lumc_sample_metadata$Sample_Overview_names
    rownames(lip_T_sample_metadata) <- lip_T_sample_metadata$Sample_Overview_names
    # measurement data
    my_assay = data.frame(t(read.csv(a$assay_data, row.names = 1)))
    # Add universal sample names to lipidomics data
    colnames(my_assay)  <-  gsub("\\.", "-", colnames(my_assay))
    lip_T_sample_metadata <- lip_T_sample_metadata[ match(colnames(my_assay), lip_T_sample_metadata$Sample.name),]
    colnames(my_assay) <- lip_T_sample_metadata$Sample_Overview_names
    my_assay_list <- list(my_assay)
    names(my_assay_list) <- c(a$assay_name)
    
    feature_metadata <- read.csv(a$assay_feature_metadata)
    rownames(feature_metadata) <- feature_metadata$feature
    
    # SummarizedExperiment    
    se <- SummarizedExperiment(
      assays = my_assay_list,
      # rows: features
      rowData = feature_metadata, 
      # columns: samples
      colData = lip_T_sample_metadata, 
      checkDimnames = TRUE)
  })


experiment_list <- append(experiment_list, lipidomics_targeted_experiments_list)
```

### Lipidomics data (Untargeted)

```{r prepare_lipidomics_untargeted}

# Sample metadata
lip_UT_sample_metadata <- read.csv(params$lipidomics_UT_sample_metadata, row.names = 1)

# Order based on Kevin nr. 
lip_UT_sample_metadata$Sample_Overview_names <- lumc_sample_metadata[order(lumc_sample_metadata$Kevin.nr),]$Sample_Overview_names
rownames(lip_UT_sample_metadata) <- lip_UT_sample_metadata$Sample_Overview_names

# measurement data
my_assay = data.frame(t(read.csv(params$lipidomics_UT_data, row.names = 1)))

# Add universal sample names to lipidomics data
colnames(my_assay)  <-  gsub("\\.", "-", colnames(my_assay))
colnames(my_assay) <- lumc_sample_metadata[order(lumc_sample_metadata$Kevin.nr),]$Sample_Overview_names


# Feature metadata
feature_metadata <- read.csv(params$lipidomics_UT_feature_metadata, row.names = 1)
rownames(feature_metadata) <- feature_metadata$my_id

## Add lipidomics without outlier sample IPS22-00151-009
my_assay <- my_assay
my_assay$IPS22_00151_1 <- NULL
lip_UT_sample_metadata <- lip_UT_sample_metadata[-c(9), ]

my_assay_list <- list(my_assay)

# SummarizedExperiment    
se <- SummarizedExperiment(
    assays = my_assay_list,
    # rows: features
    rowData = feature_metadata, 
    # columns: samples
    colData = lip_UT_sample_metadata, 
    checkDimnames = TRUE)

experiment_list <- append(experiment_list, (list(`Lipidomics, Untargeted` = se)))




## Function to select to fat
select_top_sd_cols <- function(x, n = 250) {
  std_devs <- sapply(x, sd)
  top_columns <- names(sort(std_devs, decreasing = TRUE)[1:n])
  df_sub <-  x[,top_columns]
  return(df_sub)
}

## Select subset of most variable lipids
lipidomics <- data.frame(t(my_assay))
lipidomics_sub <- select_top_sd_cols(lipidomics)

## Feature metadata
feature_metadata_sub <- feature_metadata[colnames(lipidomics_sub),]

my_assay_list <- list(data.frame(t(lipidomics_sub)))

# SummarizedExperiment    
se <- SummarizedExperiment(
    assays = my_assay_list,
    # rows: features
    rowData = feature_metadata_sub, 
    # columns: samples
    colData = lip_UT_sample_metadata, 
    checkDimnames = TRUE)

experiment_list <- append(experiment_list, (list(`Lipidomics, Untargeted | Variable` = se)))





## Unfiltered (old) Lipidomics data
# Sample metadata
lip_UT_sample_metadata <- read.csv(params$lipidomics_UT_sample_metadata_filtered, row.names = 1)

# Order based on Kevin nr. 
lip_UT_sample_metadata$Sample_Overview_names <- lumc_sample_metadata[order(lumc_sample_metadata$Kevin.nr),]$Sample_Overview_names
rownames(lip_UT_sample_metadata) <- lip_UT_sample_metadata$Sample_Overview_names

# measurement data
my_assay = data.frame(t(read.csv(params$lipidomics_UT_data_filtered, row.names = 1)))

# Add universal sample names to lipidomics data
colnames(my_assay)  <-  gsub("\\.", "-", colnames(my_assay))
colnames(my_assay) <- lumc_sample_metadata[order(lumc_sample_metadata$Kevin.nr),]$Sample_Overview_names

# Feature metadata
feature_metadata <- read.csv(params$lipidomics_UT_feature_metadata_filtered, row.names = 1)
rownames(feature_metadata) <- feature_metadata$FeatureID

# Remove outlier sample IPS22-00151-009
#my_assay <- my_assay
my_assay$IPS22_00151_1 <- NULL
lip_UT_sample_metadata <- lip_UT_sample_metadata[-c(9), ]

my_assay_list <- list(my_assay)

# SummarizedExperiment    
se <- SummarizedExperiment(
    assays = my_assay_list,
    # rows: features
    rowData = feature_metadata, 
    # columns: samples
    colData = lip_UT_sample_metadata, 
    checkDimnames = TRUE)

experiment_list <- append(experiment_list, (list(`Lipidomics, Untargeted | filtered` = se)))

```




### Transcriptomics data and metadata

#### Sample names matching
```{r}
isa_assay_rna <- readxl::read_excel(params$isa_files, sheet = 'Assay - transcriptomics')

## Read RNAseq data 
my_assay = data.frame(t(read.csv(params$mRNAseq_data_cpm, row.names = 1)))

source('createMAE_functions/match_rna_names')
universal_names_rna <- match_rna_names(colnames(my_assay), isa_assay_rna = isa_assay_rna)
```

#### Match to samples overview data frame
```{r}
match_order_tml <- lapply(sample_names_overview$Mtblmcs_TML, function(x) {
     if (x %in% universal_names_rna) {
         x
     }
     else {
         x <- NA
     }
     x
 })

sample_names_overview$RNAseq <- as.character(match_order_tml)
sample_names_overview
```


```{r}
transcriptomics_targeted_experiments_list <-  lapply(list(
    `RNAseq | Deseq` = list(
      assay_name = "RNAseq | Deseq",
      assay_data = params$mRNAseq_data_deseq,
      #assay_sample_metadata =
      assay_feature_metadata = params$mRNAseq_feature_metadata
    )
    # `RNAseq | CPM` = list(
    #   assay_name = "RNAseq | RTC",
    #   assay_data = params$mRNAseq_data_rtc_vst,
    #   #assay_sample_metadata =
    #   assay_feature_metadata = params$mRNAseq_feature_metadata
    #   )
  ), function(a) {
    # measurement data
    my_assay = data.frame(t(read.csv(a$assay_data, row.names = 1)))
    universal_names_rna_x <- match_rna_names(colnames(my_assay), isa_assay_rna = isa_assay_rna)
    colnames(my_assay) <- universal_names_rna_x
    my_assay_list <- list(my_assay)
    
    # feature metadata
    feature_metadata <- read.csv(a$assay_feature_metadata, row.names=1)
    
    # SummarizedExperiment    
    se <- SummarizedExperiment(
      assays = my_assay_list,
      # rows: features
      #rowData = feature_metadata, 
      # columns: samples
      #colData = , 
      checkDimnames = TRUE)
  })


experiment_list <- append(experiment_list, transcriptomics_targeted_experiments_list)
```

## RNAseq RTC data
```{r}
transcriptomics_targeted_experiments_list1 <-  lapply(list(
    `RNAseq | RTC` = list(
      assay_name = "RNAseq | RTC",
      assay_data = params$mRNAseq_data_rtc_vst,
      #assay_sample_metadata =
      assay_feature_metadata = params$mRNAseq_feature_metadata_rt
      )
  ), function(a) {
    # measurement data
    my_assay = data.frame(t(read.csv(a$assay_data, row.names = 1)))
    universal_names_rna_x <- match_rna_names(colnames(my_assay), isa_assay_rna = isa_assay_rna)
    colnames(my_assay) <- universal_names_rna_x
    my_assay_list <- list(counts = as.matrix(my_assay))
    
    # feature metadata
    feature_metadata <- read.csv(a$assay_feature_metadata)
    rownames(feature_metadata) <- make.unique(feature_metadata$feature)
    
    
    # SummarizedExperiment    
    se <- SummarizedExperiment(
      assays = my_assay_list,
      # rows: features
      rowData = feature_metadata, 
      # columns: samples
      #colData = , 
      checkDimnames = TRUE
      )
  })


experiment_list <- append(experiment_list, transcriptomics_targeted_experiments_list1)
```



### Epigenomics data and metadata
```{r}
source('support_functions/ReadMethrix2Df.R')

## Feature metadata variable set
meth <- readRDS(params$EMseq_filt_sd)
meth_sd <- read_meth2df_OLD(meth, isa_assay_prot)
beta_values_sd <- meth_sd[[1]]
feature_metadata_sd <- meth_sd[[2]]

## Map CpGs to genes (from mapping table)
emseq_mapping <- read.csv(params$EMseq_mapping_table, row.names=1) 
gene_symbols <- emseq_mapping$Gene_symbol[match(rownames(feature_metadata_sd), emseq_mapping$feature) ]
feature_metadata_sd$Gene_symbol <- as.character(gene_symbols)
feature_metadata_sd$Gene_symbol[feature_metadata_sd$Gene_symbol == 'character(0)'] <- NA

## Feature metadata random set
meth <- readRDS(params$EMseq_filt_rand)
meth_rand <- read_meth2df_OLD(meth, isa_assay_prot)
beta_values_rand <- meth_rand[[1]]
feature_metadata_rand <- meth_rand[[2]]


## Append to MAE
assay_list <- list( `EMseq | 100,000 most variable CpG sites` = beta_values_sd )

EMseq_experiments <- list()
EMseq_experiments$`EMseq | 100,000 most variable CpG sites`<- SummarizedExperiment(
  assays = assay_list,
  rowData = feature_metadata_sd,
  checkDimnames = TRUE
)

experiment_list <- append(experiment_list, EMseq_experiments)


## Append random selection to MAE
assay_list <- list(  `EMseq | 100,000 random CpG sites` = beta_values_rand)

EMseq_experiments2 <- list()
EMseq_experiments2$`EMseq | 100,000 random CpG sites`<- SummarizedExperiment(
  assays = assay_list,
  rowData = feature_metadata_rand,
  checkDimnames = TRUE
)

experiment_list <- append(experiment_list, EMseq_experiments2)
```


## EMseq M-value data
```{r}
read_meth2df_M <- function(meth, isa_assay) {
  
  ## Measurement data
  beta_values_df <- as.data.frame(meth)
  
  ## Re-format sample names
  new_samples_names <- gsub( "WI-3791-", "", colnames(beta_values_df))
  new_samples_names <- gsub( "Epigenomics", "Proteomics", new_samples_names)
  new_samples_names <- gsub( "_1_val_1_bismark_bt2_pe", "", new_samples_names)
  new_samples_names <- gsub( "-", "_", new_samples_names)
  
  ## Get universal names by mapping to ISA file
  isa_assay <- isa_assay[!duplicated(isa_assay$`assay identifier2`),]
  universal_names_meth <- lapply(new_samples_names, function(x) {
    name_uni <- isa_assay[isa_assay$`assay identifier2` == x, "sample identifier"]
    bio_rep <- isa_assay[isa_assay$`assay identifier2` == x, "biological replicate"]
    name_uni <- paste0(name_uni, "_", bio_rep)
    name_uni
  })
  beta_values_df <- data.frame(beta_values_df)
  colnames(beta_values_df) <- as.character(universal_names_meth)
  colnames(beta_values_df) <- gsub( "-", "_", colnames(beta_values_df))
  colnames(beta_values_df) <- gsub( "409B", "X409B", colnames(beta_values_df))
  
  return(beta_values_df) 
}

input_file <- 'Z:/data/iPSC_data/iPSC_MAE/meth_filt_sd_Mval.RDS'
meth <- readRDS(input_file)
meth_M <- read_meth2df_M(meth, isa_assay_prot)


## Append random selection to MAE
assay_list <- list(  `EMseq | 100,000 variable Mval` = meth_M)

EMseq_experimentsM <- list()
EMseq_experimentsM$`EMseq | 100,000 variable Mval` <- SummarizedExperiment(
  assays = assay_list,
  rowData = feature_metadata_sd,
  checkDimnames = TRUE
)

experiment_list <- append(experiment_list, EMseq_experimentsM)
```





## Create MultiAssayExperiment

The `SummarizedExperiment`s are now added to a `MultiAssayExperiment`. In addition to the `ExperimentList`, phenotype information on individuals is added (colData). General metadata about the cohort can be added here.

```{r create_multiassayexperiment}
mae <- MultiAssayExperiment(
  experiments = experiment_list,
  colData = pheno_df_replicates,
  metadata = list(
    description = "Multi-omics data of iNeurons"
  )
)
mae
```


## Remove mixup sample
```{r}
sample_names <- rownames(colData(mae))
sample_names_NoMixup <- sample_names[sample_names  != "IPS20_00123_3"]

mae <- subsetByColData(mae, y = sample_names_NoMixup)
mae

```  

Save MultiAssayExperiment to HDF5 file.

```{r save_multiassayexperiment}
saveHDF5MultiAssayExperiment(mae, dir = params$h5_outdir, replace = TRUE)
```


## Session information

```{r session_info}
sessionInfo()
```

## References
