---
title: "Lipidomics2Tabular"
output: html_document
params:
  lipidomics_data: Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Targeted_lipidomics_(lipidyzer)/Results/Lipidyzer_Protein content normalized.xlsx
  lipidomics_data_TA: Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Targeted_lipidomics_(lipidyzer)/Results/Lipidyzer_results.xlsx
  sample_matching: "Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Sample list.xlsx"
  isa_files: 'ipsc-data-analysis/20240514_Xomics iPSC Neurons_new_READ-ONLY.xlsx'
  outdir: Z:/data/iPSC_data/iPSC_MAE
 
---

## Read processed lipidomics (targeted) data and save to tabular formats for the MultiAssayExperiment object

#### Script parameters
```{r}
print(Sys.Date())
print(params)
```


## Libraries
```{r}
library(openxlsx)
library(dplyr)
```

## Read data
```{r}
lipidomics <- openxlsx::read.xlsx(params$lipidomics_data, rowNames = TRUE)

## Sample metadata in separate data frame
sample_metadata <- lipidomics[, 1:6]
lipidomics$ExpNum <- NULL
lipidomics$GroupNum <- NULL
lipidomics$GroupName <- NULL
lipidomics$SampleNorm <- NULL
lipidomics$NormType <- NULL
```


## Remove QC samples from measurement data and sample metadata
```{r}
samples <- sample_metadata[sample_metadata$NormType == 'Samples' , 'SampleID']

lipidomics_measurement_data <- lipidomics[lipidomics$SampleID %in% samples,]
sample_metadata <- sample_metadata[sample_metadata$SampleID %in% samples,]
```


```{r}
lipidomics_measurement_data_TA <- openxlsx::read.xlsx(params$lipidomics_data_TA, rowNames = TRUE, sheet = 7)
lipidomics_measurement_data_TA <- lipidomics_measurement_data_TA[lipidomics_measurement_data_TA$SampleID %in% samples,]
```


## Add sample IDs from LUMC sample mapping table
```{r}
## Read excel table
lumc_sample_matching <- read.xlsx(params$sample_matching)

## Check if sample order is equal to mapping table
stopifnot(identical(lumc_sample_matching$Radboud.code, lipidomics_measurement_data$SampleID))

## Add correct sample identifiers
lipidomics_measurement_data$SampleID <- lumc_sample_matching$Sample.name 
lipidomics_measurement_data_TA$SampleID <- lumc_sample_matching$Sample.name 
sample_metadata$SampleID <- lumc_sample_matching$Sample.name 

## Merge with LUMC excel table
lumc_sample_matching$SampleID <- lumc_sample_matching$Sample.name
sample_metadata <- merge(sample_metadata, lumc_sample_matching, by = 'SampleID')
```


## Read ISA file
```{r}
isa_sample <- read.xlsx("C:/Users/Casper.deVisser/Documents/ipsc-data-analysis/20240514_Xomics iPSC Neurons_new_READ-ONLY.xlsx", sheet = 'Assay - lipidomics_UT')

# Map Sample namings, without biological replicate numbers
UniSampleIDs <- lapply(isa_sample$assay.identifier, function(x) {
    name <- strsplit(x, split = '_Lipidomics')[[1]][[1]]
})

## Add the Sample IDs with bio replicate
SampleID_withrep <- lapply(seq(1, 44), function(x) {
    name <- isa_sample[x, 6]
    num <- isa_sample[x, 8]
    name <- paste0(name, '_', num)
    name
})
sample_matching <- data.frame(isa_sample$sample.identifier, 
                              as.character(UniSampleIDs),
                              as.character(SampleID_withrep))
```


## Check if Sample IDs now match
```{r}
GroupName_withrep <- lapply(seq(1, 44), function(x) {
    name <- sample_metadata[x, 4]
    num <- sample_metadata[x, 7]
    name <- paste0(name, '_', num)
    name
})
sample_metadata$SampleID_withrep <- as.character(GroupName_withrep)

match(sample_metadata$SampleID_withrep, sample_matching$as.character.SampleID_withrep.)
```


### Measurement data
```{r}
head(lipidomics_measurement_data[, 1:5])
dim(lipidomics_measurement_data)
```


### Sample metadata
```{r}
#sample_metadata <- cbind(sample_metadata, isa_sample) TODO: Combine with Assay file?
head(sample_metadata)
```


### Filiter lipid features with more than 20% missingness
```{r}
source("support_functions/filter_impute.R")

### Remove columns with 20% missingngess
lipidomics_measurement_data_filtered <- remove_column_missingness(lipidomics_measurement_data)
lipidomics_measurement_data_TA_filtered <- remove_column_missingness(lipidomics_measurement_data_TA)
```

### Heatmap NA values after filtering
```{r}
source("support_functions/heatmap_missingness.R")

heatmap_missing(lipidomics_measurement_data, 'Lipidomics Targeted (LUMC)')
heatmap_missing(lipidomics_measurement_data_filtered, 'Lipidomics Targeted (LUMC) filtered')
heatmap_missing(lipidomics_measurement_data_TA_filtered, 'Lipidomics Targeted (LUMC) filtered')
```

## Impute missing values by half of the minimum
```{r}
## Protein content normalized data
lipidomics_measurement_data_imputed <- lipidomics_measurement_data_filtered
rownames(lipidomics_measurement_data_imputed) <- lipidomics_measurement_data_imputed$SampleID
lipidomics_measurement_data_imputed$SampleID <- NULL
lipidomics_measurement_data_imputed <- lipidomics_measurement_data_imputed %>% mutate_if(is.numeric, ~replace(., is.na(.), min(., na.rm = TRUE)/2))

## Total are normalized data
lipidomics_measurement_data_TA_imputed <- lipidomics_measurement_data_TA_filtered
rownames(lipidomics_measurement_data_TA_imputed) <- lipidomics_measurement_data_TA_imputed$SampleID
lipidomics_measurement_data_TA_imputed$SampleID <- NULL
lipidomics_measurement_data_TA_imputed <- lipidomics_measurement_data_TA_imputed %>% mutate_if(is.numeric, ~replace(., is.na(.), min(., na.rm = TRUE)/2))
```


## Log transformation
```{r}
lipidomics_measurement_data_Log <- log2(lipidomics_measurement_data_imputed)
lipidomics_measurement_data_TA_Log <- log2(lipidomics_measurement_data_TA_imputed)
```


## Plot distributions
```{r}
source("support_functions/PlotDistributions2.R")

PlotDistributions(lipidomics_measurement_data, type = 'hist', main_title = 'With missing values' )
PlotDistributions(lipidomics_measurement_data_imputed, type = 'hist', main_title = 'Imputed with half of minimum' )
PlotDistributions(lipidomics_measurement_data_Log, type = 'hist', main_title = 'Log-transformed' )
PlotDistributions(lipidomics_measurement_data_TA_Log, type = 'hist', main_title = 'TA Log-transformed' )

```

### Boxplots per sample
```{r}
source('support_functions/BoxplotSamples.R')

lipidomics_measurement_data_Log$Sample_ID <- rownames(lipidomics_measurement_data_Log)
boxplot_samples(lipidomics_measurement_data_Log, 'Lipidomics (Targeted)')
lipidomics_measurement_data_Log$Sample_ID <- NULL

lipidomics_measurement_data_TA_Log$Sample_ID <- rownames(lipidomics_measurement_data_TA_Log)
boxplot_samples(lipidomics_measurement_data_TA_Log, 'Lipidomics (Targeted)')
lipidomics_measurement_data_TA_Log$Sample_ID <- NULL
```

## Create feature metadata table with classes
```{r}
feature_metadata <- data.frame(colnames(lipidomics_measurement_data_Log))
feature_metadata$Class <- sub("\\..*", "",  names(lipidomics_measurement_data_Log))
colnames(feature_metadata) <- c('Feature', 'Class')
```



## Save data and metadata to .CSV files
```{r}
write.csv(lipidomics_measurement_data_Log, file =  file.path(params$outdir, 'lipidomics_T_imputed_log_data.csv'), row.names = T)
write.csv(lipidomics_measurement_data_imputed, file =  file.path(params$outdir, 'lipidomics_T_imputed_data.csv'), row.names = T)
write.csv(lipidomics_measurement_data_TA_imputed, file =  file.path(params$outdir, 'lipidomics_T_TA_imputed_data.csv'), row.names = T)

write.csv(sample_metadata, file =  file.path(params$outdir, 'lipidomics_T_sample_metadata.csv'), row.names = F)
write.csv(feature_metadata, file =  file.path(params$outdir, 'lipidomics_T_feature_metadata.csv'))
```


```{r}
sessionInfo()   
```
