---
title: "Lipidomics Untargeted with Unknown lipids"
output: html_document
params:
  lipidomics_data_pos_unfiltered: Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/New information requested/Untargeted met+lip/20231123_134746_lipidomics_pos_CURATED.txt
  lipidomics_data_neg_unfiltered: Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/New information requested/Untargeted met+lip/20231123_135559_lipidomics_neg_CURATED.txt
  metadata: 'Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/New information requested/Untargeted met+lip/KB_2023_029_ALain Gool_IPSC cells_sample list lipid.xlsx'
  bca_results: "Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/New information requested/Untargeted met+lip/KB_2023_029_Alain Gool_IPSC cells_BCA results.xlsx"
  sample_matching: "Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Sample list.xlsx"
 
  outdir: Z:/data/iPSC_data/iPSC_MAE
---

# Read un-processed lipidomics (untargeted) data and filter for RSD and blank-sample ratio

##
```{r}
library(here)
library(sessioninfo)
library(openxlsx)
library(tidyr)
library(RColorBrewer)
library(grid)
library(ggplot2)
```

#### Script parameters
```{r}
print(Sys.Date())
print(params)
```

### Function for scaling rows
```{r}
scale_rows <- function(x){
  m <- apply(x, 1, mean, na.rm = TRUE)
  s <- apply(x, 1, stats::sd, na.rm = TRUE)
  
  return((x - m) / s)
}
```


### Read files
```{r}
lip_pos <- read.csv(params$lipidomics_data_pos_unfiltered, sep = '\t')
lip_neg <- read.csv(params$lipidomics_data_neg_unfiltered, sep = '\t')

# Add my_id column
lip_pos$my_id <- paste0('pos_', lip_pos$X)
lip_neg$my_id <- paste0('neg_', lip_neg$X)

## Feature metadata
feature_metadata_pos <- lip_pos[4:2358,c(1:35, 101)]
feature_metadata_neg <- lip_neg[4:701,c(1:35, 101)]
feature_metadata <- rbind(feature_metadata_neg, feature_metadata_pos)
rownames(feature_metadata) <- feature_metadata$my_id
# Change column names
colnames(feature_metadata) <- feature_metadata[1,]
feature_metadata <- feature_metadata[2:3053,]
names(feature_metadata)[names(feature_metadata) == "neg_Alignment ID"] <- 'my_id'

# Select only measurement data
lip_pos <- lip_pos[4:2358,36:101]
lip_neg <- lip_neg[4:701,36:101]


# Merge pos and neg
all_data <- rbind(lip_neg,lip_pos)

# Change column names
colnames(all_data) <- all_data[1,]
all_data <- all_data[2:3053,]
names(all_data)[names(all_data) == "neg_Alignment ID"] <- 'my_id'
```

### Read metadata
```{r}
# extract meta data
meta_data <- read.xlsx(xlsxFile = params$metadata,
                       sheet = "Sheet4")
```


### Transform the data in long format.
```{r}
# extract just data
all_data_long <- all_data |> 
  pivot_longer(cols = -c(my_id),
               names_to = "sampleId",
               values_to = "peakArea")
#all_data_long$LipidClass <- as.factor(all_data_long$LipidClass)
all_data_long$peakArea <- as.numeric(all_data_long$peakArea)

```

## BCA Results
```{r}
bca1 <- read.xlsx(xlsxFile = params$bca_results, 
                  sheet = 2, 
                  cols = 1:6, 
                  rows = 23:41)


bca2 <- read.xlsx(xlsxFile = params$bca_results,
                  sheet = 3,
                  cols = 1:6,
                  rows = 23:41)

bca <- rbind.data.frame(bca1, bca2)
colnames(bca) <- c("kevinNr", "absAvg1", "absAvg2", "conc1", "conc2", "avgProteinContent")
```


### Merge with metadata
```{r}
meta_data <- merge(
  x = meta_data,
  y = bca[, c("kevinNr", "avgProteinContent")],
  by.x = "kevinNr",
  by.y = "kevinNr",
  all.x = TRUE
)
```

```{r}
all_data_long <- merge(
  x = all_data_long,
  y = meta_data,
  by.x = "sampleId",
  by.y = "SampleName",
  all.x = TRUE
)

all_data_long$SampleID <- NULL
```

```{r}
all_data_long$polarity <- as.factor(
  gsub(x = all_data_long$my_id,
       pattern = "(pos|neg).*",
       replacement = "\\1")
)
```

## QC

```{r}
# get qc  data
qc_data <- all_data_long[all_data_long$SampleType == "qcpool", ]
# get reference data
ref_data <- qc_data[qc_data$sampleId == "Qcpool_006", ]
```



```{r}
library(dplyr)
library(tidyr)
```


```{r}
# calculate the RSD's
rsd_data <- data.frame(rsd = tapply(qc_data, qc_data$my_id, function(x) {
  sd(x$peakArea, na.rm = TRUE) / mean(x$peakArea, na.rm = TRUE)
}))
rsd_data$my_id <- rownames(rsd_data)
rsd_data$polarity <- as.factor(
  gsub(x = rsd_data$my_id,
       pattern = "(pos|neg).*",
       replacement = "\\1")
)

rsd_data |> 
  ggplot(aes(x = rsd,
             fill = polarity)) +
  geom_histogram(binwidth = 0.005,
                 alpha = 0.5) +
  labs(x = "Relative standard deviation") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
## Remove features with RSD higher than 0.3
```{r}
rsd_data_highRSD <- rsd_data[rsd_data$rsd > 0.3,]
lipids_remove <- unique(rsd_data_highRSD$my_id) 

all_data_long <- all_data_long[!all_data_long$my_id %in% lipids_remove,]
```

```{r}
# get qc  data
qc_data <- all_data_long[all_data_long$SampleType == "qcpool", ]
# get reference data
ref_data <- qc_data[qc_data$sampleId == "Qcpool_006", ]
```


```{r}
# calculate the RSD's
rsd_data <- data.frame(rsd = tapply(qc_data, qc_data$my_id, function(x) {
  sd(x$peakArea, na.rm = TRUE) / mean(x$peakArea, na.rm = TRUE)
}))
rsd_data$my_id <- rownames(rsd_data)
rsd_data$polarity <- as.factor(
  gsub(x = rsd_data$my_id,
       pattern = "(pos|neg).*",
       replacement = "\\1")
)

rsd_data |> 
  ggplot(aes(x = rsd,
             fill = polarity)) +
  geom_histogram(binwidth = 0.005,
                 alpha = 0.5) +
  labs(x = "Relative standard deviation") +
  theme_minimal() +
  theme(legend.position = "bottom")
```


## Normalization

### Total are norm.
```{r}
all_data_long_totnorm <- all_data_long
all_data_long_totnorm <- all_data_long_totnorm[all_data_long_totnorm$SampleType == "sample", ]
#all_data_long_totnorm <- all_data_long_totnorm[!apply(all_data_long_totnorm, 1, function(row) all(is.na(row))),] 
tot_sum_data <- tapply(all_data_long_totnorm$peakArea, all_data_long_totnorm$sampleId, sum, simplify = FALSE, na.rm=TRUE)
tot_sum_data <- as.data.frame(do.call("rbind", tot_sum_data))
colnames(tot_sum_data)[1] <- "totalPeakArea"
tot_sum_data$sampleId <- rownames(tot_sum_data)

all_data_long_totnorm <- merge(
  x = all_data_long_totnorm,
  y = tot_sum_data,
  by = "sampleId"
)

all_data_long_totnorm$peakAreaNorm <- all_data_long_totnorm$peakArea / all_data_long_totnorm$totalPeakArea
```

### Protein content norm.
```{r}
all_data_long_norm <- all_data_long
all_data_long_norm$peakAreaNorm <- all_data_long_norm$peakArea / all_data_long_norm$avgProteinContent
# all_data_long_norm <- all_data_long_norm[!(all_data_long_norm$kevinNr %in% c(9)), ]
```


## Prepare for export
```{r}
library(tidyverse)
# Transpose the meta data for export
t_meta <- meta_data |> 
  # remove blanks
  filter(SampleType != "blank") |> 
  # select the columns
  select(SampleName, SampleType, Group, InjOrder) |> 
  # sort by filename
  arrange(SampleName) |> 
  # everything needs to be character in order to transpose
  mutate(across(everything(), as.character)) |> 
  pivot_longer(-SampleName) |> 
  pivot_wider(names_from = SampleName,
              values_from = value) |> 
  mutate(LipidClass = NA,
         ShortLipidName = NA,
         LongLipidName = NA) |> 
  relocate(LipidClass, ShortLipidName, LongLipidName, .after = "name") |> 
  rename(LipidID = name)
```

```{r}
export_raw <- all_data_long |> 
  filter(SampleType != "blank") |> 
  pivot_wider(id_cols = c(my_id),
              names_from = sampleId,
              values_from = peakArea) |> 
  rename(LipidID = my_id) 


export_totnorm <- all_data_long_totnorm |>
  filter(SampleType == "sample") |>
  pivot_wider(id_cols = c(my_id),
              names_from = sampleId,
              values_from = peakAreaNorm) |> 
  rename(LipidID = my_id) 

export_norm <- all_data_long_norm |>
  filter(SampleType == "sample") |>
  pivot_wider(id_cols = c(my_id),
              names_from = sampleId,
              values_from = peakAreaNorm) |> 
  rename(LipidID = my_id) 
```



## Save data and metadata to .CSV files
```{r}
wb <- createWorkbook()
addWorksheet(wb, "Raw")
writeData(wb, "Raw", export_raw)

addWorksheet(wb, "Total area normalized")
writeData(wb, "Total area normalized", export_totnorm)

addWorksheet(wb, "Protein content normalized")
writeData(wb, "Protein content normalized", export_norm)

addWorksheet(wb, "Feature metadata")
writeData(wb, "Feature metadata", feature_metadata)

saveWorkbook(wb, file =  "Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Lipidomics_results_unfiltered.xlsx", overwrite = T)
```


```{r}
sessionInfo()   
```


