---
title: "Lipidomics Untargeted 2 Tabular"
output: html_document
params:
  lipidomics_data: Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Untargeted_lipidomics/Results/Lipidomics_results.xlsx
  isa_files: Z:/data/iPSC_data/Xomics iPSC Neurons_new.xlsx
  lumc_samples: Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Samples_match_ISA.xlsx
  sample_matching: "Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Sample list.xlsx"
 
  outdir: Z:/data/iPSC_data/iPSC_MAE
---

## Read processed lipidomics (untargeted) data and save to tabular formats for the MultiAssayExperiment object

#### Script parameters
```{r}
print(Sys.Date())
print(params)
```


### Libraries
```{r}
library(openxlsx)
library(dplyr)
```

## Read Excel file
```{r}
lipidomics_excel <- openxlsx::read.xlsx(params$lipidomics_data, rowNames = TRUE,   sheet = 'Total area normalized')
```


### Feature metadata
```{r}
## feature metadata
feature_metadata <- lipidomics_excel[4:264, 1:3]
head(feature_metadata)
```

### Sample metadata #TODO
```{r}
sample_metadata <- lipidomics_excel[1:3,4:47]
sample_metadata <- data.frame(t(sample_metadata))
head(sample_metadata)
dim(sample_metadata)
```



### Measurement data
```{r}
lipidomics_data <- lipidomics_excel[4:264,4:47]
lipidomics_data <- data.frame(t(lipidomics_data))
lipidomics_data[] <- lapply(lipidomics_data, as.numeric)
head(lipidomics_data[, 1:5])
```

### Visualize missingness in data
```{r}
source("support_functions/heatmap_missingness.R")
lipidomics_data_heatmap <- lipidomics_data
lipidomics_data_heatmap[lipidomics_data_heatmap==0] <- NA
heatmap_missing(lipidomics_data, 'Lipidomics Untargeted (LUMC)')
```



## Sample metadata (ISA file)
```{r}
# library(readxl)
# isa_sample <- read.xlsx("C:/Users/Casper.deVisser/Documents/ipsc-data-analysis/20240514_Xomics iPSC Neurons_new_READ-ONLY.xlsx", sheet = 'Assay - lipidomics_UT')
```

## Add column for sample names as written in ISA file using the LUMC Sample Sheet
```{r}
lumc_samples <- read.xlsx(params$lumc_samples)

lumc_sheet_order <- lapply(rownames(lipidomics_data), function(x) {
  num <- strsplit(x, "_")[[1]][[3]]
  num
  
})
lipidomics_data$sampleOrder <- as.numeric(as.character(lumc_sheet_order))
lipidomics_data <- lipidomics_data[order(lipidomics_data$sampleOrder, decreasing=F),]
lipidomics_data$sampleOrder <- NULL
```


## Sample metadata in same order
```{r}
sample_metadata <- sample_metadata[rownames(lipidomics_data),]
```


## Remove last digit code in sample namings to match ISA files
```{r}
final_sample_names <- lapply(rownames(lipidomics_data), function(x) {
  end_char <- nchar(x) - 4
  x <- substr(x, 1, end_char)
  x
})
## Add to measurement data and sample metadata
rownames(lipidomics_data) <- as.character(final_sample_names)
#rownames(sample_metadata) <- as.character(final_sample_names)
```


## Replace zero's by half of minimum
```{r}
lipidomics_data_noZero <- lipidomics_data %>% mutate(across(everything(), ~if_else(. == 0, min(.[. != 0])/2, .)))
```


## Log transformation
```{r}
epsilon <- 0.01 * min(lipidomics_data_noZero[lipidomics_data_noZero> 0], na.rm = T)
lipidomics_data_Log <- log10(lipidomics_data_noZero + epsilon)

#shifted_data <- lipidomics_data_Log - min(lipidomics_data_Log, na.rm=T) + 1e-6
```


## Plot distributions
```{r}
source("support_functions/PlotDistributions2.R")

PlotDistributions(lipidomics_data, type = 'hist', main_title = 'Lipidomics untargeted' )
PlotDistributions(lipidomics_data_noZero, type = 'hist', main_title = 'Zero values replaced by half of minimum' )
PlotDistributions(lipidomics_data_Log, type = 'hist', main_title = 'Log-transformed' )
PlotDistributions(shifted_data, type = 'hist', main_title = 'Log-transformed Shifted' )
```


## Ensure feature metadata has same features
```{r}
feature_metadata <- feature_metadata[colnames(lipidomics_data_Log),]
feature_metadata <- na.omit(feature_metadata)

lipidomics_data_Log <- lipidomics_data_Log[, rownames(feature_metadata)]
```


## Save data and metadata to .CSV files
```{r}
write.csv(lipidomics_data_Log, file =  file.path(params$outdir, 'lipidomics_UT_data_filtered.csv'))

write.csv(sample_metadata, file =  file.path(params$outdir, 'lipidomics_UT_sample_metadata_filtered.csv'))
write.csv(feature_metadata, file =  file.path(params$outdir, 'lipidomics_UT_feature_metadata_filtered.csv'))
```


```{r}
sessionInfo()   
```


