---
title: "Metablomics_LUMC_untargeted2Tabular"
output: html_document
params:
  metabolomics_raw_pos: Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/New information requested/Untargeted met+lip/20231128_095100_metabolomics_pos.txt
  metabolomics_raw_neg: Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/New information requested/Untargeted met+lip/20231128_095736_metabolomics_neg.txt
  metadata1: "Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/New information requested/Untargeted met+lip/KB_2023_029_Alain Gool_IPSC cells_BCA results.xlsx"
  metadata2: "Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/New information requested/Untargeted met+lip/KB_2023_029_ALain Gool_IPSC cells_sample list _metab.xlsx"
  lumc_samples: Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Samples_match_ISA.xlsx
  outdir: Z:/data/iPSC_data/iPSC_MAE

---

## Read processed untargeted metabolomics (LUMC) data and save to tabular formats for the MultiAssayExperiment object

#### Script parameters
```{r}
print(Sys.Date())
print(params)
```

### Libraries
```{r}
library(here)
library(sessioninfo)
library(tidyverse)
library(Rfast)
library(openxlsx)
#library(ComplexHeatmap)
source("support_functions/LUMC_mtblmcs_functions.R")
```


```{r}
work_dir <- here()

#### folders
meta_dir <- file.path(work_dir, "Meta")
result_dir <- file.path(work_dir, "Results")

#### files
# Define the file names of the MS-DIAL result files, first positive mode, 
# next negative mode.
msdial_files <- c(params$metabolomics_raw_pos, 
                  params$metabolomics_raw_neg)

# Name of the meta data file.
meta_file <- "KB_2023_029_ALain Gool_IPSC cells_sample list _metab.xlsx"

#### variables / parameters
# rsd limit
rsd_limit <- 0.3
# sample / blank ratio limit
sample_blank_limit <- 5
# sample threshold
sample_threshold_limit <- 0.8
# group threshold
group_threshold_limit <- 0.8
```



```{r}
scale_rows <- function(x){
  m <- apply(x, 1, mean, na.rm = TRUE)
  s <- apply(x, 1, stats::sd, na.rm = TRUE)
  
  return((x - m) / s)
}
```


```{r}
raw_data <- tibble(polarity = c("pos", "neg"),
                   filename = file.path(msdial_files),
                   raw_data = map(.x = filename,
                                  .f = ~ read_msdial(filename = .x)))
```


```{r}
clean_data_wide <- clean_up(met_data = raw_data)
```


```{r}
clean_data_wide |> 
  select(-MSMSspectrum)
```

## Replace 'Unknown' by new feature_id
```{r}
replace_unknowns <- function(data, target_col, source_col) {
  data[[target_col]][data[[target_col]] == "Unknown"] <- paste0('Unknown_', data[[source_col]][data[[target_col]] == 'Unknown'])
  return(data)
}

clean_data_wide <- replace_unknowns(clean_data_wide, 'MetaboliteName', 'my_id')
```

```{r}
clean_data_long <- tidy_data(met_data = clean_data_wide)
```

```{r}
cur_pos <- read.table(
  quote = "",
  sep = "\n",
  col.names = "comments",
  text = "- PYROCATECHOL (329), not a great match
- CYTOSINE (335), not a great match
- 3-HYDROXYBENZOATE (426), not a great match
- N-ACETYLPUTRESCINE (553), not a great match
- HYPOXANTHINE (690), not a great match
- N-ACETYLASPARTATE (942), not a great match
- INDOLE-3-ETHANOL (990), not a great match
- AMINOADIPATE (989), not a great match
- 5-HYDROXYLYSINE (1024), not a great match
- 4-PYRIDOXATE (1053), not a great match
- HOMOGENTISATE (975), not a great match
- N-ACETYLASPARAGINE (1182), not a great match
- TRYPTAMINE (1307), not a great match
- 4-PYRIDOXATE (1311), not a great match
- 3,4 DIHYDROXYMANDELATE (1331), not a great match
- FERULATE (1499), not a great match
- OMEGA-HYDROXYDODECANOATE (1562), not a great match
- N-ACETYLPHENYLALANINE (1704), not a great match
- 2-ACETAMIDO-2-DEOXY-BETA-D-GLUCOSYLAMINE (1899), not a great match
- CYSTATHIONINE (1933), not a great match
- URIDINE (2294), not a great match
- 5'-DEOXYADENOSINE (2414), not a great match
- MELATONIN (2676), not a great match
- OPHTHALMATE (2938), not a great match
- INOSINE (3162), not a great match
- ADENOSINE 2',3'-CYCLIC PHOSPHATE (3400), not a great match
- DEOXYCYTIDINE MONOPHOSPHATE (3571), not a great match
- CORTISOL (3736), not a great match
- RIBOFLAVIN (3843), not a great match
- MELIBIOSE (3868), not a great match
- CORTISONE (4002), not a great match
- CORTISONE (4001), not a great match"
)

pattern <- "^[ ]{0,2}- ([-:/ ',A-Z0-9wo\\(\\)]*) \\(([0-9]{3,4})\\), (no MSMS|not a great match)$"

cur_pos$MetaboliteName <- gsub(x = cur_pos$comments,
                               pattern = pattern,
                               replacement = "\\1")
cur_pos$my_id <- paste0("pos_", gsub(x = cur_pos$comments,
                                     pattern = pattern,
                                     replacement = "\\2"))
cur_pos$reason <- gsub(x = cur_pos$comments,
                       pattern = pattern,
                       replacement = "\\3")
cur_pos$comments <- NULL
```


```{r}
cur_neg <- read.table(
  quote = "",
  sep = "\n",
  col.names = "comments",
  text = "- ETHYLMALONATE (1427), not a great match
  - RIBOFLAVIN (7321), not a great match
  - w/o MS2: N-FORMYLGLYCINE (833), not a great match
  - w/o MS2: MALONATE (863), not a great match
  - w/o MS2: MALEATE (1128), not a great match
  - N-ACETYLGLYCINE (1148), not a great match
  - N-ACETYLGLYCINE (1147), not a great match
  - SUCCINATE (1190), not a great match
  - THREONINE (1208), not a great match
  - 4-HYDROXYBENZALDEHYDE (1257), not a great match
  - OXOPROLINE (1368), not a great match
  - ASPARTATE (1445), not a great match
  - ADENINE (1481), not a great match
  - ADENINE (1475), not a great match
  - N-ACETYLSERINE (1670), not a great match
  - N-ACETYLSERINE (1667), not a great match
  - 2',4'-DIHYDROXYACETOPHENONE (1748), not a great match
  - w/o MS2: MANDELATE (1749), not a great match
  - w/o MS2: 3-SULFINOALANINE (1754), not a great match
  - w/o MS2: N-ACETYLASPARAGINE (1795), not a great match
  - w/o MS2: 3-(2-HYDROXYPHENYL)PROPANOATE (1992), not a great match
  - w/o MS2: THIOPURINE S-METHYLETHER (1978), not a great match
  - w/o MS2: THIOPURINE S-METHYLETHER (1977), not a great match
  - w/o MS2: URATE (2013), not a great match
  - w/o MS2: 3-DEHYDROSHIKIMATE (2064), not a great match
  - w/o MS2: L-ARGININE (2120), not a great match
  - w/o MS2: N-ACETYLASPARAGINE (2113), not a great match
  - w/o MS2: TRANS-ACONITATE (2109), not a great match
  - w/o MS2: TRANS-ACONITATE (2108), not a great match
  - w/o MS2: SUBERATE (2114), not a great match
  - w/o MS2: SUBERATE (2115), not a great match
  - w/o MS2: N-ACETYLASPARTATE (2134), not a great match
  - w/o MS2: CAFFEATE (2209), not a great match
  - HYDROXYPHENYLLACTATE (2234), not a great match
  - AZELATE (2332), not a great match
  - AZELATE (2330), not a great match
  - N-ACETYLGLUTAMATE (2352), not a great match
  - N-ACETYLGLUTAMATE (2351), not a great match
  - QUINATE (2409), not a great match
  - CITRATE (2405), not a great match
  - FERULATE (2428), not a great match
  - L-DOPA (2489), not a great match
  - N-ACETYLPHENYLALANINE (2726), not a great match
  - D-SEDOHEPTULOSE (2787), not a great match
  - SORBOSE (2909), not a great match
  - CYSTATHIONINE (3053), not a great match
  - D-RIBOSE 5-PHOSPHATE (3234), not a great match
  - CYSTINE (3461), not a great match
  - THYMIDINE (3521), not a great match
  - CYTIDINE (3549), not a great match
  - URIDINE (3581), not a great match
  - BIOTIN (3582), not a great match
  - GLUCOSAMINE 6-PHOSPHATE (3942), not a great match
  - N-ACETYLSEROTONIN (4045), not a great match
  - DEOXYGUANOSINE (4123), not a great match
  - DEOXYCYTIDINE (4436), not a great match
  - MELATONIN (4573), not a great match
  - THYMIDINE (4799), not a great match
  - CYTIDINE (4822), not a great match
  - URIDINE (4848), not a great match
  - EICOSAPENTAENOATE (5138), not a great match
  - DEOXYURIDINE-MONOPHOSPHATE (5254), not a great match
  - URIDINE MONOPHOSPHATE (5621), not a great match
  - ADENOSINE 2',3'-CYCLIC PHOSPHATE (5749), not a great match
  - AICAR (5937), not a great match
  - INOSINE-MONOPHOSPHATE (6166), not a great match
  - CORTISOL (6435), not a great match
  - LITHOCHOLATE (6710), not a great match
  - S-ADENOSYLHOMOCYSTEINE (6845), not a great match
  - DEOXYCHOLATE (6984), not a great match
  - DEOXYCHOLATE (6985), not a great match
  - CORTISONE (7051), not a great match
  - CORTISONE (7234), not a great match
  - CORTISOL 21-ACETATE (7733), not a great match
  - CHOLATE (7951), not a great match
  - CITICOLINE (8499), not a great match
  - RAFFINOSE (8706), not a great match
  - RAFFINOSE (9125), not a great match"
)

pattern <- "^[ ]{0,2}- ([-:/ ',A-Z0-9wo\\(\\)]*) \\(([0-9]{3,4})\\), (no MSMS|not a great match)$"

cur_neg$MetaboliteName <- gsub(x = cur_neg$comments,
                               pattern = pattern,
                               replacement = "\\1")
cur_neg$my_id <- paste0("neg_", gsub(x = cur_neg$comments,
                                     pattern = pattern,
                                     replacement = "\\2"))
cur_neg$reason <- gsub(x = cur_neg$comments,
                       pattern = pattern,
                       replacement = "\\3")
cur_neg$comments <- NULL
```


```{r}
# remove the bad matches
cur <- rbind(
  cur_pos,
  cur_neg
)

clean_data_long <- clean_data_long[!(clean_data_long$my_id %in% cur$my_id), ]
```


```{r}
clean_data_long <- clean_data_long[!is.na(clean_data_long$MSMSspectrum), ]
```


```{r}
clean_data_long <- clean_data_long |> 
  group_by(MetaboliteName, polarity, SampleName, SampleType) |> 
  summarise(PeakArea = sum(PeakArea),
            .groups = "drop") |> 
  # add ID's
  group_by(polarity, SampleName) |> 
  mutate(my_id = 1:n(),
         my_id = paste0(polarity, "_", my_id)) |> 
  ungroup()
```


```{r}
meta_data <- read.xlsx(params$metadata2, 
                       sheet = "Sheet4") |> 
  mutate(SampleType = as.factor(SampleType),
         Sequence = as.factor(Sequence),
         Group = as.factor(Group),
         kevinNr = as.numeric(kevinNr))



# If there are multiple sequences some extra things need to be calculated. Determine
# here if this needed.
incl_seq <- ifelse(length(unique(meta_data$Sequence)) > 1,
                   TRUE,
                   FALSE)
```


```{r}
bca1 <- read.xlsx(xlsxFile = params$metadata1, 
                  sheet = 2, 
                  cols = 1:6, 
                  rows = 23:41)


bca2 <- read.xlsx(xlsxFile = params$metadata1,
                  sheet = 3,
                  cols = 1:6,
                  rows = 23:41)

bca <- rbind.data.frame(bca1, bca2)
colnames(bca) <- c("kevinNr", "absAvg1", "absAvg2", "conc1", "conc2", "avgProteinContent")
```

```{r}
meta_data <- merge(
  x = meta_data,
  y = bca[, c("kevinNr", "avgProteinContent")],
  by.x = "kevinNr",
  by.y = "kevinNr",
  all.x = TRUE
)
```



```{r}
clean_data <- clean_data_long |> 
  left_join(y = meta_data,
            by = c("SampleName" = "SampleName"),
            suffix = c("", ".y"))
```


```{r}
blank_data <- clean_data |> 
  filter(SampleType == "blank") |> 
  select(my_id, polarity, MetaboliteName, Sequence, PeakArea) |> 
  group_by(my_id, Sequence) |> 
  summarise(AvgBlank = mean(PeakArea, na.rm = TRUE),
            .groups = "drop")

# calculate the ratio sample / blank
ratio_data <- clean_data |> 
  filter(SampleType == "sample") |> 
  left_join(y = blank_data,
            by = c("my_id" = "my_id",
                   "Sequence" = "Sequence")) |> 
  # determine the sample / blank ratio
  mutate(sample_blank_ratio = PeakArea / AvgBlank,
         sample_blank_keep = ifelse(sample_blank_ratio < sample_blank_limit,
                                    FALSE,
                                    TRUE))
sample_data_threshold <- ratio_data |> 
  group_by(my_id) |> 
  summarise(sample_threshold = mean(sample_blank_keep),
            .groups = "drop") |> 
  filter(sample_threshold >= sample_threshold_limit) |> 
  pull(my_id)

# group_data_threshold <- ratio_data |> 
#   group_by(my_id, Group) |> 
#   summarise(group_threshold = mean(sample_blank_keep),
#             .groups = "drop") |> 
#   group_by(my_id) |> 
#   summarise(group_threshold_max = max(group_threshold, na.rm = TRUE),
#             .groups = "drop") |> 
#   filter(group_threshold_max >= group_threshold_limit) |> 
#   pull(my_id)
# 
# keep_met_sampleblank <- base::union(sample_data_threshold, group_data_threshold)
keep_met_sampleblank <- sample_data_threshold
```



```{r}
set.seed(123)
rnd_metabolites <- sample(x = unique(clean_data$my_id),
                          size = 10)
```


## Find possible trend in QC samples and filter if necessary

```{r}
trend_data <- clean_data |> 
  filter(SampleType == "qcpool",
         my_id %in% rnd_metabolites)
```


```{r, fig.height=10}
plot_trend(data = trend_data)
```


```{r}
reference_data <- clean_data |> 
  filter(SampleName == "Qcpool_006") |> 
  select(my_id, Sequence, PeakArea) |> 
  rename(Reference = PeakArea)

trend_all_data <- clean_data |> 
  filter(SampleType == "qcpool") |> 
  left_join(y = reference_data,
            by = c("my_id" = "my_id",
                   "Sequence" = "Sequence")) |> 
  mutate(log2fc = log2(PeakArea / Reference))

plot_all_trends(data = trend_all_data)
```


## Remove samples with infinite log-fold change
```{r}
trend_all_data_noInfite <- trend_all_data[trend_all_data$log2fc != -Inf ,] # Removes 697 rows
trend_all_data_noInfite <- trend_all_data_noInfite[trend_all_data_noInfite$log2fc != Inf ,] # Removes 339 rows
plot_all_trends(data = trend_all_data_noInfite)
```
No trend correction needed


## Remove features with infinite values in first trend plot
```{r}
metabolite_keep <- unique(trend_all_data_noInfite$MetaboliteName)
clean_data <- clean_data[clean_data$MetaboliteName %in% metabolite_keep, ]
```


```{r}
# get the qc data for rsd calculations
qc_data <- clean_data |> 
  filter(SampleType == "qcpool")
```


```{r}
rsd_data_overall <- qc_data |> 
  group_by(polarity, MetaboliteName, my_id) |> 
  summarise(rsd = sd(PeakArea, na.rm = TRUE) / mean(PeakArea, na.rm = TRUE),
            .groups = "drop")

keep_met_rsd <- rsd_data_overall$my_id[rsd_data_overall$rsd <= rsd_limit]

rsd_data_seq <- qc_data |> 
  group_by(polarity, Sequence, MetaboliteName, my_id) |> 
  summarise(rsd = sd(PeakArea, na.rm = TRUE) / mean(PeakArea, na.rm = TRUE),
            .groups = "drop")
```


```{r}
rsd_plot(data = rsd_data_overall,
         type = "overall",
         rsd_limit = rsd_limit)
```



```{r}
# keep_metabolites <- rsd_data_overall_cor |> 
keep_metabolites <- rsd_data_overall |> 
  group_by(MetaboliteName) |> 
  mutate(keep_metabolite = if_else(rsd == min(rsd), TRUE, FALSE)) |> 
  filter(keep_metabolite == TRUE) |> 
  pull(my_id)

# clean_data_cor <- clean_data_cor |> 
#   filter(my_id %in% keep_metabolites)

# also remove from the raw data
clean_data <- clean_data |> 
  filter(my_id %in% keep_metabolites)
```


```{r}
keep_metabolites <- base::intersect(
  keep_met_rsd,
  keep_met_sampleblank
)
```


```{r}
# cleaned_data <- clean_data_cor |> 
cleaned_data <- clean_data |> 
  filter(my_id %in% keep_metabolites) |> 
  select(my_id, polarity, MetaboliteName, SampleName, everything()) |> 
  arrange(SampleName)

cleaned_data_raw <- clean_data |> 
  filter(my_id %in% keep_metabolites) |> 
  select(my_id, polarity, MetaboliteName, SampleName, everything()) |> 
  arrange(SampleName)
```

```{r}
# get the qc data for rsd calculations
qc_data_cleaned <- cleaned_data |> 
  filter(SampleType == "qcpool")
```

```{r}
rsd_data_overall_cleaned <- qc_data_cleaned |> 
  group_by(polarity, MetaboliteName, my_id) |> 
  # summarise(rsd = sd(PeakAreaCor, na.rm = TRUE) / mean(PeakAreaCor, na.rm = TRUE),
  summarise(rsd = sd(PeakArea, na.rm = TRUE) / mean(PeakArea, na.rm = TRUE),
            .groups = "drop")
```


```{r}
cleaned_data_totnorm <- cleaned_data
cleaned_data_totnorm <- cleaned_data_totnorm[cleaned_data_totnorm$SampleType == "sample", ]
tot_sum_data <- tapply(cleaned_data_totnorm$PeakArea, cleaned_data_totnorm$SampleName, sum, simplify = FALSE)
tot_sum_data <- as.data.frame(do.call("rbind", tot_sum_data))
colnames(tot_sum_data)[1] <- "totalPeakArea"
tot_sum_data$SampleName <- rownames(tot_sum_data)

cleaned_data_totnorm <- merge(
  x = cleaned_data_totnorm,
  y = tot_sum_data,
  by = "SampleName"
)

cleaned_data_totnorm$peakAreaNorm <- cleaned_data_totnorm$PeakArea / cleaned_data_totnorm$totalPeakArea
```


```{r}
cleaned_data_norm <- cleaned_data
cleaned_data_norm$peakAreaNorm <- cleaned_data_norm$PeakArea / cleaned_data_norm$avgProteinContent
```


```{r}
# Transpose the meta data for export
t_meta <- meta_data |> 
  # remove blanks
  filter(SampleType != "blank") |> 
  # select the columns
  select(SampleName, SampleType, Group, InjOrder) |> 
  # sort by filename
  arrange(SampleName) |> 
  # everything needs to be character in order to transpose
  mutate(across(everything(), as.character)) |> 
  pivot_longer(-SampleName) |> 
  pivot_wider(names_from = SampleName,
              values_from = value) |> 
  mutate(MetaboliteName = NA) |> 
  relocate(MetaboliteName, .after = "name") |> 
  rename(MetID = name)
```


```{r}
export_raw <- cleaned_data_raw |> 
  filter(SampleType != "blank") |> 
  pivot_wider(id_cols = c(my_id, MetaboliteName),
              names_from = SampleName,
              values_from = PeakArea) |> 
  rename(MetID = my_id) |> 
  arrange(MetaboliteName)

export_raw <- rbind(t_meta, export_raw)

# export_cor <- cleaned_data |> 
#   filter(SampleType != "blank") |> 
#   pivot_wider(id_cols = c(my_id, MetaboliteName),
#               names_from = SampleName,
#               values_from = PeakAreaCor) |> 
#   rename(MetID = my_id) |> 
#   arrange(MetaboliteName)
# 
# export_cor <- rbind(t_meta, export_cor)

export_totnorm <- cleaned_data_totnorm |>
  filter(SampleType == "sample") |>
  pivot_wider(id_cols = c(my_id, MetaboliteName),
              names_from = SampleName,
              values_from = peakAreaNorm) |>
  rename(MetID = my_id)

export_totnorm <- rbind(t_meta[, !grepl(x = colnames(t_meta),
                                        pattern = "Qcpool")], 
                        export_totnorm)

export_norm <- cleaned_data_norm |>
  filter(SampleType == "sample") |>
  pivot_wider(id_cols = c(my_id, MetaboliteName),
              names_from = SampleName,
              values_from = peakAreaNorm) |>
  rename(MetID = my_id)

export_norm <- rbind(t_meta[, !grepl(x = colnames(t_meta),
                                     pattern = "Qcpool")], 
                     export_norm)
```

## Feature metadata export
```{r}
feature_metadata <- clean_data_wide[,1:14]
## Subset with metabolites in export
feature_metadata <- feature_metadata[feature_metadata$MetaboliteName %in% export_norm$MetaboliteName,]
## Put in the same order as measurement data
feature_metadata <- feature_metadata[match(export_norm$MetaboliteName, feature_metadata$MetaboliteName),]
## Add same my_id column
feature_metadata$my_id <- export_norm$MetID
```

```{r}
wb <- createWorkbook()
addWorksheet(wb, "Raw")
writeData(wb, "Raw", export_raw)

addWorksheet(wb, "Total area normalized")
writeData(wb, "Total area normalized", export_totnorm)

addWorksheet(wb, "Protein content normalized")
writeData(wb, "Protein content normalized", export_norm)

addWorksheet(wb, "Feature metadata")
writeData(wb, "Feature metadata", feature_metadata)

saveWorkbook(wb, file =  "Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Metabolomics_results_unfiltered.xlsx")
```


```{r}
sessionInfo()   
```