---
title: "Metabolomics  to tabular format"
output: html_document
params:
  
  metabomics_feature_neg_metadata: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231128_XCMS/XCMSR.annotated.Report_231128_ESIneg.tsv
  metabomics_feature_pos_metadata: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231120_XCMS/XCMSR.annotated.Report_231120_ESIpos.tsv
  metabolomics_pos_data: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231120_UDA/231120_ControlsvsPatients_2024-03-13_ESIpos.tsv
  metabolomics_neg_data: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231128_UDA/231128_ControlsvsPatients_2024-03-13_ESIneg.tsv
  
  outdir: Z:/data/iPSC_data/iPSC_MAE
---

## Read processed metabolomics data and save to tabular formats for the MultiAssayExperiment object

### Libraries
```{r}
library(fuzzyjoin)
library(dplyr)
```

#### Script parameters
```{r}
print(Sys.Date())
print(params)
```


#### Functions to read data
```{r}

## Function to read metabolomics data from TML
read_mtblcs_data_tml <- function(filepath, id_string) {
  
  ### Read data
  mtblmcs_data <- read.delim(filepath, header = FALSE)
  mtblmcs_data$V105 <- NULL
  
  ### Remove columns
  colnames(mtblmcs_data) <- mtblmcs_data[3,]
  rows <- nrow(mtblmcs_data)
  mtblmcs_data <- mtblmcs_data[4:rows,]
  
  mtblmcs_data$Mass <- NULL
  mtblmcs_data$R <- NULL
  
  ## Rename metabolomics features
  feature_names <- lapply(mtblmcs_data$F, function(x) {
    y <- paste0(id_string, x)
    y
  })
  rownames(mtblmcs_data) <- feature_names
  mtblmcs_data$F <- NULL
  
  return(mtblmcs_data)
}

### Save sample/filename mapping dataframe
create_sample_metadata_df <- function(filepath) {
  sample_mapping <- read.delim(filepath, header = FALSE)
  sample_mapping <- sample_mapping[1:3,4:104]
  sample_mapping <- data.frame(t(sample_mapping))
  sample_mapping$X2 <- NULL
  colnames(sample_mapping) <- c('File_name', 'Sample_name')
  return(sample_mapping)
}


### Store Mass and Retention time in other dataframe
create_feature_metadata_df  <- function(filepath) {
  data <- read.delim(filepath, header = FALSE)
  colnames(data) <- data[3,]
  rows <- nrow(data)
  data <- data[4:rows,]
  feature_metadata_df <- data[, 2:3]
  return(feature_metadata_df)
}  
```

## Read data for positive ionization mode
```{r}
## measurement data
metabolomics_pos <- read_mtblcs_data_tml(params$metabolomics_pos_data, 'TML_pos')

## sample metadata
sample_mapping <- create_sample_metadata_df(params$metabolomics_pos_data)

## feature metadata
metabolomics_features_pos_ids <- create_feature_metadata_df(params$metabolomics_pos_data)
```


## Read data for negative ionization mode
```{r}
## measurement data
metabolomics_neg <- read_mtblcs_data_tml(params$metabolomics_neg_data, 'TML_neg')

## feature metadata
metabolomics_features_neg_ids <- create_feature_metadata_df(params$metabolomics_neg_data)
```


#### Feature metadata (positive)
```{r }
metabolomics_features_pos <- read.delim(params$metabomics_feature_pos_metadata, row.names = 1)
columns_keep <- c("mzmed", "mzmin", "mzmax", "rtmed", 
                  "rtmin", "rtmax", "npeaks", "X1", "ms_level",
                  "isotopes", "adduct", "pcgroup")
metabolomics_features_pos <- metabolomics_features_pos[ , names(metabolomics_features_pos) %in% columns_keep]

## Add mass and retention time from UDA file
metabolomics_features_pos <- cbind(metabolomics_features_pos_ids, metabolomics_features_pos)

## Add feature IDs (temporarily)
rownames(metabolomics_features_pos) <- rownames(metabolomics_pos)
head(metabolomics_features_pos)
print(dim(metabolomics_features_pos))
```


### Filter features with too much missing values
```{r}
library(dplyr)
filter_na_columns <- function(data, threshold = 0.2) {
  
  # calculate NA values per column
  na_proportion <- colSums((data==0) / nrow(data))
  # Filter out columns about threshold
  data_out <- data %>% select(which(na_proportion <= threshold))
  return(data_out)
}

# Remove features
metabolomics_pos <- data.frame(t(metabolomics_pos))
metabolomics_filtered_pos <- filter_na_columns(metabolomics_pos)
dim(metabolomics_filtered_pos)


# Remove these features from feature metadata table
metabolomics_features_pos <- metabolomics_features_pos[colnames(metabolomics_filtered_pos),]
```
# Filter for metabolomics panel

### Read Metabolomics panel
```{r}
mtblmcs_panel <- read.csv('Z:/data/iPSC_data/iPSC_metabolomics_TML/2024-10-17_Metabolic_Package_NGMS_v0.8.csv')

## Adjust column names
mtblmcs_panel$RTpos <- mtblmcs_panel$RT._NMG
mtblmcs_panel$RTneg <- mtblmcs_panel$RT._NMG.1

## Filter RT -1 and -2
mtblmcs_panel <- mtblmcs_panel[mtblmcs_panel$RTpos != -2,]
mtblmcs_panel <- mtblmcs_panel[mtblmcs_panel$RTpos != -1,]

mtblmcs_panel <- mtblmcs_panel[mtblmcs_panel$RTneg != -2,]
mtblmcs_panel <- mtblmcs_panel[mtblmcs_panel$RTneg != -1,]
```


## Prepare metabolomics data for matching
```{r}
metabolomics_features_pos$RTpos <- as.numeric(metabolomics_features_pos$R)
metabolomics_features_pos$Mass <- as.numeric(metabolomics_features_pos$Mass)

#adduct shift
metabolomics_features_pos$Adduct_num <- sapply(strsplit(as.character(metabolomics_features_pos$adduct), "\\+ "), function(x) {
  if (length(x) < 2) return(0) else return(as.numeric(x[2]))
})
metabolomics_features_pos$Adduct_num <- as.numeric(metabolomics_features_pos$Adduct_num)

#metabolomics_features_pos$Mass <- metabolomics_features_pos$Mass + metabolomics_features_pos$Adduct_num
```


## Match RT and MZ
```{r}
# Function to calculate ppm tolerance
ppm_tolerance <- function(mz1, mz2) {
  if (is.na(mz1) || is.na(mz2)) return(NA)  # Handle NA values
  abs(mz1 - mz2) / mz2 * 1e6  # Return difference in ppm
}

# Function to match metabolomics data
match_metabolites <- function(data, reference, ppm_tol = 5.5, rt_tol = 0.10, mz_col = "Mass", rt_column = "RT") {
  matched_results <- data.frame(m_z = numeric(0), RT = numeric(0), match_m_z = numeric(0), match_RT = numeric(0), ppm_diff = numeric(0), rt_diff = numeric(0))
  
  for (i in 1:nrow(data)) {
    mz1 <- data[[mz_col]][i]
    rt1 <- data[[rt_column]][i]  
    
    for (j in 1:nrow(reference)) {
      mz2 <- reference[[mz_col]][j]
      rt2 <- reference[[rt_column]][j]
      
      # Calculate ppm difference and RT difference
      ppm_diff <- ppm_tolerance(mz1, mz2)
      rt_diff <- abs(rt1 - rt2)
      
      # Skip if ppm_diff or rt_diff is NA
      if (is.na(ppm_diff) || is.na(rt_diff)) next
      
      # Check if both m/z and RT fall within the specified tolerances
      if (ppm_diff <= ppm_tol && rt_diff <= rt_tol) {
      #if (rt_diff <= rt_tol) {
      #if (ppm_diff <= ppm_tol) {
        matched_results <- rbind(matched_results, data.frame(m_z = mz1, RT = rt1, match_m_z = mz2, match_RT = rt2, ppm_diff = ppm_diff, rt_diff = rt_diff))
      }
    }
  }
  
  return(matched_results)
}

match_pos <- match_metabolites(metabolomics_features_pos, mtblmcs_panel, mz_col = "Mass", rt_col = "RTpos")
```




# Negative ionization mode

### Filter features with too much missing values
```{r}
library(dplyr)
filter_na_columns <- function(data, threshold = 0.2) {
  
  # calculate NA values per column
  na_proportion <- colSums((data==0) / nrow(data))
  # Filter out columns about threshold
  data_out <- data %>% select(which(na_proportion <= threshold))
  return(data_out)
}

# Remove features
metabolomics_neg <- data.frame(t(metabolomics_neg))
metabolomics_filtered_neg <- filter_na_columns(metabolomics_neg)
dim(metabolomics_filtered_neg)
```

#### Feature metadata (negative)
```{r }
metabolomics_features_neg <- read.delim(params$metabomics_feature_neg_metadata, row.names = 1)
columns_keep <- c("mzmed", "mzmin", "mzmax", "rtmed", 
                  "rtmin", "rtmax", "npeaks", "X1", "ms_level",
                  "isotopes", "adduct", "pcgroup")
metabolomics_features_neg <- metabolomics_features_neg[ , names(metabolomics_features_neg) %in% columns_keep]

## Add mass and retention time from UDA file
metabolomics_features_neg <- cbind(metabolomics_features_neg_ids, metabolomics_features_neg)

## Add feature IDs (temporarily)
rownames(metabolomics_features_neg) <- colnames(metabolomics_neg)
head(metabolomics_features_neg)
print(dim(metabolomics_features_neg))

# Remove filtered features from feature metadata tablw
metabolomics_features_neg <- metabolomics_features_neg[colnames(metabolomics_filtered_neg),]
```


### Match with metabolomics panel, prepare feature table
```{r}
metabolomics_features_neg$RTneg <- as.numeric(metabolomics_features_neg$R)
metabolomics_features_neg$Mass <- as.numeric(metabolomics_features_neg$Mass)

#adduct shift
metabolomics_features_neg$Adduct_num <- sapply(strsplit(as.character(metabolomics_features_neg$adduct), "\\- "), function(x) {
  if (length(x) < 2) return(0) else return(as.numeric(x[2]))
})
metabolomics_features_neg$Adduct_num <- as.numeric(metabolomics_features_neg$Adduct_num)

metabolomics_features_neg$Mass <- metabolomics_features_neg$Mass - metabolomics_features_neg$Adduct_num
```


```{r}
match_neg <- match_metabolites(metabolomics_features_neg, mtblmcs_panel, mz_col = "Mass", rt_col = "RTneg")
```



```{r}
sessionInfo()   
```

