---
title: "Metablomics_LUMC_untargeted2Tabular"
output: html_document
params:
  metabolomics_data: Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Metabolomics_results_unfiltered.xlsx
  lumc_samples: Z:/data/iPSC_data/iPSC_metabolomics_lipidomics_LUMC/Samples_match_ISA.xlsx
  outdir: Z:/data/iPSC_data/iPSC_MAE

---

## Read processed untargeted metabolomics (LUMC) data and save to tabular formats for the MultiAssayExperiment object

#### Script parameters
```{r}
print(Sys.Date())
print(params)
```

### Libraries
```{r}
library(openxlsx)
```

### Read data
```{r}
# Total area normalized
metabolomics_TA <- openxlsx::read.xlsx(params$metabolomics_data, rowNames = TRUE, sheet = 'Total area normalized')

# Protein content normalized
metabolomics_PC <- openxlsx::read.xlsx(params$metabolomics_data, rowNames = TRUE, sheet = 'Protein content normalized')
```



### Feature metadata
```{r}
feature_metadata <- openxlsx::read.xlsx(params$metabolomics_data, rowNames = TRUE, sheet = 'Feature metadata')
feature_metadata <- feature_metadata[4:2325,]
```

### Sample metadata
```{r}
sample_metadata <- metabolomics_TA[1:3,2:45]
sample_metadata <- data.frame(t(sample_metadata))
head(sample_metadata)
dim(sample_metadata)
```


### Measurement data
```{r}
## Function to extract relevant measurement data
process_measurement_data <- function(data) {
  data <- data[4:2325, 2:45]
  data <- data.frame(t(data))
  data[] <- lapply(data, as.numeric)
  return(data)
}

metabolomics_TA_data <- process_measurement_data(metabolomics_TA)
metabolomics_PC_data <- process_measurement_data(metabolomics_PC)

dim(metabolomics_TA_data)
dim(metabolomics_PC_data)
```

### Visualize missingness in data
```{r}
source("support_functions/heatmap_missingness.R")
metabolomics_TA_data <- metabolomics_TA_data
metabolomics_TA_data[metabolomics_TA_data==0] <- NA
heatmap_missing(metabolomics_TA_data, 'Filtered Metabolomics (LUMC)')
```



## Add column for sample names as written in ISA file using the LUMC Sample Sheet
```{r}
lumc_samples <- read.xlsx(params$lumc_samples)

lumc_sheet_order <- lapply(rownames(metabolomics_TA_data), function(x) {
  num <- strsplit(x, "_")[[1]][[3]]
  num
  
})
## IMPORTANT!! THIS IS THE ORDER NEEDED TO MATHC THE SAMPLE NAMES
lumc_samples_order <- as.numeric(as.character(lumc_sheet_order))

sample_metadata_ordered <- sample_metadata[order(lumc_samples_order),]
```


## Match Sample IDs across sample metadata and LUMC excel sheet
```{r}
## Merge with LUMC excel table
lumc_samples$SampleID <- lumc_samples$Sample.name
sample_metadata_ordered$SampleID <- lumc_samples$Sample.name 
sample_metadata_ordered <- merge(sample_metadata_ordered, lumc_samples, by = 'SampleID')

GroupName_withrep <- lapply(seq(1, 44), function(x) {
    name <- sample_metadata_ordered[x, 10]
    num <- sample_metadata_ordered[x, 5]
    name <- paste0(name, '_', num)
    name
})

## Add Sample ID with bio replicate number
sample_metadata_ordered$SampleID_withrep <- as.character(GroupName_withrep)
```


## Add new Sample IDs to measurement data based on the orderd samples metadata table
```{r}
rownames(metabolomics_TA_data) <- sample_metadata_ordered$SampleID_withrep
rownames(metabolomics_PC_data) <- sample_metadata_ordered$SampleID_withrep
```

## Impute NA values with half of the minimum
```{r}
library(dplyr)
metabolomics_TA_data_imputed <- metabolomics_TA_data %>% mutate_if(is.numeric, ~replace(., is.na(.), min(., na.rm = TRUE)/2))

metabolomics_PC_data_imputed <- metabolomics_PC_data %>% mutate_if(is.numeric, ~replace(., is.na(.), min(., na.rm = TRUE)/2))
```


## Log-transformation
```{r}
## Log-transform on data, where minimum is added to all values
metabolomics_TA_data_log <- log2(metabolomics_TA_data_imputed)
metabolomics_PC_data_log <- log2(metabolomics_PC_data_imputed)
```

## Create subset with only annotated metabolites (from TA normalized data)
```{r}
feature_metadata_ann <- feature_metadata[!is.na(feature_metadata$DotProduct),]
metabolomics_TA_data_log_ann <- metabolomics_TA_data_log[, rownames(feature_metadata_ann)]
```



## Plot distributions
```{r}
source("support_functions/PlotDistributions2.R")

PlotDistributions(metabolomics_TA_data, type = 'hist', main_title = 'Total area norm' )
PlotDistributions(metabolomics_PC_data, type = 'hist', main_title = 'Protein content norm' )
PlotDistributions(metabolomics_TA_data_log, type = 'hist', main_title = 'Total area norm Log' )
PlotDistributions(metabolomics_PC_data_log, type = 'hist', main_title = 'Protein content norm Log' )
```



### Boxplots per sample
```{r}
source('support_functions/BoxplotSamples.R')

metabolomics_TA_data_log$Sample_ID <- rownames(metabolomics_TA_data_log)
boxplot_samples(metabolomics_TA_data_log, 'Metabolomics (LUMC)')
metabolomics_TA_data_log$Sample_ID <- NULL
```


## Save data and metadata to .CSV files
```{r}
write.csv(metabolomics_TA_data_log, file =  file.path(params$outdir, 'metabolomics_lumc_data_total.csv'))
#write.csv(metabolomics_PC_data_log, file =  file.path(params$outdir, 'metabolomics_lumc_data_protein.csv'))
write.csv(feature_metadata, file =  file.path(params$outdir, 'metabolomics_lumc_feature_metadata.csv'))
write.csv(sample_metadata_ordered, file =  file.path(params$outdir, 'metabolomics_lumc_sample_metadata.csv'))

## Only Annotated data
write.csv(metabolomics_TA_data_log_ann, file =  file.path(params$outdir, 'metabolomics_lumc_data_total_Ann.csv'))
write.csv(feature_metadata_ann, file =  file.path(params$outdir, 'metabolomics_lumc_feature_metadata_Ann.csv'))
```


```{r}
sessionInfo()   
```