---
title: "Metabolomics  to tabular format"
output: html_document
params:
  
  metabomics_feature_neg_metadata: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231128_XCMS/XCMSR.annotated.Report_231128_ESIneg.tsv
  metabomics_feature_pos_metadata: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231120_XCMS/XCMSR.annotated.Report_231120_ESIpos.tsv
  metabolomics_pos_data: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231120_UDA/231120_ControlsvsPatients_2024-03-13_ESIpos.tsv
  metabolomics_neg_data: Z:/data/iPSC_data/iPSC_metabolomics_TML/Processed_untargeted_metabolomics_data/231128_UDA/231128_ControlsvsPatients_2024-03-13_ESIneg.tsv
  
  outdir: Z:/data/iPSC_data/iPSC_MAE
---

## Read processed metabolomics data and save to tabular formats for the MultiAssayExperiment object

### Libraries
```{r}
library(dplyr)
```

#### Script parameters
```{r}
print(Sys.Date())
print(params)
```


#### Functions to read data
```{r}

## Function to read metabolomics data from TML
read_mtblcs_data_tml <- function(filepath, id_string) {
  
  ### Read data
  mtblmcs_data <- read.delim(filepath, header = FALSE)
  mtblmcs_data$V105 <- NULL
  
  ### Remove columns
  colnames(mtblmcs_data) <- mtblmcs_data[3,]
  rows <- nrow(mtblmcs_data)
  mtblmcs_data <- mtblmcs_data[4:rows,]
  
  mtblmcs_data$Mass <- NULL
  mtblmcs_data$R <- NULL
  
  ## Rename metabolomics features
  feature_names <- lapply(mtblmcs_data$F, function(x) {
    y <- paste0(id_string, x)
    y
  })
  rownames(mtblmcs_data) <- feature_names
  mtblmcs_data$F <- NULL
  
  return(mtblmcs_data)
}

### Save sample/filename mapping dataframe
create_sample_metadata_df <- function(filepath) {
  sample_mapping <- read.delim(filepath, header = FALSE)
  sample_mapping <- sample_mapping[1:3,4:104]
  sample_mapping <- data.frame(t(sample_mapping))
  sample_mapping$X2 <- NULL
  colnames(sample_mapping) <- c('File_name', 'Sample_name')
  return(sample_mapping)
}


### Store Mass and Retention time in other dataframe
create_feature_metadata_df  <- function(filepath) {
  data <- read.delim(filepath, header = FALSE)
  colnames(data) <- data[3,]
  rows <- nrow(data)
  data <- data[4:rows,]
  feature_metadata_df <- data[, 2:3]
  return(feature_metadata_df)
}  
```

## Read data for positive ionization mode
```{r}
## measurement data
metabolomics_pos <- read_mtblcs_data_tml(params$metabolomics_pos_data, 'TML_pos')

## sample metadata
sample_mapping <- create_sample_metadata_df(params$metabolomics_pos_data)

## feature metadata
metabolomics_features_pos_ids <- create_feature_metadata_df(params$metabolomics_pos_data)
```


## Read data for negative ionization mode
```{r}
## measurement data
metabolomics_neg <- read_mtblcs_data_tml(params$metabolomics_neg_data, 'TML_neg')

## feature metadata
metabolomics_features_neg_ids <- create_feature_metadata_df(params$metabolomics_neg_data)
```


#### Feature metadata (positive)
```{r }
metabolomics_features_pos <- read.delim(params$metabomics_feature_pos_metadata, row.names = 1)
columns_keep <- c("mzmed", "mzmin", "mzmax", "rtmed", 
                  "rtmin", "rtmax", "npeaks", "X1", "ms_level",
                  "isotopes", "adduct", "pcgroup")
metabolomics_features_pos <- metabolomics_features_pos[ , names(metabolomics_features_pos) %in% columns_keep]

## Add mass and retention time from UDA file
metabolomics_features_pos <- cbind(metabolomics_features_pos_ids, metabolomics_features_pos)

## Add feature IDs (temporarily)
rownames(metabolomics_features_pos) <- rownames(metabolomics_pos)
head(metabolomics_features_pos)
print(dim(metabolomics_features_pos))
```


### Filter features with too much missing values
```{r}
filter_na_columns <- function(data, threshold = 0.2) {
  
  # calculate NA values per column
  na_proportion <- colSums((data==0) / nrow(data))
  # Filter out columns about threshold
  data_out <- data %>% select(which(na_proportion <= threshold))
  return(data_out)
}


# Remove QC samples
metabolomics_pos <- data.frame(t(metabolomics_pos))
metabolomics_pos <- metabolomics_pos[1:82,]

# Remove features with missing values
metabolomics_filtered_pos <- filter_na_columns(metabolomics_pos)
dim(metabolomics_filtered_pos)


# Remove these features from feature metadata table
metabolomics_features_pos <- metabolomics_features_pos[colnames(metabolomics_filtered_pos),]
```

### Get list of metabolites for complete data
```{r}
metabolomics_pos[metabolomics_pos ==  0] <- NA
metabolomics_pos_complete <- metabolomics_pos[ , colSums(is.na(metabolomics_pos)) == 0]
```


### Replace zero's by half of the minimum value
```{r}
### Make columns numericals
metabolomics_filtered_pos[] <- lapply(metabolomics_filtered_pos, as.numeric)

## Replace zero's by half of minimum
metabolomics_pos_noZero <- metabolomics_filtered_pos %>% mutate(across(everything(), ~if_else(. == 0, min(.[. != 0])/2, .)))
```

### Log-transform data
```{r}
metabolomics_pos_noZero_log <- log(metabolomics_pos_noZero + 1)
```


### Print measurement data
```{r}
head(metabolomics_pos_noZero_log[, 1:6])
print(dim(metabolomics_pos_noZero_log))
```


### Sample names matching
```{r}
sample_mapping_noQC <- sample_mapping[1:82,]

sample_ids_temp <- sub(".*?-.*?-(.*)", "\\1", sample_mapping_noQC$Sample_name)
sample_ids <- sub("^([^_]*)_", "\\1", sample_ids_temp)

sample_mapping_noQC$Sample_name <- sample_ids


# New column for ISA names
sample_identifiers_ISA <- lapply(sample_mapping_noQC$Sample_name, function(x){
    name <- substr(x, start = 1, stop = (nchar(x) - 1))
    name
})
sample_mapping_noQC$Sample_name_ISA <- as.character(sample_identifiers_ISA)
```


### Add correct sample IDs
```{r}
metabolomics_pos_noZero_log$SampleName <- sample_ids
```


### Merge technical replicates by taking mean
```{r}
metabolomics_pos_noZero_log_noQC_merged <- metabolomics_pos_noZero_log

## Function to rename sample IDs to isa names
rename_samples <- function(sample_id) {
  
  ## Remove ".1", indicating technical replicate
  if (substring(sample_id, nchar(sample_id)-1, nchar(sample_id)) == ".1") {
    temp_name <- substring(sample_id, 1, nchar(sample_id)-2)
  } else {
    temp_name <- sample_id
  }
  
  ## Split based on - char
  temp_name <- substr(temp_name, 21, nchar(temp_name))
  temp_name <- gsub("\\.", "-", temp_name)
  parts <- strsplit(temp_name, "-")

  ## Select parts that are relevent for naming  
  if (length(parts[[1]]) > 2) {
    new_sample_id <- paste0(parts[[1]][[2]], '-', parts[[1]][[3]])
  }
  else {
    new_sample_id <- parts[[1]][[2]]
  }
  return(new_sample_id)
}

# Rename Sample IDs without .1
new_names <- lapply(rownames(metabolomics_pos_noZero_log_noQC_merged), rename_samples)
metabolomics_pos_noZero_log_noQC_merged$Sample_ID <- as.character(new_names)
#metabolomics_pos_noZero_log_noQC_merged$SampleName <- NULL

# Aggregate technical replicates with mean
metabolomics_pos_noZero_log_noQC_merged <- metabolomics_pos_noZero_log_noQC_merged %>%
  group_by(Sample_ID)  %>%
  summarize(across(everything(), mean))
```


### Plot histogram of metabolomics data after different pre-processsings
```{r}
source("support_functions/PlotDistributions2.R")
PlotDistributions(metabolomics_pos_noZero, type = 'hist', main_title = 'Zero as half of minimum' )
PlotDistributions(metabolomics_pos_noZero_log, type = 'hist', main_title = 'Log transformed' )
PlotDistributions(metabolomics_pos_noZero_log_noQC_merged, type = 'hist', main_title = 'Replicates merged' )
```




# Negative ionization mode

### Filter features with too much missing values
```{r}
filter_na_columns <- function(data, threshold = 0.2) {
  
  # calculate NA values per column
  na_proportion <- colSums((data==0) / nrow(data))
  # Filter out columns about threshold
  data_out <- data %>% select(which(na_proportion <= threshold))
  return(data_out)
}


# Remove QC samples
metabolomics_neg <- data.frame(t(metabolomics_neg))
metabolomics_neg <- metabolomics_neg[1:82,]

# Remove features with missing values
metabolomics_filtered_neg <- filter_na_columns(metabolomics_neg)
dim(metabolomics_filtered_neg)
```

### Get list of metabolites for complete data
```{r}
metabolomics_neg[metabolomics_neg ==  0] <- NA
metabolomics_neg_complete <- metabolomics_neg[ , colSums(is.na(metabolomics_neg)) == 0]
```

### Replace zero's by half of the minimum value
```{r}
### Make columns numericals
metabolomics_filtered_neg[] <- lapply(metabolomics_filtered_neg, as.numeric)

## Replace zero's by half of minimum
metabolomics_neg_noZero <- metabolomics_filtered_neg %>% mutate(across(everything(), ~if_else(. == 0, min(.[. != 0])/2, .)))
```


### Log-transform data
```{r}
metabolomics_neg_noZero_log <- log2(metabolomics_neg_noZero)
```


### Print measurement data
```{r}
head(metabolomics_neg_noZero_log[, 1:6])
print(dim(metabolomics_neg_noZero_log))
```


#### Feature metadata (negative)
```{r }
metabolomics_features_neg <- read.delim(params$metabomics_feature_neg_metadata, row.names = 1)
columns_keep <- c("mzmed", "mzmin", "mzmax", "rtmed", 
                  "rtmin", "rtmax", "npeaks", "X1", "ms_level",
                  "isotopes", "adduct", "pcgroup")
metabolomics_features_neg <- metabolomics_features_neg[ , names(metabolomics_features_neg) %in% columns_keep]

## Add mass and retention time from UDA file
metabolomics_features_neg <- cbind(metabolomics_features_neg_ids, metabolomics_features_neg)

## Add feature IDs (temporarily)
rownames(metabolomics_features_neg) <- colnames(metabolomics_neg)
head(metabolomics_features_neg)
print(dim(metabolomics_features_neg))

# Remove filtered features from feature metadata table
metabolomics_features_neg <- metabolomics_features_neg[colnames(metabolomics_filtered_neg),]
```


### Add correct sample IDs
```{r}
metabolomics_neg_noZero_log$Sample_ID <- sample_ids
```


## Save data before merging technical replicates
```{r}
## Merge measurement data
metabolomics_tech_reps <- cbind(metabolomics_pos_noZero_log, metabolomics_neg_noZero_log)


## Function to rename metabolomics sample IDs to isa names
rename_samples <- function(sample_id) {
  
  ## Remove ".1", indicating technical replicate
  if (substring(sample_id, nchar(sample_id)-1, nchar(sample_id)) == ".1") {
    temp_name <- substring(sample_id, 1, nchar(sample_id)-2)
  } else {
    temp_name <- sample_id
  }
  
  ## Split based on - char
  temp_name <- substr(temp_name, 21, nchar(temp_name))
  temp_name <- gsub("\\.", "-", temp_name)
  parts <- strsplit(temp_name, "-")

  ## Select parts that are relevent for naming  
  if (length(parts[[1]]) > 2) {
    new_sample_id <- paste0(parts[[1]][[2]], '-', parts[[1]][[3]])
  }
  else {
    new_sample_id <- parts[[1]][[2]]
  }
  return(new_sample_id[[1]])
}

tech_rep_names <- lapply(rownames(metabolomics_tech_reps), rename_samples)
metabolomics_tech_reps$group <- as.character(tech_rep_names)
metabolomics_tech_reps$SampleName <- NULL
metabolomics_tech_reps$Sample_ID <- NULL

## Remove sample with only 1 tech rep
single_tech_reps <- names(table(metabolomics_tech_reps$group)[table(metabolomics_tech_reps$group) < 2])
metabolomics_tech_reps<- subset(metabolomics_tech_reps, !(group %in% single_tech_reps))
```

Save to CSV
```{r}
### Subset technical replicates for complete data
metabolomics_tech_reps_complete <- metabolomics_tech_reps[, c(colnames(metabolomics_pos_complete), colnames(metabolomics_neg_complete), 'group')]

write.csv(metabolomics_tech_reps, file.path(params$outdir, 'metabolomics_tml_data_tech_reps.csv'))
write.csv(metabolomics_tech_reps_complete, file.path(params$outdir, 'metabolomics_tml_data_tech_rep_complete.csv'))
```


### Merge technical replicates by taking mean
```{r}
metabolomics_neg_noZero_log$Sample_ID <- sample_ids

metabolomics_neg_noZero_log_noQC_merged <- metabolomics_neg_noZero_log

# Rename Sample IDs without .1
new_names <- lapply(rownames(metabolomics_neg_noZero_log_noQC_merged), rename_samples)
metabolomics_neg_noZero_log_noQC_merged$Sample_ID <- as.character(new_names)
#metabolomics_neg_noZero_log_noQC_merged$SampleName <- NULL

# Aggregate technical replicates with mean
metabolomics_neg_noZero_log_noQC_merged <- metabolomics_neg_noZero_log_noQC_merged %>%
  group_by(Sample_ID)  %>%
  summarize(across(everything(), mean))
```


### Plot histogram of metabolomics data after different pre-processsings
```{r}
source("support_functions/PlotDistributions2.R")

#PlotDistributions(metabolomics_neg, type = 'hist', main_title = 'No pre-processing' )
PlotDistributions(metabolomics_neg_noZero, type = 'hist', main_title = 'Zero as half of minimum' )
PlotDistributions(metabolomics_neg_noZero_log, type = 'hist', main_title = 'Log transformed' )
PlotDistributions(metabolomics_neg_noZero_log_noQC_merged, type = 'hist', main_title = 'Replicates merged' )
```


## Rename tech reps for plotting
```{r}
## Function to rename metabolomics sample IDs to isa names
rename_samples <- function(sample_id) {
  
  ## Remove ".1", indicating technical replicate
  if (substring(sample_id, nchar(sample_id)-1, nchar(sample_id)) == ".1") {
    temp_name <- substring(sample_id, 1, nchar(sample_id)-2)
  } else {
    temp_name <- sample_id
  }
  
  ## Split based on - char
  temp_name <- substr(temp_name, 21, nchar(temp_name))
  temp_name <- gsub("\\.", "-", temp_name)
  parts <- strsplit(temp_name, "-")

  ## Select parts that are relevent for naming  
  if (length(parts[[1]]) > 2) {
    new_sample_id <- paste0(parts[[1]][[2]], '-', parts[[1]][[3]])
  }
  else {
    new_sample_id <- parts[[1]][[2]]
  }
  return(new_sample_id[[1]])
}



tech_rep_names <- lapply(rownames(metabolomics_neg_noZero_log), rename_samples)
metabolomics_neg_noZero_log$group <- as.character(tech_rep_names)
## Also add to metabolomics_noZero
metabolomics_neg_noZero <- metabolomics_neg_noZero[rownames(metabolomics_neg_noZero_log),]
metabolomics_neg_noZero$group <- as.character(tech_rep_names)


## Remove sample with only 1 tech rep
single_tech_reps <- names(table(metabolomics_neg_noZero_log$group)[table(metabolomics_neg_noZero_log$group) < 2])
metabolomics_neg_noZero_log <- subset(metabolomics_neg_noZero_log, !(group %in% single_tech_reps))
```



## Merge ionization modes
```{r}
## Merge measurement data
metabolomics_final <- cbind(metabolomics_pos_noZero_log_noQC_merged, metabolomics_neg_noZero_log_noQC_merged)

## Merge feature metadata
feature_metadata <- rbind(metabolomics_features_pos, metabolomics_features_neg)

```


### Boxplots per sample
```{r}
source('support_functions/BoxplotSamples.R')

boxplot_samples(metabolomics_final, 'Metabolomics (TML)')
```


## Save to CSV
```{r}
write.csv(metabolomics_final, file =  file.path(params$outdir, 'metabolomics_tml_data.csv'), row.names = F)
write.csv(feature_metadata, file =  file.path(params$outdir, 'metabolomics_tml_feature_metadata.csv'))
write.csv(sample_mapping_noQC, file =  file.path(params$outdir, 'metabolomics_tml_sample_metadata.csv'))
```


```{r}
sessionInfo()   
```

