---
title: "Proteomics2Tabular"
output: html_document
params:
  proteomics_data: Z:/data/iPSC_data/iPSC_proteomics/PTX471b_EATRIS_iPSC_DIA/_DIANN_results/DIA-MBR_iPSC_20231110_results.pg_matrix.tsv
  #proteomics_data_spectro: Z:/data/iPSC_data/iPSC_proteomics/PTX471b_EATRIS_iPSC_DIA_Spectronaut/iPSC_Xomics_spectronaut_20241104.xlsx
  proteomics_data_spectro: 'Z:/data/iPSC_data/iPSC_proteomics/DIA_Spectronaut_Complete/summary-results.proteinpgquantity (1).parquet'
  
  outdir: Z:/data/iPSC_data/iPSC_MAE
---

# Read processed proteomics data and save to tabular formats for the MultiAssayExperiment

```{r}
library(dplyr)
```

## Script parameters

```{r}
print(Sys.Date())
print(params)
```

## Read DIANN data (TSV)

```{r}
proteomics <- read.delim(params$proteomics_data)
print(colnames(proteomics))
```

## Feature metadata (DIANN)

```{r }
feature_headers <- colnames(proteomics)[1:5]
feature_metadata <- proteomics[, feature_headers]
head(feature_metadata)
```

## Measurement data (DIANN)

```{r }
measurement_data <- proteomics[, !names(proteomics) %in% feature_headers]

# Protein IDs as rownames
rownames(measurement_data) <- feature_metadata$Protein.Ids

#transpose matrix
measurement_data <- data.frame(t(measurement_data))
  
# Add a new column to indicate sample categories
# The categories have been extracted from the rownames, as the first word before row names
# NANS_p3c1_Proteomics_2_S2_A11_1 is NANS
sampleType <- sub("^(.*?)_.*","\\1", rownames(measurement_data))
measurement_data$group <- sampleType
```


## Read Spectronaut Matrix (.xlsx)

```{r}
#proteomics_sp <- readxl::read_xlsx(params$proteomics_data_spectro)
proteomics_sp <- arrow::read_parquet(params$proteomics_data_spectro)
print(colnames(proteomics_sp))
```

## Feature metadata (Spectronaut)

```{r }
feature_headers <- colnames(proteomics_sp)[1:5]
feature_metadata_sp <- proteomics_sp[, feature_headers]
```

## Measurement data (Spectronaut)

```{r }
measurement_data_sp <- proteomics_sp[, !names(proteomics_sp) %in% feature_headers]

# Protein IDs as rownames
rownames(measurement_data_sp) <- feature_metadata_sp$protein_accession

#transpose matrix
measurement_data_sp <- data.frame(t(measurement_data_sp))
  
# Add a new column to indicate sample categories
# The categories have been extracted from the rownames, as the first word before row names
# NANS_p3c1_Proteomics_2_S2_A11_1 is NANS
sampleType <- sub("^(.*?)_.*","\\1", rownames(measurement_data_sp))
measurement_data_sp$group <- sampleType
head(measurement_data_sp)
```


# Data pre-processing

## Remove features with more than 20% missingness

```{r}
## Function to remove feature with more than 20% missing
##
##
filter_na_columns <- function(data, threshold = 0.2) {
  
  # calculate NA values per column
  na_proportion <- colSums(is.na(data)) / nrow(data)
  # Filter out columns about threshold
  data %>% select(which(na_proportion <= threshold))
  
}


## Function to remove feature with more than 20% missing, in each pheno group
##
##
filter_na_groupwise <- function(data, threshold = 0.2) {
  
  group_col <- 'group'
  # Groups by pheno groups and calculate NA percentages per peptide
  missingness <- data %>%
    group_by(across(all_of(group_col))) %>%
    summarise(across(everything(), ~ mean(is.na(.)), .names = "missing_{col}")) %>%
    summarise(across(starts_with("missing_"), max))
    
  cols_keep <- names(missingness)[missingness <= 0.2]
  cols_keep <- gsub("missing_", "", cols_keep)
    
  df_clean <- data %>%
    select(all_of(c(cols_keep, group_col)))
    
  return(df_clean)
}
```


### Filter features in DIANN data
```{r}
measurement_filtered <- filter_na_columns(measurement_data)
#measurement_filtered<- filter_na_groupwise(measurement_data)
dim(measurement_filtered)

# Remove these features from feature metadata table
feature_metadata <- feature_metadata[feature_metadata$Protein.Ids %in% colnames(measurement_filtered),]

# Convert NA to 0
measurement_filtered_NA <- measurement_filtered
measurement_filtered[is.na(measurement_filtered)] <- 0
```



### Get list of proteins with complete measuremetns
```{r}
measurement_data_sp_noNA <- measurement_data_sp[ , colSums(is.na(measurement_data_sp)) == 0]
#write.csv(measurement_data_sp_noNA, file =  file.path(params$outdir, 'proteomics_data_noNA.csv'), row.names = F)
```



### Filter features in Spectronaut data
```{r}
measurement_filtered_sp <- filter_na_columns(measurement_data_sp)
#measurement_filtered_sp <- filter_na_groupwise(measurement_data_sp)
dim(measurement_filtered_sp)

# Remove these features from feature metadata table
feature_metadata_sp <- feature_metadata_sp[feature_metadata_sp$protein_accession %in% colnames(measurement_filtered_sp),]

# Convert NA to 0
measurement_filtered_sp_NA <- measurement_filtered_sp
measurement_filtered_sp[is.na(measurement_filtered_sp)] <- 0
```

## Visualize missingness in data

```{r}
source("support_functions/heatmap_missingness.R")
heatmap_missing(measurement_data, 'Proteomics')
heatmap_missing(measurement_filtered_NA, 'Proteomics , Filter 20% NA columns')
heatmap_missing(measurement_data_sp, 'Proteomics SP')
heatmap_missing(measurement_filtered_sp_NA, 'Proteomics SP , Filter 20% NA columns')
```


## Replace zero's by half of the minimum value
```{r}
### Make columns numericals
measurement_filtered[] <- lapply(measurement_filtered, as.numeric)
measurement_filtered_sp[] <- lapply(measurement_filtered_sp, as.numeric)

## Replace zero's by half of minimum
proteomics_noZero <- measurement_filtered %>% mutate(across(everything(), ~if_else(. == 0, min(.[. != 0])/2, .)))
proteomics_sp_noZero <- measurement_filtered_sp %>% mutate(across(everything(), ~if_else(. == 0, min(.[. != 0])/2, .)))
```


## Log-transform data
```{r}
## DIANN
proteomics_noZero_log <- log2(proteomics_noZero)

## Spectronaut
proteomics_sp_noZero_log <- log2(proteomics_sp_noZero)
```



# Merge technical replicates

## Add Sample IDs column
```{r}
# Add column wich indiciates bio replicate
sampleIDs_temp <- strsplit(rownames(proteomics_noZero_log),"_Proteomics_")
sample_IDs <- lapply(sampleIDs_temp, function(x) {
    sample_name <- x[[1]]
    sample_No <- x[[2]]
    sample_No <- strsplit(sample_No, "_")[[1]][[1]]
    sample_name <- paste0(sample_name, "_Proteomics_", sample_No)
    sample_name
    
})
# Add as column 
proteomics_noZero_log$Sample_ID <- as.character(sample_IDs)
```


## Add Sample IDs column (Spectronaut)
```{r}
# Add column wich indiciates bio replicate
sampleIDs_temp <- strsplit(rownames(proteomics_sp_noZero_log),"_Proteomics_")
sample_IDs_sp <- lapply(sampleIDs_temp, function(x) {
    sample_name <- x[[1]]
    sample_No <- x[[2]]
    sample_No <- strsplit(sample_No, "_")[[1]][[1]]
    sample_name <- paste0(sample_name, "_Proteomics_", sample_No)
    sample_name
    
})
# Add as column 
proteomics_sp_noZero_log$Sample_ID <- as.character(sample_IDs_sp)
```


## Scatter plot of good examples

```{r}
library(ggplot2)
library(ggpubr)
library(stats)


## Function to plot
scatter_plot_cor <- function(df, col='pink', x_axis, y_axis, title =  "Scatter plot") {
  
  ggp <- ggscatter(df, 
          x = colnames(df)[[1]], 
          y = colnames(df)[[2]],
          color = col,
          add = 'reg.line',
          add.params = list(color = "blue")) +
    labs(title = title,
         x = paste0(x_axis, " protein levels"), 
         y = paste0(y_axis, " protein levels"), ) +
    theme(axis.title.x =  element_text(size = 14),
          axis.title.y =  element_text(size = 14)) +
    stat_cor(aes(label = ..rr.label..), method = "pearson")
  
  return(ggp)
}

## Function to get data frame and use the plotting function above
make_scatter_plot_cor <- function(df, title, id_list) {
  df$Sample_ID <- as.character(id_list)
  scatter_plot_df <- df[df$Sample_ID == 'WTC_Proteomics_3',]
  scatter_plot_df <- data.frame(t(scatter_plot_df))
  scatter_plot_df[,1] <- as.numeric(scatter_plot_df[,1])
  scatter_plot_df[,2] <- as.numeric(scatter_plot_df[,2])
  ggp <- scatter_plot_cor(scatter_plot_df, 'dodgerblue1', 'Tech. rep 1', 'Tech. rep 2', title=title)
  return(ggp)  
  
}

## Function to get data frame and use the plotting function above
make_scatter_plot_cor_cell_line <- function(df, title, id_list) {
  df$Sample_ID <- as.character(id_list)
  scatter_plot_df <- df[df$Sample_ID == 'WTC_Proteomics_1' | df$Sample_ID == 'WTC_Proteomics_2',]
  scatter_plot_df <- scatter_plot_df[!duplicated(scatter_plot_df$Sample_ID),]
  scatter_plot_df <- data.frame(t(scatter_plot_df))
  scatter_plot_df[,1] <- as.numeric(scatter_plot_df[,1])
  scatter_plot_df[,2] <- as.numeric(scatter_plot_df[,2])
  ggp <- scatter_plot_cor(scatter_plot_df, 'dodgerblue1', 'WTC line bio rep 1', 'WTC line bio rep 2', title=title)
  return(ggp)  
  
}
```

## Compare tech reps of DIANN and Spectronaut
```{r}
make_scatter_plot_cor(proteomics_sp_noZero_log, 'Log-transformation, Spectronaut', sample_IDs_sp)
make_scatter_plot_cor(proteomics_noZero_log, 'Log-transformation, DIANN', sample_IDs)
```

## Merge technical replicates by taking median

```{r}
proteomics_noZero_log_merged <- proteomics_noZero_log %>%
  group_by(Sample_ID)  %>%
  summarize(across(everything(), mean))

proteomics_sp_noZero_log_merged <- proteomics_sp_noZero_log %>%
  group_by(Sample_ID)  %>%
  summarize(across(everything(), mean))
```

### Sample names DIANN
```{r}
proteomics_noZero_log_merged$Sample_ID <- gsub('X', '', proteomics_noZero_log_merged$Sample_ID)
proteomics_noZero_log_merged$Sample_ID
```

## Sample names Spectronaut
```{r}
proteomics_sp_noZero_log_merged$Sample_ID <- gsub('X', '', proteomics_sp_noZero_log_merged$Sample_ID)
proteomics_sp_noZero_log_merged$Sample_ID
```

# Select only proteins with complete measurements for technical replicates
```{r}
proteomics_sp_noZero_log_complete <- proteomics_sp_noZero_log[, c(colnames(measurement_data_sp_noNA), "Sample_ID")]
```


# Histograms after different pre-processing steps

## DIAN data
```{r}
source("support_functions/PlotDistributions2.R")

PlotDistributions(measurement_filtered, type = 'hist', main_title = 'No pre-processing' )
PlotDistributions(proteomics_noZero, type = 'hist', main_title = 'Zero as half of minimum' )
PlotDistributions(proteomics_noZero_log, type = 'hist', main_title = 'Log-transformation base 2' )
PlotDistributions(proteomics_noZero_log_merged, type = 'hist', main_title = 'Merged replicates' )
```

## Spectronaut data
```{r}
PlotDistributions(measurement_filtered_sp, type = 'hist', main_title = 'No pre-processing' )
PlotDistributions(proteomics_sp_noZero, type = 'hist', main_title = 'Zero as half of minimum' )
PlotDistributions(proteomics_sp_noZero_log, type = 'hist', main_title = 'Log-transformation base 2' )
PlotDistributions(proteomics_sp_noZero_log_complete, type = 'hist', main_title = 'Log-transformation base 2 (complete)' )
PlotDistributions(proteomics_sp_noZero_log_merged, type = 'hist', main_title = 'Merged replicates' )
```


### Boxplots per sample
```{r}
source('support_functions/BoxplotSamples.R')

boxplot_samples(proteomics_noZero_log_merged, 'DIANN')
boxplot_samples(proteomics_sp_noZero_log_merged, 'Spectronaut')
```


## Save positive to CSV

### DIANN
```{r}
# write.csv(proteomics_noZero_log_merged, file =  file.path(params$outdir, 'proteomics_data_di.csv'), row.names = F)
# write.csv(proteomics_noZero_log, file =  file.path(params$outdir, 'proteomics_data_tech_reps_di.csv'), row.names = F)
# write.csv(feature_metadata, file =  file.path(params$outdir, 'proteomics_feature_metadata_di.csv'))
```

### Spectronaut
```{r}
write.csv(proteomics_sp_noZero_log_merged, file =  file.path(params$outdir, 'proteomics_data.csv'), row.names = F)
write.csv(proteomics_sp_noZero_log, file =  file.path(params$outdir, 'proteomics_data_tech_reps.csv'), row.names = F)
write.csv(proteomics_sp_noZero_log_complete, file =  file.path(params$outdir, 'proteomics_data_tech_reps_complete.csv'), row.names = F)
write.csv(feature_metadata_sp, file =  file.path(params$outdir, 'proteomics_feature_metadata.csv'), row.names = F)
```


```{r}
sessionInfo()   
```
