---
title: "QC Proteomics"
output: 
  html_notebook:
    toc: true
    theme: united
author: "Purva Kulkarni"
email: "purva.kulkarni@radboudumc.nl"
#Date:`r Sys.Date()`
params:
  proteomics_data_stats: Z:/data/iPSC_data/iPSC_proteomics/PTX471b_EATRIS_iPSC_DIA/_DIANN_results/DIA-MBR_iPSC_20231110_results.stats.tsv
---

# Introduction
This notebook contains the QC report for the proteomics data.

# Read data

```{r}
print(params)
proteomics_data_stats <- read.delim(params$proteomics_data_stats)
print(colnames(proteomics_data_stats))
print(head(proteomics_data_stats[,1:7]))
```
# Quality control metrics

## Total signal, median RT and median mass across samples
Plot the total ion count observed for MS1 acquisition
```{r fig.width=14}
library(ggplot2)

# Make plots wider
#options(repr.plot.width = 20, repr.plot.height = 8)

ggplot(proteomics_data_stats, aes(x = File.Name, y=MS1.Signal)) + geom_point() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```
Plot the total ion count observed for MS1 acquisition
```{r fig.width=14}
library(ggplot2)

ggplot(proteomics_data_stats, aes(x = File.Name, y=MS2.Signal)) + geom_point() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) 
```

Plot the median RT predicted across samples
```{r fig.width=14}
library(ggplot2)

ggplot(proteomics_data_stats, aes(x = File.Name, y=Median.RT.Prediction.Acc)) + geom_point() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) 
```

Plot the median MS1 mass across samples
```{r fig.width=14}
library(ggplot2)

ggplot(proteomics_data_stats, aes(x = File.Name, y=Median.Mass.Acc.MS1.Corrected)) + geom_point() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) 
```

## Principal Component analysis
```{r  fig.width=14}
source("support_functions/performPCA.R")
proteomics_data <- read.delim("Z:/data/iPSC_data/iPSC_proteomics/PTX471b_EATRIS_iPSC_DIA/_DIANN_results/DIA-MBR_iPSC_20231110_results.pg_matrix.tsv")
performPCA(proteomics_data, "CDH2_hetWTC_Proteomics_1_S1_C12_1_13378", "NANS_p1c1_Proteomics_2_S2_A5_1_13436") 

```