---
title: "RNAseq2Tabular"
output: html_document
params:
  mrna_deseq_data: Z:/data/iPSC_data/iPSC_RNAseq/iPSC_USEQ_processed/KUL9276_processed/featureCounts/KUL9276_processed_featureCounts_deseq2.csv
  mrna_cpm_data: Z:/data/iPSC_data/iPSC_RNAseq/iPSC_USEQ_processed/KUL9276_processed/featureCounts/KUL9276_processed_featureCounts_cpm.csv
  mrna_rpkm_data: Z:/data/iPSC_data/iPSC_RNAseq/iPSC_USEQ_processed/KUL9276_processed/featureCounts/KUL9276_processed_featureCounts_RPKM.csv
  mrna_read_data: Z:/data/iPSC_data/iPSC_RNAseq/iPSC_USEQ_processed/KUL9276_processed/featureCounts/KUL9276_processed_exon_featureCounts.raw.txt
  outdir: Z:/data/iPSC_data/iPSC_MAE
 
---

## Read processed RNAseq data and save to tabular formats for the MultiAssayExperiment object

#### Script parameters
```{r}
print(Sys.Date())
print(params)
```


### Read data
```{r}
mrna_deseq <- read.csv(params$mrna_deseq_data, row.names = 1)
mrna_cpm <- read.csv(params$mrna_cpm_data, row.names = 1)
#mrna_rkpm <- read.csv(params$mrna_rkpm_data, row.names = 1)
```


### Read count data with feature metadata
```{r}
mrna_read_data <- read.csv(params$mrna_read_data, row.names = 1, sep = '\t')
feature_metadata <- mrna_read_data[, c(1:6)]
```



## CPM
### Remove mRNA with only 0 (rows)
```{r}
mrna_cpm_filtered <- mrna_cpm[rowSums(mrna_cpm != 0) > 0, , drop=FALSE] 
mrna_cpm_filtered <- data.frame(t(mrna_cpm_filtered))
head(mrna_cpm_filtered[, 1:5])
dim(mrna_cpm_filtered)
``` 

### Log transformation
```{r}
mrna_cpm_filtered_log <- log(mrna_cpm_filtered + 1)
```


### Visualize missingness in data
```{r}
source("support_functions/heatmap_missingness.R")
heatmap_missing(mrna_cpm_filtered_log, 'RNAseq')
```


## DESeq2
### Remove mRNA with only 0 (rows)
```{r}
mrna_deseq_filtered <- mrna_deseq[rowSums(mrna_deseq != 0) > 0, , drop=FALSE] 
mrna_deseq_filtered <- data.frame(t(mrna_deseq_filtered))
head(mrna_deseq_filtered[, 1:5])
dim(mrna_deseq_filtered)


# Log transform DeSeq data
mrna_deseq_filtered_log <- log(mrna_deseq_filtered + 1)

``` 

## Additional filter step to remove features that are 0 in more than 50% of the samples
```{r}
filter_zeros <- function(data, threshold = 0.1) {

  zero_counts <- colSums(as.matrix(data) == 0)
  total_samples <- nrow(data)
  proportion_zeros <- zero_counts / total_samples
  
  data_filtered <- data[, proportion_zeros <= threshold]
  return(data_filtered)
}


# Filter features for both CPM and DESeq data
mrna_cpm_filtered_log_filtered <- filter_zeros(mrna_cpm_filtered_log)
mrna_deseq_filtered_log_filtered <- filter_zeros(mrna_deseq_filtered_log)
```



## Plot distributions
```{r}
source("support_functions/PlotDistributions2.R")

PlotDistributions(mrna_cpm_filtered, type = 'hist', main_title = 'CPM' )
PlotDistributions(mrna_deseq_filtered, type = 'hist', main_title = 'DeSEQ' )
PlotDistributions(mrna_cpm_filtered_log, type = 'hist', main_title = 'CPM Log' )
PlotDistributions(mrna_deseq_filtered_log, type = 'hist', main_title = 'DeSEQ Log' )
PlotDistributions(mrna_cpm_filtered_log_filtered, type = 'hist', main_title = 'CPM Log filtered' )
PlotDistributions(mrna_deseq_filtered_log_filtered, type = 'hist', main_title = 'DeSEQ Log filtered' )


```


## Subset feature metadata on features that are still present after filtering
```{r}
feature_metadata <- feature_metadata[colnames(mrna_cpm_filtered_log_filtered),]
```



## Save data to .CSV
```{r}
write.csv(mrna_deseq_filtered_log_filtered, file.path(params$outdir, 'mrna_dseq_log.csv'))
write.csv(mrna_cpm_filtered_log_filtered, file.path(params$outdir, 'mrna_cpm_log.csv'))
write.csv(feature_metadata, file.path(params$outdir, 'mrna_feature_metadata.csv'))
```


```{r}
sessionInfo()
```

