---
title: "RNAseq pre-processing comparison"
output: html_document
params:
  mrna_read_data: Z:/data/iPSC_data/iPSC_RNAseq/iPSC_USEQ_processed/KUL9276_processed/featureCounts/KUL9276_processed_exon_featureCounts.raw.txt
  outdir: Z:/data/iPSC_data/iPSC_MAE
 
---

## Normalize read count with edgeR data according to UMCUGenetics/RNASeq and custom scripts.

#### Script parameters
```{r}
#print(Sys.Date())
print(params)
```


### Libraries
```{r}
library(edgeR)
suppressMessages(library(DESeq2))
```


### Read data
```{r}
mrna_read_data <- read.csv(params$mrna_read_data, row.names = 1, sep = '\t')
mrna_read_data <- mrna_read_data[ , c(7:ncol(mrna_read_data))]
```


## EdgeR pre-processing from UMCUGenetics/RNASeq-NF
```{r}
tpm3 <- function(counts,len) {
  x <- counts/len
  return(t(t(x)*1e6/colSums(x)))
}


fc.df <- read.delim(params$mrna_read_data, comment.char="#")
# Normalize 
counts <- data.frame(fc.df[,8:ncol(fc.df)])
colnames(counts) <- colnames(fc.df)[8:ncol(fc.df)]
rownames(counts) <- fc.df$Geneid
x <- DGEList(counts=counts, genes=fc.df[,c("Geneid","Length")] )
# RPKM/CPM normalization
df.rpkm <- rpkm(x,x$genes$Length, normalized.lib.sizes=F)
df.cpm <- cpm(x)
df.tpm = tpm3(counts, x$genes$Length)
```


### Make data frame of CPM data
```{r}
df.cpm <- data.frame(df.cpm)
dim(df.cpm)
```


## Custom pre-procssing with CPM
```{r}
# Obtain CPMs
myCPM <- cpm(counts)
head(myCPM)
```


#### Filter lowly expressed genes based on CPM
```{r}
# Determine which in the CPM data frame are larger than 0.5
threshold <- myCPM > 0.5

# Keep genes have 2 TRUE in each row
keep <- rowSums(threshold) >= 3
counts.keep <- counts[keep,]
dim(counts.keep)

# Filter feature metadata as well
rownames(fc.df) <- fc.df$Geneid
fc.df.keep <- fc.df[rownames(counts.keep),]
fc.df.keep <- fc.df.keep[,c(1:6)]
```


### Add mapped gene symbols
```{r}
rna_results <- openxlsx::read.xlsx('C:/Users/Casper.deVisser/Documents/All_Diseases_vsControls.xlsx', sheet = 'RNAseq')
gene_symbols <- lapply(rownames(fc.df.keep), function(x) {
  sym <- rna_results[rna_results$feature == x, 10]
})
fc.df.keep$Gene_symbol <- as.character(gene_symbols)
fc.df.keep$Gene_symbol[fc.df.keep$Gene_symbol == 'character(0)'] <- NA
```


#### Create DGEList object
```{r}
dgeObj <- DGEList(counts= counts.keep, genes=fc.df.keep[,c("Geneid","Length")] )
names(dgeObj)
dgeObj$samples
```


## Log-transformations for both CPM data sets
```{r}
logcounts_cpm_umcu <- data.frame(cpm(x, log = TRUE))
logcounts_cpm_custom <- data.frame(cpm(dgeObj, log = TRUE))
```


## Transpose dataframe 
```{r}
logcounts_cpm_custom <- data.frame(t(logcounts_cpm_custom))
```


## Plot distributions
```{r, warning=FALSE}
source("support_functions/PlotDistributions2.R")

PlotDistributions(mrna_read_data, type = 'hist', main_title = 'Raw read counts' )
PlotDistributions(df.cpm, type = 'hist', main_title = 'CPM UMCU' )
PlotDistributions(logcounts_cpm_umcu, type = 'hist', main_title = 'CPM Log UMCU' )
PlotDistributions(logcounts_cpm_custom, type = 'hist', main_title = 'CPM Log Custom (filtered)' )
```

### Boxplots per sample
```{r}
source('support_functions/BoxplotSamples.R')

logcounts_cpm_custom$Sample_ID <- rownames(logcounts_cpm_custom)
boxplot_samples(logcounts_cpm_custom, 'RNAseq')
logcounts_cpm_custom$Sample_ID <- NULL
```


### Save custom to .csv
```{r}
write.csv(logcounts_cpm_custom, file.path(params$outdir, 'mrna_cpm_log.csv'))
write.csv(fc.df.keep, file.path(params$outdir, 'mrna_feature_metadata.csv'), row.names = F)
```



## Normalize read count with DESeq2, again compare UMCUGenetics and custom script

#### Pre-processing from UMCUGenetics/RNASeq-NF
```{r}
counts_raw <- read.delim(params$mrna_read_data, comment.char="#", row.names = 1)

cds = as.matrix(counts_raw[,7:ncol(counts_raw)])
colnames(cds) <- colnames(counts_raw)[7:ncol(counts_raw)]
#Get column data and set dummy condition
(colData <- data.frame(row.names=colnames(cds),condition=factor(c(rep("exp", ncol(cds))))))
#Create DEseq2 dataset
dds <- DESeqDataSetFromMatrix(countData = cds[, rownames(colData)], colData = colData, design = ~ 1)
#Estimate size factors
size.factors <- estimateSizeFactors(dds)
#Normalize counts
cds.norm <- counts(size.factors, normalized = T)

# Dimension
cds.norm <- data.frame(cds.norm)
dim(cds.norm)
```

### Sample metadata from ISA files
```{r}
#isa_assay_rna <- readxl::read_excel(paste0(here::here(), params$isa_files), sheet = 'Assay - transcriptomics')
isa_assay_rna <- readxl::read_xlsx('C:/Users/Casper.deVisser/Documents/ipsc-data-analysis/20240514_Xomics iPSC Neurons_new_READ-ONLY.xlsx', sheet = 'Assay - transcriptomics')
isa_observation <- readxl::read_xlsx('C:/Users/Casper.deVisser/Documents/ipsc-data-analysis/20240514_Xomics iPSC Neurons_new_READ-ONLY.xlsx', sheet = 4)

## Rewrite sample names to match ISA file
isa_matching_names <- gsub("\\.", "-", rownames(colData))
isa_matching_names <- paste0(isa_matching_names, "_1")
isa_matching_names <- gsub("X", "", isa_matching_names)
## Add to colData
colData$isa_name <- isa_matching_names

## Add sample idetifier to colData
sample_identifiers <- lapply(isa_matching_names, function(x) {
  sample_id <- isa_assay_rna[isa_assay_rna$`assay identifier` == x, 6]
  sample_id[[1]]
})
colData$Sample_ID <- as.factor(as.character(sample_identifiers))
```


#### Custom script
```{r}
#Create DEseq2 dataset
dds <- DESeqDataSetFromMatrix(countData = counts.keep, colData = colData, design = ~ Sample_ID)
dds <- estimateSizeFactors(dds, type='ratio')
dds <- estimateDispersions(dds)
df.vst <- as.data.frame(assay(vst(dds, blind = F)))
```

## Transpose dataframe
```{r}
df.vst <- data.frame(t(df.vst))
```


### Plot distributions
```{r}
#PlotDistributions(cds.norm, type = 'hist', main_title = 'DESeq UMCU' )
PlotDistributions(df.vst, type = 'hist', main_title = 'DESeq VST (custom)' )
```

### Boxplots per sample
```{r}
source('support_functions/BoxplotSamples.R')

df.vst$Sample_ID <- rownames(df.vst)
boxplot_samples(df.vst, 'RNAseq')
df.vst$Sample_ID <- NULL
```


### Save custom DESeq to .csv
```{r}
write.csv(df.vst, file.path(params$outdir, 'mrna_deseq_log.csv'))
```


```{r}
sessionInfo()
```

