---
title: "RNAseq pre-processing comparison"
output: html_document
params:
  mrna_read_data: Z:/data/iPSC_data/iPSC_RNAseq/iPSC_RTC/RawCounts/
  outdir: Z:/data/iPSC_data/iPSC_MAE
 
---

## Normalize read count with edgeR data according to UMCUGenetics/RNASeq and custom scripts.

#### Script parameters
```{r}
#print(Sys.Date())
print(params)
```

## Read .txt files
```{r}
setwd(params$mrna_read_data)
txt_files <- list.files( pattern = "\\counts.txt$")
counts_tables <- lapply(txt_files, function(x) {
 df <- read.table(x, sep = '\t', header = F, stringsAsFactors = F)
 colnames(df) <- c("Key", "Value")
 
 named_vec <- setNames(df$Value, df$Key)
})

mrna_read_data <- as.data.frame(do.call(rbind, counts_tables))
rownames(mrna_read_data) <- sub("\\.counts.txt$", "", txt_files)

## Select actual count value columns
cols_data <- startsWith(x = colnames(mrna_read_data), prefix = 'ENS')
mrna_read_data <- mrna_read_data[,cols_data]
```


### Libraries
```{r}
library(edgeR)
suppressMessages(library(DESeq2))
```


### Sample metadata from ISA files
```{r}
isa_sample <- readxl::read_xlsx('C:/Users/Casper.deVisser/Documents/ipsc-data-analysis/20240514_Xomics iPSC Neurons_new_READ-ONLY.xlsx', sheet = 7)

## Fix wrong CRISPR line sample names
isa_sample[isa_sample$`assay identifier` == 'DM1-p2c1rep-RNA-1_1',1] <- 'DM1-p1c1rep-RNA-1_1'
isa_sample[isa_sample$`assay identifier` == 'DM1-p2c1rep-RNA-2_1',1] <- 'DM1-p1c1rep-RNA-2_1'

isa_sample$temp_col <- isa_sample$`assay identifier`
temp_values <- gsub("_1", "", isa_sample$temp_col)
isa_sample$temp_col <- temp_values

## Fix typo CDH2 --> CHD2
temp_values <- gsub("CDH2", "CHD2", rownames(mrna_read_data))

## Fix wrong CRISPR line sample names
temp_values[temp_values == 'DM1-p2c1rep-RNA-1'] <- 'DM1-p1c1rep-RNA-1'
temp_values[temp_values == 'DM1-p2c1rep-RNA-2'] <- 'DM1-p1c1rep-RNA-2'
rownames(mrna_read_data) <- temp_values

isa_sample <- isa_sample[isa_sample$temp_col %in% rownames(mrna_read_data),]
isa_sample$SampleID <- isa_sample$temp_col
isa_sample$ObsID <- isa_sample$`sample identifier`

## Add SampleId as rownames
rownames(isa_sample) <- isa_sample$SampleID

## Match order or rows
isa_sample <- isa_sample[match(rownames(mrna_read_data), rownames(isa_sample)),]
rownames(isa_sample) <- isa_sample$SampleID
```


## Remove features with only 0
```{r}
no_zero_columns <- colSums(mrna_read_data) !=0
mrna_read_data <- mrna_read_data[, no_zero_columns]
```


## Normalize with CPM
```{r}
mrna_read_data <- data.frame(t(mrna_read_data))
colnames(mrna_read_data) <- temp_values
cpm_data <- cpm(mrna_read_data)
```


#### Filter lowly expressed genes based on CPM
```{r}
# Determine which in the CPM data frame are larger than 0.5
threshold <- cpm_data > 0.5

# Keep genes have 2 TRUE in each row
keep <- rowSums(threshold) >= 3
counts.keep <- mrna_read_data[keep,]
dim(counts.keep)
```


## Normalize read count with DESeq2, again compare UMCUGenetics and custom script

#### Pre-processing from UMCUGenetics/RNASeq-NF
```{r}
dds = DESeqDataSetFromMatrix(counts.keep, colData = isa_sample, design = ~ ObsID)
dds <- estimateSizeFactors(dds, type='ratio')
dds <- estimateDispersions(dds)
vst_data <- as.data.frame(assay(vst(dds, blind = F)))

dim(vst_data)
```


### Plot distributions
```{r}
source('support_functions/PlotDistributions2.R')
#PlotDistributions(cds.norm, type = 'hist', main_title = 'DESeq UMCU' )
PlotDistributions(vst_data, type = 'hist', main_title = 'DESeq VST (custom)' )
```

### Boxplots per sample
```{r}
source('support_functions/BoxplotSamples.R')
vst_data <- data.frame(t(vst_data))
vst_data$Sample_ID <- rownames(vst_data)
boxplot_samples(vst_data, 'RNAseq')
vst_data$Sample_ID <- NULL
```

## Save RTC data to MAE folder
```{r}
write.csv(vst_data, 'Z:/data/iPSC_data/iPSC_MAE/mrna_deseq_rt_vst.csv')
```


## Compare UMCU and RTC pipelines
```{r}
mrna_umcu <- read.csv('Z:/data/iPSC_data/iPSC_MAE/mrna_deseq_log.csv', row.names = 1)
intersection_samples <- intersect(rownames(mrna_umcu), rownames(vst_data))
intersection_feature <- intersect(colnames(mrna_umcu), colnames(vst_data))

mrna_umcu_other <- mrna_umcu[intersection_samples, !colnames(mrna_umcu) %in% intersection_feature]
mrna_umcu <- mrna_umcu[intersection_samples, intersection_feature]
vst_data_compare <- vst_data[intersection_samples, intersection_feature]


plot_df <- data.frame(
  umcu = as.vector(as.matrix(mrna_umcu)),
  rtc = as.vector(as.matrix(vst_data_compare))
)

library(viridis)
library(ggpointdensity)
ggplot(plot_df, aes(x = umcu, y = rtc)) +
  #geom_pointdensity() +
  geom_hex(bins=50) +
  #geom_abline(slope = 1, intercept = 0, color='red', linetype='dashed') +
  scale_fill_viridis()
```

## Save Ensembl IDs which ae not in RTC data
```{r}
write.csv(mrna_umcu_other, 'Z:/users/devissercasper/mrna_other_features.csv')
```





```{r}
sessionInfo()
```

