---
title: "Compare cell lines to reference methylation of neurons"
output: html_document
params:
  outdir: Z:/data/iPSC_data/results/DMR
  methrix_dir: Z:/data/iPSC_data/iPSC_EMseq/Methrix/h5_raw/
  reference_dir: Z:/data/iPSC_data/Markers_files/Kaplan2023_ref_material
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
---


## Libraries
```{r}
library(methrix)
library(MultiAssayExperiment)
library(HDF5Array)
library(openxlsx)
library(ggplot2)
library(tidyr)
library(dplyr)
```

## Params
```{r}
print(params)
```


## Load MAE data
```{r}
mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')

phenotype_data <- as.data.frame(colData(mae))
```


## Read in Methrix object from shared disk
```{r}
meth <- methrix::load_HDF5_methrix(params$methrix_dir)
```


## Select ISA file to get correct sample IDs
```{r}
isa_assay_prot <- readxl::read_xlsx('../../../20240822_Xomics_iPSC_Neurons_new_READ-ONLY.xlsx', sheet = 'Assay - proteomics')
```


## Read cell-type specific unmethylated regions
```{r}
ref_mat <- openxlsx::read.xlsx(file.path(params$reference_dir, '41586_2022_5580_MOESM4_ESM.xlsx'), sheet = 'Table S4A')
colnames(ref_mat) <- ref_mat[1,]

ref_mat_neurons <- ref_mat[ref_mat$Type == 'Neuron',]
ref_mat_oligodend <- ref_mat[ref_mat$Type == 'Oligodend',]
ref_mat_adipocytes <- ref_mat[ref_mat$Type == 'Adipocytes',]
ref_mat_lung_bron <- ref_mat[ref_mat$Type == 'Lung-Ep-Bron',]
```


## Add liftover data (hg38)
## Liftover made with genome.ucsc.edu/cgi-bin/hgLiftOver
```{r}
hg38_KaplanCpGs <- read.table(file.path(params$reference_dir, 'hglft_genome_25904c_Neurons_withOld.bed'))
oligo_hg38_KaplanCpGs <- read.table(file.path(params$reference_dir, 'hglft_genome_2635de_Oligo_withOld.bed'))
adipocytes_hg38_KaplanCpGs <- read.table(file.path(params$reference_dir, 'hglft_genome_274f35_Adip_wihtOld.bed'))
lung_hg38_KaplanCpGs <- read.table(file.path(params$reference_dir, 'hglft_genome_26a1fb_Lung_withOld.bed'))

add_liftover_coord <- function(liftover_df) {
  cpg_parts <- do.call(rbind, strsplit(liftover_df$V1, ":"))
  liftover_df$chr <- cpg_parts[, 1]
  liftover_df$pos <- cpg_parts[, 2]
  
  pos_parts <- do.call(rbind, strsplit(liftover_df$pos, "-"))
  liftover_df$start <- as.numeric(pos_parts[, 1])
  liftover_df$end <- as.numeric(pos_parts[, 2])
  return(liftover_df)
}

neurons_coord_hg38 <- add_liftover_coord(hg38_KaplanCpGs)
oligo_coord_hg38 <- add_liftover_coord(oligo_hg38_KaplanCpGs)
adipocytes_coord_hg38 <- add_liftover_coord(adipocytes_hg38_KaplanCpGs)
lung_coord_hg38 <- add_liftover_coord(lung_hg38_KaplanCpGs)
```


## Select cell-specific CpGs
```{r}
subset_meth_coord <- function(meth, coord_df) {
  
  df <- apply(coord_df, 1, function(row){
    chr = row['chr']
    start = row['start']
    end = row['end']
    meth_subset <- try(methrix::subset_methrix(m = meth, regions = data.table(chr=chr, start = start, end = end)))
    meth_subset
  })
  return(df)
}

meth_neurons <- subset_meth_coord(meth, neurons_coord_hg38)
meth_adipo <- subset_meth_coord(meth, adipocytes_coord_hg38)
meth_oligo <- subset_meth_coord(meth, oligo_coord_hg38)
meth_lung <- subset_meth_coord(meth, lung_coord_hg38)
```


## Make dataframe of averages per cell line for the 'neuronal hypomethylation blocks'
```{r}
source('../../support_functions/ReadMethrix2Df.R')

make_avr <- function(meths, region_names) {
  meth_regions <- lapply(meths, function(meth) {
    df <- read_meth2df(meth, isa_assay_prot)
    df <- df[[1]]
    df <- data.frame(t(df))
    means <- rowMeans(df)
    means
  })
 
  df <- as.data.frame(do.call(cbind, meth_regions))
  colnames(df) <- region_names
  df <- data.frame(t(df))

  return(df)
}

neuron_meth_regions <- make_avr(meth_neurons, hg38_KaplanCpGs$V1)
adipo_meth_regions <- make_avr(meth_adipo, adipocytes_hg38_KaplanCpGs$V1)
oligo_meth_regions <- make_avr(meth_oligo, oligo_hg38_KaplanCpGs$V1)
lung_meth_regions <- make_avr(meth_lung, lung_hg38_KaplanCpGs$V1)
```


## Add reference beta-values
```{r}
ref_mat_neurons_sub <- ref_mat_neurons[ref_mat_neurons$Gene != 'SPATA31A6', ]
neuron_meth_regions$Kaplan <- ref_mat_neurons_sub$`Target meth. `

adipo_meth_regions$Kaplan <- ref_mat_adipocytes$`Target meth. `
oligo_meth_regions$Kaplan <- ref_mat_oligodend$`Target meth. `
lung_meth_regions$Kaplan <- ref_mat_lung_bron$`Target meth. `
```


## Pairwise correlations
```{r}
correlations_with_Kaplan <- function(df) {
  
  df[] <- lapply(df, as.numeric)

  correlations <- apply(df, 2, function(col) cor(col, df[,ncol(df) ], use = "complete.obs"))

  cor_df <- data.frame(
    feature =  colnames(df),
    correlations= as.numeric(correlations)
    )
  nrows <- nrow(cor_df) -1
  cor_df <- cor_df[1:nrows,]
  return(cor_df)
}

cor_neuro <- correlations_with_Kaplan(neuron_meth_regions)
cor_adip <- correlations_with_Kaplan(adipo_meth_regions)
cor_oligo <- correlations_with_Kaplan(oligo_meth_regions)
cor_lung <- correlations_with_Kaplan(lung_meth_regions)

```


## visualize
```{r}
df_combined <- data.frame(cbind(cor_neuro$correlations,  cor_oligo$correlations, cor_adip$correlations, cor_lung$correlations))
colnames(df_combined) <- c("Neurons", "Oligodend", "Adipocytes", "Lung-bron")
df_combined$SampleID <- cor_adip$feature

df_plot <- reshape2::melt(df_combined)

df_label <- df_plot[startsWith(df_plot$SampleID, 'IPS23_00089'), ] # DM patient 2, clone 2
df_label$SampleLabel <- gsub("IPS23_00089", "DM1_P2", df_label$SampleID)

ggp <- ggplot(df_plot, aes(x=variable, y =value, color = variable)) +
  geom_boxplot() +
  geom_jitter(size = 1) +
  geom_text(data = df_label, aes(label = SampleLabel), size = 3) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(title = "Correlations with cell-type specific methylation regions",
       x = "Cell type", 
       y = "Pearson correlation")

ggp
ggsave(file.path(params$outdir, 'AtlasCorrelations.pdf'), ggp, width=5, height = 4)
```


