---
title: "Cluster_samples_with_sign_hits"
output: html_document
params:
  outdir: Z:/data/iPSC_data/results/MixedModels/Plots
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
---

# Visualize results of linear modelling

## Parameters

```{r}
print(params)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!dir.exists(outdir)){
    outdir <- getwd()
}
```


## Read data and metadata: MultiAssayExperiment
```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 

mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
```


## Libraries
```{r}
library(dplyr)
library(tidyr)
library(pheatmap)
library(viridis)
library(ggplot2)
```


## Read nested mixed models (disease-specific associations)
```{r}
MixedModel <-  read.csv('MixedModels2/Model_phenotype.csv', row.names=1)
MixedModel <- MixedModel[MixedModel$outcome != 'phenotypeDiseased: DM1',]
MixedModel_DM1 <- read.csv('MixedModels2/MixedModels_DM1_Rescue.csv', row.names = 1)
MixedModel <- rbind(MixedModel, MixedModel_DM1)
```


## Sample metadata 
```{r}
pheno_df <- as.data.frame(colData(mae))
pheno_df <- pheno_df %>% arrange(phenotype)
sorted_samples <- pheno_df$sample.identifier
sorted_samples <- gsub("-", "_", sorted_samples)
```


## Get significant omics featues in any disease contrast
```{r}
# Find significant proteins
Features_sign <- MixedModel[MixedModel$p.adj < 0.05,]
Features_sign <- Features_sign[Features_sign$outcome != 'phenotypeRescue: repeat excised CRISPR DM1', ]
```


## Subset significant proteins across all comparisons
```{r, fig.width=10, fig.height=10}
## Clustering heatmap based on selection of omcis features, color by disease group
##
## @param assay_name Omics assay name
## @param omics_features List of omics features to used
## @return heatmap plot object
heatmap_sign_features <- function(assay_name, row_names =  TRUE, omics_features, sub_dir) {
  
  # Select these proteins in measurement data
  se <- experiments(mae)[[assay_name]]
  df_assay <- as.data.frame(t(assays(se)[[1]]))
  df_assay <- df_assay[,names(df_assay) %in% omics_features]
  # Remove X in 409B sample names
  rownames(df_assay) <- gsub("X", "", rownames(df_assay))
  sorted_samples_sub <- sorted_samples[sorted_samples %in% rownames(df_assay)]
  df_assay <- df_assay[sorted_samples_sub,]
  
  annotation_col <- data.frame(Group = pheno_df$phenotype)
  rownames(annotation_col) <- sorted_samples
  
  # Clustering of proteins
  row_clustering <- hclust(dist(t(df_assay)))
  
  color_palette <- palette.colors(palette = 'Okabe-Ito')
  
  annotation_colors <- list(
    Group = c('Diseased: CHD2' = color_palette[[3]],
              'Diseased: CHD2 CRISPR' = color_palette[[2]],
              'Diseased: DM1' = color_palette[[4]],
              'Diseased: NANS-CDG' = color_palette[[7]],
              'Healthy' = color_palette[[5]],
              'Rescue: repeat excised CRISPR DM1' = color_palette[[6]])
  )
  
  distance_matrix <- as.matrix(as.dist(df_assay))
  heatmap <- pheatmap(t(df_assay),
           clustering_distance_rows = "euclidean",
           clustering_method = "ward.D2",
           cluster_rows = TRUE,
           cluster_cols = TRUE,
           annotation_col = annotation_col,
           #annotation_row = annotation_row,
           annotation_colors = annotation_colors,
           scale='row',
           show_rownames = row_names,
           show_colnames = TRUE,
           main = paste0(assay_name))
  
  # Check which features cluster together
  row_clust <- heatmap$tree_row
  
  cut_height <- 15
  
  cluster_assignments <- cutree(row_clust, h = cut_height)
  
  annotation_row <- data.frame(Cluster = as.factor(cluster_assignments))
  rownames(annotation_row) <- names(cluster_assignments)
  
  num_clusters <- length(unique(cluster_assignments))
  cluster_colors <- setNames(palette.colors(num_clusters, palette = "Okabe-Ito"),
                             as.character(sort(unique(cluster_assignments))))
  annotation_row_colors <- list(Cluster=cluster_colors)
  
  heatmap <- pheatmap(t(df_assay),
             clustering_distance_rows = "euclidean",
             clustering_method = "ward.D2",
             cluster_rows = row_clust,
             cluster_cols = TRUE,
             annotation_col = annotation_col,
             annotation_row = annotation_row,
             annotation_colors = c(annotation_colors, annotation_row_colors),
             scale='row',
             show_rownames = row_names,
             show_colnames = TRUE,
             main = paste0(assay_name))
  
  
  assay_name <- strsplit(assay_name, "\\|")[[1]][1]
  assay_name <- gsub(" ", "", assay_name)
  full_path <- file.path(outdir, 'Heatmap', sub_dir)
  if (!dir.exists(full_path)) {
    dir.create(full_path, recursive = T)
  }
  file_path <- file.path(full_path, paste0(assay_name, '.png'))
  ggsave(file_path, heatmap,
      width = 25.04, height = 17.46, units = "cm", dpi = 300)
  
  
  # Write cluste features to .csv
  cluster_features <- data.frame(annotation_row)
  file_path <- file.path(full_path, paste0('Clusters_', assay_name, '.csv'))
  write.csv(cluster_features, file_path)
  
  return(heatmap)
  
}
  
```


```{r, fig.width=10}
heatmap_sign_features('EMseq | 100,000 most variable CpG sites', row_names = FALSE, Features_sign$feature, sub_dir = 'Sign')
```


```{r, fig.width=10}
heatmap_sign_features('Proteomics', row_names = T, Features_sign$feature, sub_dir = 'Sign')
```


```{r, fig.width=10}
heatmap_sign_features('Metabolomics, TML', row_names = T, Features_sign$feature, sub_dir = 'Sign')
```


```{r, fig.width=10}
heatmap_sign_features('RNAseq | Deseq', row_names = T, Features_sign$feature, sub_dir = 'Sign')
```


## Same heatmaps, but with top 50 ranked features (even if they are not significant)

```{r, fig.width=10}
get_top_features <- function(assay_name, top_n) {
  MixedModel_assay <- MixedModel[MixedModel$Assay == assay_name,]
  MixedModel_assay <- MixedModel_assay[order(MixedModel_assay$p.value),]
  features <- MixedModel_assay[c(1:top_n), 'feature']
  return(features)
}

assay_names <- unique(MixedModel$Assay)
lapply(assay_names, function(x) {
  print(heatmap_sign_features(x, row_names = T, get_top_features(x, 50), sub_dir = 'Top50'))
})
```


## Top 100 ranked features
```{r, fig.width=10}
lapply(assay_names, function(x) {
  print(heatmap_sign_features(x, row_names = T, get_top_features(x, 100), sub_dir = 'Top100'))
})
```


## Heatmap of Lipidomics data including Class annotations
```{r}
assay_name <- "Lipidomics, Targeted | imputed | log" 
omics_features <- get_top_features(assay_name, 50)

# Select these proteins in measurement data
se <- experiments(mae)[[assay_name]]
df_assay <- as.data.frame(t(assays(se)[[1]]))
df_assay <- df_assay[,names(df_assay) %in% omics_features]
# Remove X in 409B sample names
rownames(df_assay) <- gsub("X", "", rownames(df_assay))
sorted_samples_sub <- sorted_samples[sorted_samples %in% rownames(df_assay)]
df_assay <- df_assay[sorted_samples_sub,]
  
annotation_col <- data.frame(Group = pheno_df$phenotype)
rownames(annotation_col) <- sorted_samples
  
color_palette <- palette.colors(palette = 'Okabe-Ito')
  
annotation_colors <- list(
  Group = c('Diseased: CHD2' = color_palette[[3]],
            'Diseased: CHD2 CRISPR' = color_palette[[2]],
            'Diseased: DM1' = color_palette[[4]],
            'Diseased: NANS-CDG' = color_palette[[7]],
            'Healthy' = color_palette[[5]],
            'Rescue: repeat excised CRISPR DM1' = color_palette[[6]])
)
  
distance_matrix <- as.matrix(as.dist(df_assay))
heatmap <- pheatmap(t(df_assay),
         clustering_distance_rows = "euclidean",
         clustering_method = "ward.D2",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         annotation_col = annotation_col,
         #annotation_row = annotation_row,
         annotation_colors = annotation_colors,
         scale='row',
         show_rownames = row_names,
         show_colnames = TRUE,
         main = paste0(assay_name))
  
# Check which features cluster together
row_clust <- heatmap$tree_row
  
cut_height <- 15
  
cluster_assignments <- cutree(row_clust, h = cut_height)
  
annotation_row <- data.frame(Cluster = as.factor(cluster_assignments))
rownames(annotation_row) <- names(cluster_assignments)
  
num_clusters <- length(unique(cluster_assignments))
cluster_colors <- setNames(palette.colors(num_clusters, palette = "Okabe-Ito"),
                           as.character(sort(unique(cluster_assignments))))
annotation_row_colors <- list(Cluster=cluster_colors)
  
heatmap <- pheatmap(t(df_assay),
           clustering_distance_rows = "euclidean",
           clustering_method = "ward.D2",
           cluster_rows = row_clust,
           cluster_cols = TRUE,
           annotation_col = annotation_col,
           annotation_row = annotation_row,
           annotation_colors = c(annotation_colors, annotation_row_colors),
           scale='row',
           show_rownames = row_names,
           show_colnames = TRUE,
           main = paste0(assay_name))
```


