---
title: 'ANCOVA models'
author:
  - Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  mofa_scores: Z:/data/iPSC_data/Results/MOFA/All_Samples/SampleScores.csv
  outdir: Z:/data/iPSC_data/results/RLM/
---

## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir
```

## Read data and metadata: MultiAssayExperiment

```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
mae
```

The MultiAssayExperiment contains individual omics assays

```{r}
experiment_names <- names(experiments(mae))
experiment_names
```



Technical variables are available per assay

```{r}
tech_vars_factor <- lapply(experiments(mae), function(x) {
  num_levels <- apply(colData(x), 2, function(x) {length(levels(as.factor(x)))})
  t <- names(colData(x))[num_levels <= 9 & num_levels > 1]
  t <- t[t!="subject.id"]
  t <- t[!t %in% names(colData(mae))]
})
tech_vars_factor

tech_vars_continuous <- lapply(experiments(mae), function(x) {
  num_levels <- apply(colData(x), 2, function(x) {length(levels(as.factor(x)))})
  t <- names(colData(x))[num_levels > 9]
  t <- t[t!="subject.id"]
  t <- t[t!="measurement.id"]
  t <- t[t!="measurement_names"]
  t <- t[t!="ID.Patient"]
  t <- t[!t %in% c("SampleID", "Radboud.code", "Kevin.nr", "Kevin.CPM", 
                   "Sample.name", "ISA_name", "SampleID_withrep", "Sample_Overview_names")] #TODO, remove these from sample metadata sheet
  t <- t[!t %in% c("GroupNum", "GroupName", "SampleNorm", "Radboud.code.1")] 

  t <- t[!t %in% names(colData(mae))]
})
tech_vars_continuous
```


Phenotype information shared across all assays can be accessed via `colData()`

```{r sample_information_overview}
names(colData(mae))

pheno_df <- as.data.frame(colData(mae))

## Add column for Disease vs Healthy comparison
new_column <- lapply(pheno_df$phenotype, function(x) {
    if (startsWith(x, 'Diseas')) {
        value <- 'Disease'
    }
    else{
        value <- 'Healthy'
    }
})
pheno_df$Disease <- as.character(new_column)
```


## Linear models with all healthy vs all disease
```{r}
run_lm_se <- function(se, outcomes, covariates = NULL) {
  # create data frame with omics feature values
  df <- as.data.frame(t(assays(se)[[1]]))
  # remove non-variable features
  df <- df[, apply(df, 2, function(column) {
    var(column, na.rm = T) != 0 })]
  # remove extremely sparse features
  xpercent <- nrow(df)*0.5
  df <- df[, apply(df, 2, function(column) {
    sum(column == 0, na.rm = T) < xpercent })]
  features <- colnames(df)
  # add biological variables to data frame
  df <- cbind(df, pheno_df[rownames(df), ]) 
  # --> TODO: add mae as parameter and use experiment name to select se
  # add assay-specific technical variables to data frame
  #df <- cbind(df, colData(se))
  
  #' Apply robust linear modeling to one feature and one outcome variable
  #'
  #' @param feature A molecule identifier present in rowData(SummarizedExperiment).
  #' @param df A data frame containing assay data and biological variables.
  #' @param outcome A character string indicating biological variable column name.
  #' @param covariates A character string indicating covariate column name.
  #' @returns A data frame.
  #' @examples
  #' run_rlm_df("hmdb:HMDB0000062", df, outcome = "Age", covariates = c("Sex"))
  run_rlm_df <- function(feature, df, outcome, covariates = NULL) {
    # run rlm for one feature and one outcome
    df$y <- df[[feature]]
    # define formula
    covar_term <- ""
    if (!is.null(covariates)) {
      covar_term <- paste0(" + ", paste0(covariates, collapse = " + "))}
    str_formula <- paste0("y ~ ", outcome, covar_term)
    my_formula <- as.formula(str_formula)
    # fit robust regression model
    lm_fit <- try(lm(formula = my_formula, data = df))
    if (length(lm_fit) < 2) {
      rlm_results = data.frame(
        feature = feature,
        outcome = outcome,
        formula = str_formula,
        covariates = paste0(covariates, collapse = "+"),
        coefficient = NA, 
        std.error = NA,
        t.value = NA,
        p.value = NA)
    } else {
      # robust F-test - Wald test for multiple coefficients of rlm object
      #f_test <- f.robftest(rlm_fit, var = 2) # 1=intercept, 3:=covariates
      #f_test <- waldtest(lm_fit, terms = c(outcome))
      # results
      rlm_results = data.frame(
        feature = feature,
        outcome = outcome,
        formula = str_formula,
        covariates = paste0(covariates, collapse = "+"),
        coefficient = summary(lm_fit)$coefficients[2, "Estimate"], 
        std.error = summary(lm_fit)$coefficients[2, "Std. Error"],
        t.value = summary(lm_fit)$coefficients[2, "t value"],
        p.value = summary(lm_fit)$coefficients[2, 4],
        BIC = BIC(lm_fit),
        AIC = AIC(lm_fit)
        #p.value = f_test[2,4]
        )
    }
  }
  rlm_results <- do.call(rbind, lapply(outcomes, function(outcome) {
    rlm_result_ <- do.call(rbind, lapply(
      features,
      run_rlm_df,
      df = df, outcome = outcome, covariates = covariates))
    # adjust p-values
    rlm_result_$p.adj <- p.adjust(rlm_result_$p.value, method = "BH")
    rlm_result_$rank <- rank(rlm_result_$p.value)
    rlm_result_
  }))
  rlm_results
}
```

Assay selection
```{r}
assay_names <- names(experiments(mae))[c(1,2,3,6,8,11)]
assay_names
```


## Execute Linear model 
```{r}
outcomes <- c(
  "phenotype",
  #"Disease",
  "sex")

# fit robust linear models using different formulas
mae_rlm_woCovariate <- lapply( # without any covariates
   experiments(mae)[assay_names], run_lm_se,
   outcomes = outcomes)

```


```{r}
mae_rlm_all <- list(
   `w/o covariates` = mae_rlm_woCovariate)

## Convert into data frame
mae_rlm_all_disease_merged <- do.call(
  rbind, lapply(names(mae_rlm_all), function(method_name) {
    rlm_res_method <- do.call(rbind, lapply(assay_names, function(assay_name) {
      print(paste0(method_name, assay_name))
      rlm_res_ <- mae_rlm_all[[method_name]][[assay_name]]
      rlm_res_$assay <- assay_name #strsplit(assay_name, "_")[[1]][[2]]

      # label only significant; limit to top x significant
      rlm_res_$label <- ifelse((rlm_res_$p.adj<=0.05) & (rlm_res_$rank < 4),
                               rlm_res_$feature, NA)
      rlm_res_
    }))
  }))

## Write to csv
time <- gsub(" ", "_", paste0(Sys.time()))
time <- substr(time, 1, 16)
time <- gsub(":", "_", time)
write.csv(mae_rlm_all_disease_merged, file.path(paste0('rlm_all_disease_merged_', time, '.csv')))
```


## Run linear models on corrected data
## Correct omics values with MOFA factor 2 scores
```{r}
mofa_scores_df <- read.csv(params$mofa_scores, row.names = 1)
```


## Load functions for Nested Mixed Effect model
```{r, warning=FALSE, message=FALSE}
source('Mixed_Model_functions.R')
```

```{r}
run_lm_se_corrected <- function(se, outcomes, covariates = NULL, confound_factors, mofa_scores) {
  # create data frame with omics feature values
  df <- as.data.frame(t(assays(se)[[1]]))
  # remove non-variable features
  df <- df[, apply(df, 2, function(column) {
    var(column, na.rm = T) != 0 })]
  # remove extremely sparse features
  xpercent <- nrow(df)*0.5
  df <- df[, apply(df, 2, function(column) {
    sum(column == 0, na.rm = T) < xpercent })]
  features <- colnames(df)
  # add biological variables to data frame
  # Correct for MOFA factors 1 and 2
  df <- add_confounding_factors(df, mofa_scores_df, confound_factors)
  df <- regress_out_factors(df, features, confound_factors)
  df <- cbind(df, pheno_df[rownames(df), ]) 
  # --> TODO: add mae as parameter and use experiment name to select se
  # add assay-specific technical variables to data frame
  #df <- cbind(df, colData(se))
  
  #' Apply robust linear modeling to one feature and one outcome variable
  #'
  #' @param feature A molecule identifier present in rowData(SummarizedExperiment).
  #' @param df A data frame containing assay data and biological variables.
  #' @param outcome A character string indicating biological variable column name.
  #' @param covariates A character string indicating covariate column name.
  #' @returns A data frame.
  #' @examples
  #' run_rlm_df("hmdb:HMDB0000062", df, outcome = "Age", covariates = c("Sex"))
  run_rlm_df <- function(feature, df, outcome, covariates = NULL) {
    # run rlm for one feature and one outcome
    df$y <- df[[feature]]
    # define formula
    covar_term <- ""
    if (!is.null(covariates)) {
      covar_term <- paste0(" + ", paste0(covariates, collapse = " + "))}
    str_formula <- paste0("y ~ ", outcome, covar_term)
    my_formula <- as.formula(str_formula)
    # fit robust regression model
    lm_fit <- try(lm(formula = my_formula, data = df))
    if (length(lm_fit) < 2) {
      rlm_results = data.frame(
        feature = feature,
        outcome = outcome,
        formula = str_formula,
        covariates = paste0(covariates, collapse = "+"),
        coefficient = NA, 
        std.error = NA,
        t.value = NA,
        p.value = NA)
    } else {
      # robust F-test - Wald test for multiple coefficients of rlm object
      #f_test <- f.robftest(rlm_fit, var = 2) # 1=intercept, 3:=covariates
      #f_test <- waldtest(lm_fit, terms = c(outcome))
      # results
      rlm_results = data.frame(
        feature = feature,
        outcome = outcome,
        formula = str_formula,
        covariates = paste0(covariates, collapse = "+"),
        coefficient = summary(lm_fit)$coefficients[2, "Estimate"], 
        std.error = summary(lm_fit)$coefficients[2, "Std. Error"],
        t.value = summary(lm_fit)$coefficients[2, "t value"],
        p.value = summary(lm_fit)$coefficients[2, 4],
        BIC = BIC(lm_fit),
        AIC = AIC(lm_fit)
        #p.value = f_test[2,4]
        )
    }
  }
  rlm_results <- do.call(rbind, lapply(outcomes, function(outcome) {
    rlm_result_ <- do.call(rbind, lapply(
      features,
      run_rlm_df,
      df = df, outcome = outcome, covariates = covariates))
    # adjust p-values
    rlm_result_$p.adj <- p.adjust(rlm_result_$p.value, method = "BH")
    rlm_result_$rank <- rank(rlm_result_$p.value)
    rlm_result_
  }))
  rlm_results
}
```


```{r}
# fit linear models after correction
mae_rlm_woCovariate <- lapply( # without any covariates
   experiments(mae)[assay_names], run_lm_se_corrected,
   outcomes = outcomes,
   confound_factors = c("Factor1", "Factor2"),
   mofa_scores = mofa_scores_df)


mae_rlm_all <- list(
   `w/o covariates` = mae_rlm_woCovariate)

## Convert into data frame
mae_rlm_all_corrected_merged <- do.call(
  rbind, lapply(names(mae_rlm_all), function(method_name) {
    rlm_res_method <- do.call(rbind, lapply(assay_names, function(assay_name) {
      print(paste0(method_name, assay_name))
      rlm_res_ <- mae_rlm_all[[method_name]][[assay_name]]
      rlm_res_$assay <- assay_name #strsplit(assay_name, "_")[[1]][[2]]

      # label only significant; limit to top x significant
      rlm_res_$label <- ifelse((rlm_res_$p.adj<=0.05) & (rlm_res_$rank < 4),
                               rlm_res_$feature, NA)
      rlm_res_
    }))
  }))
```

## Compare models
```{r}
source('../support_functions/Feature_plots.R')

cor_plot(mae_rlm_all_disease_merged$BIC, mae_rlm_all_corrected_merged$AIC, x_lab ='data', y_lab='corrected data')
```













## Compare disease groups invidually with healthy 

```{r}
pheno_df$phenotype[pheno_df$phenotype == 'Diseased: CHD2 CRISPR'] <- 'Diseased: CHD2'
#pheno_df$phenotype <- droplevels(pheno_df$phenotype)
pheno_df$phenotype <- relevel(as.factor(pheno_df$phenotype), ref = 'Healthy')

run_rlm_se <- function(se, outcomes, covariates = NULL) {
  
  # create data frame with omics feature values
  df <- as.data.frame(t(assays(se)[[1]]))
  # remove non-variable features
  df <- df[, apply(df, 2, function(column) {
    var(column, na.rm = T) != 0 })]
  # remove extremely sparse features
  xpercent <- nrow(df)*0.5
  df <- df[, apply(df, 2, function(column) {
    sum(column == 0, na.rm = T) < xpercent })]
  # remove features with missing values
  df <- df[,colSums(is.na(df))==0]
  features <- colnames(df)
  # add biological variables to data frame
  df <- cbind(df, pheno_df[rownames(df), ])

  # --> TODO: add mae as parameter and use experiment name to select se
  # add assay-specific technical variables to data frame
  #df <- cbind(df, colData(se))
    #' Apply robust linear modeling to one feature and one outcome variable
  #'
  #' @param feature A molecule identifier present in rowData(SummarizedExperiment).
  #' @param df A data frame containing assay data and biological variables.
  #' @param outcome A character string indicating biological variable column name.
  #' @param covariates A character string indicating covariate column name.
  #' @returns A data frame.
  #' @examples
  #' run_rlm_df("hmdb:HMDB0000062", df, outcome = "Age", covariates = c("Sex"))
  run_rlm_df <- function(feature, df, outcome, covariates = NULL) {
    # run rlm for one feature and one outcome
    df$y <- df[[feature]]
    # define formula
    covar_term <- ""
    if (!is.null(covariates)) {
      covar_term <- paste0(" + ", paste0(covariates, collapse = " + "))}
    str_formula <- paste0("y ~ ", outcome, covar_term)
    my_formula <- as.formula(str_formula)
    # fit robust regression model
    df_CHD2 <- df[df$phenotype %in% c('Healthy', 'Diseased: CHD2'),]
    lm_fit_CHD2 <- try(lm(formula = my_formula, data = df_CHD2))
    lm_fit_df_CHD2 <- as.data.frame(summary(lm_fit_CHD2)$coefficients)
    
    df_DM1 <- df[df$phenotype %in% c('Healthy', 'Diseased: DM1', 'Rescue: repeat excised CRISPR DM1'),]
    df_DM1$phenotype[df_DM1$phenotype == 'Rescue: repeat excised CRISPR DM1'] <- 'Healthy'
    lm_fit_DM1 <- try(lm(formula = my_formula, data = df_DM1))
    lm_fit_df_DM1 <- as.data.frame(summary(lm_fit_DM1)$coefficients)
    
    df_NANS <- df[df$phenotype %in% c('Healthy', 'Diseased: NANS-CDG'),]
    lm_fit_NANS <- try(lm(formula = my_formula, data = df_NANS))
    lm_fit_df_NANS <- as.data.frame(summary(lm_fit_NANS)$coefficients)

    rlm_result <- rbind(lm_fit_df_CHD2[2,],
                    lm_fit_df_DM1[2,],
                    lm_fit_df_NANS[2,])
    colnames(rlm_result) <- c('Estimate', 'std.error', 't.value', 'p.value')
    rlm_result$feature = feature
    rlm_result$outcome = rownames(rlm_result)
    rlm_result$formula = str_formula
    
    return(rlm_result)
    # lm_fit <- try(lm(formula = my_formula, data = df))
    # lm_fit_df <- as.data.frame(summary(lm_fit)$coefficients)
    # results
    # rlm_result <- do.call(rbind, lapply(
    #   c(2,3,4), function(x) {
    #     rlm_results = data.frame(
    #     feature = feature,
    #     outcome = rownames(lm_fit_df)[[x]],
    #     covariates = paste0(covariates, collapse = "+"),
    #     formula = str_formula,
    #     coefficient = lm_fit_df[x, 1], 
    #     std.error = lm_fit_df[x, 2],
    #     t.value = lm_fit_df[x, 3],
    #     p.value = lm_fit_df[x, 4]
    #   )}
    # ))
    }
    rlm_results <- do.call(rbind, lapply(outcomes, function(outcome) {
    rlm_result_ <- do.call(rbind, lapply(
      features,
      run_rlm_df,
      df = df, outcome = outcome, covariates = covariates))
    # adjust p-values
    rlm_result_$p.adj <- p.adjust(rlm_result_$p.value, method = "BH")
    rlm_result_$rank <- rank(rlm_result_$p.value)
    rlm_result_
  }))
  rlm_results
}

#assay_names <- experiment_names[c(6)]
assay_names <- experiment_names[c(1,2,3,6,9,10)]


# fit robust linear models using different formulas
mae_rlm_woCovariate <- lapply( # without any covariates
  experiments(mae)[assay_names], run_rlm_se,
    outcomes = c('phenotype'))

# mae_rlm_wCovarObsUnit <- lapply( # include sex as covariate
#   experiments(mae)[assay_names], run_rlm_se,
#   outcomes = c('phenotype'),
#   covariates = c("observation.unit.identifier"))

mae_rlm_all <- list(
  `w/o covariates` = mae_rlm_woCovariate)
  #`with sex as covariates` = mae_rlm_wCovarGender,
  #`with ObsUnit as covariate` = mae_rlm_wCovarObsUnit)

# Make into data frame
mae_rlm_all_merged_phenotypes <- do.call(
  rbind, lapply(names(mae_rlm_all), function(method_name) {
    rlm_res_method <- do.call(rbind, lapply(assay_names, function(assay_name) {
      print(paste0(method_name, assay_name))
      rlm_res_ <- mae_rlm_all[[method_name]][[assay_name]]
      rlm_res_$assay <- assay_name #strsplit(assay_name, "_")[[1]][[2]]
      
      # label only significant; limit to top x significant
      rlm_res_$label <- ifelse((rlm_res_$p.adj<=0.05) & (rlm_res_$rank < 4),
                               rlm_res_$feature, NA)
      rlm_res_
    })) 
  }))

## Write to csv
time <- gsub(" ", "_", paste0(Sys.time()))
time <- substr(time, 1, 16)
time <- gsub(":", "_", time)
write.csv(mae_rlm_all_merged_phenotypes, file.path(paste0('rlm_all_phenotypes_merged_', time, '.csv')))
```

Print top results

```{r}
# top 10 features across all outcomes
# lapply(names(mae_rlm_all), function(method_name) {
#   lapply(assay_names, function(assay_name) {
#     rlm_res <- mae_rlm_all[[method_name]][[assay_name]]
#     head(rlm_res[order(rlm_res$p.value, decreasing = FALSE),], n = 3)
#   })
# })
```


## Session information

```{r}
sessionInfo()
```


