---
title: "RLM_plot_results"
output: html_document
date: "2024-11-15"
params:
  outdir: Z:/data/iPSC_data/results/MixedModels/Plots
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  config_file: 'config.yml'
---

# Visualize results of linear modelling

## Parameters
```{r}
print(params)
outdir <- params$outdir
```

## Read data and metadata: MultiAssayExperiment
```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
```


## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

## Assay selection
```{r}
assay_names <- config$experiment_names$general_selection
assay_names
```


## Phenotypic data
```{r}
pheno_df <- as.data.frame(colData(mae))
```


## Read nested mixed models (disease-specific associations)
```{r}
MixedModel_phenotype <-  read.csv('MixedModels2/Model_phenotype.csv', row.names=1)
MixedModel_phenotype$df <- NULL
```


# Proteomics

## Get Proteomics data per disease group (no MOFA correction)
```{r}
MixedModel_phenotype_prot <- MixedModel_phenotype[MixedModel_phenotype$Assay == 'Proteomics',]
MixedModel_phenotype_prot_sign <- MixedModel_phenotype_prot[MixedModel_phenotype_prot$p.adj < 0.05,]

MixedModel_phenotype_prot_NANS <- MixedModel_phenotype_prot_sign[MixedModel_phenotype_prot_sign$outcome == 'phenotypeDiseased: NANS-CDG',]
#MixedModel_phenotype_prot_DM1 <- MixedModel_phenotype_prot_sign[MixedModel_phenotype_prot_sign$outcome == 'phenotypeDiseased: DM1',]
MixedModel_phenotype_prot_CHD2 <- MixedModel_phenotype_prot_sign[MixedModel_phenotype_prot_sign$outcome == 'phenotypeDiseased: CHD2',]

## DM1 including vs CRISPR line
MixedModel_phenotype_prot_DM1 <- read.csv('MixedModels_DM1_CRISPR_onlySampleRenames/Model_phenotype_DM1_Proteomics.csv', row.names=1)
MixedModel_phenotype_prot_DM1 <- MixedModel_phenotype_prot_DM1[MixedModel_phenotype_prot_DM1$p.adj < 0.05,]
```


## Read Proteomics (after MOFA correction)
```{r}
MixedModel_CORRECTED_Prot <- read.csv('MixedModels_CORRECTED/Model_phenotype_Proteomics.csv', row.names=1)
MixedModel_CORRECTED_Prot_SIGN <- MixedModel_CORRECTED_Prot[MixedModel_CORRECTED_Prot$p.adj < 0.05,]

MixedModel_CORRECTED_Prot_NANS <- MixedModel_CORRECTED_Prot_SIGN[MixedModel_CORRECTED_Prot_SIGN$outcome == 'phenotypeDiseased: NANS-CDG',]
#MixedModel_CORRECTED_Prot_DM1 <- MixedModel_CORRECTED_Prot_SIGN[MixedModel_CORRECTED_Prot_SIGN$outcome == 'phenotypeDiseased: DM1',]
MixedModel_CORRECTED_Prot_CHD2 <- MixedModel_CORRECTED_Prot_SIGN[MixedModel_CORRECTED_Prot_SIGN$outcome == 'phenotypeDiseased: CHD2',]

## DM1 including vs CRISPR line
MixedModel_CORRECTED_Prot_DM1 <- read.csv('MixedModels_CORRECTED/Model_phenotype_DM1_Proteomics.csv', row.names=1)
MixedModel_CORRECTED_Prot_DM1 <- MixedModel_CORRECTED_Prot_DM1[MixedModel_CORRECTED_Prot_DM1$p.adj < 0.05,]
```


```{r}
## Function to make data frame of overlapping sign hits
get_overlap_df <- function(v1, v2) {
  
  Overlap <- union(v1, v2)
  df <- data.frame(
    Feature = Overlap, 
    Uncorrected = Overlap %in% v1,
    Corrected = Overlap %in% v2
  )
  df <- df[order(-(df$Uncorrected & df$Corrected), df$Feature),]
  return(df)
}

Prot_NANS_overlap <- get_overlap_df(MixedModel_phenotype_prot_NANS$feature, MixedModel_CORRECTED_Prot_NANS$feature)
Prot_DM1_overlap <- get_overlap_df(MixedModel_phenotype_prot_DM1$feature, MixedModel_CORRECTED_Prot_DM1$feature)
Prot_CHD2_overlap <- get_overlap_df(MixedModel_phenotype_prot_CHD2$feature, MixedModel_CORRECTED_Prot_CHD2$feature)

```


## Check overlapping significant proteomics hits per disease
```{r}
library(VennDiagram)

venn.plot <- venn.diagram(
  x = list(Corrected = MixedModel_CORRECTED_Prot_CHD2$feature, Uncorrected = MixedModel_phenotype_prot_CHD2$feature),
  filename = 'CHD2_Prot.png',
  scaled=F,
  fill = c("skyblue", "pink"), 
  cex=2,
  cat.cex=2,
  cat.dist = -0.8
)
```


```{r}
library(VennDiagram)

venn.plot <- venn.diagram(
  x = list(Corrected = MixedModel_CORRECTED_Prot_NANS$feature, Uncorrected = MixedModel_phenotype_prot_NANS$feature),
  filename = 'NANS_Prot.png',
  scaled=F,
  fill = c("skyblue", "pink"), 
  cex=2,
  cat.cex=2,
  cat.dist = -0.8
)
```


```{r}
library(VennDiagram)

venn.plot <- venn.diagram(
  x = list(Corrected = MixedModel_CORRECTED_Prot_DM1$feature, Uncorrected = MixedModel_phenotype_prot_DM1$feature),
  filename = 'DM1_Prot.png',
  scaled=F,
  fill = c("skyblue", "pink"), 
  cex=2,
  cat.cex=2,
  cat.dist = -0.8
)
```





# Transcriptomics

## Get RNAseq data per disease group
```{r}
read_MixedModel_results <- function(file_path, assay_name) {
  results_df <- read.csv(file_path)
  results_df$X <- NULL
  ## Remove the CRISPR comparisons
  results_df <- results_df[!grepl('excised', results_df$outcome), ]
  if (!("nest_effect" %in% names(results_df))) {
    results_df$nest_effect <- 0
    results_df <- results_df[,c(1:8, 11,9,10)]
  }
  results_df$p.adj <- p.adjust(results_df$p.value, method = "BH")
  results_df$Assay <- assay_name
  results_df$Feature_plot <- results_df$feature #If not other feature ID available
  return(results_df)
}

MixedModel_phenotype_RNAseq <- MixedModel_phenotype[MixedModel_phenotype$Assay == 'RNAseq | RTC',]
MixedModel_phenotype_RNAseq_sign <- MixedModel_phenotype_RNAseq[MixedModel_phenotype_RNAseq$p.adj < 0.05,]

MixedModel_phenotype_RNAseq_NANS <- MixedModel_phenotype_RNAseq_sign[MixedModel_phenotype_RNAseq_sign$outcome == 'phenotypeDiseased: NANS-CDG',]
#MixedModel_phenotype_RNAseq_DM1 <- MixedModel_phenotype_RNAseq_sign[MixedModel_phenotype_RNAseq_sign$outcome == 'phenotypeDiseased: DM1',]
MixedModel_phenotype_RNAseq_CHD2 <- MixedModel_phenotype_RNAseq_sign[MixedModel_phenotype_RNAseq_sign$outcome == 'phenotypeDiseased: CHD2',]

## DM1 including vs CRISPR line
MixedModel_phenotype_RNAseq_DM1 <- read_MixedModel_results('MixedModels_DM1_CRISPR_onlySampleRenames/Model_Results_mergedRTC_DM.csv', assay_name = 'RNAseq | RTC')
MixedModel_phenotype_RNAseq_DM1 <- MixedModel_phenotype_RNAseq_DM1[MixedModel_phenotype_RNAseq_DM1$p.adj < 0.05,]
```


## Read RNA-seq data (after MOFA correction)
```{r}
MixedModel_CORRECTED_RNAseq <- read.csv('MixedModels_CORRECTED/Model_phenotype_RNAseq__RTC.csv', row.names=1)
MixedModel_CORRECTED_RNAseq_SIGN <- MixedModel_CORRECTED_RNAseq[MixedModel_CORRECTED_RNAseq$p.adj < 0.05,]

MixedModel_CORRECTED_RNAseq_NANS <- MixedModel_CORRECTED_RNAseq_SIGN[MixedModel_CORRECTED_RNAseq_SIGN$outcome == 'phenotypeDiseased: NANS-CDG',]
#MixedModel_CORRECTED_RNAseq_DM1 <- MixedModel_CORRECTED_RNAseq_SIGN[MixedModel_CORRECTED_RNAseq_SIGN$outcome == 'phenotypeDiseased: DM1',]
MixedModel_CORRECTED_RNAseq_CHD2 <- MixedModel_CORRECTED_RNAseq_SIGN[MixedModel_CORRECTED_RNAseq_SIGN$outcome == 'phenotypeDiseased: CHD2',]

## DM1 including vs CRISPR line
MixedModel_CORRECTED_RNAseq_DM1 <- read.csv('MixedModels_CORRECTED/Model_phenotype_DM1_RNAseq__Deseq.csv', row.names=1)
MixedModel_CORRECTED_RNAseq_DM1 <- MixedModel_CORRECTED_RNAseq_DM1[MixedModel_CORRECTED_RNAseq_DM1$p.adj < 0.05,]
```

## Analyze intersections of significant hits
```{r}
RNA_NANS_overlap <- get_overlap_df(MixedModel_phenotype_RNAseq_NANS$feature, MixedModel_CORRECTED_RNAseq_NANS$feature)
RNA_DM1_overlap <- get_overlap_df(MixedModel_phenotype_RNAseq_DM1$feature, MixedModel_CORRECTED_RNAseq_DM1$feature)
```



## Session information
```{r}
sessionInfo()
```
