---
title: "RLM_plot_results"
output: html_document
date: "2024-11-15"
params:
  outdir: Z:/data/iPSC_data/results/MixedModels_CORRECTED/Plots
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  config_file: 'config.yml'
---

# Visualize results of linear modelling

## Parameters
```{r}
print(params)
outdir <- params$outdir
```

## Directories of Mixed Models results
```{r}
## Normal results
MixedModels_dir <- 'MixedModelsMerged_Uncorrected'


## Corrected for MOFA factors 1 and 2
Corrected_MixedModels_dir <- 'MixedModelsMerged_Corrected'
```


## Read data and metadata: MultiAssayExperiment
```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
```


## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

## Assay selection
```{r}
assay_names <- config$experiment_names$general_selection
assay_names
```


## Phenotypic data
```{r}
pheno_df <- as.data.frame(colData(mae))
```


## Uncorrected data

Read nested mixed models (disease-specific associations)
```{r}
MixedModel_phenotype <-  read.csv(file.path(MixedModels_dir, 'Model_phenotype.csv'), row.names=1)
MixedModel_phenotype$df <- NULL
```

DM1 separate model with rescue line
```{r}
MixedModel_DM1 <- read.csv(file.path(MixedModels_dir, 'MixedModels_DM1_Rescue.csv'), row.names = 1)
MixedModel_DM1 <- MixedModel_DM1[  , match(colnames(MixedModel_phenotype), colnames(MixedModel_DM1))]
```


## Corrected data

Read nested mixed models (disease-specific associations)
```{r}
MixedModel_CORRECTED_phenotype <-  read.csv(file.path(Corrected_MixedModels_dir, 'Model_phenotype.csv'), row.names=1)
MixedModel_CORRECTED_phenotype$df <- NULL
```

DM1 separate model with rescue line
```{r}
MixedModel_CORRECTED_DM1 <- read.csv(file.path(Corrected_MixedModels_dir, 'MixedModels_DM1_Rescue.csv'), row.names = 1)
MixedModel_CORRECTED_DM1 <- MixedModel_CORRECTED_DM1[  , match(colnames(MixedModel_phenotype), colnames(MixedModel_DM1))]
```


## Compare ranks of models
```{r}
source('../support_functions/Feature_plots.R')
```
```{r}
MixedModel_DM1_Prot <- MixedModel_DM1[MixedModel_DM1$Assay == 'Proteomics',]
MixedModel_CORRECTED_DM1_Prot <- MixedModel_CORRECTED_DM1[MixedModel_CORRECTED_DM1$Assay == 'Proteomics',]

merged_df <-merge(MixedModel_DM1_Prot, MixedModel_CORRECTED_DM1_Prot, by = 'feature')

cor_plot(merged_df$rank.x, merged_df$rank.y, x_lab = 'Uncorrected_pvalue', y_lab = 'Corrected_pvalue', 'DM1')
```

```{r}
MixedModel_CHD2_Prot <- MixedModel_phenotype[MixedModel_phenotype$Assay == 'Proteomics' & MixedModel_phenotype$outcome == 'phenotypeDiseased: CHD2',]
MixedModel_CORRECTED_CHD2_Prot <- MixedModel_CORRECTED_phenotype[MixedModel_CORRECTED_phenotype$Assay == 'Proteomics' & MixedModel_CORRECTED_phenotype$outcome == 'phenotypeDiseased: CHD2',]

merged_df <-merge(MixedModel_CHD2_Prot, MixedModel_CORRECTED_CHD2_Prot, by = 'feature')

cor_plot(merged_df$rank.x, merged_df$rank.y, x_lab = 'Uncorrected_pvalue', y_lab = 'Corrected_pvalue', title = 'CHD2')
```
```{r}
MixedModel_NANS_Prot <- MixedModel_phenotype[MixedModel_phenotype$Assay == 'Proteomics' & MixedModel_phenotype$outcome == 'phenotypeDiseased: NANS-CDG',]
MixedModel_CORRECTED_NANS_Prot <- MixedModel_CORRECTED_phenotype[MixedModel_CORRECTED_phenotype$Assay == 'Proteomics' & MixedModel_CORRECTED_phenotype$outcome == 'phenotypeDiseased: NANS-CDG',]

merged_df <-merge(MixedModel_NANS_Prot, MixedModel_CORRECTED_NANS_Prot, by = 'feature')

cor_plot(merged_df$rank.x, merged_df$rank.y, x_lab = 'Uncorrected_pvalue', y_lab = 'Corrected_pvalue', title = 'NANS-CDG')
```

## Compare BIC/AIC of Corrected and Uncorrected data (Only for phenotype col)
```{r}
MixedModel_Prot <- read.csv('BIC_AIC_UNCORRECTED/Model_phenotype_Proteomics.csv', row.names=1)
MixedModel_Prot_CORRECTED <- read.csv('BIC_AIC_CORRECTED/Model_phenotype_Proteomics.csv', row.names=1)

MixedModel_MTBL_RUMC <- read.csv('BIC_AIC_UNCORRECTED/Model_phenotype_Metabolomics_TML.csv', row.names=1)
MixedModel_MTBL_RUMC_CORRECTED <- read.csv('BIC_AIC_CORRECTED/Model_phenotype_Metabolomics_TML.csv', row.names=1)

MixedModel_MTBL_LUMC <- read.csv('BIC_AIC_UNCORRECTED/Model_phenotype_Metabolomics_LUMC__total_area_normalized.csv', row.names=1)
MixedModel_MTBL_LUMC_CORRECTED <- read.csv('BIC_AIC_CORRECTED/Model_phenotype_Metabolomics_LUMC__total_area_normalized.csv', row.names=1)

MixedModel_LipT <- read.csv('BIC_AIC_UNCORRECTED/Model_phenotype_Lipidomics_Targeted__imputed__log.csv', row.names=1)
MixedModel_LipT_CORRECTED <- read.csv('BIC_AIC_CORRECTED/Model_phenotype_Lipidomics_Targeted__imputed__log.csv', row.names=1)

MixedModel_LipUnt <- read.csv('BIC_AIC_UNCORRECTED/Model_phenotype_Lipidomics_Untargeted.csv', row.names=1)
MixedModel_LipUnt_CORRECTED <- read.csv('BIC_AIC_CORRECTED/Model_phenotype_Lipidomics_Untargeted.csv', row.names=1)

MixedModel_RNA <- read.csv('BIC_AIC_UNCORRECTED/Model_phenotype_RNAseq__RTC.csv', row.names=1)
MixedModel_RNA_CORRECTED <- read.csv('BIC_AIC_CORRECTED/Model_phenotype_RNAseq__RTC.csv', row.names=1)



cor_plot_AIC_BIC <- function(uncor_data, cor_data, col1, col2, pheno, title) {
  cor_data <- cor_data[cor_data$outcome == pheno,]
  uncor_data <- uncor_data[uncor_data$outcome == pheno,]
  merged_df <-merge(cor_data, uncor_data, by = 'feature')
  plt <- cor_plot(merged_df[, col1], merged_df[, col2], x_lab = 'Uncorrected data AIC', y_lab = 'Corrected data AIC', title = paste0(title, "\n", pheno))
  return(plt)
} 


cor_plot_AIC_BIC(MixedModel_Prot, MixedModel_Prot_CORRECTED, 'AIC.x', 'AIC.y', 'phenotypeDiseased: CHD2', title = 'Proteomics')
cor_plot_AIC_BIC(MixedModel_MTBL_RUMC, MixedModel_MTBL_RUMC_CORRECTED, 'AIC.x', 'AIC.y', 'phenotypeDiseased: CHD2', title ='Metabolomics RUMC')
cor_plot_AIC_BIC(MixedModel_MTBL_LUMC, MixedModel_MTBL_LUMC_CORRECTED, 'AIC.x', 'AIC.y', 'phenotypeDiseased: CHD2', title ='Metabolomics LUMC')
cor_plot_AIC_BIC(MixedModel_LipT, MixedModel_LipT_CORRECTED, 'AIC.x', 'AIC.y', 'phenotypeDiseased: CHD2', title ='Lipidomics Targeted')
cor_plot_AIC_BIC(MixedModel_LipUnt, MixedModel_LipUnt_CORRECTED, 'AIC.x', 'AIC.y', 'phenotypeDiseased: CHD2', title ='Lipidomics Untargeted')
cor_plot_AIC_BIC(MixedModel_RNA, MixedModel_RNA_CORRECTED, 'AIC.x', 'AIC.y', 'phenotypeDiseased: CHD2', title ='RNAseq')
```

Same plots with BIC

```{r}
cor_plot_AIC_BIC(MixedModel_Prot, MixedModel_Prot_CORRECTED, 'BIC.x', 'BIC.y', 'phenotypeDiseased: CHD2', title = 'Proteomics')
cor_plot_AIC_BIC(MixedModel_MTBL_RUMC, MixedModel_MTBL_RUMC_CORRECTED, 'BIC.x', 'BIC.y', 'phenotypeDiseased: CHD2', title ='Metabolomics RUMC')
cor_plot_AIC_BIC(MixedModel_MTBL_LUMC, MixedModel_MTBL_LUMC_CORRECTED, 'BIC.x', 'BIC.y', 'phenotypeDiseased: CHD2', title ='Metabolomics LUMC')
cor_plot_AIC_BIC(MixedModel_LipT, MixedModel_LipT_CORRECTED, 'BIC.x', 'BIC.y', 'phenotypeDiseased: CHD2', title ='Lipidomics Targeted')
cor_plot_AIC_BIC(MixedModel_LipUnt, MixedModel_LipUnt_CORRECTED, 'BIC.x', 'BIC.y', 'phenotypeDiseased: CHD2', title ='Lipidomics Untargeted')
cor_plot_AIC_BIC(MixedModel_RNA, MixedModel_RNA_CORRECTED, 'BIC.x', 'BIC.y', 'phenotypeDiseased: CHD2', title ='RNAseq')
```

Check top metabolites in corrected data that are not significant in uncorrected data
```{r}
MixedModel_CHD2_MTBL_RUMC <- MixedModel_phenotype[MixedModel_phenotype$Assay == 'Metabolomics, TML' & MixedModel_phenotype$outcome == 'phenotypeDiseased: CHD2',]
MixedModel_CORRECTED_CHD2_MTBL_RUMC <- MixedModel_CORRECTED_phenotype[MixedModel_CORRECTED_phenotype$Assay == 'Metabolomics, TML' & MixedModel_CORRECTED_phenotype$outcome == 'phenotypeDiseased: CHD2',]

merged_df <-merge(MixedModel_CHD2_MTBL_RUMC, MixedModel_CORRECTED_CHD2_MTBL_RUMC, by = 'feature')
merged_df <- merged_df[merged_df$p.adj.y < 0.05,]

cor_plot(merged_df$coefficient.x, merged_df$coefficient.y, x_lab = 'Uncorrected_Coeff', y_lab = 'Corrected_Coeff', title = 'CHD2 Metabolomics coefficients')
```


# Session information
```{r}
sessionInfo()
```
