---
title: "RLM_plot_results"
output: html_document
date: "2024-11-15"
params:
  outdir: Z:/data/iPSC_data/results/RLM/
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
---

# Visualize results of linear modelling

## Parameters

```{r}
print(params)
outdir <- params$outdir
```

## Read data frames with RLM results

```{r}
mae_rlm_all_merged_phenotypes <- read.csv('RobustModels/rlm_all_phenotypes_merged_2025-01-20_22_55.csv', row.names=1)
```

## Add DNA Methylation data
```{r}
# mae_rlm_all_merged_phenotypes_Var <- read.csv('RobustModels/EMseq_VariableSet_Pheno.csv', row.names=1)
# 
# # Merge with other data frames
# mae_rlm_all_merged_phenotypes <- rbind(mae_rlm_all_merged_phenotypes, mae_rlm_all_merged_phenotypes_Var)
# 
# rm(mae_rlm_all_merged_phenotypes_Var)
```


### Remove result with covariate
```{r}
RLM_noCov <- mae_rlm_all_merged_phenotypes[mae_rlm_all_merged_phenotypes$covariates != 'observation.unit.identifier',]
```



## Read in Mixed Models results
```{r}
MixedModel_mtblmcs_tml <- read.csv('MixedModels/Model_Metabolomics_TML.csv', row.names=1)
MixedModel_prot <- read.csv('MixedModels/Model_Proteomics.csv', row.names=1)
MixedModel_mtlmbcs_lumc <- read.csv('MixedModels/Model_Metabolomics_LUMC__total_area_normalized.csv', row.names=1)
MixedModel_lip_unt <- read.csv('MixedModels/Model_Lipidomics_Untargeted__no_outlier.csv', row.names=1)
MixedModel_lip_tar <- read.csv('MixedModels/Model_Lipidomics_Targeted__imputed__log.csv', row.names=1)
MixedModel_RNAseq <- read.csv('MixedModels/Model_RNAseq.csv')
MixedModel_RNAseq$X <- NULL
MixedModel_RNAseq$p.adj <- p.adjust(MixedModel_RNAseq$p.value, method = "BH")

## EMseq data (TODO: Add to Nextflow?)
MixedModel_EMseq <- read.csv('MixedModels/Model_EMseq.csv')
MixedModel_EMseq$X <- NULL
MixedModel_EMseq$p.adj <- p.adjust(MixedModel_EMseq$p.value, method = "BH")
```


## Intersetion of the two metabolomics (TML) models
```{r}
rlm_mtblmcs_TML <- RLM_noCov[RLM_noCov$assay == "Metabolomics, TML", ] 

cor(rlm_mtblmcs_TML$coefficient, MixedModel_mtblmcs_tml$coefficient)
```


```{r}
plot(rlm_mtblmcs_TML$coefficient, MixedModel_mtblmcs_tml$coefficient)
```

## Intersection of significants
```{r}
MixedModel_mtblmcs_TML_signif <- MixedModel_mtblmcs_tml[MixedModel_mtblmcs_tml$p.adj < 0.05,]
rlm_mtblmcs_TML_signif <- rlm_mtblmcs_TML[rlm_mtblmcs_TML$p.adj < 0.05,]

MixedModel_mtblmcs_TML_signif$model_outcome <- paste0(MixedModel_mtblmcs_TML_signif$feature, MixedModel_mtblmcs_TML_signif$outcome)
rlm_mtblmcs_TML_signif$model_outcome <- paste0(rlm_mtblmcs_TML_signif$feature, rlm_mtblmcs_TML_signif$outcome)

intersection_signif <- rlm_mtblmcs_TML_signif[rlm_mtblmcs_TML_signif$model_outcome %in% MixedModel_mtblmcs_TML_signif$model_outcome,]

library(VennDiagram)

grid.newpage()

draw.pairwise.venn(area1=609, cross.area=609, area2=10087,
                   category = c('Mixed_Model', 'Simple_model'),
                   fill = c("Red", "Yellow"),)
```


## Compare Proteomics data per disease group
```{r}

analyze_ranks <- function(disease, df_RLM, df_MM){
  
  ## RLM results
  df_RLM <- df_RLM[df_RLM$outcome == disease,]
  
  ## MixedModel results
  df_MM <- df_MM[df_MM$outcome == disease,]
  
  ## Determine ranks (again)
  df_RLM$rank <- rank(df_RLM$p.value)
  df_MM$rank <- rank(df_MM$p.value)
  
  ## Plot ranks
  data_combined <- data.frame(df_RLM$rank, df_RLM$p.adj, df_MM$rank, df_MM$p.adj)
  rownames(data_combined) <- df_RLM$feature
  
  return(data_combined)
  
}

RLM_prot <- RLM_noCov[RLM_noCov$assay == 'Proteomics',]
Proteomics_CHD2 <- analyze_ranks('phenotypeDiseased: CHD2', RLM_prot, MixedModel_prot)
Proteomics_NANS <- analyze_ranks('phenotypeDiseased: NANS-CDG', RLM_prot, MixedModel_prot)
Proteomics_DM1 <- analyze_ranks('phenotypeDiseased: DM1', RLM_prot, MixedModel_prot)


RLM_mtblmcs_tml <- RLM_noCov[RLM_noCov$assay == 'Metabolomics, TML',]
MetabolomicsTML_CHD2 <- analyze_ranks('phenotypeDiseased: CHD2', RLM_mtblmcs_tml, MixedModel_mtblmcs_tml)
MetabolomicsTML_NANS <- analyze_ranks('phenotypeDiseased: NANS-CDG', RLM_mtblmcs_tml, MixedModel_mtblmcs_tml)
MetabolomicsTML_DM1 <- analyze_ranks('phenotypeDiseased: DM1', RLM_mtblmcs_tml, MixedModel_mtblmcs_tml)
```


```{r}
library(ggplot2)

plot_ranks <- function(data_combined, color_dots= "blue") {
  
  cutoff_rank_RLM <- sum(data_combined$df_RLM.p.adj <= 0.05)
  cutoff_rank_MixedModel <- sum(data_combined$df_MM.p.adj <= 0.05)
  
  title <- deparse(substitute(data_combined))
  omics <- strsplit(title, split = '_')[[1]][[1]]
  disease <- strsplit(title, split = '_')[[1]][[2]]
  
  ggp <- ggplot(data_combined, aes(x = df_RLM.rank, y = df_MM.rank)) +
    geom_point(color = color_dots, alpha =0.7) +
    geom_hline(yintercept = cutoff_rank_MixedModel, color = "black", linetype = 'dashed') +
    geom_vline(xintercept = cutoff_rank_RLM, color = "black", linetype = 'dashed') +
    labs(
      title = paste0("Compare ", omics, " feature ranks based on p-values\nDisease: ", disease),
      x = 'Robust linear model ranks',
      y = 'Mixed Model ranks'
    ) +
    theme_bw()
  return(ggp)
  
}

plot_ranks(Proteomics_CHD2)
plot_ranks(Proteomics_NANS, "red")
plot_ranks(Proteomics_DM1, "darkgreen")

plot_ranks(MetabolomicsTML_CHD2)
plot_ranks(MetabolomicsTML_NANS, "red")
plot_ranks(MetabolomicsTML_DM1, "darkgreen")
```

```{r}
plot_ranks(Proteomics_CHD2)
```


## Functions to plot interesting features 
```{r}
library(ggplot2)
library(ggrepel)
library(ggokabeito) # Okabe-Ito
library(ggpubr)


## Function to prepare data frame for boxlots of top 10 features
prepare_boxplots <- function(assay_name) {
  
  se <- experiments(mae)[assay_name]
  df <- as.data.frame(t(assays(se)[[1]]))
  df <- cbind(df, pheno_df[rownames(df),])
  return(df)
}


## Function to make boxplot of top associated -omics feature
boxplot_top_feature <- function(df_input, feature, assay_name) {
  #print(feature)
  df2plot <- df_input[, c(feature, 'phenotype')]
  colnames(df2plot) <- c("y_feature", 'group')
  
  ggp <- ggplot(df2plot, 
             aes(x = group, y = y_feature, fill = group)) +
        geom_boxplot(position = position_dodge(width = 0.9))  +
        scale_fill_okabe_ito() + 
        stat_compare_means() +
        ggtitle(stringr::str_wrap(paste0(
          feature, "\t", assay_name), 
          width = 60)) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1))
  ggsave(file.path(outdir, paste0("MixedModel_top_hit_boxplot_all_groups", feature, ".png")), ggp,
      width = 25.04, height = 17.46, units = "cm", dpi = 300)
  return(ggp)
}
```


## Read data and metadata: MultiAssayExperiment
```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')


pheno_df <- as.data.frame(colData(mae))

experiment_names <- names(experiments(mae))
assay_names <- experiment_names[c(1,2,3,9,10,11)]
```


### Which features are significant in RLMs, but the least in Mixed Models?
```{r}
highest_difference_df <- Proteomics_NANS[Proteomics_NANS$df_RLM.rank <= 30,]
features_diff <- rownames(highest_difference_df[highest_difference_df$df_MM.rank >300,])
```


## Plot interesting features
```{r}
assay_name <- "Proteomics"
df <- prepare_boxplots(assay_name)
lapply(features_diff, function(x) boxplot_top_feature(df, x, assay_name))
```


## Session information
```{r}
sessionInfo()
```
