---
title: "Pathway enrichment enrichment analysis on Linear Mixed Models results"
author:
  - Casper de Visser^[Radboud university medical center, Casper.devisser@radboudumc.nl]
output:
  html_document:
    df_print: paged
params:
    indir: Z:/data/iPSC_data/results/MixedModels/
---

## Libraries
```{r}
library(dplyr)
library(GenomicRanges)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(biomaRt)
require(ggplot2)
require(stringr)
library(multiGSEA)
library(org.Hs.eg.db)

max_num_pathways <- 20
```


## Load data
Loadings (ranks) and feature metadata
```{r}
MOFA_Loadings <- params$indir

#RNAseq_loadings <- openxlsx::read.xlsx(file.path(MOFA_Loadings, 'factorweights_RNAseq___RTC.xlsx'), rowNames = F)
#Proteomics_loadings <- openxlsx::read.xlsx(file.path(MOFA_Loadings, 'factorweights_Proteomics.xlsx'), rowNames = F)
#EMseq_loadings <- openxlsx::read.xlsx(file.path(MOFA_Loadings, 'factorweights_EMseq___100_000_most_variable_CpG_sites.xlsx'), rowNames = F)

RNAseq_loadings <- read.csv(file.path(MOFA_Loadings, 'factorweights_RNAseq___RTC.csv'), row.names = 1)
Proteomics_loadings <- read.csv(file.path(MOFA_Loadings, 'factorweights_Proteomics.csv'), row.names = 1)
EMseq_loadings <- read.csv(file.path(MOFA_Loadings, 'factorweights_EMseq___100_000_most_variable_CpG_sites.csv'), row.names = 1)
```


## Functions
```{r}
source('../support_functions/GSEA_functions.R')
```

## Map CpG sites to nearest genes (ENSEMBL ids)
```{r}
# Map gene IDs to Ensembl IDs
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
mapping <- getBM(
    attributes = c('ensembl_gene_id', 'hgnc_symbol'),
    filters = 'hgnc_symbol',
    values = unique(EMseq_loadings$Gene_symbol),
    mart = ensembl
)

# Clean mapping: keep only 1:1 hgnc -> ensembl mappings
mapping_clean <- mapping %>%
  group_by(hgnc_symbol) %>%
  dplyr::filter(n() == 1) %>%
  ungroup()

# Map using the clean 1:1 mapping (match returns NA automatically if not found)
EMseq_loadings$Ensembl <- mapping_clean$ensembl_gene_id[
  match(EMseq_loadings$Gene_symbol, mapping_clean$hgnc_symbol)
]
```


## Get pathways

Pathways for different omics are loaded with `multiGSEA` [@Canzler2020].
https://bioconductor.org/packages/release/bioc/vignettes/multiGSEA/inst/doc/multiGSEA.html

```{r, message=FALSE, warning=FALSE}
#metabolomics_id_type <- "HMDB"
proteomics_id_type <- "UNIPROT"
transcriptomics_id_type <- "ENSEMBL"

# get pathways
pathways <- getMultiOmicsFeatures(
  dbs = c('kegg', 'wikipathways', 'reactome'), #'panther', 'pharmgkb', 'reactome'
  #layer = omics_layers,
  returnTranscriptome = transcriptomics_id_type,
  returnProteome = proteomics_id_type,
  #returnMetabolome = metabolomics_id_type, 
  organism = "hsapiens",
  useLocal = F)

# see first few pathways
lapply(1:length(pathways), function(l) { head(pathways[[l]], 1) })
```


```{r} 
EMseq_loadings_clean <- as.data.frame(EMseq_loadings)

EMseq_loadings_summed <- as.data.frame(EMseq_loadings) %>%
  dplyr::filter(!is.na(Ensembl)) %>%
  group_by(Ensembl) %>%
  summarise(
    Factor2 = weighted.mean(Factor2, w = abs(Factor2)),  # weight by strength of signal
    n_CpGs  = n(),
    .groups = "drop"
  ) %>%
  as.data.frame()

# Join Gene_symbol back
gene_map <- EMseq_loadings %>%
  filter(!is.na(Ensembl)) %>%
  dplyr::select(Ensembl, Gene_symbol) %>%
  distinct(Ensembl, .keep_all = TRUE)

EMseq_loadings_summed <- EMseq_loadings_summed %>%
  left_join(gene_map, by = "Ensembl")

EMseq_loadings_summed$feature <- EMseq_loadings_summed$Ensembl
EMseq_loadings_ranks <- get_ranks_vector(EMseq_loadings_summed, "Factor2", col = 'Ensembl')
EMseq_enrichments <- multiGSEA_new(pathways, EMseq_loadings_ranks, scoreType = "std")
EMseq_enrichments_df <-EMseq_enrichments$transcriptome
```


### Add feature column
```{r}
RNAseq_loadings$feature <- rownames(RNAseq_loadings)
Proteomics_loadings$feature <- rownames(Proteomics_loadings)
EMseq_loadings_summed$feature <- EMseq_loadings_summed$Ensembl
```





## Perform GSEA for RNAseq loadings
```{r}
# Create ranks for all factors
factors <- paste0("Factor", 1:4)
RNAseq_ranks <- lapply(factors, function(f) {
  get_ranks_vector(RNAseq_loadings, f)
})
names(RNAseq_ranks) <- factors

# Run enrichment analysis for all factors
RNAseq_enrichments <- lapply(RNAseq_ranks, function(ranks) {
  multiGSEA_new(pathways, ranks)
})

# Extract transcriptome data
df_enrichments_rna <- lapply(RNAseq_enrichments, function(x) x[["transcriptome"]])
names(df_enrichments_rna) <- factors
```


## Perform GSEA for protein loadings
```{r, warning=FALSE}
# Create ranks for all factors
factors <- paste0("Factor", 1:4)
Prot_ranks <- lapply(factors, function(f) {
  get_ranks_vector(Proteomics_loadings, f)
})
names(Prot_ranks) <- factors

# Run enrichment analysis for all factors
Prot_enrichments <- lapply(Prot_ranks, function(ranks) {
  multiGSEA_new(pathways, ranks)
})

# Extract transcriptome data
df_enrichments_prot <- lapply(Prot_enrichments, function(x) x[["proteome"]])
names(df_enrichments_prot) <- factors
```



## Plot significantly enriched pathways per omics and component

Significant pathways
```{r}
get_df_signif_enrich <- function(enrichments) {
  
  ## Plot significantly enriched pathways per omics and component
  ## Extract transcriptome data and filter for significant pathways
  enrichments_signif <- lapply(names(enrichments), function(factor) {
    df <- enrichments[[factor]]
    df_signif <- prepare_sign_hits_df(df)
    if (nrow(df_signif) > 0) {
      df_signif$Factor <- factor
    }
    return(df_signif)
  })

  ## Remove empty dataframes and combine
  enrichments_signif <- enrichments_signif[sapply(enrichments_signif, nrow) > 0]

  ## Use bind_rows to handle different structures
  df_signif <- bind_rows(enrichments_signif)
  return(df_signif)
}

df_signif_rna <- get_df_signif_enrich(df_enrichments_rna)
df_signif_prot <- get_df_signif_enrich(df_enrichments_prot)
```


```{r}
## Prepare plot data for all databases
databases <- c("KEGG", "REACTOME", "WIKIPATHWAYS")
```


RNAseq
```{r}
plot_enrichments_factor <- function(df_enrich, Factor_n, assay_name) {
  
  plots <- lapply(databases, function(db) {
    prepare_plot_df(df_enrich[df_enrich$Factor == Factor_n,], db = db)
  })
  names(plots) <- databases
  
  ## Plot top pathways per database
  lapply(plots, function(plot_df) {
    plot_patwhays(plot_df, assay_name)
  })
}


plot_enrichments_factor(df_signif_rna, 'Factor1', 'RNAseq')
plot_enrichments_factor(df_signif_rna, 'Factor2', 'RNAseq')
```


Proteomics
```{r}
plot_enrichments_factor(df_signif_prot, 'Factor1', 'Proteomics')
plot_enrichments_factor(df_signif_prot, 'Factor2', 'Proteomics')
```

EMseq data
```{r}
plot_enrichments_factor(df_signif_emseq, 'Factor1', 'EMseq')
plot_enrichments_factor(df_signif_emseq, 'Factor2', 'EMseq')
```



```{r}
df_signif_rna <- translateLeadingEdgeGenes(df_signif_rna, RNAseq_loadings, 12)
df_signif_prot_F1 <- translateLeadingEdgeGenes(df_signif_prot_F1, Proteomics_loadings, 9)
df_signif_prot_F2 <- translateLeadingEdgeGenes(df_signif_prot_F2, Proteomics_loadings, 9)


EMseq_feature_metadata <- EMseq_loadings %>%
  distinct(feature, .keep_all = TRUE) %>%
  filter((!is.na(feature)))
df_signif_meth <- translateLeadingEdgeGenes(df_signif_meth, EMseq_feature_metadata, 10)
```



## Save to .csv
```{r}
df_signif_prot$leadingEdge <- NULL
df_signif_rna$leadingEdge <- NULL
df_signif_meth$leadingEdge <- NULL


write.csv(df_signif_prot, 'GSEA_MOFA_Factor2_Proteomics.csv')
write.csv(df_signif_rna, 'GSEA_MOFA_Factor2_RNAseq.csv')
write.csv(df_signif_meth, 'GSEA_MOFA_Factor2_EMseq.csv')
```


## Session information

```{r}
sessionInfo()
```


## References
