---
title: "Pathway enrichment enrichment analysis on Linear Mixed Models results"
author:
  - Casper de Visser^[Radboud university medical center, Casper.devisser@radboudumc.nl]
output:
  html_document:
    df_print: paged
params:
    indir: Z:/data/iPSC_data/results/MixedModels/
---


## Directory of Mixed Model results
```{r}
MixedModels_dir <- params$indir
```


## R libraries
```{r}
library(openxlsx)
library(dplyr)
```


## Load Mixed Models results
```{r}
dm1_rna <- openxlsx::read.xlsx(file.path(MixedModels_dir, 'DM1_P2_P3_MixedModel_Results.xlsx'), sheet = 'RNAseq')
dm1_prot <- openxlsx::read.xlsx(file.path(MixedModels_dir, 'DM1_P2_P3_MixedModel_Results.xlsx'), sheet = 'Proteomics')

nans_rna <- openxlsx::read.xlsx(file.path(MixedModels_dir, 'NANS_MixedModel_Results.xlsx'), sheet = 'RNAseq')
nans_prt <- openxlsx::read.xlsx(file.path(MixedModels_dir, 'NANS_MixedModel_Results.xlsx'), sheet = 'Proteomics')

chd2_rna <- openxlsx::read.xlsx(file.path(MixedModels_dir, 'CHD2_MixedModel_Results.xlsx'), sheet = 'RNAseq')
chd2_prt <- openxlsx::read.xlsx(file.path(MixedModels_dir, 'CHD2_MixedModel_Results.xlsx'), sheet = 'Proteomics')
```


## Get pathways

Pathways for different omics are loaded with `multiGSEA` [@Canzler2020].
https://bioconductor.org/packages/release/bioc/vignettes/multiGSEA/inst/doc/multiGSEA.html

```{r}
library(multiGSEA)
library(org.Hs.eg.db)
```


```{r, message=FALSE, warning=FALSE}
metabolomics_id_type <- "HMDB"
proteomics_id_type <- "UNIPROT"
transcriptomics_id_type <- "ENSEMBL"

# get pathways
pathways <- getMultiOmicsFeatures(
  dbs = c('kegg', 'wikipathways', 'reactome' 
          #'panther', 
          #'pharmgkb',
          
          ),
  #layer = omics_layers,
  returnTranscriptome = transcriptomics_id_type,
  returnProteome = proteomics_id_type,
  returnMetabolome = metabolomics_id_type, 
  organism = "hsapiens",
  useLocal = F)

# see first few pathways
#lapply(1:length(pathways), function(l) { head(pathways[[l]], 1) })
```


## Create named lists of ranking metric
```{r}
get_ranks_vector <- function(df) {
  # For duplicates, keep the row with the largest absolute t-value
  df_unique <- df[order(df$feature, -abs(df$t.value)), ] |>
    dplyr::distinct(feature, .keep_all = TRUE)
  
  ranks <- df_unique$t.value
  names(ranks) <- df_unique$feature
  return(ranks)
}
```


```{r}
RNAseq_ranks <- get_ranks_vector(dm1_rna)
Proteomics_ranks <- get_ranks_vector(dm1_prot)

NANS_RNAseq_ranks <- get_ranks_vector(nans_rna)
NANS_Proteomics_ranks <- get_ranks_vector(nans_prt)

CHD2_RNAseq_ranks <- get_ranks_vector(chd2_rna)
CHD2_Proteomics_ranks <- get_ranks_vector(chd2_prt)

```


## Pathway enrichment analysis

Perform pathway enrichment analysis per omics
The function `multiGSEA::multiGSEA()` [@Canzler2020] which employs
`fgsea::fgseaMultilevel()` [@Korotkevich2019] is modified to be able to change
parameters of `fgsea::fgseaMultilevel()`.

```{r, warning=FALSE}
multiGSEA_new <- function(pathways, ranks, eps = 1e-100, scoreType = "std") {
  # function modifed from multiGSEA::multiGSEA 
  # Canzler, Sebastian and Hackermüller, Jörg: multiGSEA: A GSEA-based pathway 
  # enrichment analysis for multi-omics data. BMC Bioinformatics 21, 561 (2020).
  # https://doi.org/10.1186/s12859-020-03910-x
  # allows to change parameters of fgsea::fgseaMultilevel because
  # eps=0 (boundary to calculate p-value) can result in very long running times
  # note that scoreType = "std" test both directions of ranked list
  require(fgsea)

  # Go through all omics layers
  es <- lapply(names(pathways), function(omics) {
      fgseaMultilevel(pathways = pathways[[omics]], stats = ranks, 
                      eps = eps, scoreType = scoreType)
  })

  names(es) <- names(pathways)

  return(es)
}
```


## Perofrm GSEA per omics
```{r}
RNAseq_enrichments <- multiGSEA_new(pathways, RNAseq_ranks)
Proteomics_enrichments <- multiGSEA_new(pathways, Proteomics_ranks)

NANS_RNAseq_enrichments <- multiGSEA_new(pathways, NANS_RNAseq_ranks)
NANS_Proteomics_enrichments <- multiGSEA_new(pathways, NANS_Proteomics_ranks)

CHD2_RNAseq_enrichments <- multiGSEA_new(pathways, CHD2_RNAseq_ranks)
CHD2_Proteomics_enrichments <- multiGSEA_new(pathways, CHD2_Proteomics_ranks)
```


## Function to write GSEA output to Excel
```{r, warning=FALSE}
write_filtered_excel <- function(data, file_path, patterns) {
  wb <- createWorkbook()
  
  for (pattern in names(patterns)) {
    sheet_name <- patterns[[pattern]]
    subset_data <- data[startsWith(data[[1]], pattern), ]
    
    # Skip if no matching rows
    if (nrow(subset_data) == 0) next
    
    addWorksheet(wb, sheet_name)
    writeData(wb, sheet_name, subset_data)
  }
  
  saveWorkbook(wb, file_path, overwrite = TRUE)
}

# Usage:
patterns <- c(
  "(REAC" = "REACT",
  "(WIKI" = "WIKI",
  "(KEGG" = "KEGG"
)
```


## Get dataframes of significant pathways

CHD2 vs Controls
```{r}
CHD2_Prot_Enrichments <- CHD2_Proteomics_enrichments[["proteome"]]
CHD2_Prot_Enrichments_Signif <- CHD2_Prot_Enrichments[CHD2_Prot_Enrichments$padj < 0.05,]

CHD2_RNA_Enrichments <- CHD2_RNAseq_enrichments[["transcriptome"]]
CHD2_RNA_Enrichments_Signif <- CHD2_RNA_Enrichments[CHD2_RNA_Enrichments$padj < 0.05,]
```

NANS vs Controls
```{r}
NANS_Prot_Enrichments <- NANS_Proteomics_enrichments[["proteome"]]
NANS_Prot_Enrichments_Signif <- NANS_Prot_Enrichments[NANS_Prot_Enrichments$padj < 0.05,]

NANS_RNA_Enrichments <- NANS_RNAseq_enrichments[["transcriptome"]]
NANS_RNA_Enrichments_Signif <- NANS_RNA_Enrichments[NANS_RNA_Enrichments$padj < 0.05,]
```

DM1 vs Controls
```{r}
DM1_Prot_Enrichments <- Proteomics_enrichments[["proteome"]]
DM1_Prot_Enrichments_Signif <- DM1_Prot_Enrichments[DM1_Prot_Enrichments$padj < 0.05,]

DM1_RNA_Enrichments <- RNAseq_enrichments[["transcriptome"]]
DM1_RNA_Enrichments_Signif <- DM1_RNA_Enrichments[DM1_RNA_Enrichments$padj < 0.05,]
```


### Plot significantly enriched pathways per omics and component
```{r}
require(ggplot2)
require(stringr)


max_num_pathways <- 20


prepare_sign_hits_df <- function(df) {
  # extract significantly enriched pathways
  df$rank <- -log10(df$pval)*abs(df$NES)
  df.signif <- df[df$padj < 0.05, ]

    
  if (nrow(df.signif) > 0) {
  # subset results for plotting if too many significant pathways
  
      # Remove pathways that share the same leading edge genes
      df.signif <- df.signif[!duplicated((as.character(df.signif$leadingEdge))),]
      
      # Add column indicating negative or positive enrichment
      df.signif$Enrichment <- NA
      enrichment_col <- lapply(df.signif$NES, function(x){
      if (x > 0) {
         value <- 'Positively associated'
      }
      else {
         value <- 'Negatively associated'
      }
      value
      })
      df.signif$Enrichment <- as.character(enrichment_col)
      
      if (nrow(df.signif) > max_num_pathways) {
      # Make separate df for negative and positive enrichments
        df.signif_pos <- df.signif[df.signif$NES > 0, ]
        df.signif_neg <- df.signif[df.signif$NES < 0, ]
        
        # Order based on NES
        df.signif_pos <- df.signif_pos[order(NES, decreasing = T),][1:5,]
        df.signif_neg <- df.signif_neg[order(NES, decreasing = F),][1:5,]
        
        # Bind data frames
        df.signif <- rbind(df.signif_pos, df.signif_neg)
        df.signif <- na.omit(df.signif)
      }
      
      # Add Gene ratio column
      gene_ratios <- lapply(seq(1, length(df.signif$leadingEdge)), function(x) { 
      leadingedge <- df.signif[x,8]
      leadingedge_n <- length(leadingedge[[1]][[1]])
      size <- df.signif[x,7]
      gene_ratio <- as.numeric(leadingedge_n) / as.numeric(size) 
      round(gene_ratio, digits = 3)
      })
      df.signif$GeneRatio <- as.numeric(as.character(gene_ratios))
  

      #order pathways by normalized enrichment score
      df.signif$pathway <- factor(
        stringr::str_wrap(df.signif$pathway, width = 50),
        levels = stringr::str_wrap(df.signif$pathway, width = 50)[order(df.signif$NES)],
        ordered = T)
  return(df.signif)
  }
}
  
  
df_signif_rna <- prepare_sign_hits_df(RNAseq_enrichments[["transcriptome"]])  
df_signif_prot <- prepare_sign_hits_df(Proteomics_enrichments[["proteome"]])  

NANS_df_signif_rna <- prepare_sign_hits_df(NANS_RNAseq_enrichments[["transcriptome"]])  
NANS_df_signif_prot <- prepare_sign_hits_df(NANS_Proteomics_enrichments[["proteome"]])  

CHD2_df_signif_rna <- prepare_sign_hits_df(CHD2_RNAseq_enrichments[["transcriptome"]])  
CHD2_df_signif_prot <- prepare_sign_hits_df(CHD2_Proteomics_enrichments[["proteome"]])  
```


## Plot significant pathways
```{r}
plot_patwhays <- function(df.signif, title) {

  # Plot NES
  ggp <- ggplot(df.signif) +
      geom_segment(aes(y = pathway, x = GeneRatio, xend = 0, yend = pathway)) +
      geom_point(aes(y = pathway, x = GeneRatio, fill=padj, size = size), shape = 21) +
      scale_fill_viridis_c() +
      facet_grid(Enrichment ~ ., scales = "free", space = "free") +
      ggtitle(title)
      
  return(ggp)
}

plot_patwhays(df_signif_prot, 'Proteomics (DM1)')
plot_patwhays(df_signif_rna, 'RNAseq (DM1)')

plot_patwhays(NANS_df_signif_prot, 'Proteomics (NANS-CDG)')
plot_patwhays(NANS_df_signif_rna, 'RNAseq (NANS-CDG)')

plot_patwhays(CHD2_df_signif_prot, 'Proteomics (CHD2)')
plot_patwhays(CHD2_df_signif_rna, 'RNAseq (CHD2)')

```



## LeadingEdge genes, translate ENSEMBL to Gene Symbols (mRNA-seq)
```{r}
#proteomics_feature_metadata <- read.csv('C:/Users/Z896216/Documents/proteomics_feature_metadata.csv', row.names=1)
#proteomics_feature_metadata$feature <- proteomics_feature_metadata$protein_accession

translateLeadingEdgeGenes <- function(df, feature_metadata, col_nr) {
  leadingEdge_col <- df$leadingEdge

  leadingEdge_symb <- lapply(leadingEdge_col, function(x) {
          
      ## Remove "\n" that is present in some ENSEMBL IDs
      x <- (gsub("\n", "", x))
          
      leading_genes <- x
      symbols <- lapply(leading_genes, function(x) {
        sym <- feature_metadata[feature_metadata$feature==x, col_nr] # Use result dataframe with gene symbols
        sym <- (gsub("\n", "", sym))
        sym
      }) 
      list(as.character(symbols))
        
      })
  df$leadingEdge_Symbols <- as.character(leadingEdge_symb)
  return(df)
}


DM1_RNA_Enrichments_Signif <- translateLeadingEdgeGenes(DM1_RNA_Enrichments_Signif, chd2_rna, 10)
DM1_Prot_Enrichments_Signif <- translateLeadingEdgeGenes(DM1_Prot_Enrichments_Signif, chd2_prt, 11)

NANS_RNA_Enrichments_Signif <- translateLeadingEdgeGenes(NANS_RNA_Enrichments_Signif, chd2_rna, 10)
NANS_Prot_Enrichments_Signif <- translateLeadingEdgeGenes(NANS_Prot_Enrichments_Signif, chd2_prt, 11)

CHD2_RNA_Enrichments_Signif <- translateLeadingEdgeGenes(CHD2_RNA_Enrichments_Signif, chd2_rna, 10)
CHD2_Prot_Enrichments_Signif <- translateLeadingEdgeGenes(CHD2_Prot_Enrichments_Signif, chd2_prt, 11)
```



## Save to .csv
```{r}
library(openxlsx)

out_dir <- 'C:/Users/Z896216/Documents/GSEA_MixedModels/'
dir.create(out_dir)

write_filtered_excel(CHD2_Prot_Enrichments_Signif, paste0(out_dir, 'CHD2_Proteomics.xlsx'), patterns)
write_filtered_excel(CHD2_RNA_Enrichments_Signif, paste0(out_dir, 'CHD2_RNAseq.xlsx'), patterns)

write_filtered_excel(DM1_RNA_Enrichments_Signif, paste0(out_dir, 'CHD2_RNAseq.xlsx'), patterns)
write_filtered_excel(DM1_Prot_Enrichments_Signif, paste0(out_dir, 'DM1_Proteomics.xlsx'), patterns)

write_filtered_excel(NANS_Prot_Enrichments_Signif, paste0(out_dir, 'NANS_Proteomics.xlsx'), patterns)
write_filtered_excel(NANS_RNA_Enrichments_Signif, paste0(out_dir, 'NANS_RNAseq.xlsx'), patterns)
```




### Aggregate p-values across omics

`MultiGSEA::combinePvalues()` aggregates p-values across all omics types.
```{r}
aggregated_pval <- lapply(components, function(comp) {
  df <- extractPvalues(enrichmentScores = enrichment_scores[[comp]],
                       pathwayNames = names(pathways[[1]])) # first omics

  df$combined_pval <- combinePvalues(df)
  df$combined_padj <- p.adjust(df$combined_pval, method = "BH")
  
  df <- cbind(data.frame(pathway = names(pathways[[1]])), df)
  df
})

# top 10 pathways based on aggregated p-values 
lapply(aggregated_pval, function(df) {
  df[order(df$combined_pval), ][1:max_num_pathways, ] 
})
```


# Find genes nearest to methylation sites
```{r, warning=FALSE}
library(dplyr)

ensembl_genes <- read.csv('ensembl_genes.csv', row.names = 1)

mofa_factor_sub <- read.csv('MOFA/factorweights_EM-seq___cell-type_adjusted___100_000_most_variable_CpG_sites.csv', row.names=1)


mofa_factor_per_chromosome <- lapply(seq(1:22), function(x) {
     sub <- mofa_factor_sub[mofa_factor_sub$chr==x,]
 })

genes_per_chromosome <- lapply(seq(1:22), function(x) {
     df_sub <- ensembl_genes[ensembl_genes$chromosome_name==x,]
 })


mofa_factors_with_ensembl <- lapply(seq(1:22), function(x) {
    genes <- genes_per_chromosome[[x]]
    mofa_sub <- mofa_factor_per_chromosome[[x]]
    
    test <- lapply(mofa_sub$start, function(x) {
      nearest_row <-which(abs(genes$start_pos - x) == min(abs(genes$start_pos - x)))
      nearest_gene <- genes[nearest_row, 1]
      #return(nearest_gene)
    })
    mofa_sub$ENSEMBL <- test
    return(mofa_sub)
})
rm(ensembl_genes)
rm(mofa_factor_sub)
rm(genes_per_chromosome)
rm(mofa_factor_per_chromosome)

combined <- bind_rows(mofa_factors_with_ensembl)
rm(mofa_factors_with_ensembl)
combined <- combined[1:1000,] # Smaller set to test
combined$Accession <- unlist(combined$ENSEMBL)
combined$ENSEMBL <- combined$Accession

input_data$`EM-seq` <- combined
```


## Session information

```{r}
sessionInfo()
```


## References
