---
title: 'Prepare chunks to anallyze large omics data sets with mixed linear models'
author:
  - Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  config_file: config.yml
  outdir: Z:/data/iPSC_data/results/RLM/
---

## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!dir.exists(outdir)){
    outdir <- getwd()
}
```

## Read data and metadata: MultiAssayExperiment
```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
mae
```


## Select assay from config file
```{r}
config <- yaml::read_yaml(params$config_file)
assay_name <- config$experiment_names$chunks_mixed_model
```


Technical variables are available per assay

```{r}
tech_vars_factor <- lapply(experiments(mae), function(x) {
  num_levels <- apply(colData(x), 2, function(x) {length(levels(as.factor(x)))})
  t <- names(colData(x))[num_levels <= 9 & num_levels > 1]
  t <- t[t!="subject.id"]
  t <- t[!t %in% names(colData(mae))]
})
tech_vars_factor

tech_vars_continuous <- lapply(experiments(mae), function(x) {
  num_levels <- apply(colData(x), 2, function(x) {length(levels(as.factor(x)))})
  t <- names(colData(x))[num_levels > 9]
  t <- t[t!="subject.id"]
  t <- t[t!="measurement.id"]
  t <- t[t!="measurement_names"]
  t <- t[t!="ID.Patient"]
  t <- t[!t %in% c("SampleID", "Radboud.code", "Kevin.nr", "Kevin.CPM", 
                   "Sample.name", "ISA_name", "SampleID_withrep", "Sample_Overview_names")] #TODO, remove these from sample metadata sheet
  t <- t[!t %in% c("GroupNum", "GroupName", "SampleNorm", "Radboud.code.1")] 

  t <- t[!t %in% names(colData(mae))]
})
tech_vars_continuous
```


Phenotype information shared across all assays can be accessed via `colData()`

```{r sample_information_overview}
names(colData(mae))

pheno_df <- as.data.frame(colData(mae))

## Add column for Disease vs Healthy comparison
new_column <- lapply(pheno_df$phenotype, function(x) {
    if (startsWith(x, 'Diseas')) {
        value <- 'Disease'
    }
    else{
        value <- 'Healthy'
    }
})
pheno_df$Disease <- as.character(new_column)
```


## Load functions for Nested Mixed Effect model
```{r}
source_paths <- c(
  "../support_functions/Mixed_Model_functions.R",      # when running in Rstudio
  "Mixed_Model_functions.R"                    # when running in Nextflow
)

for (path in source_paths) {
  if (file.exists(path)) {
    source(path)
    break
  }
}
```    


## Split data into chunks, add phenotypic columns and save to .csv files
```{r}
se <- experiments(mae)[assay_name]

## Prepare data for splitting
# Prepare data
data_prep <- prepare_feature_data(assay_name, mae)
df <- data_prep$df
features <- data_prep$features

## Split in chunks
chunks_df <- split_in_chunks(df, 100)

## Save chunks
lapply(1:length(chunks_df), function(x) {
  chunk <- add_pheno_cols(chunks_df[[x]])
  write.csv(chunk, paste0('Chunk_', x, '.csv'))
})
```


## Session information
```{r}
sessionInfo()
```


