---
title: 'Linear Mixed models'
author:
  - Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  mofa_scores: Z:/data/iPSC_data/Results/MOFA/All_Samples/SampleScores.csv
  outdir: Z:/data/iPSC_data/results/RLM/
  config_file: "config.yml"
---

## Libraries
```{r}
library(lme4)
library(dplyr)
library(openxlsx)
```


## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir
```

## Read data and metadata: MultiAssayExperiment

```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
#mae
```


The MultiAssayExperiment contains individual omics assays

## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```


## Assay selection
```{r}
experiment_names <- config$experiment_names$general_selection
```


Phenotype information shared across all assays can be accessed via `colData()`

```{r sample_information_overview}
names(colData(mae))

pheno_df <- as.data.frame(colData(mae))

## Add column for Disease vs Healthy comparison
new_column <- lapply(pheno_df$phenotype, function(x) {
    if (startsWith(x, 'Diseas')) {
        value <- 'Disease'
    }
    else{
        value <- 'Healthy'
    }
})
pheno_df$Disease <- as.character(new_column)
```


## Correct omics values with MOFA factor 2 scores
```{r}
mofa_scores_df <- read.csv(params$mofa_scores, row.names = 1)
```


## Load functions for Nested Mixed Effect model
```{r, warning=FALSE, message=FALSE}
source('Mixed_Model_functions.R')
```


```{r}
experiment_names
```


## Run Mixed Effect model, comparing individual disease groups with the control lines
```{r}
Predictor <- 'phenotype'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model_CORRECTED(a, pred=Predictor, confound_factors = c("Factor1", "Factor2"))
  model_result <- model_result[model_result$outcome == 'phenotypeDiseased: CHD2',]
  model_result$p.adj <- p.adjust(model_result$p.value, method = "BH")
  model_result
})
merged_result_CHD2 <- do.call(rbind, mae_rlm_RandomEffect)
```

```{r}
Predictor <- 'phenotype'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model_CORRECTED(a, pred=Predictor, confound_factors = c("Factor1", "Factor2"))
  model_result <- model_result[model_result$outcome == 'phenotypeDiseased: NANS-CDG',]
  model_result$p.adj <- p.adjust(model_result$p.value, method = "BH")
  model_result
})
merged_result_NANS <- do.call(rbind, mae_rlm_RandomEffect)
```


## Run Mixed Effect model, comparing all disease with all controls
```{r}
Predictor <- 'Disease'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model_CORRECTED(a, pred=Predictor, confound_factors = c("Factor1", "Factor2"))
  model_result
})
merged_result_Disease <- do.call(rbind, mae_rlm_RandomEffect)
```


## Run Mixed Effect model, comparing female and male samples
```{r}
Predictor <- 'sex'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model_CORRECTED(a, pred=Predictor, confound_factors = c("Factor1", "Factor2"))
  model_result
})
merged_result_Sex <- do.call(rbind, mae_rlm_RandomEffect)
```


## Compare DM1 with control lines inculding CRISPR Rescue line
```{r}
Predictor <- 'phenotype_DM1'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model_CORRECTED(a, pred=Predictor, confound_factors = c("Factor1", "Factor2"), DM1_ONLY=T)
  model_result
})
merged_result_DM1 <- do.call(rbind, mae_rlm_RandomEffect)
```


## Only compare patient 2 & 3 for DM1
Remove DM patient 1 from MAE object
```{r}
DM1_p1_names <- c(
  'IPS20_00122_1',
  'IPS20_00122_2',
  'IPS20_00122_3',
  'IPS23_00090_1', #Rescue
  'IPS23_00090_2', #Rescue
  'IPS23_00090_3' #Rescue
)

sample_names <- rownames(colData(mae))
sample_names_NoDMP1 <- sample_names[!sample_names  %in% DM1_p1_names]

mae <- subsetByColData(mae, y = sample_names_NoDMP1)
```

```{r}
Predictor <- 'phenotype_DM1'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model_CORRECTED(a, pred=Predictor, confound_factors = c("Factor1", "Factor2"), DM1_ONLY=T)
  model_result
})
merged_result_DM1_P2P3 <- do.call(rbind, mae_rlm_RandomEffect)
```


Merge all results and save to .csv
```{r}
merged_results_all <- rbind(
  merged_result_CHD2,
  merged_result_NANS,
  merged_result_DM1,
  merged_result_DM1_P2P3,
  merged_result_Disease,
  merged_result_Sex
  )

write.csv(merged_results_all, 'MixedModelsResults_Corrected.csv')
```


Save tables per contrast per omics including feature metadata
```{r}
save2excel <- function(df, namefile) {
  split_df <- split(df, df$Assay)
  wb <- createWorkbook()
  names(split_df) <- experiment_names
  for (omics in names(split_df)) {
    assay_name <- omics
    omics <- strsplit(omics, " \\| ")[[1]][1]
    addWorksheet(wb, sheetName = omics)
    data <- split_df[[assay_name]][, c( "feature","coefficient","std.error","t.value","p.value","rand_effect","nest_effect","p.adj")]
    data$rank <- rank(data$p.value)
    ## Add feature metadata
    feature_metadata <- data.frame(rowData(mae[[assay_name]]))
    feature_metadata$feature <- rownames(feature_metadata)
    data <- data %>%
        left_join(feature_metadata, by = 'feature')
    writeData(wb, sheet = omics, data)
  }
  saveWorkbook(wb, paste0(namefile, "_MixedModel_Results_CORRECTED.xlsx"), overwrite = T)
}


save2excel(merged_result_CHD2, 'CHD2')
save2excel(merged_result_NANS, 'NANS')
save2excel(merged_result_DM1_P2P3, 'DM1_P2_P3')
save2excel(merged_result_DM1, 'DM1')
save2excel(merged_result_Disease, 'All_diseases_combined')
save2excel(merged_result_Sex, 'Sex')
```

## Session information

```{r}
sessionInfo()
```


