---
title: "Features_overview"
author:
- Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  outdir: Z:/data/iPSC_data/results/PCA/
  config_file: 'config.yml'
---


## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!dir.exists(outdir)){
    outdir <- getwd()
}
```

## Libraries
```{r}
library(gridExtra)
library(ggplot2)
library(dplyr)
suppressMessages(library(MultiAssayExperiment))
suppressMessages(library(HDF5Array))
```


## Read input data 

Input data is a MultiAssayExperiment (MAE) object. 
```{r}
mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
mae
```


## Functions used to retrieve data from MAE
```{r}
# Function to read omics data from MAE into dataframe
read_omics_from_mae <- function(mae_object, name_assay){
  se <- mae_object[[name_assay]]
  df_wide <- data.frame(t(assay(se)))
  df_wide <- df_wide[ , colSums(is.na(df_wide))==0]
  df_wide
  
}
```


## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

## Assay selection
```{r}
experiment_names <- config$experiment_names$general_selection
experiment_names
```


## Make list of assays
```{r}
assay_list <- lapply(experiment_names, function(assay_name) {
  df_wide <- read_omics_from_mae(mae, assay_name)
  }) 
names(assay_list) <- experiment_names
```


## Make phenotype dataframe
```{r}
pheno_df <- as.data.frame(colData(mae))

## Add column to compare only disease vs. healthy
disease_healthy_col <- lapply(pheno_df$phenotype, function(x) {
    
    if (startsWith(x, 'Healthy')) {
        x <- x
    }
    else if (startsWith(x, 'Diseased')) {
        x <- 'Diseased'
    }
    else if (startsWith(x, 'Rescue')) {
        x <- 'Healthy'
    }
    x
})
pheno_df$Disease <- as.character(disease_healthy_col)
```


## Functions used for plotting the CVs
```{r, fig.width = 20}

## Function that calculates the Coefficient of Variation (%)
cv <- function(x) {
  if (mean(x) != 0) {
    sd(x) / mean(x) * 100
  } else {
    NA
  }
}


# Function to make density plot without sample groups
density_plot_CV <- function(item) {
  
  # -omics data
  input_df <- assay_list[[item]]
  
  # add minimum value to all values + 1, preventing negative values
  input_df <- input_df + (abs(min(input_df)) + 1)
  
  # Calculate Coefficients of Variance
  CV_list <- sapply(input_df, cv)
  CVs_df <- data.frame(CV_list)
  colnames(CVs_df) <- c('CV')
  
  p <- ggplot(CVs_df, aes(x=CV)) +
    geom_density(color = 'darkblue', fill='lightblue') +
    ggtitle(paste0(item))+
    theme_grey(base_size=22)
  return(p)
}


## Function to sort phenotype column according to -omics df order
sort_pheno_group_col <- function(df_pheno, df_omics, pheno_col) {
  
  pheno_values <- lapply(rownames(df_omics), function(x){
    value <- df_pheno[x, pheno_col]
  })
  pheno_values <- as.character(pheno_values)
  return(pheno_values)
}


## Function to calculate CV per group
compute_cv_by_category <- function(df){
    df %>%
        group_by(Group) %>%
        summarise(across(where(is.numeric), cv))
}


# Function to plot CVs per phenotype group
density_plot_CV_groups <- function(assay_name, df_pheno, pheno_col) {
  
  # Get -omics data frame
  input_df <- assay_list[[assay_name]]
  
  # add minimum value to all values + 1, preventing negative values
  input_df <- input_df + (abs(min(input_df)) + 1)
  
  # Add column for pheno group
  pheno_group_col <- sort_pheno_group_col(df_pheno, input_df, pheno_col)
  input_df$Group <- pheno_group_col
  
  # Calculate CV
  df_grouped <- compute_cv_by_category(input_df)
  df_melted <- reshape2::melt(df_grouped)
  
  # Number of features
  n_features <- ncol(df_grouped)
  
  p <- ggplot(df_melted, aes(x=value, fill=Group)) +
     geom_density(color = 'darkblue', alpha=0.4) +
     ggtitle(paste0(assay_name, '\n Number of features: ', n_features)) +
    theme_grey(base_size=22)
  return(p)
}
```


## Plot CVs without groups
```{r, fig.width=20}
lapply(experiment_names, density_plot_CV)
```


## Plot CVs per phenotype group
```{r, fig.width=20}
lapply(experiment_names, function(x) density_plot_CV_groups(x, pheno_df, 'phenotype') )
```


## Plot CVs for males/females
```{r, fig.width=20}
plot_list <- lapply(experiment_names, function(x) density_plot_CV_groups(x, pheno_df, 'sex') )
plot_list
```

## Plot CVs comparing diseased and healthy groups
```{r, fig.width=20}
plot_list_pheno <- lapply(experiment_names, function(x) density_plot_CV_groups(x, pheno_df, 'Disease') )
plot_list_pheno
```


## Grid arrange all plots together into PDF
```{r}
n <- length(plot_list)
nrow <- ceiling(n /2)
pdf("arranged_CVplot_Sex.pdf", width = 20, height = 3 * nrow)
do.call(grid.arrange, c(plot_list, ncol = 2, padding=unit(20, "cm")))
dev.off()
```


```{r}
n <- length(plot_list_pheno)
nrow <- ceiling(n /2)
pdf("arranged_CVplot_Pheno.pdf", width = 20, height = 3 * nrow)
do.call(grid.arrange, c(plot_list_pheno, ncol = 2, padding=unit(20, "cm")))
dev.off()
```
