---
title: 'Multi-omics Factor Analysis (MOFA)'
author:
- Casper de Visser^[Radboud university medical center, casper.devisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  outdir: Z:/data/iPSC_data/results/MOFA
  config_file: config.yml
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
---

This notebook performs Multi-omics Factor Analysis (MOFA) [@Argelaguet2018,@Argelaguet2020] on preprocessed data. 

MOFA is an unsupervised method that uncovers main sources of variation in multi-omics data. It allows to identify variation shared across omics data types or specific to individual omics data types. 
Subsequent to MOFA, correlation between MOFA factors and phenotypic information is determined to identify possible biological factors contributing to the observed variation.


```{r show_parameters}
if (paste0(R.Version()$major, ".", R.Version()$minor) != "4.2.1") {
  warning("This R notebook was implemented using R version 4.2.1")
}
print(params)
dir.create(params$outdir)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!dir.exists(outdir)){
    outdir <- getwd()
}
```


## Libraries
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(ggokabeito)
```


## Read input data

Input data is a MultiAssayExperiment (MAE) object. All or specified assays in the MAE will be used for analysis Multi-omics Factor Analysis (MOFA).

```{r read_input_data, message=FALSE}
library(MultiAssayExperiment)
library(HDF5Array)

mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')

```


Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

Assay selection
```{r}
assay_names <- config$experiment_names$general_selection
#assay_names <- config$experiment_names$Annotated_data
names(assay_names) <- assay_names
assay_names
```


## Phenotype data
```{r}
phenotype_data <- as.data.frame(colData(mae))
phenotype_data$sample <- rownames(phenotype_data)
## Add column for Disease vs Healthy comparison
new_column <- lapply(phenotype_data$phenotype, function(x) {
    if (startsWith(x, 'Diseas')) {
        value <- 'Disease'
    }
    else{
        value <- 'Healthy'
    }
})
## Add 2nd column of phenotype with CHD2 CRISPR line in CRISPR group
phenotype_data$phenotype_CHD2_CRISPR <- phenotype_data$phenotype
phenotype_data$phenotype_CHD2_CRISPR[phenotype_data$phenotype_CHD2_CRISPR == 'Diseased: CHD2 CRISPR'] <- 'Diseased: CHD2'

## Add 3rd column of phenotype with CRISPR Rescue line in Healthy group
phenotype_data$phenotype_CRISPRS_both <- phenotype_data$phenotype_CHD2_CRISPR
phenotype_data$phenotype_CRISPRS_both[phenotype_data$phenotype_CRISPRS_both == 'Rescue: repeat excised CRISPR DM1'] <- 'Healthy'

phenotype_data$Disease <- as.character(new_column)


## Add columns as factor
phenotype_data <- phenotype_data %>%
  mutate(across(where(is.character), as.factor))
```


## Select samples (for disease specific comparisons)
```{r}
select_samples <- phenotype_data[phenotype_data$phenotype_CRISPRS_both %in% c('Healthy', 'Diseased: DM1'),]
mae_subset <- mae[, rownames(select_samples)]
mae_subset
```

## Multi-omics factor analysis (MOFA)

For running MOFA, it is recommended to first remove known batch effects. 
Make sure to use input data that were adjusted for batch effect if necessary.

```{r load_mofa, message=FALSE}
library(MOFA2)
library(data.table)
```

Prepare MOFA input data: create a long data frame with sample, feature, view, 
[group], and value. 

```{r functions_for_mofa_input, message=FALSE}
#' Convert omics data to long format
wide_to_long_keep_samples <- function(df) {
  df$sample <- rownames(df)
  gather(df, key = "feature", value = "value", -sample, factor_key = TRUE)
}

#' Prepare MultiAssayExperiment for MOFA
#' If both hematology and phenotype data are selected, they are merged to one data modality.
#'
#' @param mae A MultiAssayExperiment.
#' @param assay_selection An optional character vector with assay names to be included in analysis.
#' @returns A long data frame.
mae_to_MOFAinput <- function(mae) {


  # get omics data from SummarizedExperiments
  assay_list <- lapply(assay_names, function(assay_name) {
    se <- mae[[assay_name]]
    df_wide <- data.frame(t(assay(se)))
    # standardize data
    df_wide <- as.data.frame(t(scale(t(df_wide), center = TRUE, scale = TRUE)))
    # convert wide to long data frame
    df_long <- wide_to_long_keep_samples(df_wide)
    # shorten assay name of targeted metabolomics assays
    assay_shortname <- ifelse(
      grepl("_metabolite_profiling", assay_name),
      strsplit(assay_name, "_")[[1]][2],
      assay_name)
    df_long$view <- assay_shortname
    df_long
  }) 
  merged_assay_data <- do.call(rbind, assay_list)

}
```


```{r prepare_mofa_input}
mofa_input <- mae_to_MOFAinput(mae_subset)

# shorter names
shorter_names <- lapply(mofa_input$view, function(x) {
  x <-strsplit(x, " \\| ", x)[[1]][1]
})

mofa_input$view <- as.character(shorter_names)
#mofa_input$view <- as.character(assay_names)

mofa_input$value <- as.numeric(mofa_input$value)
```

Create the MOFA object

```{r create_mofa_object}
MOFAobject <- create_mofa(mofa_input)
print(MOFAobject)
```


Plot views available in MOFA object (N = number of samples, D = number of 
features per view)

```{r show_mofa_modalities}
require(ggplot2)
ggp <- plot_data_overview(MOFAobject)

ggsave(file.path(outdir, "MOFA_modality_overview.pdf"), 
       ggp, width = 15, height = 10.46, units = "cm", dpi = 300)

ggp
```


Get parameters of MOFA object

```{r}
data_opts <- get_default_data_options(MOFAobject)
data_opts
```

Set number of factors. This parameter can have a large impact on the results.

```{r}
model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 6
model_opts
```


```{r}
train_opts <- get_default_training_options(MOFAobject)
train_opts
```

Prepare MOFA object for training using the parameters specified above

```{r}
MOFAobject <- prepare_mofa(
  object = MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
```

The conda environment with Python can now be used by `reticulate` in R.

```{r define_python_version}
#reticulate::use_condaenv(condaenv = "mofapy") #Rstudio
reticulate::use_python("/usr/bin/python3", required = T) #Nextflow/Singulatiy
```

Train the MOFA model

```{r run_mofa, message=FALSE}
outdir <- params$outdir
outfile = file.path(outdir, "model_DM1.hdf5")
MOFAobject.trained <- run_mofa(MOFAobject, outfile)
```


## Visualizations of MOFA results 

### Variance decomposition

Amount of variance explained per modality (= view, = assay)

```{r}
head(MOFAobject.trained@cache$variance_explained$r2_total)
```

Amount of variance explained by each factor per modality

```{r}
head(MOFAobject.trained@cache$variance_explained$r2_per_factor)
```

Plot amount of variance explained by each factor per modality

```{r, fig.width=10, fig.height=5}
expl_var <- c(
  get_variance_explained(MOFAobject.trained)$r2_per_factor$single_group)

ggp <- plot_variance_explained(MOFAobject.trained) +
  geom_text(
    label = round(expl_var, 2),
    color = ifelse(expl_var < 0.6*max(expl_var), "black", "white")) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )
      

ggsave(file.path(outdir, "MOFA_explained_variance.pdf"), 
       ggp, width = 14, height = 10.46, units = "cm", dpi = 300)

ggp
```

###  Factor correlations

Plot correlations between factors. When factors are correlated (containing 
redundant information) it is recommended to reduce the number of factors

```{r}
plot_factor_cor(MOFAobject.trained)
```

### Factor weights

Plot feature weights per factor and assay

```{r}
num_factors <- nrow(
  MOFAobject.trained@cache$variance_explained$r2_per_factor[[1]])
view_names <- names(MOFAobject.trained@data)

lapply(1:5, function(f) {
  lapply(view_names, function(a) {
    plot_weights(
      MOFAobject.trained,
      view = a,
      factor = f,
      nfeatures = 10
      )+ ggtitle(paste0(a))
  }) 
})
```

Top weights per factor and assay

```{r}
lapply(1:5, function(f) { 
  lapply(view_names, function(a) {
    plot_top_weights(
      MOFAobject.trained,
      view = a,
      factor = f,
      nfeatures = 10
    ) + ggtitle(paste0(a))
  })
})
```


# Compare disease groups on MOFA scores

## Add phenotype data to MOFA object
```{r}
samples_metadata(MOFAobject.trained) <- select_samples
```


## Plot MOFA scores per category
```{r}
plot_factor_category <- function(MOFAobject.trained, 
                                 var, 
                                 excluded_categories = c(),
                                 num_factors = NULL) {
  
    
  if (is.factor(MOFAobject.trained@samples_metadata[[var]])) {
    original_categories <- levels(MOFAobject.trained@samples_metadata[[var]])
  } else {
    original_categories <- levels(
      as.factor(MOFAobject.trained@samples_metadata[[var]]))
  }
  
  categories <- original_categories[
    !original_categories %in% excluded_categories]
  if (length(categories) > 1) {
    comparisons <- as.list(as.data.frame(combn(categories, 2)))
  }
  
  num_factors <- ifelse(
    is.null(num_factors),
    nrow(MOFAobject.trained@cache$variance_explained$r2_per_factor[[1]]),
    num_factors)
  
  ggp <- plot_factor(
    MOFAobject.trained,
    factor = 1:num_factors,
    color_by = var,
    dodge = TRUE,
    add_boxplot = TRUE) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          panel.spacing = unit(0, "lines")) 
  
  #ggp$data$color_by <- as.factor(ggp$data$color_by)
  ggp$data$group_by <- ggp$data$color_by
  
  add_stat <- TRUE
  if (length(categories) > 2) {
    # kruskal wallis
    stat_method <- "Kruskal-Wallis"
    stat_test <- ggp$data %>%
      filter(!color_by %in% excluded_categories) %>%
      group_by(factor) %>%
      kruskal_test(value ~ group_by) %>%
      adjust_pvalue(method = "fdr") 
    
    stat_test_pairwise <- ggp$data %>%
      filter(!color_by %in% excluded_categories) %>%
      group_by(factor) %>%
      dunn_test(value ~ group_by, p.adjust.method = "fdr") %>%
      add_y_position()
  } else if (length(categories) == 2) {
    # wilcoxon
    stat_method <- "Wilcoxon"
    stat_test <- stat_test_pairwise <- ggp$data %>%
      group_by(factor) %>%
      wilcox_test(value ~ group_by, comparisons = comparisons) %>%
      adjust_pvalue(method = "fdr") %>%
      add_significance() %>%
      add_y_position()
  } else {
    add_stat <- FALSE
  }
  
  ggp <- ggp + scale_fill_okabe_ito(
    breaks = original_categories,
    labels = original_categories)
  ggp
}
```

```{r}
var = 'phenotype'
ggp <- plot_factor(
    MOFAobject.trained,
    factor = 1:num_factors,
    color_by = var,
    dodge = TRUE,
    add_boxplot = TRUE) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          panel.spacing = unit(0, "lines")) 
data<- ggp$data


mofa_factor <- 'Factor2'

df_factor <- data %>% filter(factor == mofa_factor)
ggplot(df_factor, aes(x= color_by,y=value, fill =color_by)) + 
  geom_boxplot() +
  theme_bw() +
  labs(x = 'Phenotype group', y = "MOFA scores", title = paste0(mofa_factor, ' scores boxplot')) +
  stat_compare_means(method = 'kruskal.test')
```

```{r}
var = 'sex'
ggp <- plot_factor(
    MOFAobject.trained,
    factor = 1:num_factors,
    color_by = var,
    dodge = TRUE,
    add_boxplot = TRUE) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          panel.spacing = unit(0, "lines")) 
data<- ggp$data


mofa_factor <- 'Factor2'

df_factor <- data %>% filter(factor == mofa_factor)
ggplot(df_factor, aes(x= color_by,y=value, fill =color_by)) + 
  geom_boxplot() +
  theme_bw() +
  labs(x = 'Phenotype group', y = "MOFA scores", title = paste0(mofa_factor, ' scores boxplot')) +
  stat_compare_means(method = 'wilcox.test')
```

# Recode variables as character for labels in plots below
```{r}
phenotype_data$observation.unit.identifier <- as.character(phenotype_data$observation.unit.identifier)

var = 'phenotype'
ggp <- plot_factor(
    MOFAobject.trained,
    factor = 1:num_factors,
    color_by = var,
    dodge = TRUE,
    add_boxplot = TRUE) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          panel.spacing = unit(0, "lines")) 
data<- ggp$data
obs_units <- lapply(data$sample, function(x) {
    obsunit <- phenotype_data[x, 7]
})
data$ObsUnit <- obs_units
```


### Factor 1 vs 2
```{r}
library(ggrepel)
data_f1_f2 <- data[data$factor %in% c('Factor1', 'Factor2'),]
df_wider <- data_f1_f2 %>%
pivot_wider(names_from = factor, values_from = value)
ggp <- ggplot(df_wider, aes(x= Factor1,y=Factor2, color =color_by)) + 
  geom_point() +
  geom_text_repel(aes(label = ObsUnit), size =3)

ggsave(file.path(outdir, "MOFA_DM1_F1_F2_scores.pdf"), 
       ggp, width = 16, height = 10.46, units = "cm", dpi = 300)

ggp
```


### Factor 3 vs 4
```{r}
data_f3_f4 <- data[data$factor %in% c('Factor3', 'Factor4'),]
df_wider <- data_f3_f4 %>%
pivot_wider(names_from = factor, values_from = value)
ggplot(df_wider, aes(x= Factor3,y=Factor4, color =color_by)) + 
  geom_point() +
  geom_text_repel(aes(label = ObsUnit), size =3)
```


### Factor 5 vs 6
```{r}
data_f5_f6 <- data[data$factor %in% c('Factor5', 'Factor6'),]
df_wider <- data_f5_f6 %>%
pivot_wider(names_from = factor, values_from = value)
ggplot(df_wider, aes(x= Factor5,y=Factor6, color =color_by)) + 
  geom_point() +
  geom_text_repel(aes(label = ObsUnit), size =3)
```


```{r, fig.width=15, fig.height=5}
ggp <- plot_factor_category(MOFAobject.trained, "phenotype")

ggsave(file.path(outdir, "MOFA_factors_vs_Age.group.pdf"), 
        ggp, width = 28.04, height = 10.46, units = "cm", dpi = 300)
ggp
```


```{r}
ggp <- plot_factor_category(MOFAobject.trained, "sex")

ggsave(file.path(outdir, "MOFA_factors_vs_Age.group.pdf"), 
       ggp, width = 28.04, height = 10.46, units = "cm", dpi = 300)
ggp
```
```{r}
MOFA_weights <- get_weights(MOFAobject.trained)
names(MOFA_weights) <- assay_names

lapply(assay_names, function(a) {
  df <- cbind(MOFA_weights[[a]], data.frame(rowData(mae[[a]])))
  f <- gsub("[| ,]", "_", a)
  outdur_disease <- file.path(outdir, 'DM1')
  dir.create(outdur_disease)
  write.csv(df, file = file.path(outdur_disease,  paste0("factorweights_", f, ".csv")))
})
```



```{r}
sessionInfo()
```

