---
title: 'Multi-omics Factor Analysis (MOFA)'
author:
- Casper de Visser^[Radboud university medical center, casper.devisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  outdir: Z:/data/iPSC_data/results/MOFA
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  mofa_model: Z:/data/iPSC_data/results/MOFA/All_samples/model_ALL_SAMPLES_Prot_Glob_Min.hdf5
  mofa_scores: Z:/data/iPSC_data/Results/MOFA/All_Samples/SampleScores.csv
  lipid_class_data: Z:/data/iPSC_data/iPSC_MAE/lipidomics_T_Class_data.csv
  config_file: 'config.yml'
---

This notebook performs Multi-omics Factor Analysis (MOFA) [@Argelaguet2018,@Argelaguet2020] on preprocessed data. 

MOFA is an unsupervised method that uncovers main sources of variation in multi-omics data. It allows to identify variation shared across omics data types or specific to individual omics data types. 
Subsequent to MOFA, correlation between MOFA factors and phenotypic information is determined to identify possible biological factors contributing to the observed variation.


```{r show_parameters}
print(params)
dir.create(params$outdir)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!exists(outdir)){
    outdir <- getwd()
}
```


## Libraries
```{r}
library(tidyr)
library(dplyr)
library(stringr)
library(purrr)
library(ggplot2)
library(MOFA2)
library(data.table)
library(glue)
```


## Read input data

Input data is a MultiAssayExperiment (MAE) object. All or specified assays in the MAE will be used for analysis Multi-omics Factor Analysis (MOFA).

```{r read_input_data, message=FALSE}
suppressMessages(library(MultiAssayExperiment))
suppressMessages(library(HDF5Array))

mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')

```


Omics to be included in MOFA

## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

## Assay selection
```{r}
assay_names <- config$experiment_names$general_selection
names(assay_names) <- assay_names
assay_names
```


## Prepare phenotype data
```{r}
phenotype_data <- as.data.frame(colData(mae)) %>%
  tibble::rownames_to_column("sample") %>%
  mutate(
    Disease = if_else(startsWith(phenotype, "Diseas"), "Disease", "Healthy"),
    phenotype_CHD2_CRISPR = recode(
      phenotype,
      'Diseased: CHD2 CRISPR' = 'Diseased: CHD2'
    ),
    phenotype_CRISPRS_both = recode(
      phenotype_CHD2_CRISPR,
      'Rescue: repeat excised CRISPR DM1' = 'Healthy'
  )
  ) %>%
  mutate(across(where(is.character), as.factor))
```


## Multi-omics factor analysis (MOFA)

For running MOFA, it is recommended to first remove known batch effects. 
Make sure to use input data that were adjusted for batch effect if necessary.

Include shorter readable omics names
```{r}
shoter_names <- config$experiment_names$shorter_names
```

Prepare MOFA input data: create a long data frame with sample, feature, view, 
[group], and value. 
```{r prepare_mofa_input}
# Function files
required_files <- c(
  "MOFA_functions.R",
  "LipidClassFunctions.R",
  "Feature_plots.R"
)

# possible directories
possible_dirs <- c(
  "support_functions", # when running in Nextflow
  "../support_functions" # when running in Rstudio
)

# Source each required file from the first location where it exists
for (file in required_files) {
  found <- FALSE
  for (dir in possible_dirs) {
    path <- file.path(dir, file)
    if (file.exists(path)) {
      source(path)
      message("Sourced: ", path)
      found <- TRUE
      break  # Move to next file once found
    }
  }
  if (!found) {
    warning("Could not find: ", file)
  }
}
```

Omics to data to correct input format
```{r}
mofa_input <- mae_to_MOFAinput(mae)
mofa_input$value <- as.numeric(mofa_input$value)
```

Create the MOFA object
```{r create_mofa_object}
MOFAobject <- create_mofa(mofa_input)
```


## Read in MOFA model
```{r}
MOFAobject.trained <- MOFA2::load_model(params$mofa_model)
```


## Visualizations of MOFA results 

## Plot amount of variance explained by each factor per modality

```{r, fig.width=10, fig.height=5}
expl_var <- c(
  get_variance_explained(MOFAobject.trained)$r2_per_factor$single_group)

ggp <- plot_variance_explained(MOFAobject.trained) +
  geom_text(
    label = round(expl_var, 2),
    color = ifelse(expl_var < 0.6*max(expl_var), "black", "white")) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )
      
ggp
```


## Factor weights

Lipidomics features can be grouped in lipid classes. And the features within a class (species) have different chain lengths and number of double bonds.
```{r}
lipidomics_feature_metadata <- as.data.frame(rowData(mae[["Lipidomics, Targeted | imputed | log"]]))
table(lipidomics_feature_metadata$Class)
```

Each Lipid Class has different nomenclature for structural properties
```{r}
lipidomics_feature_metadata[!duplicated(lipidomics_feature_metadata$Class), c("Feature", "Class")]
```


### Color lipidomics feature weights per class

```{r, fig.width=7}
p <- plot_feature_weights(
  MOFAobject.trained,
  "Lipidomics_Targeted",
  1,
  lipidomics_feature_metadata,
  color_col = "Class", 
  top_n = 20, 
  dot_size = 3
)
p
ggsave(file.path(outdir, 'Lipid_F1_top_loadings_Class.pdf'), p)
```



```{r, fig.width=7}
p <- plot_feature_weights(
  MOFAobject.trained,
  view = "Lipidomics_Targeted",
  factor = 2,
  feature_metadata = lipidomics_feature_metadata,
  color_col = "Class", 
  top_n = 20,
  dot_size = 3
)

p 
ggsave(file.path(outdir, 'Lipid_F2_top_loadings_Class.pdf'), p)
```

### Gene set enrichment style analysis to test overrepresentation of lipid classes
```{r}
lipid_sets <- split(lipidomics_feature_metadata$Feature, lipidomics_feature_metadata$Class)
lipid_sets[1:3]
```


### Lipidomics weights and chain length
```{r}
## 'Groups' of lipid classes with the same nomenclature
SingleChains <- c("CE", "PA", "FA", "LPC", "LPE")
TwoChains <- c("PC", "PE", "PS", "PI", "PG")
Sphingolipids <- c("Cer", "HexCER", "LacCER", "SM")
Glyercols <- c("TG", "DG")

SingleChains_df <- lipidomics_feature_metadata[lipidomics_feature_metadata$Class %in% SingleChains,]
TwoChains_df <- lipidomics_feature_metadata[lipidomics_feature_metadata$Class %in% TwoChains,]
Sphingolipids_df <- lipidomics_feature_metadata[lipidomics_feature_metadata$Class %in% Sphingolipids,]
Glycerols_df <- lipidomics_feature_metadata[lipidomics_feature_metadata$Class %in% Glyercols,]

SingleChains_df <- parse_single_chain_df(SingleChains_df)
TwoChains_df <- pparse_two_chain_df(TwoChains_df)
Sphingolipids_df <- parse_sphingo_df(Sphingolipids_df)
Glycerols_df <- parse_glycerolipid_df(Glycerols_df)

lipidomics_feature_metadata <- rbind(SingleChains_df, TwoChains_df, Sphingolipids_df, Glycerols_df)
```


## Add lipid structural properties to MOFA weights
```{r}
MOFA_weights <- get_weights(MOFAobject.trained)
names(MOFA_weights) <- assay_names

lipid_weights <- as.data.frame(MOFA_weights$`Lipidomics, Targeted | imputed | log`)

lipid_weights <- cbind(lipid_weights, lipidomics_feature_metadata[rownames(lipid_weights),])
```


## Combined plot
```{r}
correlation_plot_all_groups(lipid_weights, 
                            target_column = 'Factor1',
                            filter_column = 'Class', 
                            fixed_column = 'Total_Carbons')
```
## Combined plot double bonds
```{r}
correlation_plot_all_groups(lipid_weights, 
                            target_column = 'Factor1',
                            filter_column = 'Class', 
                            fixed_column = 'Total_DBs')
```


## Combined plot for Factor 2
```{r}
correlation_plot_all_groups(lipid_weights, 
                            target_column = 'Factor2',
                            filter_column = 'Class', 
                            fixed_column = 'Total_Carbons')
```

## Combined plot double bonds
```{r}
correlation_plot_all_groups(lipid_weights, 
                            target_column = 'Factor2',
                            filter_column = 'Class', 
                            fixed_column = 'Total_DBs')
```




## Factor 3
Combined plot
```{r}
correlation_plot_all_groups(lipid_weights, 
                            target_column = 'Factor3',
                            filter_column = 'Class', 
                            fixed_column = 'Total_Carbons')
```

## Combined plot
```{r}
correlation_plot_all_groups(lipid_weights, 
                            target_column = 'Factor4',
                            filter_column = 'Class', 
                            fixed_column = 'Total_Carbons')
```


## One-chain lipids

### CE
```{r}
lipid_class <- 'CE'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```

### PA
```{r}
lipid_class <- 'PA'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```

### FA
```{r}
lipid_class <- 'FA'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```

### LPC
```{r}
lipid_class <- 'LPC'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```

### LPE
```{r}
lipid_class <- 'LPE'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```


## Two-chain lipids

### PC
```{r}
lipid_class <- 'PC'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```

### PE
```{r}
lipid_class <- 'PE'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```

### PS
```{r}
lipid_class <- 'PS'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```

### PG
```{r}
lipid_class <- 'PG'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```

### PI
```{r}
lipid_class <- 'PI'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```



## Sphingolipids

### Cer
```{r}
lipid_class <- 'Cer'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```

### SM
```{r}
lipid_class <- 'SM'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```

### HexCER
```{r}
lipid_class <- 'HexCER'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```

### LacCER
```{r}
lipid_class <- 'LacCER'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```


## Glyerols
## TriGlycerides (TG)
```{r}
correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = 'TG', lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = 'TG', lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = 'TG', lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = 'TG', lipid_property = "Total_Carbons", method='spearman')
```
TG double bonds
```{r}
correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = 'TG', lipid_property = "Total_DBs", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = 'TG', lipid_property = "Total_DBs", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = 'TG', lipid_property = "Total_DBs", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = 'TG', lipid_property = "Total_DBs", method='spearman')
```


## DiGlycerides (DG)
```{r}
lipid_class <- 'DG'

correlation_plot(lipid_weights, target_column = 'Factor1', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor2', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor3', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
correlation_plot(lipid_weights, target_column = 'Factor4', filter_value = lipid_class, lipid_property = "Total_Carbons", method='spearman')
```


## Lipid Class total values correlation with MOFA scores
```{r}
MOFA_scores <- read.csv(params$mofa_scores, row.names=1)
LipidClassData <- read.csv(params$lipid_class_data, row.names=1)

LipidClassData <- data.frame(t(LipidClassData))

# Add Class info to mofa scores 
MOFA_scores <- MOFA_scores[rownames(LipidClassData),]
MOFA_scoresLipidClass <- cbind(MOFA_scores, LipidClassData)
```


```{r}
ClassVector <- as.numeric(MOFA_scoresLipidClass$PE)
ClassName <- "PE"

cor_plot(MOFA_scoresLipidClass$Factor1, y=ClassVector, x_lab = "Factor 1", y_lab=ClassName)
cor_plot(MOFA_scoresLipidClass$Factor2, y=ClassVector, x_lab = "Factor 2", y_lab=ClassName)
cor_plot(MOFA_scoresLipidClass$Factor3, y=ClassVector, x_lab = "Factor 3", y_lab=ClassName)
cor_plot(MOFA_scoresLipidClass$Factor4, y=ClassVector, x_lab = "Factor 4", y_lab=ClassName)
```

```{r}
ClassVector <- as.numeric(MOFA_scoresLipidClass$HexCER)
ClassName <- "HexCER"

cor_plot(x = MOFA_scoresLipidClass$Factor1, y = ClassVector, x_lab = "Factor1", y_lab = ClassName)
cor_plot(x = MOFA_scoresLipidClass$Factor2, y = ClassVector, x_lab = "Factor2", y_lab = ClassName)
cor_plot(x = MOFA_scoresLipidClass$Factor3, y = ClassVector, x_lab = "Factor3", y_lab = ClassName)
cor_plot(x = MOFA_scoresLipidClass$Factor4, y = ClassVector, x_lab = "Factor4", y_lab = ClassName)
```


```{r}
ClassVector <- as.numeric(MOFA_scoresLipidClass$SM)
ClassName <- "SM"

cor_plot(x = MOFA_scoresLipidClass$Factor1, y = ClassVector, x_lab = "Factor1", y_lab = ClassName)
cor_plot(x = MOFA_scoresLipidClass$Factor2, y = ClassVector, x_lab = "Factor2", y_lab = ClassName)
cor_plot(x = MOFA_scoresLipidClass$Factor3, y = ClassVector, x_lab = "Factor3", y_lab = ClassName)
cor_plot(x = MOFA_scoresLipidClass$Factor4, y = ClassVector, x_lab = "Factor4", y_lab = ClassName)
```


```{r}
ClassVector <- as.numeric(MOFA_scoresLipidClass$TG)
ClassName <- "TG"

cor_plot(x = MOFA_scoresLipidClass$Factor1, y = ClassVector, x_lab = "Factor1", y_lab = ClassName)
cor_plot(x = MOFA_scoresLipidClass$Factor2, y = ClassVector, x_lab = "Factor2", y_lab = ClassName)
cor_plot(x = MOFA_scoresLipidClass$Factor3, y = ClassVector, x_lab = "Factor3", y_lab = ClassName)
cor_plot(x = MOFA_scoresLipidClass$Factor4, y = ClassVector, x_lab = "Factor4", y_lab = ClassName)
```



```{r}
sessionInfo()
```

