---
title: 'Linear Mixed models'
author:
  - Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  outdir: Z:/data/iPSC_data/results/MixedModels
  config_file: "config.yml"
---

## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!dir.exists(outdir)){
    outdir <- getwd()
}
```

## Read data and metadata: MultiAssayExperiment

```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
mae
```

The MultiAssayExperiment contains individual omics assays


## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

## Assay selection
```{r}
experiment_names <- config$experiment_names$general_selection
experiment_names
```


Technical variables are available per assay

```{r}
tech_vars_factor <- lapply(experiments(mae), function(x) {
  num_levels <- apply(colData(x), 2, function(x) {length(levels(as.factor(x)))})
  t <- names(colData(x))[num_levels <= 9 & num_levels > 1]
  t <- t[t!="subject.id"]
  t <- t[!t %in% names(colData(mae))]
})
tech_vars_factor

tech_vars_continuous <- lapply(experiments(mae), function(x) {
  num_levels <- apply(colData(x), 2, function(x) {length(levels(as.factor(x)))})
  t <- names(colData(x))[num_levels > 9]
  t <- t[t!="subject.id"]
  t <- t[t!="measurement.id"]
  t <- t[t!="measurement_names"]
  t <- t[t!="ID.Patient"]
  t <- t[!t %in% c("SampleID", "Radboud.code", "Kevin.nr", "Kevin.CPM", 
                   "Sample.name", "ISA_name", "SampleID_withrep", "Sample_Overview_names")] #TODO, remove these from sample metadata sheet
  t <- t[!t %in% c("GroupNum", "GroupName", "SampleNorm", "Radboud.code.1")] 

  t <- t[!t %in% names(colData(mae))]
})
tech_vars_continuous
```


Phenotype information shared across all assays can be accessed via `colData()`

```{r sample_information_overview}
names(colData(mae))

pheno_df <- as.data.frame(colData(mae))

## Add column for Disease vs Healthy comparison
new_column <- lapply(pheno_df$phenotype, function(x) {
    if (startsWith(x, 'Diseas')) {
        value <- 'Disease'
    }
    else{
        value <- 'Healthy'
    }
})
pheno_df$Disease <- as.character(new_column)
```

# Mixed effect model with cell line as random effect, and bio rep nested in the cell line

```{r}
library(lme4)
library(dplyr)
library(openxlsx)
```


## Load functions for Nested Mixed Effect model
```{r, warning=FALSE, message=FALSE}
source_paths <- c(
  "../support_functions/Mixed_Model_functions.R",      # when running in Rstudio
  "Mixed_Model_functions.R"                    # when running in Nextflow
)

for (path in source_paths) {
  if (file.exists(path)) {
    source(path)
    break
  }
}
```


```{r}
run_rand_model <- function(assay_name, pred, DM1_ONLY = F) {
  se <- experiments(mae)[assay_name]
  df <- get_df_input_lm(se)
  features <- colnames(df)
  features <- features[1:100]
  features <- features[features != "ENSG00000129824"] #TODO
  # add biological variables to data frame
  df <- cbind(df, pheno_df[rownames(df), ])
  df <- add_nested_cols(df)
  df <- code_phenotype_column(df)
  # Select only DM1 and control samples if DM1_ONLY = TRUE
  if (DM1_ONLY) {
    df <-  df[df$phenotype_DM1 %in% c('Diseased: DM1', 'Healthy'),]
  }
  # fit linear mixed model
  model_result <- do.call(rbind, lapply(
    features,
    run_MixedModel_df,
    df=df,
    predictor=pred))
  #adjust p-values
  model_result$p.adj <- p.adjust(model_result$p.value, method = "BH")
  model_result$rank <- rank(model_result$p.value)
  ## Remove the CRISPR comparisons
  model_result <- model_result[!grepl('excised', model_result$outcome), ]
  if (!("nest_effect" %in% names(model_result))) {
    model_result$nest_effect <- 0
    model_result <- model_result[,c(1:8, 11,9,10)]
  }
  model_result$Assay <- assay_name
  model_result$Feature_plot <- model_result$feature #If not other feature ID available
  return(model_result)
}
```


## Run Mixed Effect model, comparing individual disease groups with the control lines
```{r}
# Select assays
#experiment_names <- experiment_names[c(5,6)]

Predictor <- 'phenotype'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model(a, pred=Predictor)
  model_result <- model_result[model_result$outcome == 'phenotypeDiseased: CHD2',]
  model_result$p.adj <- p.adjust(model_result$p.value, method = "BH")
  model_result
})
merged_result_CHD2 <- do.call(rbind, mae_rlm_RandomEffect)
```


## Run Mixed Effect model, comparing individual disease groups with the control lines
```{r}
Predictor <- 'phenotype'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model(a, pred=Predictor)
  model_result <- model_result[model_result$outcome == 'phenotypeDiseased: NANS-CDG',]
  model_result$p.adj <- p.adjust(model_result$p.value, method = "BH")
  model_result
})
merged_result_NANS <- do.call(rbind, mae_rlm_RandomEffect)
```


## Run Mixed Effect model, comparing all disease with all controls
```{r}
Predictor <- 'Disease'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model(a, pred=Predictor)
  model_result
})
merged_result_Disease <- do.call(rbind, mae_rlm_RandomEffect)
```


## Run Mixed Effect model, comparing female and male samples
```{r}
Predictor <- 'sex'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model(a, pred=Predictor)
  model_result
})
merged_result_Sex <- do.call(rbind, mae_rlm_RandomEffect)
```


## Compare DM1 with control lines inculding CRISPR Rescue line
```{r}
Predictor <- 'phenotype_DM1'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model(a, pred=Predictor, DM1_ONLY = T)
  model_result
})
merged_result_DM1 <- do.call(rbind, mae_rlm_RandomEffect)
```


## Only compare patient 2 & 3 for DM1
Remove DM patient 1 from MAE object
```{r}
DM1_p1_names <- c(
  'IPS20_00122_1',
  'IPS20_00122_2',
  'IPS20_00122_3',
  'IPS23_00090_1', #Rescue
  'IPS23_00090_2', #Rescue
  'IPS23_00090_3' #Rescue
)

sample_names <- rownames(colData(mae))
sample_names_NoDMP1 <- sample_names[!sample_names  %in% DM1_p1_names]

mae <- subsetByColData(mae, y = sample_names_NoDMP1)
```

```{r}
Predictor <- 'phenotype_DM1'

mae_rlm_RandomEffect <- lapply(experiment_names, function(a) {
  model_result <- run_rand_model(a, pred=Predictor, DM1_ONLY = T)
  model_result
})
merged_result_DM1_P2P3 <- do.call(rbind, mae_rlm_RandomEffect)
```


Merge all results and save to .csv
```{r}
merged_results_all <- rbind(
  merged_result_CHD2,
  merged_result_NANS,
  merged_result_DM1,
  merged_result_DM1_P2P3,
  merged_result_Disease,
  merged_result_Sex
  )

write.csv(merged_results_all, 'MixedModelsResults.csv')
```


Save tables per contrast per omics including feature metadata
```{r}
save2excel <- function(df, namefile) {
  split_df <- split(df, df$Assay)
  wb <- createWorkbook()
  names(split_df) <- experiment_names
  for (omics in names(split_df)) {
    assay_name <- omics
    omics <- strsplit(omics, " \\| ")[[1]][1]
    addWorksheet(wb, sheetName = omics)
    data <- split_df[[assay_name]][, c( "feature","coefficient","std.error","t.value","p.value","rand_effect","nest_effect","p.adj")]
    data$rank <- rank(data$p.value)
    ## Add feature metadata
    feature_metadata <- data.frame(rowData(mae[[assay_name]]))
    feature_metadata$feature <- rownames(feature_metadata)
    data <- data %>%
        left_join(feature_metadata, by = 'feature')
    writeData(wb, sheet = omics, data)
  }
  saveWorkbook(wb, paste0(namefile, "_MixedModel_Results.xlsx"), overwrite = T)
}


save2excel(merged_result_CHD2, 'CHD2')
save2excel(merged_result_NANS, 'NANS')
save2excel(merged_result_DM1_P2P3, 'DM1_P2_P3')
save2excel(merged_result_DM1, 'DM1')
save2excel(merged_result_Disease, 'All_diseases_combined')
save2excel(merged_result_Sex, 'Sex')
```


## Session information

```{r}
sessionInfo()
```


