---
title: 'Linear Mixed models'
author:
  - Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  mofa_scores: Z:/data/iPSC_data/Results/MOFA/All_Samples/SampleScores.csv
  outdir: Z:/data/iPSC_data/results/RLM/
  config_file: "config.yml"
---

## Libraries
```{r}
library(lme4)
```


## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir
```

## Read data and metadata: MultiAssayExperiment

```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
#mae
```


The MultiAssayExperiment contains individual omics assays

## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```


## Assay selection
```{r}
experiment_names <- config$experiment_names$general_selection
```


Phenotype information shared across all assays can be accessed via `colData()`

```{r sample_information_overview}
names(colData(mae))

pheno_df <- as.data.frame(colData(mae))

## Add column for Disease vs Healthy comparison
new_column <- lapply(pheno_df$phenotype, function(x) {
    if (startsWith(x, 'Diseas')) {
        value <- 'Disease'
    }
    else{
        value <- 'Healthy'
    }
})
pheno_df$Disease <- as.character(new_column)
```


## Correct omics values with MOFA factor 2 scores
```{r}
mofa_scores_df <- read.csv(params$mofa_scores, row.names = 1)
```


## Load functions for Nested Mixed Effect model
```{r, warning=FALSE, message=FALSE}
source('Mixed_Model_functions.R')
```


```{r}
experiment_names
```



Compare Corrected and non-corrected for single feature
```{r}
assay_name <- 'Proteomics'
confound_factors <- c("Factor1", "Factor2")
 # 1. Prepare data
data_prep <- prepare_feature_data(assay_name, mae)
df <- data_prep$df
features <- data_prep$features
  
# 2. Add confounding factors
df <- add_confounding_factors(df, mofa_scores_df, confound_factors)
  
# 3. Regress out confounding factors
df_corrected <- regress_out_factors(df, features, confound_factors)
  
# 4. Prepare final modeling dataframe
df_input_cor <- prepare_modeling_data(df_corrected, df, pheno_df, DM1_ONLY=F)
df_input <- prepare_modeling_data(df, df, pheno_df, DM1_ONLY=F)
  
# 5. Run models and save
featueres_check <- c('Q9Y3U8')
model_result <- run_models_and_save(df_input, features= featueres_check, 'phenotype', assay_name)
model_result_cor <- run_models_and_save(df_input_cor, features= featueres_check, 'phenotype', assay_name)
```
RPL336 associates with Factor 1
```{r}
col <- 'Q9Y3U8'
factor_columns <- c("Factor1", "Factor2")
formula_str <- paste(col, "~", paste(factor_columns, collapse = " + "))
model <- lm(as.formula(formula_str), data = df)
plot(model)
summary(model)
```
NEUROG2 associates with Factor 2
```{r}
col <- 'Q9H2A3'
factor_columns <- c("Factor1", "Factor2")
formula_str <- paste(col, "~", paste(factor_columns, collapse = " + "))
model <- lm(as.formula(formula_str), data = df)
plot(model)
summary(model)
```


## Run model on few features, to check diagnostics
Proteomics
```{r}
assay_name <- 'Proteomics'
data_prep <- prepare_feature_data(assay_name, mae)
df <- data_prep$df
df <- prepare_modeling_data(df, df, pheno_df)
feature <- data_prep$features[1]
predictor = 'Disease'

df$y <- df[[feature]]
str_formula <- paste0("y ~ ", predictor, " +  (1 | sample.renames/clone_ids)")
my_formula <- as.formula(str_formula)
lm_fit <- try(lmerTest::lmer(formula = my_formula, data=df))

simulated_residuals <- DHARMa::simulateResiduals(lm_fit)
plot(simulated_residuals)
```




EMseq
```{r}
#assay_name <- experiment_names[c(1)]
assay_name <- 'EMseq | 100,000 variable Mval'
data_prep <- prepare_feature_data(assay_name, mae)
df <- data_prep$df
df <- prepare_modeling_data(df, df, pheno_df)
feature <- data_prep$features[3]
predictor = 'phenotype'

df$y <- df[[feature]]
## Apply tranformation, to remove 0 and 1 values for beta distribution
#df$y <- pmax(pmin(df$y, 1 - 1e-6), 1e-6)


str_formula <- paste0("y ~ ", predictor, " +  (1 | sample.renames/clone_ids)")
my_formula <- as.formula(str_formula)

#lm_fit <- try(lmerTest::lmer(formula = my_formula, data=df))
library(glmmTMB)
lm_fit <- try(glmmTMB::glmmTMB(formula = my_formula, family = t_family(link = "identity"), data=df))
#library(robustlmm)
#lm_fit <- try(rlmer(formula = my_formula,  data=df))
library(flexmix)
lm_fit <- try(flexmix(formula = my_formula,  data=df))


simulated_residuals <- DHARMa::simulateResiduals(lm_fit)
plot(simulated_residuals)
```

Lipidomics T
```{r}
assay_name <- experiment_names[[2]]
data_prep <- prepare_feature_data(assay_name, mae)
df <- data_prep$df
df <- prepare_modeling_data(df, df, pheno_df)
feature <- data_prep$features[1]
predictor = 'Disease'

df$y <- df[[feature]]
str_formula <- paste0("y ~ ", predictor, " +  (1 | sample.renames/clone_ids)")
my_formula <- as.formula(str_formula)
lm_fit <- try(lmerTest::lmer(formula = my_formula, data=df))

simulated_residuals <- DHARMa::simulateResiduals(lm_fit)
plot(simulated_residuals)
```

Lipitomics Unt
```{r}
assay_name <- experiment_names[[3]]
data_prep <- prepare_feature_data(assay_name, mae)
df <- data_prep$df
df <- prepare_modeling_data(df, df, pheno_df)
feature <- data_prep$features[1]
predictor = 'Disease'

df$y <- df[[feature]]
str_formula <- paste0("y ~ ", predictor, " +  (1 | sample.renames/clone_ids)")
my_formula <- as.formula(str_formula)
lm_fit <- try(lmerTest::lmer(formula = my_formula, data=df))

simulated_residuals <- DHARMa::simulateResiduals(lm_fit)
plot(simulated_residuals)
```

Metabolomics LUMC
```{r}
assay_name <- experiment_names[[4]]
data_prep <- prepare_feature_data(assay_name, mae)
df <- data_prep$df
df <- prepare_modeling_data(df, df, pheno_df)
feature <- data_prep$features[1]
predictor = 'Disease'

df$y <- df[[feature]]
str_formula <- paste0("y ~ ", predictor, " +  (1 | sample.renames/clone_ids)")
my_formula <- as.formula(str_formula)
lm_fit <- try(lmerTest::lmer(formula = my_formula, data=df))

simulated_residuals <- DHARMa::simulateResiduals(lm_fit)
plot(simulated_residuals)
```

Metabolomics TML
```{r}
assay_name <- experiment_names[[5]]
data_prep <- prepare_feature_data(assay_name, mae)
df <- data_prep$df
df <- prepare_modeling_data(df, df, pheno_df)
feature <- data_prep$features[1]
predictor = 'Disease'

df$y <- df[[feature]]
str_formula <- paste0("y ~ ", predictor, " +  (1 | sample.renames/clone_ids)")
my_formula <- as.formula(str_formula)
lm_fit <- try(lmerTest::lmer(formula = my_formula, data=df))

simulated_residuals <- DHARMa::simulateResiduals(lm_fit)
plot(simulated_residuals)
```


RNAseq
```{r}
assay_name <- experiment_names[[7]]
data_prep <- prepare_feature_data(assay_name, mae)
df <- data_prep$df
df <- prepare_modeling_data(df, df, pheno_df)
feature <- data_prep$features[1]
predictor = 'Disease'

df$y <- df[[feature]]
str_formula <- paste0("y ~ ", predictor, " +  (1 | sample.renames/clone_ids)")
my_formula <- as.formula(str_formula)
lm_fit <- try(lmerTest::lmer(formula = my_formula, data=df))

simulated_residuals <- DHARMa::simulateResiduals(lm_fit)
plot(simulated_residuals)
```

## Session information

```{r}
sessionInfo()
```


