---
title: 'Pairwise correlations (with correction of MOFA factor 2)'
author:
  - Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  isa_file: C:/Users/Casper.deVisser/Documents/ipsc-data-analysis/20240822_Xomics_iPSC_Neurons_new_READ-ONLY.xlsx
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  outdir: Z:/data/iPSC_data/results/RLM/
---

## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!dir.exists(outdir)){
    outdir <- getwd()
}
```

## Libraries
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggpubr)
```

## Read data and metadata: MultiAssayExperiment

```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
mae
```


The MultiAssayExperiment contains individual omics assays
```{r}
experiment_names <- names(experiments(mae))
experiment_names
```


## Read Assay files
```{r}
read_isa_file <- function(tab_name){
  file_name <- params$isa_file
  file <- readxl::read_xlsx(file_name, sheet = tab_name)
  isa_df <- file[rowSums(is.na(file)) != ncol(file),]
  isa_df<- isa_df[isa_df$`quality accessment` != 'FALSE', ] ## Sample QC filter
  isa_df <- isa_df[isa_df$`quality accessment` != 'Failed', ]
  isa_df
}

## Get Assay sheet names
sheet_names <- openxlsx::getSheetNames(params$isa_file)
assay_names <- sheet_names[startsWith(sheet_names, 'Assay')]
assay_names <- assay_names[assay_names != 'Assay - genomics']
#assay_names <- assay_names[assay_names != 'Assay - epigenomics']
#assay_names <- assay_names[assay_names != 'Assay - transcriptomics']

## Read all Assay files
isa_assay_files <- lapply(assay_names, function(x) read_isa_file(x))
names(isa_assay_files) <- assay_names
```


## Experiment names
```{r}
experiments <- names(experiments(mae))
experiments

df_match_isa_mae <- data.frame(assay_names, experiment_names[c(13, 12, 2, 1, 3, 6, 8)]) #TODO
colnames(df_match_isa_mae) <- c('Assay_names', 'MAE_names')
df_match_isa_mae
```


## Function to rename sample names of control lines in MAE Assays
```{r}
## Function to re-write the Control sample names, to be improved
rewrite_names <- function(x) {
  if (startsWith(x, "WTC")) {
    x <- 'WTC'
  }
  else if (startsWith(x, "X409")) {
    x <- '409B'
  }
  else if (startsWith(x, "Sendai")) {
    x <- 'Sendai'
  }
  else {
    x<- x
  }
  x
}
```


## Read ISA Sample tab
```{r}
isa_assay_Sample <- readxl::read_xlsx(params$isa_file, sheet = 4)
```


## Correct omics values with MOFA factor 2 scores
```{r}
mofa_scores_df <- read.csv('Z:/data/iPSC_data/Results/MOFA/All_Samples/SampleScores.csv', row.names = 1)
```


## Iterate over experiment names to create input data frames
```{r}
NumExp <- nrow(df_match_isa_mae)

isa_mae_list <- lapply(seq(2,3), function(x) {
  
  ## Get names of Assay/Experiment
  isa_assay_name <- df_match_isa_mae[x, 1]
  mae_exper_name <- df_match_isa_mae[x, 2]
  
  isa_assay <- isa_assay_files[[isa_assay_name]]
  
  ## Add phenotype column of Sample tab to Assay tab
  phenotypes <- lapply(isa_assay$`sample identifier`, function(x) {
    phenotype <- isa_assay_Sample[isa_assay_Sample$`sample identifier` == x, 20]
    phenotype[[1]]
  })
  isa_assay$phenotype <- as.character(phenotypes)
  isa_assay

  se <- experiments(mae)[[mae_exper_name]]
  df_assay <- as.data.frame(t(assays(se)[[1]]))
  measurement_columns <- colnames(df_assay)
  df_assay$sample_id <- rownames(df_assay)
  
  ## Add column with sample names as group
  new_names <- gsub("(^[^_]*_[^_]*)_.*", "\\1", rownames(df_assay))
  new_names <- lapply(new_names, function(x) rewrite_names(x))
  new_names <- gsub("_", "-", new_names) 
  df_assay$group <- new_names
  
  
  ## Correct for MOFA sample score
  mofa_scores_df_match <- mofa_scores_df[match(df_assay$sample_id, rownames(mofa_scores_df)),]
  df_assay$Factor2 <- mofa_scores_df_match$Factor2
  
  df_corrected <- as.data.frame(
    lapply(measurement_columns, function(col) {
      resid(lm(df_assay[[col]] ~ df_assay$Factor2))
    })
  )
  df_corrected <- as.data.frame(df_corrected)
  colnames(df_corrected) <- measurement_columns
  df_corrected$sample_id <- rownames(df_assay)
  df_assay$group <- new_names
  
  ## IMPORTANT, MATCH ORDER OF ISA ASSAY AND MAE ASSAY
  isa_assay <- isa_assay[match(df_assay$group, isa_assay$`sample identifier`),]
  
  list(isa_assay, df_corrected)

})
names(isa_mae_list ) <- df_match_isa_mae$MAE_names[c(2,3)]
```



## Load function to calculate R-squared for samples in the same sample group
```{r}
source('../support_functions/PairwiseCorrelationsFunctions.R')
```



## Phenotype groups vs. Cell line groups
```{r}
lapply(names(isa_mae_list), function(x) {
  ## Get pairwise correlations
  df_input <- get_pairwise_cor_df(x)
  ## Change label names
  levels(df_input$Level) <- c("Unrelated samples",  "Samples in same disease group", "Biological replicates")
  # Make boxplot
  #ylim_min <- floor(min(df_input$r_squared * 100)) / 100 - 0.02
  
  ggplot(df_input, aes(x=Level, y=r_squared, fill = Level)) +
      geom_boxplot() +
      ggtitle(paste0(x)) +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
      ylim(0.0, 1) +
      stat_compare_means()  +
      stat_summary(fun.data = function(x){
              return(data.frame(y=mean(x), label = paste0(length(x))))
          }, geom = "text")

})
```


## DM1 group only, with clones of patient 2
```{r}
clones <- c('IPS20_00035',
            'IPS23_00088', 
            'IPS23_00089'
            ) 

lapply(names(isa_mae_list), function(omics_name) {
  title <- strsplit(omics_name, split = '\\|')[[1]]
  ## Get pairwise correlations
  df_input <- get_pairwise_cor_df_clones(omics_name, clones, "Diseased: DM1")
  ## Change label names
  levels(df_input$Level) <- c("DM1 patients", "P2 clones", "Biological repl.")
  colors <- c("DM1 patients" = "green4", "P2 clones" = "gold3", "Biological repl." = "skyblue1")
  ymin <- min(df_input$r_squared)
  ymax <- max(df_input$r_squared) + 0.02
  # Make boxplot
  ggp <- ggplot(df_input, aes(x=Level, y=r_squared, fill = Level)) +
      geom_boxplot() +
      ggtitle(paste0(title)) +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
      ylim(ymin, ymax) +
      scale_fill_manual(values = colors) +
      stat_compare_means()  +
      stat_summary(fun.data = function(x){
              return(data.frame(y=mean(x), label = paste0(length(x))))
          }, geom = "text")
  ggsave(paste0('DM1_clones_', title, '.png'), plot = ggp, width = 8, height= 4)
  ggp
})
```


## DM1 group only, with clones of patient 2
```{r}
clones <- c('IPS20_00035',
            'IPS23_00088',
            'IPS23_00089'
            ) ## TODO: HARDCODED

lapply(names(isa_mae_list), function(omics_name) {
  title <- strsplit(omics_name, split = '\\|')[[1]]
  ## Get pairwise correlations
  df_input <- get_pairwise_cor_df_clones(omics_name, clones, "Diseased: DM1")
  ## Select only clones
  df_input <- df_input[df_input$Level == 'DM1_P2_clones',]
  df_input$Clones <- NA
  df_input$Clones[grepl('IPS23_00089', df_input$pair) & grepl("IPS20_00035", df_input$pair)] <- 'C1 vs C2'
  df_input$Clones[grepl('IPS23_00088', df_input$pair) & grepl("IPS20_00035", df_input$pair)] <- 'C1 vs C3'
  df_input$Clones[grepl('IPS23_00088', df_input$pair) & grepl("IPS23_00089", df_input$pair)] <- 'C2 vs C3'
  ymin <- min(df_input$r_squared)
  ymax <- 1
  # Make boxplot
  ggp <- ggplot(df_input, aes(x=Clones, y=r_squared, fill = Clones)) +
      geom_boxplot() +
      ggtitle(paste0(title)) +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
      ylim(ymin, ymax) +
      stat_compare_means()  +
      stat_summary(fun.data = function(x){
              return(data.frame(y=mean(x), label = paste0(length(x))))
          }, geom = "text")
  ggsave(paste0('DM1_clones_only_', title, '.png'), plot = ggp, width = 8, height= 4)

})
```


## CHD2 group only, with clones of patient 3
```{r}
clones <- c('IPS23_00025','IPS22_00063')

lapply(names(isa_mae_list), function(x) {
  ## Get pairwise correlations
  df_input <- get_pairwise_cor_df_clones(x, clones, "Diseased: CHD2")
  ## Change label names
  levels(df_input$Level) <- c("CHD2 patients", "P3 clones", "Cell lines")
  colors <- c("CHD2 patients" = "green4", "P3 clones" = "gold3", "Cell lines" = "skyblue1")
  # Make boxplot
  ylim_min <- floor(min(df_input$r_squared * 100)) / 100 - 0.02
  
  ggplot(df_input, aes(x=Level, y=r_squared, fill = Level)) +
      geom_boxplot() +
      ggtitle(paste0(x, '\nPairwise correlations between phenotype/bio replicate')) +
      theme_bw() +
      scale_fill_manual(values = colors) +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
      ylim(ylim_min, 1) +
      stat_compare_means()  +
      stat_summary(fun.data = function(x){
              return(data.frame(y=mean(x), label = paste0(length(x))))
          }, geom = "text")

})
```


## Session information

```{r}
sessionInfo()
```


