---
title: 'Linear mixed models'
author:
  - Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  chunk: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  mae_object: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  outdir: Z:/data/iPSC_data/results/RLM/
---

## Params
```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir
```


## Run MAE for phenotype data
```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 

## Multi Assay Experiment object
mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_object, 
  prefix = 'mae')

## Phenotype data frame
pheno_df <- as.data.frame(colData(mae))
```


## Mixed effect model with cell line as random effect
```{r, warning=FALSE, message=FALSE}

## Custom function for running linear model on a chunk
run_rand_model_chunk <- function(df, features, pred, DM1_ONLY=F) {
  
  
  
  # Select only DM1 and control samples if paramater DM1_ONLY used
  if (DM1_ONLY)  {
    df <- df[df$phenotype_DM1 %in% c("Diseased: DM1", "Healthy"), ]
  }

  # Run mixed models on all available features
  model_result <- do.call(rbind, lapply(
                           features,
                           run_MixedModel_df,
                           df=df,
                           predictor=pred))
  #adjust p-values
  model_result$p.adj <- p.adjust(model_result$p.value, method = "BH")
  model_result$rank <- rank(model_result$p.value)

  return(model_result)
}
```    


## Read chunk csv file to dataframe
```{r}
chunk_df <- read.csv(params$chunk, row.names=1)
``` 


## Remove features that are non-variable in the nested groups
```{r}
require(dplyr)

## Remove columns that are phenotype data    
features <- setdiff(names(chunk_df), union(colnames(pheno_df), c('sample.renames', 'clone_ids', 'bio_reps', 'Disease'))) #TODO
features <- features[features != "ENSG00000129824"] #TODO
    
## Remove non-variable features 
chunk_df %>%
    group_by(phenotype, sample.renames, bio_reps) %>%
    summarize(across(all_of(features),
                     ~var(.x, na.rm = TRUE),
                     .names = "var_{col}")) %>%
    ungroup() -> variance_summary

variance_summary_cleaned <- variance_summary %>%
  select(which(colSums(select(., starts_with("var_")), na.rm = TRUE) != 0))
features <- colnames(variance_summary_cleaned[, 4:ncol(variance_summary_cleaned)])
features <- gsub("var_", "", features)
```


## Set phenotypic columns as predictors
```{r}
## Add column as factor
chunk_df$sample.renames <- as.factor(as.character(chunk_df$sample.renames))

## Make other columns factors too
chunk_df$bio_reps <- as.factor(as.character(chunk_df$bio_reps))

## Add phenotype as factor
chunk_df$phenotype <- as.factor(chunk_df$phenotype)
chunk_df$phenotype_DM1 <- as.factor(chunk_df$phenotype_DM1)

## Add Healthy as first level
chunk_df$phenotype <- relevel(as.factor(chunk_df$phenotype), ref = 'Healthy')
chunk_df$phenotype_DM1 <- relevel(as.factor(chunk_df$phenotype_DM1), ref = "Healthy")
chunk_df$Disease <- relevel(as.factor(chunk_df$Disease), ref = "Healthy")
```


## Chunk Name
```{r}
chunk_name <- paste0(params$chunk)
chunk_name <- gsub(".csv", "", chunk_name)
```



## Run Mixed Effect model for comparing individual disease groups
```{r}
#model <- run_rand_model_chunk(chunk_df, features, 'phenotype')
# Save to .csv
#write.csv(model, file = paste0('Model_', chunk_name, '.csv')) ##Can this be the same for every comparison?
```


## Run Mixed Effect model for comparing female and male lines
```{r}
model <- run_rand_model_chunk(chunk_df, features, 'sex')
# Save to .csv
write.csv(model, file = paste0('Model_Sex__', chunk_name, '.csv'))
```


## Run Mixed Effect model for comparing Healty and Patient lines
```{r} 
#model <- run_rand_model_chunk(chunk_df, features, 'Disease')
## Save to CSV
#write.csv(model, file = paste0('Model_AllDisease__', chunk_name, '.csv'))
```

## OPTIONAL
## Only select DM patient 2 and 3
```{r}

 DM_p1_names <- c(
   'IPS20_00122_1',
   'IPS20_00122_2',
   'IPS20_00122_3',
   'IPS23_00090_1',
   'IPS23_00090_2',
   'IPS23_00090_3'
)
 
sample_names <- rownames(colData(mae))
sample_names_NoDMP1 <- sample_names[!sample_names %in% DM_p1_names]


chunk_df_DM_P2_P3 <- chunk_df[sample_names_NoDMP1,]

```



## Run Mixed Effect model, comparing only DM1 group to all control group (including Rescue line)
```{r}
#model <- run_rand_model_chunk(chunk_df_DM_P2_P3, features, 'phenotype_DM1', DM1_ONLY=TRUE)
## Save to .csv
#write.csv(model, file = paste0('Model_DM1_CRISPR_', chunk_name, '.csv'))
```


## Session information

```{r}
sessionInfo()
```


