---
title: "Linear Mixed models"
author: "Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    df_print: paged
params:
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  mofa_scores: Z:/data/iPSC_data/Results/MOFA/All_Samples/SampleScores.csv
  CHD2_results: Z:/data/iPSC_data/Results/MixedModels/CHD2_MixedModel_Results.xlsx
  CHD2_results_cor: Z:/data/iPSC_data/Results/MixedModels_CORRECTED/CHD2_MixedModel_CORRECTED_Results.xlsx
  NANS_results: Z:/data/iPSC_data/Results/MixedModels/NANS_MixedModel_Results.xlsx
  NANS_results_cor: Z:/data/iPSC_data/Results/MixedModels_CORRECTED/NANS_MixedModel_CORRECTED_Results.xlsx
  outdir: Z:/data/iPSC_data/Results/MixedModels/Comparison
  config_file: config.yml
---


## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir
```

## Read data and metadata: MultiAssayExperiment

```{r}
suppressMessages(require(SummarizedExperiment))
suppressMessages(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
#mae
```


The MultiAssayExperiment contains individual omics assays

## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```


## Assay selection
```{r}
experiment_names <- config$experiment_names$general_selection
```


Phenotype information shared across all assays can be accessed via `colData()`

```{r sample_information_overview}
names(colData(mae))

pheno_df <- as.data.frame(colData(mae))

## Add column for Disease vs Healthy comparison
new_column <- lapply(pheno_df$phenotype, function(x) {
    if (startsWith(x, 'Diseas')) {
        value <- 'Disease'
    }
    else{
        value <- 'Healthy'
    }
})
pheno_df$Disease <- as.character(new_column)
```


## Correct omics values with MOFA factor scores
```{r}
mofa_scores_df <- read.csv(params$mofa_scores, row.names = 1)
```


## Load functions for Nested Mixed Effect model
```{r, warning=FALSE, message=FALSE}
source('Mixed_Model_functions.R')
```


```{r}
experiment_names
```


```{r}
source('../support_functions/Feature_plots.R')
```

## Proteomics

Compare Corrected and non-corrected for single features
```{r}
assay <- 'Proteomics'
confound_factors <- c("Factor1", "Factor2")
 # 1. Prepare data
data_prep <- prepare_feature_data(assay, mae)
df <- data_prep$df
features <- data_prep$features
  
# 2. Add confounding factors
df <- add_confounding_factors(df, mofa_scores_df, confound_factors)

# 3. Regress out confounding factors
df_corrected <- regress_out_factors(df, features, confound_factors)

# 4. Prepare final modeling dataframe
df_corrected <- prepare_modeling_data(df_corrected, df, pheno_df, DM1_ONLY=F)
df <- prepare_modeling_data(df, df, pheno_df, DM1_ONLY=F)
```

Check CHD2 proteomics features with highest difference 
```{r}
CHD2vsControls_proteomics <- readxl::read_xlsx(params$CHD2_results, sheet = 'Proteomics')
CHD2vsControls_proteomics_CORRECTED <- readxl::read_xlsx(params$CHD2_results_cor, sheet = 'Proteomics')

CHD2vsControls_proteomics_CORRECTED_sign <- CHD2vsControls_proteomics_CORRECTED[CHD2vsControls_proteomics_CORRECTED$p.adj < 0.05, ]
CHD2vsControls_proteomics_overlap <- CHD2vsControls_proteomics[CHD2vsControls_proteomics$feature %in% CHD2vsControls_proteomics_CORRECTED_sign$feature,]
```


```{r}
df1 <- CHD2vsControls_proteomics_overlap
df2 <- CHD2vsControls_proteomics_CORRECTED_sign

cor_plot(df1$coefficient, 
         df2$coefficient, 
         x_lab = 'Normal data', 
         y_lab =  'Corrected data', 
         title = 'Correlation of coefficients')
cor_plot(df1$std.error, 
         df2$std.error, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Standard Deviations')
cor_plot(df1$rand_effect, 
         df2$rand_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Random effects')
cor_plot(df1$nest_effect, 
         df2$nest_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Nested effects')
cor_plot(df1$p.value, 
         df2$p.value, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of nominal p-values')
```

Rank features based on highest difference in std.dev
```{r}
CHD2vsControls_proteomics_overlap$Delta_stdv <- CHD2vsControls_proteomics_overlap$std.error - CHD2vsControls_proteomics_CORRECTED_sign$std.error
head(CHD2vsControls_proteomics_overlap[ order(CHD2vsControls_proteomics_overlap$Delta_stdv, decreasing = T), c('feature', 'protein_name', 'Delta_stdv')])
```


```{r}
# S39AE
feature <- 'Q15043'

df_CHD2 <- df[df$phenotype %in% c('Healthy', 'Diseased: CHD2'),]
df_corrected_CHD2 <- df_corrected[df_corrected$phenotype %in% c('Healthy', 'Diseased: CHD2'),]

boxplot_per_group(df_input = df_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = '	S39AE_HUMAN (with correction)'
                  )
boxplot_per_group(df_input = df_corrected_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = '	S39AE_HUMAN (with correction)'
                  )
```

```{r}
# RPL35
boxplot_per_group(df_input = df_CHD2, feature = 'P42766', 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = '	RPL35 (no correction)')
boxplot_per_group(df_input = df_corrected_CHD2, 
                  feature = 'P42766', 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = '	RPL35 (with correction)')
```


```{r, fig.height=6, fig.width=7}
# MCM4
feature <- 'P33991'

boxplot_per_group(df_input = df_CHD2, 
                  feature = feature, 
                  assay_name = 'Proteomics', 
                  y_lab = 'Log2scaled Levels', 
                  title = '	MCM4 (no correction)')
boxplot_per_group(df_input = df_corrected_CHD2, 
                  feature = feature, 
                  assay_name = 'Proteomics', 
                  y_lab = 'Log2scaled Levels', 
                  title = '	MCM4 (with correction)')
```


```{r}
# C1GALT1
feature <- 'Q9NS00'

boxplot_per_group(df_input = df_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels', 
                  title = 'C1GALT1 (no correction)')
boxplot_per_group(df_input = df_corrected_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels', 
                  title = 'C1GALT1 (with correction)')
```


Check NANS proteomics features with highest difference 
```{r}
NANSvsControls_proteomics <- readxl::read_xlsx(params$NANS_results, sheet = 'Proteomics')
NANSvsControls_proteomics_CORRECTED <- readxl::read_xlsx(params$NANS_results_cor, sheet = 'Proteomics')

NANSvsControls_proteomics_CORRECTED_sign <- NANSvsControls_proteomics_CORRECTED[NANSvsControls_proteomics_CORRECTED$p.adj < 0.05, ]
NANSvsControls_proteomics_overlap <- NANSvsControls_proteomics[NANSvsControls_proteomics$feature %in% NANSvsControls_proteomics_CORRECTED_sign$feature,]
```


```{r}
cor_plot(NANSvsControls_proteomics_overlap$coefficient, 
         NANSvsControls_proteomics_CORRECTED_sign$coefficient, 
         x_lab = 'Normal data', 
         y_lab =  'Corrected data', 
         title = 'Correlation of coefficients')
cor_plot(NANSvsControls_proteomics_overlap$std.error, 
         NANSvsControls_proteomics_CORRECTED_sign$std.error, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Standard Deviations')
cor_plot(NANSvsControls_proteomics_overlap$rand_effect, 
         NANSvsControls_proteomics_CORRECTED_sign$rand_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Random effects')
cor_plot(NANSvsControls_proteomics_overlap$nest_effect, 
         NANSvsControls_proteomics_CORRECTED_sign$nest_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Nested effects')
cor_plot(NANSvsControls_proteomics_overlap$p.value, 
         NANSvsControls_proteomics_CORRECTED_sign$p.value, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of nominal p-values')
```


Rank features based on highest difference in std.dev
```{r}
NANSvsControls_proteomics_overlap$Delta_stdv <- NANSvsControls_proteomics_overlap$std.error - NANSvsControls_proteomics_CORRECTED_sign$std.error
head(NANSvsControls_proteomics_overlap[ order(NANSvsControls_proteomics_overlap$Delta_stdv, decreasing = T), c('feature', 'protein_name', 'Delta_stdv')])
```

Select only NANS data
```{r}
df_NANS <- df[df$phenotype %in% c('Healthy', 'Diseased: NANS-CDG'),]
df_corrected_NANS <- df_corrected[df_corrected$phenotype %in% c('Healthy', 'Diseased: NANS-CDG'),]
```


```{r}
# OLFL3_HUMAN
feature <- 'Q9NRN5'

boxplot_per_group(df_input = df_NANS, feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels', 
                  title = '	OLFL3_HUMAN (no correction)'
                  )
boxplot_per_group(df_input = df_corrected_NANS, feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels', 
                  title = '	OLFL3_HUMAN (with correction)'
                  )
```


```{r}
#	NOB1_HUMAN
feature <- 'Q9ULX3'

boxplot_per_group(df_input = df_NANS, feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels', 
                  title = '	NOB1_HUMAN (no correction)'
                  )
boxplot_per_group(df_input = df_corrected_NANS, feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels', 
                  title = '	NOB1_HUMAN (with correction)'
                  )
```
## Metabolomics RUMC

Check CHD2 metabolomics features with highest difference 
```{r}
assay <- 'Metabolomics, TML'

df1 <- readxl::read_xlsx(params$CHD2_results, sheet = assay)
df2 <- readxl::read_xlsx(params$CHD2_results_cor, sheet = assay)
df2 <- df2[df2$feature %in% df1$feature,]

df2 <- df2[df2$p.adj < 0.05, ]
df1 <- df1[df1$feature %in% df2$feature,]
```


```{r}
cor_plot(df1$coefficient, 
         df2$coefficient, 
         x_lab = 'Normal data', 
         y_lab =  'Corrected data', 
         title = 'Correlation of coefficients')
cor_plot(df1$std.error, 
         df2$std.error, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Standard Deviations')
cor_plot(df1$rand_effect, 
         df2$rand_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Random effects')
cor_plot(df1$nest_effect, 
         df2$nest_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Nested effects')
cor_plot(df1$p.value, 
         df2$p.value, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of nominal p-values')
```

Rank features based on highest difference in std.dev
```{r}
df1$Delta_stdv <- df1$std.error - df2$std.error
head(df1[ order(df1$Delta_stdv, decreasing = T), c('feature', 'Delta_stdv')])
```

Compare Corrected and non-corrected for single features
```{r}
confound_factors <- c("Factor1", "Factor2")
 # 1. Prepare data
data_prep <- prepare_feature_data(assay, mae)
df <- data_prep$df
features <- data_prep$features
  
# 2. Add confounding factors
df <- add_confounding_factors(df, mofa_scores_df, confound_factors)

# 3. Regress out confounding factors
df_corrected <- regress_out_factors(df, features, confound_factors)

# 4. Prepare final modeling dataframe
df_corrected <- prepare_modeling_data(df_corrected, df, pheno_df, DM1_ONLY=F)
df <- prepare_modeling_data(df, df, pheno_df, DM1_ONLY=F)
```


```{r}
#TML_neg483
feature <- 'TML_neg483'
assay <- 'Metabolomics RUMC'

df_CHD2 <- df[df$phenotype %in% c('Healthy', 'Diseased: CHD2'),]
df_corrected_CHD2 <- df_corrected[df_corrected$phenotype %in% c('Healthy', 'Diseased: CHD2'),]

boxplot_per_group(df_input = df_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = '	TML_neg483; L-Aspartic acid (no correction)'
                  )
boxplot_per_group(df_input = df_corrected_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = 'TML_neg483; L-Aspartic acid (with correction)'
                  )
```
```{r}
#TML_neg483
feature <- 'TML_pos4365'
assay <- 'Metabolomics RUMC'

df_CHD2 <- df[df$phenotype %in% c('Healthy', 'Diseased: CHD2'),]
df_corrected_CHD2 <- df_corrected[df_corrected$phenotype %in% c('Healthy', 'Diseased: CHD2'),]

boxplot_per_group(df_input = df_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = '	TML_pos4365 (no correction)'
                  )
boxplot_per_group(df_input = df_corrected_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = 'TML_pos4365 (with correction)'
                  )
```
Check NANS metabolomics features with highest difference 
```{r}
df1 <- readxl::read_xlsx(params$NANS_results, sheet = 'Metabolomics, TML')
df2 <- readxl::read_xlsx(params$NANS_results_cor, sheet = 'Metabolomics, TML')
df2 <- df2[df2$feature %in% df2$feature,]

df2 <- df2[df2$p.adj < 0.05, ]
df1 <- df1[df1$feature %in% df2$feature,]
```


```{r}
cor_plot(df1$coefficient, 
         df2$coefficient, 
         x_lab = 'Normal data', 
         y_lab =  'Corrected data', 
         title = 'Correlation of coefficients')
cor_plot(df1$std.error, 
         df2$std.error, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Standard Deviations')
cor_plot(df1$rand_effect, 
         df2$rand_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Random effects')
cor_plot(df1$nest_effect, 
         df2$nest_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Nested effects')
cor_plot(df1$p.value, 
         df2$p.value, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of nominal p-values')
```

Rank features based on highest difference in std.dev
```{r}
df1$Delta_stdv <- df1$std.error - df2$std.error
head(df1[order(df1$Delta_stdv, decreasing = T), c('feature', 'Delta_stdv')])
```

```{r}
#TML_pos4365
feature <- 'TML_pos4365'

df_NANS <- df[df$phenotype %in% c('Healthy', 'Diseased: NANS-CDG'),]
df_corrected_NANS <- df_corrected[df_corrected$phenotype %in% c('Healthy', 'Diseased: NANS-CDG'),]

boxplot_per_group(df_input = df_NANS, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = '	TML_pos4365 (no correction)'
                  )
boxplot_per_group(df_input = df_corrected_NANS, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = 'TML_pos4365 (with correction)'
                  )
```
## RNAseq

Check CHD2 RNA features with highest difference 
```{r}
assay <- 'RNAseq'

df1 <- readxl::read_xlsx(params$CHD2_results, sheet = assay)
df2 <- readxl::read_xlsx(params$CHD2_results_cor, sheet = assay)
df2 <- df2[df2$feature %in% df2$feature,]

df2 <- df2[df2$p.adj < 0.05, ]
df1 <- df1[df1$feature %in% df2$feature,]
```


```{r}
cor_plot(df1$coefficient, 
         df2$coefficient, 
         x_lab = 'Normal data', 
         y_lab =  'Corrected data', 
         title = 'Correlation of coefficients')
cor_plot(df1$std.error, 
         df2$std.error, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Standard Deviations')
cor_plot(df1$rand_effect, 
         df2$rand_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Random effects')
cor_plot(df1$nest_effect, 
         df2$nest_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Nested effects')
cor_plot(df1$p.value, 
         df2$p.value, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of nominal p-values')
```

Rank features based on highest difference in std.dev
```{r}
df1$Delta_stdv <- df1$std.error - df2$std.error
head(df1[ order(df1$Delta_stdv, decreasing = T), c('feature', 'Delta_stdv')])
```

Compare Corrected and non-corrected for single features
```{r}
assay_name <- 'RNAseq | RTC'
confound_factors <- c("Factor1", "Factor2")
 # 1. Prepare data
data_prep <- prepare_feature_data(assay_name, mae)
df <- data_prep$df
features <- data_prep$features
  
# 2. Add confounding factors
df <- add_confounding_factors(df, mofa_scores_df, confound_factors)

# 3. Regress out confounding factors
df_corrected <- regress_out_factors(df, features, confound_factors)

# 4. Prepare final modeling dataframe
df_corrected <- prepare_modeling_data(df_corrected, df, pheno_df, DM1_ONLY=F)
df <- prepare_modeling_data(df, df, pheno_df, DM1_ONLY=F)
```


```{r}
feature <- 'ENSG00000111799'

df_CHD2 <- df[df$phenotype %in% c('Healthy', 'Diseased: CHD2'),]
df_corrected_CHD2 <- df_corrected[df_corrected$phenotype %in% c('Healthy', 'Diseased: CHD2'),]

boxplot_per_group(df_input = df_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = paste0(feature, ' (no correction)')
                  )
boxplot_per_group(df_input = df_corrected_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = paste0(feature, ' (with correction)')
                  )
```


```{r}
feature <- 'ENSG00000105971'

df_CHD2 <- df[df$phenotype %in% c('Healthy', 'Diseased: CHD2'),]
df_corrected_CHD2 <- df_corrected[df_corrected$phenotype %in% c('Healthy', 'Diseased: CHD2'),]

boxplot_per_group(df_input = df_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = paste0(feature, ' (no correction)')
                  )
boxplot_per_group(df_input = df_corrected_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = paste0(feature, ' (with correction)')
                  )
```


```{r}
feature <- 'ENSG00000138061'

df_CHD2 <- df[df$phenotype %in% c('Healthy', 'Diseased: CHD2'),]
df_corrected_CHD2 <- df_corrected[df_corrected$phenotype %in% c('Healthy', 'Diseased: CHD2'),]

boxplot_per_group(df_input = df_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = paste0(feature, ' (no correction)')
                  )
boxplot_per_group(df_input = df_corrected_CHD2, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = paste0(feature, ' (with correction)')
                  )
```
Check NANS RNA features with highest difference 
```{r}
assay <- 'RNAseq'

df1 <- readxl::read_xlsx(params$NANS_results, sheet = assay)
df2 <- readxl::read_xlsx(params$NANS_results_cor, sheet = assay)
df2 <- df2[df2$feature %in% df2$feature,]

df2 <- df2[df2$p.adj < 0.05, ]
df1 <- df1[df1$feature %in% df2$feature,]
```


```{r}
cor_plot(df1$coefficient, 
         df2$coefficient, 
         x_lab = 'Normal data', 
         y_lab =  'Corrected data', 
         title = 'Correlation of coefficients')
cor_plot(df1$std.error, 
         df2$std.error, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Standard Deviations')
cor_plot(df1$rand_effect, 
         df2$rand_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Random effects')
cor_plot(df1$nest_effect, 
         df2$nest_effect, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of Nested effects')
cor_plot(df1$p.value, 
         df2$p.value, 
         x_lab = 'Normal data', 
         y_lab = 'Corrected data', 
         title = 'Correlation of nominal p-values')
```

Rank features based on highest difference in std.dev
```{r}
df1$Delta_stdv <- df1$std.error - df2$std.error
head(df1[ order(df1$Delta_stdv, decreasing = T), c('feature', 'Delta_stdv')])
```



```{r}
feature <- 'ENSG00000104341'

df_NANS <- df[df$phenotype %in% c('Healthy', 'Diseased: NANS-CDG'),]
df_corrected_NANS <- df_corrected[df_corrected$phenotype %in% c('Healthy', 'Diseased: NANS-CDG'),]

boxplot_per_group(df_input = df_NANS, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = paste0(feature, ' (no correction)')
                  )
boxplot_per_group(df_input = df_corrected_NANS, 
                  feature = feature, 
                  assay_name = assay, 
                  y_lab = 'Log2scaled Levels',
                  title = paste0(feature, ' (with correction)')
                  )
```






## Session information

```{r}
sessionInfo()
```


