---
title: 'Linear models'
author:
  - Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  mrna_read_data: Z:/data/iPSC_data/iPSC_RNAseq/iPSC_RTC/RawCounts/
  outdir: Z:/data/iPSC_data/results/RLM/
  config_file: "config.yml"
---

## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir
```

## Read data and metadata: MultiAssayExperiment

```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
mae
```

The MultiAssayExperiment contains individual omics assays


## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

## Assay selection
```{r}
experiment_names <- config$experiment_names$general_selection
experiment_names
```

Phenotype information shared across all assays can be accessed via `colData()`

```{r sample_information_overview}
names(colData(mae))

pheno_df <- as.data.frame(colData(mae))
```

## Read .txt files
```{r}
setwd(params$mrna_read_data)
txt_files <- list.files( pattern = "\\counts.txt$")
counts_tables <- lapply(txt_files, function(x) {
 df <- read.table(x, sep = '\t', header = F, stringsAsFactors = F)
 colnames(df) <- c("Key", "Value")
 
 named_vec <- setNames(df$Value, df$Key)
})

mrna_read_data <- as.data.frame(do.call(rbind, counts_tables))
rownames(mrna_read_data) <- sub("\\.counts.txt$", "", txt_files)
```

## Selec DM1 patient 1 lines
```{r}
dm1_lines <- c('DM1-p1c1-RNA-1', 'DM1-p1c1-RNA-2',
               'DM1-p2c1rep-RNA-1', 'DM1-p2c1rep-RNA-2')
mrna_read_data <- mrna_read_data[dm1_lines, ]

new_names <- c('IPS20_00036_1', 'IPS20_00036_2',
               'IPS23_00090_1', 'IPS23_00090_2')
rownames(mrna_read_data) <- new_names
mrna_read_data <- data.frame(t(mrna_read_data))

pheno_df <- pheno_df[new_names,]
pheno_df$phenotype <- as.factor(pheno_df$phenotype)
```


## Create DESeqdata
```{r}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(countData = mrna_read_data,
                              colData = pheno_df,
                              design = ~ phenotype)

# Filter low count reads
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

## Run DEseq
```{r}
dds <- DESeq(dds)
```

## Results
```{r}
res <- results(dds, name ="phenotype_Rescue..repeat.excised.CRISPR.DM1_vs_Diseased..DM1")
summary(res)
```
## Top DE genes
```{r}
res_ordered <- as.data.frame(res[order(res$padj),])
res_ordered[1:20,]
```

## Number of significants
```{r}
nrow(res_ordered[res_ordered$padj < 0.05,])
```

## Volcano plot
```{r}
library(ggplot2)
library(ggrepel)

resLFC <- lfcShrink(dds, coef = "phenotype_Rescue..repeat.excised.CRISPR.DM1_vs_Diseased..DM1", type= 'normal')

resLFC_df <- as.data.frame(resLFC)

resLFC_df$threshold <- as.factor(ifelse(resLFC_df$padj < 0.05 & abs(resLFC_df$log2FoldChange) > 1,
                                        "Significant", "Not Significant"))


## Map to gene symbols
rna_mapping <- read.csv('Z:/users/devissercasper/RNA_seq_RTC_maping_table.csv', sep = ';')

resLFC_df$symbols <- rna_mapping$Gene_symbol[match(rownames(resLFC_df), rna_mapping$feature)]
resLFC_df$gene <- rownames(resLFC_df)
top_genes <- resLFC_df$symbols[order(resLFC_df$padj)][1:20]

ggplot(resLFC_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = threshold), alpha = 0.8, size = 1.5) +
  geom_text_repel(data = resLFC_df[resLFC_df$symbols %in% top_genes,],
                  aes(label = symbols), size = 3, max.overlaps = 10) +
  scale_color_manual(values = c("grey", "red")) +
  xlab("Log2 Fold Change") +
  ylab("-Log10 adjusted p-value") +
  theme_bw()
```


## Session information
```{r}
sessionInfo()
```


