---
title: 'Linear models'
author:
  - Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  outdir: Z:/data/iPSC_data/results/RLM/
  config_file: "config.yml"
---

## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir
```

## Read data and metadata: MultiAssayExperiment

```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
mae
```

The MultiAssayExperiment contains individual omics assays


## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

## Assay selection
```{r}
experiment_names <- config$experiment_names$general_selection
```

Phenotype information shared across all assays can be accessed via `colData()`
```{r sample_information_overview}
names(colData(mae))

pheno_df <- as.data.frame(colData(mae))

## Add column for Disease vs Healthy comparison
new_column <- lapply(pheno_df$phenotype, function(x) {
    if (startsWith(x, 'Diseas')) {
        value <- 'Disease'
    }
    else{
        value <- 'Healthy'
    }
})
pheno_df$Disease <- as.character(new_column)
```


## Load functions for Nested Mixed Effect model
```{r, warning=FALSE, message=FALSE}
source('Mixed_Model_functions.R')
```


## Omics layers
```{r}
experiment_names
```

## Adapted function for running linear model
```{r}
run_rand_model_chd2 <- function(assay_name, pred, chd2_patient, nest_effect=FALSE) {
  se <- experiments(mae)[assay_name]
  df <- get_df_input_lm(se)
  features <- colnames(df)
  features <- features[features != "ENSG00000129824"] #TODO
  # add biological variables to data frame
  df <- cbind(df, pheno_df[rownames(df), ])
  df <- add_nested_cols(df)
  df <- code_phenotype_column(df)
  # select CHD2 patient
  df <- df[df$observation.unit.identifier.crispr %in% c('CTRL_409B',
                                                        'CTRL_WTC', 
                                                        'CTRL_Sendai', 
                                                        chd2_patient),]
  ## Remove features that are with only missing values
  df <- df[,colSums(!is.na(df)) > 0]
  df <- df[, sapply(df, function(x) length(unique(na.omit(x))) > 1)]
  features <- features[features %in% colnames(df)]
  # fit linear mixed model
  model_result <- do.call(rbind, lapply(
    features,
    run_MixedModel_df,
    df=df,
    predictor=pred,
    nest_effect=nest_effect))
  #adjust p-values
  model_result$p.adj <- p.adjust(model_result$p.value, method = "BH")
  model_result$rank <- rank(model_result$p.value)
  
  ## Save model to .csv
  assay_name <- paste0(assay_name)
  assay_name <- gsub(" ", "_", assay_name)
  assay_name <- gsub("\\|", "", assay_name)
  assay_name <- gsub(",", "", assay_name)
  
  ## file out path
  outfile <- paste0("Model_", pred, "_", chd2_patient, "_", assay_name, ".csv")
  write.csv(model_result, outfile)
  
  return(model_result)
}
```


## Run Mixed Effect model, comparing individual disease groups with the control lines
```{r}
# Select assays
assay_names <- experiment_names[c(2,3,4,5,6,7)]

mae_rlm_RandomEffect <- lapply(
     assay_names, run_rand_model_chd2, pred='phenotype', chd2_patient = 'CHD2_CRISPR')
```


## Volcano plot
```{r}

```


## Session information
```{r}
sessionInfo()
```


