---
title: 'Multi-omics Factor Analysis (MOFA)'
author:
- Casper de Visser^[Radboud university medical center, casper.devisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  outdir: Z:/data/iPSC_data/results/MOFA
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  neuron_markers: Z:/data/iPSC_data/Markers_files/Neuron_markers.csv
  config_file: 'config.yml'
---

This notebook performs Multi-omics Factor Analysis (MOFA) [@Argelaguet2018,@Argelaguet2020] on preprocessed data. 

MOFA is an unsupervised method that uncovers main sources of variation in multi-omics data. It allows to identify variation shared across omics data types or specific to individual omics data types. 
Subsequent to MOFA, correlation between MOFA factors and phenotypic information is determined to identify possible biological factors contributing to the observed variation.


```{r show_parameters}
print(params)
dir.create(params$outdir)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!dir.exists(outdir)){
    outdir <- getwd()
}

## Create subdirectory for All analyzing all samples together
outdur_disease <- file.path(outdir, 'All_Samples')
dir.create(outdur_disease)
```


## Libraries
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(ggokabeito)
library(MOFA2)
library(data.table)
library(ggrepel)
library(pheatmap)
```


## Read input data

Input data is a MultiAssayExperiment (MAE) object. All or specified assays in the MAE will be used for analysis Multi-omics Factor Analysis (MOFA).

```{r read_input_data, message=FALSE}
library(MultiAssayExperiment)
library(HDF5Array)

mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')

```


Omics to be included in MOFA

## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

## Assay selection
```{r}
assay_names <- config$experiment_names$general_selection
#assay_names <- config$experiment_names$Annotated_data
names(assay_names) <- assay_names
assay_names
```


## Prepare phenotype data
```{r}
phenotype_data <- as.data.frame(colData(mae)) %>%
  tibble::rownames_to_column("sample") %>%
  mutate(
    Disease = if_else(startsWith(phenotype, "Diseas"), "Disease", "Healthy"),
    phenotype_CHD2_CRISPR = recode(
      phenotype,
      'Diseased: CHD2 CRISPR' = 'Diseased: CHD2'
    ),
    phenotype_CRISPRS_both = recode(
      phenotype_CHD2_CRISPR,
      'Rescue: repeat excised CRISPR DM1' = 'Healthy'
  )
  ) %>%
  mutate(across(where(is.character), as.factor))
```


## Multi-omics factor analysis (MOFA)

For running MOFA, it is recommended to first remove known batch effects. 
Make sure to use input data that were adjusted for batch effect if necessary.

## Shorter readable omics names
```{r}
shoter_names <- config$experiment_names$shorter_names
shoter_names
```


Include custom MOFA functions
```{r}
source_paths <- c(
  "../support_functions/MOFA_functions.R",      # when running in Rstudio
  "support_functions/MOFA_functions.R"                    # when running in Nextflow
)

for (path in source_paths) {
  if (file.exists(path)) {
    source(path)
    break
  }
}
```


Prepare MOFA input data: create a long data frame with sample, feature, view, 
[group], and value. 

```{r prepare_mofa_input}
mofa_input <- mae_to_MOFAinput(mae)
mofa_input$value <- as.numeric(mofa_input$value)
```


Create the MOFA object

```{r create_mofa_object}
MOFAobject <- create_mofa(mofa_input)
print(MOFAobject)
```


Plot views available in MOFA object (N = number of samples, D = number of 
features per view)

```{r show_mofa_modalities}
require(ggplot2)
ggp <- plot_data_overview(MOFAobject)

ggsave(file.path(outdur_disease, "MOFA_modality_overview.pdf"), 
       ggp, width = 15, height = 10.46, units = "cm", dpi = 300)

ggp
```


Get parameters of MOFA object

```{r}
data_opts <- get_default_data_options(MOFAobject)
data_opts
```

Set number of factors. This parameter can have a large impact on the results.

```{r}
model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 6
model_opts
```


```{r}
train_opts <- get_default_training_options(MOFAobject)
train_opts
```

Prepare MOFA object for training using the parameters specified above

```{r}
MOFAobject <- prepare_mofa(
  object = MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
```


**NOTE:** `mofa2` requires a Python installation with `mofapy2`

`mofapy2` can be installed from PyPi using conda:

```
conda create -n mofapy2 python pip
conda activate mofapy2
pip install mofapy2
```

In case of computing environments without internet access, the conda environment
can be created on a computer with internet and pip dependencies for `mofapy2`
can be downloaded:

```
pip list --format=freeze > requirements.txt
pip download -r requirements.txt
```

After uploading the dependencies to the environment without internet access,
`mofapy2` can be installed with pip defining the path to the package wheel:

```
pip install --no-index --find-links Z:/software/pip Z:/software/pip/mofapy2-0.6.7-py3-none-any.whl
```

The conda environment with Python can now be used by `reticulate` in R.

```{r define_python_version}
#reticulate::use_condaenv(condaenv = "mofapy")
reticulate::use_python("/usr/bin/python3", required = T)
```

Train the MOFA model

```{r run_mofa, message=FALSE}
outdir <- params$outdir
outfile = file.path(outdir, "model_ALL_SAMPLES.hdf5")
MOFAobject.trained <- run_mofa(MOFAobject, outfile)
```


## Read in MOFA model
```{r}
#outfile = file.path(outdir, "model_ALL_SAMPLES_Prot_Glob_Min.hdf5")
#MOFAobject.trained <- MOFA2::load_model(outfile)
```



## Visualizations of MOFA results 


### Variance decomposition

Amount of variance explained per modality (= view, = assay)

```{r}
head(MOFAobject.trained@cache$variance_explained$r2_total)
```

Amount of variance explained by each factor per modality

```{r}
head(MOFAobject.trained@cache$variance_explained$r2_per_factor)
```

Plot amount of variance explained by each factor per modality

```{r, fig.width=10, fig.height=5}
expl_var <- c(
  get_variance_explained(MOFAobject.trained)$r2_per_factor$single_group)

ggp <- plot_variance_explained(MOFAobject.trained) +
  geom_text(
    label = round(expl_var, 2),
    color = ifelse(expl_var < 0.6*max(expl_var), "black", "white")) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )
      

ggsave(file.path(outdur_disease, "MOFA_explained_variance.pdf"), 
       ggp, width = 14, height = 10.46, units = "cm", dpi = 300)

ggp
```

###  Factor correlations

Plot correlations between factors. When factors are correlated (containing 
redundant information) it is recommended to reduce the number of factors

```{r}
plot_factor_cor(MOFAobject.trained)
```

### Factor weights

Plot feature weights per factor and assay

```{r}
num_factors <- nrow(
  MOFAobject.trained@cache$variance_explained$r2_per_factor[[1]])
view_names <- names(MOFAobject.trained@data)

lapply(1:5, function(f) {
 lapply(view_names, function(a) {
   plot_weights(
      MOFAobject.trained,
      view = a,
      factor = f,
      nfeatures = 10
      )+ ggtitle(paste0(a))
  }) 
})
```


Top weights per factor and assay

```{r}
lapply(1:2, function(f) { 
  lapply(view_names, function(a) {
    ggp <- plot_top_weights(
      MOFAobject.trained,
      view = a,
      factor = f,
      nfeatures = 10
    ) + ggtitle(paste0(a))
    file.out <- paste0('Factor', f, '_', a, '.png')
    ggsave(file.path(outdur_disease, file.out), ggp)
  })
})
```


# Compare disease groups on MOFA scores

Add phenotype data to MOFA object
```{r}
samples_metadata(MOFAobject.trained) <- phenotype_data
```


Plot MOFA scores per category
```{r}
plot_factor_category <- function(MOFAobject.trained, 
                                 var, 
                                 excluded_categories = c(),
                                 num_factors = NULL) {
  
    
  if (is.factor(MOFAobject.trained@samples_metadata[[var]])) {
    original_categories <- levels(MOFAobject.trained@samples_metadata[[var]])
  } else {
    original_categories <- levels(
      as.factor(MOFAobject.trained@samples_metadata[[var]]))
  }
  
  categories <- original_categories[
    !original_categories %in% excluded_categories]
  if (length(categories) > 1) {
    comparisons <- as.list(as.data.frame(combn(categories, 2)))
  }
  
  num_factors <- ifelse(
    is.null(num_factors),
    nrow(MOFAobject.trained@cache$variance_explained$r2_per_factor[[1]]),
    num_factors)
  
  ggp <- plot_factor(
    MOFAobject.trained,
    factor = 1:num_factors,
    color_by = var,
    dodge = TRUE,
    add_boxplot = TRUE) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          panel.spacing = unit(0, "lines")) 
  
  #ggp$data$color_by <- as.factor(ggp$data$color_by)
  ggp$data$group_by <- ggp$data$color_by
  
  add_stat <- TRUE
  if (length(categories) > 2) {
    # kruskal wallis
    stat_method <- "Kruskal-Wallis"
    stat_test <- ggp$data %>%
      filter(!color_by %in% excluded_categories) %>%
      group_by(factor) %>%
      kruskal_test(value ~ group_by) %>%
      adjust_pvalue(method = "fdr") 
    
    stat_test_pairwise <- ggp$data %>%
      filter(!color_by %in% excluded_categories) %>%
      group_by(factor) %>%
      dunn_test(value ~ group_by, p.adjust.method = "fdr") %>%
      add_y_position()
  } else if (length(categories) == 2) {
    # wilcoxon
    stat_method <- "Wilcoxon"
    stat_test <- stat_test_pairwise <- ggp$data %>%
      group_by(factor) %>%
      wilcox_test(value ~ group_by, comparisons = comparisons) %>%
      adjust_pvalue(method = "fdr") %>%
      add_significance() %>%
      add_y_position()
  } else {
    add_stat <- FALSE
  }
  ggp <- ggp + scale_fill_okabe_ito(
    breaks = original_categories,
    labels = original_categories)
  ggp
}
```


Get Sample scores and save as csv
```{r}
scores_df <- get_factors(MOFAobject.trained, factors = "all", as.data.frame = T)

wide_df <- scores_df |>
  select(-group) |>
  pivot_wider(
    names_from = factor,
    values_from = value
  )

wide_df<- wide_df %>%
    left_join(phenotype_data)


## Shorter name for DM1 CRISPR line
levels(wide_df$phenotype) <- c(levels(wide_df$phenotype), "CRISPR DM1")
wide_df$phenotype[wide_df$phenotype == 'Rescue: repeat excised CRISPR DM1'] <- 'CRISPR DM1'


write.csv(wide_df, file.path(outdur_disease, "SampleScores.csv"), row.names = F)
```

### Factor 1 vs 2
```{r, fig.width=12}
ggp <- Factor_scores_plot(wide_df, Col1 = 'Factor1', Col2 = 'Factor2', )
ggsave(filename = file.path(outdur_disease, "SampleScores_F1F2.pdf"), ggp, width = 10, height = 7)
ggp
```


### Factor 3 vs 4
```{r, fig.width=12}
ggp <- Factor_scores_plot(wide_df, Col1 = 'Factor3', Col2 = 'Factor4', )
ggsave(filename = file.path(outdur_disease, "SampleScores_F3F4.pdf"), ggp, width = 10, height = 7)
ggp
```


### Factor 5 vs 6
```{r}
ggp <- Factor_scores_plot(wide_df, Col1 = 'Factor5', Col2 = 'Factor6', )
ggsave(filename = file.path(outdur_disease, "SampleScores_F5F6.pdf"), ggp, width = 10, height = 7)
ggp
```


```{r, fig.width=10, fig.height=5}
ggp <- plot_factor_category(MOFAobject.trained, "phenotype", num_factors = 2)

# ggsave(file.path(outdur_disease, "MOFA_factors_vs_Age.group.png"), 
#        ggp, width = 28.04, height = 10.46, units = "cm", dpi = 300)
ggp
```

```{r, fig.width=15, fig.height=5}
# ggp <- plot_factor_category(MOFAobject.trained, "phenotype_CHD2_CRISPR")
# 
# # ggsave(file.path(outdur_disease, "MOFA_factors_vs_Age.group.png"), 
# #        ggp, width = 28.04, height = 10.46, units = "cm", dpi = 300)
# ggp
```



```{r}
ggp <- plot_factor_category(MOFAobject.trained, "sex")

# ggsave(file.path(outdur_disease, "MOFA_factors_vs_Age.group.png"), 
#        ggp, width = 28.04, height = 10.46, units = "cm", dpi = 300)
ggp
```

Save loadings to Excel or CSV
```{r}
MOFA_weights <- get_weights(MOFAobject.trained)
names(MOFA_weights) <- assay_names

lapply(assay_names, function(a) {
 df <- cbind(MOFA_weights[[a]], data.frame(rowData(mae[[a]])))
 df$feature <- rownames(df)
 f <- gsub("[| ,]", "_", a)
 write.csv(df, file = file.path(outdur_disease,  paste0("factorweights_", f, ".csv")))
})
```


## Lipidomics heatmap

```{r}
# 1. extract weights
w <- get_weights(MOFAobject.trained, views = "Lipidomics_Targeted", factors = c(1,2))[[1]]
# 2. merge with metadata
lipidomics_feature_metadata <- as.data.frame(rowData(mae[["Lipidomics, Targeted | imputed | log  | TA"]]))
df <- cbind(w, lipidomics_feature_metadata)

## Subset Triglycerides
df_noTG <- df[df$Class != "TG",] 
df_TG <- df[df$Class == "TG",]

## Function to make heatmap
make_heatmap <- function(df){
  ggp <- pheatmap(
    df[,c(1,2)],
    cluster_rows = T,
    cluster_cols = F,
    show_colnames = T,
    show_rownames = F,
    annotation_row = df[,"Class", drop = F],
    scale = "column"
  )
  return(ggp)
} 

make_heatmap(df)
make_heatmap(df_noTG)
make_heatmap(df_TG)
```


```{r}
sessionInfo()
```

