---
title: "Principal component analysis (PCA)"
author: "Casper de Visser^[Radboud university medical center, casper.devisser@radboudumc.nl]"
output:
  html_document:
    df_print: paged
  pdf_document: default
params:
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  outdir: Z:/data/iPSC_data/results/PCA/
  config_file: 'config.yml'
---


## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!dir.exists(outdir)){
    outdir <- getwd()
}
```

## Read data and metadata: MultiAssayExperiment

```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
mae
```

The MultiAssayExperiment contains individual omics assays

```{r}
experiment_names <- names(experiments(mae))
experiment_names
```



Technical variables are available per assay

```{r}
tech_vars_factor <- lapply(experiments(mae), function(x) {
  num_levels <- apply(colData(x), 2, function(x) {length(levels(as.factor(x)))})
  t <- names(colData(x))[num_levels <= 9 & num_levels > 1]
  t <- t[t!="subject.id"]
  t <- t[!t %in% names(colData(mae))]
})
tech_vars_factor

tech_vars_continuous <- lapply(experiments(mae), function(x) {
  num_levels <- apply(colData(x), 2, function(x) {length(levels(as.factor(x)))})
  t <- names(colData(x))[num_levels > 9]
  t <- t[t!="subject.id"]
  t <- t[t!="measurement.id"]
  t <- t[t!="measurement_names"]
  t <- t[t!="ID.Patient"]
  t <- t[!t %in% c("SampleID", "Radboud.code", "Kevin.nr", "Kevin.CPM", 
                   "Sample.name", "ISA_name", "SampleID_withrep", "Sample_Overview_names")] #TODO, remove these from sample metadata sheet
  t <- t[!t %in% c("GroupNum", "GroupName", "SampleNorm", "Radboud.code.1")] 

  t <- t[!t %in% names(colData(mae))]
})
tech_vars_continuous
```

**NOTE:** consider add sum (total carnitines) and ratios (vlcfa: C24/C22 and C26/C22)

Phenotype information shared across all assays can be accessed via `colData()`

```{r}
names(colData(mae))
```
## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

## Assay selection
```{r}
assay_names <- config$experiment_names$PCA_selection
assay_names
```



## Principal component analysis -- PCA

PCA performed on standardized (centered to mean and scaled to unit-variance) data.

```{r run_pca}
run_pca_se <- function(name_assay) {
  
  print(name_assay)
  se <- mae[[name_assay]]
  df_wide <- data.frame(t(assay(se)))
  
  # Remove columns that are zero
  df_wide <- df_wide[ , colSums(is.na(df_wide)) ==0]
  #df_wide$Constant <- 1
  # Remove columns that are constants
  cols_remove <- names(df_wide[, sapply(df_wide, function(v) var(v, na.rm=TRUE) == 0)])
  df_wide <- df_wide[ , !(names(df_wide) %in% cols_remove)] 
  pca <- prcomp(df_wide, center = T, scale. = T)
}


#assay_names <- experiment_names
pca <- lapply(assay_names, function(n) {run_pca_se(n)})

## Add experiment names to pca
names(pca) <- assay_names
```


### PCA -- scree plots 

Scree plots are bar plots showing the explained variation per Principal Component (PC).
Cumulative explained variation is shown as a line.

```{r pcs_scree_plots}
library(ggplot2)
#'  Create a screeplot of the variance explained by principal components
#'  
#'  @param pc PCA as input
#'  @param name_file Name of the barplot (.png file)
#'  @return The screeplot object 
make_screeplot <- function(pc, main_title) {
  require(ggplot2)
  require(ggbreak)
  # create summary of pca
  vars <- pc$sdev^2
  vars <- vars/sum(vars)
  df <- as.data.frame(cbind(
    'SD' = pc$sdev, 'Prop' = vars, 'cumulative' = cumsum(vars)))
  df$PC <- seq(nrow(df))
  df$Prop <- df$Prop * 100
  df$cumulative <- df$cumulative * 100
  df <- df[c('PC', 'SD', 'Prop', 'cumulative')]
  df <- as.data.frame(lapply(df, unlist))
  
  # Get number of PCs that explain 80%
  eighty_percent <-  min(which(df$cumulative > 80))
  scale <- round(eighty_percent/25)
  if (scale == 0) {
    scale <- 1
  }
  
  # Screeplot
  plot1 <- ggplot(data = df[df$PC <= eighty_percent+3, ], 
                  aes(x = PC, y = Prop)) + #reorder(PC, -Prop)
    geom_bar(stat = 'identity') +
    geom_line(aes(y=cumulative,group=1 ), size = 1, color = 'red') +
    xlab('Principal Component') +
    ylab('Proportion of variance explained (%)') +
    theme(plot.title = element_text(size = 12, hjust = 0.5),
          legend.position = 'none') +
    ggtitle(main_title)
  plot1 <- plot1 + 
    geom_vline(aes(xintercept = eighty_percent), size = 2) +
    geom_label(aes(x = eighty_percent, y = 0.50,
                   label = '80%', vjust = -1, size = 8)) 
  plot1 <- plot1 +
    scale_x_continuous(breaks = seq(1, length(df$PC), by = 1), position = "bottom")
  if (eighty_percent > 18) {
    plot1 <- plot1 + scale_x_break(breaks = c(13, eighty_percent-3))
  }
  return(plot1)
}


lapply(names(pca), function(x) {
  make_screeplot(pca[[x]], paste0("PCA, scree plot\n", x))
})
```


### PCA -- score plots 

#### Pairwise scatter plots of PC scores colored by biological and technical variables

```{r pca_score_plots, fig.width=8, warning=FALSE, message=FALSE}
library(ggplot2)
library(GGally)
library(ggokabeito) # Okabe-Ito

bio_vars <- c(
  #"age.at.sampling.from.host", 
  "phenotype",
  #"observation.unit.identifier",
  "tissues.from.which.cells.were.derived",
  "culturing.strategy",
  "sex")

num_PCs <- 3

lapply(names(pca), function(assay_name) { 
  lapply(c(bio_vars, tech_vars_factor[[assay_name]]), function(color_var) {
    #message("PCA score plots - ", assay_name, ", ", color_var)
    # PCA scores 
    pca_scores <- as.data.frame(pca[[assay_name]]$x[, 1:num_PCs])
    # data for coloring
    if (color_var %in% bio_vars) {
      pca_scores[[color_var]] = colData(mae)[rownames(pca_scores), color_var]
    } else {
      pca_scores[[color_var]] = colData(experiments(mae)[[assay_name]])[
        rownames(pca_scores), color_var]
    }
    num_factor_levels <- length(levels(as.factor(pca_scores[[color_var]])))
    if (num_factor_levels > 1) { # multiple factor levels
      pca_scores[[color_var]][is.na(pca_scores[[color_var]])] <- "NA"
      # pairs plot of PCs
      ggp <- ggpairs(
        pca_scores, 
        columns = 1:num_PCs,
        aes(color = .data[[color_var]]),#, shape = .data[[color_var]], alpha = 0.25),
        progress = FALSE,
        diag = list(continuous = "barDiag"),
        upper = list(continuous = "blank"),
        legend = 1
      ) +
        ggtitle(stringr::str_wrap(paste0(
          "PCA score plot, color by ", color_var, ", ", assay_name), width = 60)) +
        theme_bw() +
        # suppress showing alpha in legend
        scale_alpha(guide = "none")

      ggp + theme(legend.position = "bottom") 
    } else {
      ggp <- NULL
    }
    ggp
  })
})
```


### Biplots of PC1-3
```{r, fig.width=14, fig.height=10, warning=FALSE, message=FALSE, results='hide', fig.keep='all'}
library(ggbiplot)
library(grid)
library(gridExtra)

## Function to make biplot with larege margins and samples labeled individually
make_ggbiplot <- function(pca.object, pc_x, pc_y, groups, labels, ellipse = FALSE) {
  ggp <- ggbiplot(
        pca.object,
        choices = pc_x:pc_y,
        groups = groups,
        labels = labels,
        ellipse = ellipse,
        var.axes = FALSE
      ) + scale_color_discrete(name ='') +  theme_bw()
  return(ggp)
}

lapply(names(pca), function(assay_name) { 
  lapply(c(bio_vars, tech_vars_factor[[assay_name]]), function(color_var) {
    #message("PCA bi-plots - ", assay_name, ", ", color_var)
    # PCA scores 
    prcomp <- pca[[assay_name]]
    pca_scores <- as.data.frame(prcomp$x[, 1:num_PCs])
    # data for coloring
    if (color_var %in% bio_vars) {
      pca_scores[[color_var]] = colData(mae)[rownames(pca_scores), color_var]
    } else {
      pca_scores[[color_var]] = colData(experiments(mae)[[assay_name]])[
        rownames(pca_scores), color_var]
    }
    num_factor_levels <- length(levels(as.factor(pca_scores[[color_var]])))
    pca_scores[[color_var]][is.na(pca_scores[[color_var]])] <- "NA"
    if (num_factor_levels <= 3) { # multiple factor levels
      # pairs plot of PCs
      ellipse_truefalse = TRUE
    } else {
      ellipse_truefalse = FALSE}
    
    ggp <- make_ggbiplot(prcomp, 
                           1, 2, 
                           pca_scores[[color_var]],
                           rownames(pca_scores),
                           ellipse_truefalse)
    
    assay_name_file <- gsub(" ", "", assay_name, fixed = TRUE)
    assay_name_file <- gsub("|", "_", assay_name_file, fixed = TRUE)
    assay_name_file <- gsub(",", "_", assay_name_file, fixed = TRUE)
    
    ggsave(
      file.path(outdir, paste0("Score_plot_", assay_name_file, "_", color_var, ".png")),
      plot = ggp,
      width = 23, height = 17.46, units = "cm", dpi = 300)
    
    ggp2 <- make_ggbiplot(prcomp,
                           3, 4,
                           pca_scores[[color_var]],
                           rownames(pca_scores),
                           ellipse_truefalse)
    
    ggsave(
      file.path(outdir, paste0("Score_plot_PC34_", assay_name_file, "_", color_var, ".png")),
      plot = ggp2,
      width = 23, height = 17.46, units = "cm", dpi = 300)
    
    # ggp3 <- make_ggbiplot(prcomp, 
    #                        5, 6, 
    #                        pca_scores[[color_var]],
    #                        rownames(pca_scores),
    #                        ellipse_truefalse)
    # ggp4 <- make_ggbiplot(prcomp, 
    #                        7, 8, 
    #                        pca_scores[[color_var]],
    #                        rownames(pca_scores),
    #                        ellipse_truefalse)
    # 
    # grid.arrange(ggp, ggp2, ggp3, ggp4,
    #                             ncol = 2, nrow=2,
    #                             top = stringr::str_wrap(paste0("PCA score plot, color by ", color_var, ", ", assay_name))) 
  })
})

```

#### Box plots of PC scores colored by biological and technical variables

Outlier samples are labeled. Statistical tests are performed to determine if PCs are significantly associated with biological or technical variables

```{r pca_box_plots, results='asis', warning=F}
library(reshape2)
library(dplyr)
library(ggrepel)
library(ggpubr)
num_PCs <- 10

detect_outlier <- function(x) {
  {x < quantile(x, 0.25) - 1.5 * IQR(x)} | {
    x > quantile(x, 0.75) + 1.5 * IQR(x)}
}

box_plots <- list()
significant_associations_factor <- list()
i <- 1

for (assay_name in names(pca)) {
  for (color_var in c(bio_vars, tech_vars_factor[[assay_name]])) {
    #message("PCA scores box plots - ", assay_name, ", ", color_var)
    # PCA scores 
    num_PCs_ <- ifelse(num_PCs > ncol(pca[[assay_name]]$x),
                       ncol(pca[[assay_name]]$x), num_PCs)
    pca_scores <- as.data.frame(pca[[assay_name]]$x[, 1:num_PCs_])
    # data for coloring
    if (color_var %in% bio_vars) {
      pca_scores[[color_var]] = as.factor(
        colData(mae)[rownames(pca_scores), color_var])
    } else {
      pca_scores[[color_var]] = as.factor(
        colData(experiments(mae)[[assay_name]])[rownames(pca_scores), color_var])
    }
    if (length(levels(pca_scores[[color_var]])) > 1) { # multiple factor levels
      pca_scores$subject.id <- rownames(pca_scores)
      pca_scores_long <- pca_scores %>%
        melt(id.vars = c("subject.id", color_var)) %>%
        group_by(across(c(all_of(color_var), "variable"))) %>%
        mutate(outlier = ifelse(detect_outlier(value), 
                                gsub("CZC", "", subject.id), NA))
      # compare mean values of groups - significance tests
      stat_tests <- compare_means(
        as.formula(paste0('value~', color_var)),
        data = pca_scores_long, 
        group.by = "variable"
        #method = 't.test', #or Wilcoxon
        )
      stat_tests$assay_name <- assay_name
      stat_tests$tested_variable <- color_var
      df <- data.frame(stat_tests[stat_tests$p < 0.05, ])
      # if (nrow(df) > 0) {
      significant_associations_factor[[i]] <- local(df)
      # }
      #print(knitr::kable(stat_tests))
      cat("\n\n") # some extra lines for kable
      # box plot
      ggp <- ggplot(pca_scores_long, 
             aes(x = variable, y = value, fill = .data[[color_var]])) +
        geom_boxplot(position = position_dodge(width = 0.9)) +
        geom_text_repel( # label outliers
          aes(label = outlier), na.rm = TRUE, 
          position = position_dodge(width = 0.9)) + 
        #scale_fill_okabe_ito() + 
        #stat_compare_means() +
        ggtitle(stringr::str_wrap(paste0(
          "Box plot, PC scores, color by ", color_var, ", ", assay_name), 
          width = 60)) +
        theme_bw() +
        theme(
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          # panel.spacing.x = unit(0, "lines"),
          legend.position = "bottom") +
        facet_wrap(vars(variable), nrow = 1, scales = "free") 
      
      ## TODO
      assay_name_file <- gsub(" ", "", assay_name, fixed = TRUE)
      assay_name_file <- gsub("|", "_", assay_name_file, fixed = TRUE)
      assay_name_file <- gsub(",", "_", assay_name_file, fixed = TRUE)
      
      filename <- paste0(outdir, assay_name_file, "_", color_var, ".pdf")
      ggsave(filename, ggp, height = 5, width=20)
      box_plots[[i]] <- local(ggp)
      i <- i + 1
    }
  }
}
```


```{r, fig.width=20}
#lapply(box_plots, plot)
```

```{r}
significant_associations_factor <- do.call(rbind, significant_associations_factor)
#head(significant_associations_factor)
#dim(significant_associations_factor)
knitr::kable(significant_associations_factor[significant_associations_factor$p.adj<0.05,])
```


### Correlation of PCs with biological variables

```{r pca_corr_plots}
library(corrplot)

bio_vars_continuous <- c(
  "age.at.sampling.from.host",
  "age.at.sampling.from.host"
  #"days.in.vitro",
  #"biosafety.level"
  )

sig.level <- 0.05

significant_associations_contin <- lapply(names(pca), function(assay_name) {
  message("Correlations - ", assay_name)
  num_PCs_calculated <- dim(pca[[assay_name]]$x)[2]
  print(num_PCs_calculated)
  expl_var <- pca[[assay_name]]$sdev^2 / sum(pca[[assay_name]]$sdev^2)
  num_PCs_todisplay <- ifelse(num_PCs > num_PCs_calculated, 
                              num_PCs_calculated, num_PCs)
    #sum(expl_var>0.02)
  
    
  expl_var <- expl_var[1:num_PCs_todisplay]
  
  pca_scores <- as.data.frame(
    pca[[assay_name]]$x[, 1:num_PCs_todisplay])
  phenotype_continuous = as.data.frame(
    colData(mae)[rownames(pca_scores), bio_vars_continuous])
  phenotype_continuous <- cbind(
    phenotype_continuous,
    subset(data.frame(colData(mae[[assay_name]]))[rownames(pca_scores), ],
           select = tech_vars_continuous[[assay_name]]))
  # Remove non-variable phenotyp columns TODO
  
  
  
  # calculate correlation
  score_pheno_cor <- cor(
   pca_scores, phenotype_continuous, 
   method = "pearson", use = "pairwise.complete.obs")


  
  # significance test
  sig_test <- psych::corr.test(
    pca_scores, phenotype_continuous,
    use = "pairwise",
    method = "pearson",
    adjust = "BH",
    alpha = sig.level
  )
  
  significant_cor <- NULL
  
  if (sum(sig_test$p < 0.05) > 0) {
    significant_cor <- data.frame(sig_test$p)
    significant_cor$variable <- rownames(significant_cor)
    significant_cor <- tidyr::gather(
      significant_cor, "tested_variable", "p", -variable)
    significant_cor <- significant_cor[significant_cor$p < 0.05, ]
   
    significant_cor$p.adj <- apply(significant_cor, 1, function(row) {
      sig_test$p.adj[row[["variable"]], row[["tested_variable"]]]
    })
    significant_cor$assay_name <- assay_name
    significant_cor$method <- "Pearson"
    significant_cor$.y. <- significant_cor$group1 <- significant_cor$group2 <-
      significant_cor$p.format <- significant_cor$p.signif <- NA
  }
  
  rownames(score_pheno_cor) <- paste0(
    sprintf("%-4s", rownames(score_pheno_cor)), 
    " ",  
    sprintf("%3s", scales::percent(
      round(expl_var,2))))
  
  # correlation plot with FDR-adjusted p-values 
  cor_lim <- max(abs(c(min(score_pheno_cor), max(score_pheno_cor))))
  
  # png(file.path(outdir, 
  #               paste0("PCA_cor_", shorten_name(assay_name), ".png")),
  #   width = 28.04, height = 10.46, units = "cm", res = 300)
  # corrplot(
  #   score_pheno_cor, 
  #   is.corr = FALSE,
  #   #col.lim = c(-cor_lim, cor_lim),
  #   title = stringr::str_wrap(assay_name, width = 30),
  #   mar = c(0, 0, 3, 0),
  #   tl.col = "black",
  #   p.mat = sig_test$p.adj,
  #   sig.level = sig.level)
  # dev.off()
  corrplot(
    score_pheno_cor, 
    is.corr = FALSE,
    #col.lim = c(-cor_lim, cor_lim), #TODO
    title = stringr::str_wrap(assay_name, width = 30),
    mar = c(0, 0, 3, 0),
    tl.col = "black",
    p.mat = sig_test$p.adj,
    sig.level = sig.level)
  
  # make scatter plot of PCs and variables with significant correlation
  sig_pairs <- as.data.frame(which(sig_test$p.adj < sig.level, arr.ind = TRUE))
  sig_pairs$p.adj <- sig_test$p.adj[which(sig_test$p.adj < sig.level)]
  sig_pairs <- sig_pairs[order(sig_pairs$row), ] # sort by PCs
  if (dim(sig_pairs)[1] > 0) {
    apply(sig_pairs, 1, function(x) {
      pc_ <- rownames(sig_test$p.adj)[[x[["row"]]]]
      bio_ <- colnames(sig_test$p.adj)[[x[["col"]]]]
      df <- data.frame(pc_ = pca_scores[, pc_], 
                       bio_ = phenotype_continuous[, bio_])
      names(df) <- c(pc_, bio_)
      ggp <- ggplot(df, aes(x = .data[[pc_]], y = .data[[bio_]])) +
        geom_point() +
        stat_smooth(method = "lm") +
        stat_cor(method = "pearson") +
        ggtitle(stringr::str_wrap(assay_name, width = 60)) +
        theme_bw()
      plot(ggp)
      NULL
    })
  }
  significant_cor
})
```


```{r}
significant_associations_contin <- do.call(rbind, significant_associations_contin)
head(significant_associations_contin)
```



```{r, fig.width=10, fig.height=5}
significant_associations <- rbind(
  significant_associations_factor, 
  significant_associations_contin[, names(significant_associations_factor)])

unique(significant_associations$tested_variable)
significant_associations$tested_variable <- factor(
  significant_associations$tested_variable,
  levels =c(
  #"age.at.sampling.from.host",
  "phenotype",
  "tissues.from.which.cells.were.derived",
  "culturing.strategy",
  "observation.unit.identifier",
  "sex"))

unique(significant_associations$assay_name)
significant_associations$assay_name <- factor(
  significant_associations$assay_name,
  levels = experiment_names
)

significant_associations$expl_var <- apply(
  significant_associations, 1, function(row) {
    vars <- pca[[row[["assay_name"]]]]$sdev^2
    vars <- vars/sum(vars)
    vars[as.numeric(gsub("PC", "", row[["variable"]]))]
  })


library(dplyr)
data2plot <- significant_associations %>%
  filter(p.adj < 0.05) %>%
  distinct(assay_name, tested_variable, variable, .keep_all = T) %>%
  group_by(assay_name, tested_variable) %>%
  summarise(`Joint variation\nexplained by PCs` = sum(expl_var))


data2plot <- data2plot[!is.na(data2plot$tested_variable), ] # Remove rows with NA value for tested_variable.


# Create column for variable type, first fill with NA
data2plot$`Variable type` <- 'NA'
data2plot[data2plot$tested_variable %in% c(
   #"age.at.sampling.from.host",
   "phenotype",
   "observation.unit.identifier",
   "sex"), 
   ]$`Variable type` <- "Phenotype"

data2plot[data2plot$tested_variable %in% c(
   "tissues.from.which.cells.were.derived",
   "culturing.strategy"), 
   ]$`Variable type` <- "Technical"



ggp <- ggplot(data2plot) +
  geom_point(aes(x = factor(.data$tested_variable, levels = c(
  #age.at.sampling.from.host",
  "phenotype",
  "tissues.from.which.cells.were.derived",
  "culturing.strategy",
  "observation.unit.identifier",
  "sex")),
                 y = .data$assay_name,
                 size = .data$`Joint variation\nexplained by PCs`,
                 col = .data$`Variable type`)) +
  theme(
    axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1)
  ) +
  scale_color_okabe_ito() +
  ylab("Data set") +
  xlab("Biological/technical variable") +
  ggtitle("Significant associations between top 10\nPrincipal Components and variables")

ggsave(
  file.path(outdir, "PCA_summary.pdf"),
  plot = ggp,
  width = 20, height = 14.46, units = "cm", dpi = 300)

ggp
```



## Session information

```{r}
sessionInfo()
```


## References

