---
title: 'Pairwise correlations'
author:
  - Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  isa_file: C:/Users/Casper.deVisser/Documents/ipsc-data-analysis/20240822_Xomics_iPSC_Neurons_new_READ-ONLY.xlsx
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  config_file: 'config.yml'
  outdir: 'Z:/data/iPSC_data/Results/Pairwise correlations/'
  prot_tech_reps: Z:/data/iPSC_data/iPSC_MAE/proteomics_data_tech_reps_complete.csv
  metab_tech_reps: Z:/data/iPSC_data/iPSC_MAE/metabolomics_tml_data_tech_rep_complete.csv
---

## Params
The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
print(params)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!dir.exists(outdir)){
    outdir <- getwd()
}
```

## Libraries
```{r}
library(stats)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggpubr)
library(purrr)
```

## Read data and metadata: MultiAssayExperiment

```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
mae
```

The MultiAssayExperiment contains individual omics assays

```{r}
experiment_names <- names(experiments(mae))
experiment_names
```
## Read Assay files
```{r}
read_isa_file <- function(tab_name){
  file_name <- params$isa_file
  file <- readxl::read_xlsx(file_name, sheet = tab_name)
  isa_df <- file[rowSums(is.na(file)) != ncol(file),]
  isa_df<- isa_df[isa_df$`quality accessment` != 'FALSE', ] ## Sample QC filter
  isa_df <- isa_df[isa_df$`quality accessment` != 'Failed', ]
  isa_df
}

## Get Assay sheet names
sheet_names <- openxlsx::getSheetNames(params$isa_file)
assay_names <- sheet_names[startsWith(sheet_names, 'Assay')]
assay_names <- assay_names[assay_names != 'Assay - genomics']
## Read all Assay files
isa_assay_files <- lapply(assay_names, function(x) read_isa_file(x))
names(isa_assay_files) <- assay_names
```

## Map MAE assay names to ISA assay names
```{r}
experiments <- names(experiments(mae))
experiments

df_match_isa_mae <- data.frame(assay_names, experiment_names[c(13, 12, 2, 1, 3, 6, 8)]) #TODO, expaned mapping table
colnames(df_match_isa_mae) <- c('Assay_names', 'MAE_names')
df_match_isa_mae
```


## Assay selection from config file
```{r}
config <- yaml::read_yaml(params$config_file)
assay_names <- config$experiment_names$general_selection
df_match_isa_mae <- df_match_isa_mae[ df_match_isa_mae$MAE_names %in% assay_names,]
```


## Function to rename sample names of control lines in MAE Assays
```{r}
## Function to re-write the Control sample names, to be improved
rewrite_names <- function(x) {
  if (startsWith(x, "WTC")) {
    x <- 'WTC'
  }
  else if (startsWith(x, "X409")) {
    x <- '409B'
  }
  else if (startsWith(x, "Sendai")) {
    x <- 'Sendai'
  }
  else {
    x<- x
  }
  x
}
```

## Read ISA Sample tab
```{r}
isa_assay_Sample <- readxl::read_xlsx(params$isa_file, sheet = 4)
```

## Iterate over experiment names to create input data frames
```{r}
NumExp <- nrow(df_match_isa_mae)

isa_mae_list <- lapply(seq(1,NumExp), function(x) {
  
  ## Get names of Assay/Experiment
  isa_assay_name <- df_match_isa_mae[x, 1]
  mae_exper_name <- df_match_isa_mae[x, 2]
  
  isa_assay <- isa_assay_files[[isa_assay_name]]
  
  ## Add phenotype column of Sample tab to Assay tab
  phenotypes <- lapply(isa_assay$`sample identifier`, function(x) {
    phenotype <- isa_assay_Sample[isa_assay_Sample$`sample identifier` == x, 20]
    phenotype[[1]]
  })
  isa_assay$phenotype <- as.character(phenotypes)
  isa_assay

  se <- experiments(mae)[[mae_exper_name]]
  df_assay <- as.data.frame(t(assays(se)[[1]]))
  df_assay$sample_id <- rownames(df_assay)
  
  ## Add column with sample names as group
  new_names <- gsub("(^[^_]*_[^_]*)_.*", "\\1", rownames(df_assay))
  new_names <- lapply(new_names, function(x) rewrite_names(x))
  new_names <- gsub("_", "-", new_names) 
  df_assay$group <- new_names
  
  ## IMPORTANT, MATCH ORDER OF ISA ASSAY AND MAE ASSAY
  isa_assay <- isa_assay[match(df_assay$group, isa_assay$`sample identifier`),]
  
  list(isa_assay, df_assay)
})
names(isa_mae_list ) <- df_match_isa_mae$MAE_names
```


## Load function to calculate R-squared for samples in the same sample group
```{r}
source_paths <- c(
  "../support_functions/PairwiseCorrelationsFunctions.R",     # when running in Rstudio
  "PairwiseCorrelationsFunctions.R"                           # when running in Nextflow
)

for (path in source_paths) {
  if (file.exists(path)) {
    source(path)
    break
  }
}
```


## Calculate R-squared per assay, per group

```{r, fig.width=14, fig.height=6}
lapply(names(isa_mae_list), function(x) {
  
  isa_assay <- isa_mae_list[[x]][[1]]
  df_assay <- isa_mae_list[[x]][[2]]
  
  df_sample_identifier <- make_dataframe_rsquared_groups(df_assay, isa_assay, 'sample identifier')
  df_phenotype <- make_dataframe_rsquared_groups(df_assay, isa_assay, 'phenotype')
  
  ## IMPORTANT STEP! REMOVE PAIRS THAT WERE ALSO IN THE OTHER LEVEL
  df_phenotype <- df_phenotype[!df_phenotype$pair %in% df_sample_identifier$pair, ]
  
  plt1 <- plot_rsquared_groups_bio(df_sample_identifier, isa_assay, x, "bio replicate cultures from same cell line")
  plt2 <- plot_rsquared_groups_pheno(df_phenotype, x, "patients in the same disease group")
  ggarrange(plt1, plt2)

})
```


## Phenotype groups vs. Cell line groups
```{r}
lapply(names(isa_mae_list)[2:7], function(x) {
  ## Get pairwise correlations
  print(x)
  df_input <- get_pairwise_cor_df(x)
  ## Change label names
  levels(df_input$Level) <- c("Unrelated samples",  "Samples in same disease group", "Biological replicates")
  # Make boxplot
  #ylim_min <- floor(min(df_input$r_squared * 100)) / 100 - 0.02
  
  ggp <- ggplot(df_input, aes(x=Level, y=r_squared, fill = Level)) +
      geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.9)) +
      theme_minimal() +
      ggtitle(paste0(x)) +
      #theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
      theme(axis.text.x = element_blank()) +
      ylim(0.7, 1) +
      stat_compare_means()  +
      stat_summary(fun.data = function(x){
              return(data.frame(y=mean(x), label = paste0(length(x))))
          }, geom = "text") +
      theme(legend.position = 'none')
  name <- gsub(" ", "_", x)
  name <- gsub("_\\|_", "_", name)
  ggsave(file.path(outdir, paste0(name, '.pdf')), 
         ggp,
         width = 3, height = 4)

})
```

### EM-seq data on larger scale
```{r}
omics_name <- "EMseq | 100,000 most variable CpG sites"
## Get pairwise correlations
df_input <- get_pairwise_cor_df(omics_name)
## Change label names
levels(df_input$Level) <- c("Unrelated samples",  "Samples in same disease group", "Biological replicates")
# Make boxplot
ggp <- ggplot(df_input, aes(x=Level, y=r_squared, fill = Level)) +
    geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.9)) +
    theme_minimal() +
    ggtitle(paste0(omics_name)) +
    theme(axis.text.x = element_blank()) +
    ylim(0, 1) +
    stat_compare_means()  +
    stat_summary(fun.data = function(x){
              return(data.frame(y=mean(x), label = paste0(length(x))))
          }, geom = "text") 

name <- gsub(" ", "_", omics_name)
name <- gsub("_\\|_", "_", name)
ggsave(file.path(outdir, paste0(name, '.pdf')), 
         ggp,
         width = 3, height = 4)
```


## DM1 group only, with clones of patient 2
```{r}
clones <- c('IPS20_00035',
            'IPS23_00088', 
            'IPS23_00089'
            ) ## TODO: HARDCODED

lapply(names(isa_mae_list)[2:7], function(omics_name) {
  title <- strsplit(omics_name, split = '\\|')[[1]]
  ## Get pairwise correlations
  df_input <- get_pairwise_cor_df_clones(omics_name, clones, "Diseased: DM1")
  ## Change label names
  levels(df_input$Level) <- c("DM1 patients", "P2 clones", "Biological repl.")
  colors <- c("DM1 patients" = "green4", "P2 clones" = "gold3", "Biological repl." = "skyblue1")
  ymin <- min(df_input$r_squared)
  ymax <- max(df_input$r_squared) + 0.02
  # Make boxplot
  ggp <- ggplot(df_input, aes(x=Level, y=r_squared, fill = Level)) +
      geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.9)) +
      theme_minimal() +
      ggtitle(paste0(title)) +
      #theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
      theme(axis.text.x = element_blank()) +
      ylim(ymin, ymax) +
      scale_fill_manual(values = colors) +
      theme(legend.position = 'none') +
      stat_compare_means()  +
      stat_summary(fun.data = function(x){
              return(data.frame(y=mean(x), label = paste0(length(x))))
          }, geom = "text")
  ggsave(file.path(outdir, paste0('DM1_clones_', title, '.pdf')), plot = ggp, width = 3, height= 4)

})
```

### EM-seq data on larger scale
```{r}
omics_name <- "EMseq | 100,000 most variable CpG sites"
df_input <- get_pairwise_cor_df_clones(omics_name, clones, "Diseased: DM1")
## Change label names
levels(df_input$Level) <- c("DM1 patients", "P2 clones", "Biological rep.")
colors <- c("DM1 patients" = "green4", "P2 clones" = "gold3", "Biological rep." = "skyblue1")
# Make boxplot
ggp <- ggplot(df_input, aes(x=Level, y=r_squared, fill = Level)) +
    geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.9)) +
    theme_minimal() +
    ggtitle(paste0(omics_name, '\nPairwise correlations between phenotype/bio replicate')) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
    scale_fill_manual(values = colors) +
    ylim(0.25, 0.85) +
    stat_compare_means()  +
    stat_summary(fun.data = function(x){
            return(data.frame(y=mean(x), label = paste0(length(x))))
        }, geom = "text")

ggsave(file.path(outdir, 'DM1_clones_EMseq.pdf'), plot = ggp, width = 3, height= 4)
```



## Only clones of DM1 patient 2
```{r}
clones <- c('IPS20_00035',
            'IPS23_00088',
            'IPS23_00089'
            ) ## TODO: HARDCODED

lapply(names(isa_mae_list)[2:7], function(omics_name) {
  title <- strsplit(omics_name, split = '\\|')[[1]]
  ## Get pairwise correlations
  df_input <- get_pairwise_cor_df_clones(omics_name, clones, "Diseased: DM1")
  ## Select only clones
  df_input <- df_input[df_input$Level == 'DM1_P2_clones',]
  df_input$Clones <- NA
  df_input$Clones[grepl('IPS23_00089', df_input$pair) & grepl("IPS20_00035", df_input$pair)] <- 'C1 vs C2'
  df_input$Clones[grepl('IPS23_00088', df_input$pair) & grepl("IPS20_00035", df_input$pair)] <- 'C1 vs C3'
  df_input$Clones[grepl('IPS23_00088', df_input$pair) & grepl("IPS23_00089", df_input$pair)] <- 'C2 vs C3'
  ymin <- min(df_input$r_squared)
  ymax <- 1
  # Make boxplot
  ggp <- ggplot(df_input, aes(x=Clones, y=r_squared, fill = Clones)) +
      geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.9)) +
      theme_minimal() +
      ggtitle(paste0(title)) +
      theme(axis.text.x = element_blank()) +
      ylim(ymin, ymax) +
      theme(legend.position = 'none') +
      stat_compare_means()  +
      stat_summary(fun.data = function(x){
              return(data.frame(y=mean(x), label = paste0(length(x))))
          }, geom = "text")
  ggsave(file.path(outdir, paste0('DM1_clones_only_', title, '.pdf')), plot = ggp, width = 3, height= 4)

})
```


## CHD2 group only, with clones of patient 3
```{r}
clones <- c('IPS23_00025','IPS22_00063')

lapply(names(isa_mae_list)[2:7], function(x) {
  title <- strsplit(x, split = '\\|')[[1]]
  ## Get pairwise correlations
  df_input <- get_pairwise_cor_df_clones(x, clones, "Diseased: CHD2")
  ## Change label names
  levels(df_input$Level) <- c("CHD2 patients", "P3 clones", "Cell lines")
  colors <- c("CHD2 patients" = "green4", "P3 clones" = "gold3", "Cell lines" = "skyblue1")
  # Make boxplot
  ylim_min <- floor(min(df_input$r_squared * 100)) / 100 - 0.02
  
  ggp <-ggplot(df_input, aes(x=Level, y=r_squared, fill = Level)) +
      geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.9)) +
      theme_minimal() +
      ggtitle(paste0(x, '\nPairwise correlations between phenotype/bio replicate')) +
      scale_fill_manual(values = colors) +
      #theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
      theme(axis.text.x = element_blank()) +
      theme(legend.position = 'none') +
      ylim(ylim_min, 1) +
      stat_compare_means()  +
      stat_summary(fun.data = function(x){
              return(data.frame(y=mean(x), label = paste0(length(x))))
          }, geom = "text")
  ggsave(file.path(outdir, paste0('CHD2_clones_', title, '.pdf')), plot = ggp, width = 3, height= 4)

})
```


### EM-seq data on larger scale
```{r}
omics_name <- "EMseq | 100,000 most variable CpG sites"
df_input <- get_pairwise_cor_df_clones(omics_name, clones, "Diseased: CHD2")
## Change label names
levels(df_input$Level) <- c("CHD2 patients", "P3 clones", "Cell lines")
# Make boxplot
ggp <- ggplot(df_input, aes(x=Level, y=r_squared, fill = Level)) +
    geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.9)) +
    theme_minimal() +
    ggtitle(paste0(omics_name, '\nPairwise correlations between phenotype/bio replicate')) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
    ylim(0.0, 1) +
    #theme(legend.position = 'none') +
    stat_compare_means()  +
    stat_summary(fun.data = function(x){
            return(data.frame(y=mean(x), label = paste0(length(x))))
        }, geom = "text")

ggsave(file.path(outdir, paste0('CHD2_clones_EMseq.pdf')), plot = ggp, width = 3, height= 4)
```



# Add technical replicates

## Read technical replicates proteomics 
```{r}
proteomics_tech_reps <- read.csv(params$prot_tech_reps)
proteomics_tech_reps$group <- proteomics_tech_reps$Sample_ID
proteomics_tech_reps$Sample_ID <- NULL
```


## Read technical replicates metabolomics
```{r}
metabolomics_tech_reps <- read.csv(params$metab_tech_reps, row.names = 1)
```


## Proteomics with technical reps
```{r, fig.height=5}
df_technical <- pairwise_cor_tech(proteomics_tech_reps)

outfile <- file.path(outdir, 'Proteomics_tech_reps.pdf')
ggsave(filename = outfile,
       final_combined_plot(df_technical, 'Proteomics'),
       width = 7, height =5)
```


## Metabolomics with technical reps
```{r, fig.height=5}
df_technical <- pairwise_cor_tech(metabolomics_tech_reps)

outfile <- file.path(outdir, 'Metabolomics_tech_reps.pdf')
ggsave(filename = outfile,
        final_combined_plot(df_technical, "Metabolomics, TML" ),
        width = 7, height =5)
```


## Scatter plot of good examples
```{r}
scatter_plot_cor <- function(df, col='pink', x_axis, y_axis, title =  "Scatter plot") {
  
  ggp <- ggscatter(df, 
          x = colnames(df)[[1]], 
          y = colnames(df)[[2]],
          color = col,
          add = 'reg.line',
          add.params = list(color = "blue")) +
    labs(title = title,
         x = paste0(x_axis, " metabolite levels"), 
         y = paste0(y_axis, " metabolite levels"), ) +
    theme(axis.title.x =  element_text(size = 14),
          axis.title.y =  element_text(size = 14)) +
    stat_cor(aes(label = ..rr.label..), method = "pearson")
  
  return(ggp)
}
```


```{r}
## Select metabolomics data as example
name_assay <- "Metabolomics, TML"   
isa_assay <- isa_mae_list[[name_assay]][[1]]
df_assay <- isa_mae_list[[name_assay]][[2]]

## Samples with same phenotype
df_assay$group <- isa_assay[["phenotype"]]
scatter_plot_df <- df_assay
scatter_plot_df <- data.frame(t(scatter_plot_df))
scatter_plot_df <- scatter_plot_df[,c("IPS20_00124_1", 'IPS20_00036_1')]
scatter_plot_df[,1] <- as.numeric(scatter_plot_df[,1])
scatter_plot_df[,2] <- as.numeric(scatter_plot_df[,2])
scatter_plot_cor(scatter_plot_df, 'brown2', 'NANS sample', 'DM1 sample', 'Samples in same disease group')
```


```{r}
## Select metabolomics data as example
name_assay <- "Metabolomics, TML"   
isa_assay <- isa_mae_list[[name_assay]][[1]]
df_assay <- isa_mae_list[[name_assay]][[2]]

## Samples with same phenotype
df_assay$group <- isa_assay[["phenotype"]]
scatter_plot_df <- df_assay[df_assay$group == 'Diseased: DM1',]
scatter_plot_df <- data.frame(t(scatter_plot_df))
scatter_plot_df <- scatter_plot_df[,c("IPS20_00035_1", 'IPS20_00036_1')]
scatter_plot_df[,1] <- as.numeric(scatter_plot_df[,1])
scatter_plot_df[,2] <- as.numeric(scatter_plot_df[,2])
scatter_plot_cor(scatter_plot_df, 'darkgreen', 'DM1 patient 1', 'DM1 patient 2', 'Samples in same disease group')
```


```{r}
## Bio replicates
df_assay$group <- isa_assay[["phenotype"]]
scatter_plot_df <- df_assay[df_assay$group == 'Diseased: DM1',]
scatter_plot_df <- data.frame(t(scatter_plot_df))
scatter_plot_df <- scatter_plot_df[,c("IPS20_00035_1", 'IPS20_00035_3')]
scatter_plot_df[,1] <- as.numeric(scatter_plot_df[,1])
scatter_plot_df[,2] <- as.numeric(scatter_plot_df[,2])
scatter_plot_cor(scatter_plot_df, 'dodgerblue', 'DM1 patient 1, bio rep 1', 'DM1 patient 1, bio rep 2', 'Samples from same cell line')
```


```{r}
## Tech replicates
scatter_plot_df <- metabolomics_tech_reps[metabolomics_tech_reps$group == 'WTC_2',]
scatter_plot_df <- data.frame(t(scatter_plot_df))
scatter_plot_df[,1] <- as.numeric(scatter_plot_df[,1])
scatter_plot_df[,2] <- as.numeric(scatter_plot_df[,2])
scatter_plot_cor(scatter_plot_df, 'purple', 'Tech. rep 1', 'Tech. rep 2', 'Technical replicates')
```

## Session information

```{r}
sessionInfo()
```


