---
title: "RLM_plot_results"
output: html_document
date: "2024-11-15"
params:
  outdir: Z:/data/iPSC_data/results/RLM/
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
---

# Visualize results of linear modelling

## Parameters

```{r}
print(params)
outdir <- params$outdir
```

## Read data frames with RLM results

```{r}
mae_rlm_all_disease_merged <-  read.csv('RobustModels/rlm_all_disease_merged_2026_02_27.csv', row.names=1)
mae_rlm_all_merged_phenotypes <- read.csv('RobustModels/rlm_all_phenotype_merged_2026_02_27.csv', row.names=1)
```

## Add DNA Methylation data
```{r}
mae_rlm_all_disease_merged_Var <-  read.csv('RobustModels/OLD/EMseq_VariableSet_Disease.csv', row.names=1)
mae_rlm_all_merged_phenotypes_Var <- read.csv('RobustModels/OLD/EMseq_VariableSet_Pheno.csv', row.names=1)

# Merge with other data frames
mae_rlm_all_disease_merged <- rbind(mae_rlm_all_disease_merged, mae_rlm_all_disease_merged_Var)
mae_rlm_all_merged_phenotypes <- rbind(mae_rlm_all_merged_phenotypes, mae_rlm_all_merged_phenotypes_Var)
```

## OPTIONAL: Add random set of CpG sites
```{r}
# mae_rlm_all_disease_merged_RandomSet <-  read.csv('RobustModels/EMseq_RandomSet_Disease.csv', row.names=1)
# mae_rlm_all_merged_phenotypes_RandomSet <- read.csv('RobustModels/EMseq_RandomSet_Phenotypes.csv', row.names=1)
# 
# # Merge with other data frames
# mae_rlm_all_disease_merged <- rbind(mae_rlm_all_disease_merged, mae_rlm_all_disease_merged_RandomSet)
# mae_rlm_all_merged_phenotypes <- rbind(mae_rlm_all_merged_phenotypes, mae_rlm_all_merged_phenotypes_RandomSet)
```

```{r}
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(ggokabeito) # Okabe-Ito
```


### Combine the pairwise pheno comparisons with the disease and gender comparisons

```{r}
mae_rlm_all_merged <- rbind(mae_rlm_all_disease_merged, mae_rlm_all_merged_phenotypes)
```


## Assay names

```{r}
assay_names <- unique(mae_rlm_all_merged$assay)
```

Volcano plots per phenotypic variable

```{r volcano_plots, fig.width = 11, fig.height = 4.5}
outcomes <- unique(mae_rlm_all_merged$outcome)

lapply(outcomes, function(outcome) {
  rlm_res <- mae_rlm_all_merged[mae_rlm_all_merged$outcome == outcome, ]
  rlm_res$method <- stringr::str_wrap(paste(rlm_res$dataset, " | ", rlm_res$formula), width = 20)
  # my_levels <- unique(rlm_res$method)[order(sapply(unique(rlm_res$method), nchar))]
  my_levels <- unique(rlm_res$method)[order(unique(rlm_res$method))]
  rlm_res$method <- factor(rlm_res$method, levels = my_levels)
  rlm_res$assay <- factor(rlm_res$assay, levels = assay_names)
  
  ggp <- ggplot(rlm_res,
         aes(x = coefficient, y = -log10(p.adj))) +
    geom_point(aes(col = p.adj<=0.05, shape = p.adj<=0.05), 
               na.rm = T, size = 1) +
    scale_color_okabe_ito() + scale_fill_okabe_ito() +
    facet_grid(assay~method, scales = "free_y") +
    theme_bw() +
    theme(strip.text.y = element_text(angle = 0),
          text = element_text(size = 13),
          panel.spacing = unit(0.1, "lines")) +
    geom_text_repel(
      aes(label = label), size = 2.5, direction = "both",
      min.segment.length = unit(0.1, "lines"),
      max.time = 3, max.iter = 1000000) +
    xlab("Regression coefficient")
    # ggtitle(stringr::str_wrap(
    #   paste0("Robust linear models, ", outcome), width = 60))
  ggsave(file.path(outdir, paste0("RLMvolcano_", outcome, ".png")),
         ggp, width = 28.04, height = 10.46, units = "cm", dpi = 300)
  
  ggp
})
```

Compare number (or proportion) of significant hits per omics, outcome variable, and model.

```{r, fig.width=11, fig.height=9}
require(dplyr)
require(tidyr)

names(mae_rlm_all_merged)
mae_rlm_all_merged$significant <- mae_rlm_all_merged$p.adj <= 0.05
mae_rlm_all_merged$method <- as.factor(paste0(
  ifelse(mae_rlm_all_merged$covariates != "", 
         paste0("covariate: ", mae_rlm_all_merged$covariates), 
         "no covariate")))

# calculate proportion of significant features per omics, outcome, data set/model
df <- mae_rlm_all_merged %>%
  group_by(assay, outcome, method) %>%
  #count(significant, name = "n_signif") %>%
  summarise(
      n_signif = sum(significant == TRUE),
      total = n()
  ) %>%
  #group_by(assay, outcome, method) %>%
  mutate(freq = n_signif / total * 100) %>%
  drop_na()
df[df$method == 'NA', 'method'] <- 'no covariate'


df$assay <- factor(df$assay, levels = unique(assay_names))
#df$outcome <- factor(df$outcome, levels = outcomes)
df$`Dataset | model` <- factor(df$method, levels = unique(mae_rlm_all_merged$method))
df[df$method == 'NA', 'method'] <- 'no covariate'
df$`Dataset | model` <- 'no covariate'
df_noCov <- df[df$method == 'no covariate', ]
df_noCov$outcome <- gsub("phenotypeDiseased: ", "", df_noCov$outcome) 
# Outcome as factor and re-level
df_noCov$outcome <- as.factor(df_noCov$outcome)
df_noCov$outcome <- relevel(df_noCov$outcome, ref = 'Disease')
df_noCov$outcome <- relevel(df_noCov$outcome, ref = 'sex')

ggp <- ggplot(df_noCov) + 
  geom_col(aes_string(
    x = "`Dataset | model`", y = "freq", fill = "`Dataset | model`"),
    position = "dodge") + #position_dodge2(0.9, preserve = "total")) +
    geom_text(aes_string(x = "`Dataset | model`", y = 10, label="n_signif"),
            position = position_dodge(width =0.9),
            size=5) +
  facet_grid(rows = vars(assay), cols = vars(outcome)) +
  theme_bw() +
  theme(axis.text.x = element_blank(), #element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text.x = element_text(angle = 0),
        strip.text.y = element_text(angle = 0),
        axis.title.x = element_blank(),
        panel.spacing = unit(0, "lines"),
        legend.position = "bottom",
        text = element_text(size = 17)) +
  scale_fill_okabe_ito(order = c(1, 3, 7, 5)) +
  scale_y_continuous(
    name = stringr::str_wrap(
      "Proportion of significantly associated omics features",
      width = 30),
    limits = c(0, 41),
    breaks = c(0, 20, 40),
    labels = c("0", "20", "40"))

#dir.create(outdir)
ggsave(file.path(outdir, "RLMproportion_signif_hits_noCov.pdf"), ggp,
      width = 26.04, height = 15.46, units = "cm", dpi = 300)

ggp
```


## Session information

```{r}
sessionInfo()
```
