---
title: "RLM_plot_results"
output: html_document
date: "2024-11-15"
params:
  outdir: Z:/data/iPSC_data/results/MixedModels/Plots
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  MixedModelsResults: 'MixedModelsResults.csv'
  config_file: 'config.yml'
---

# Visualize results of linear modelling

## Parameters
```{r}
print(params)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!dir.exists(outdir)){
    outdir <- getwd()
}
```


## Read data and metadata: MultiAssayExperiment
```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
```


## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

## Assay selection
```{r}
assay_names <- config$experiment_names$general_selection
assay_names
```


## Phenotypic data
```{r}
pheno_df <- as.data.frame(colData(mae))
```


## Read nested mixed models
```{r}
MixedModel_complete <- read.csv(params$MixedModelsResults, row.names=1)

## Corrected data
#MixedModel_complete <- read.csv("MixedModelsResults_Corrected.csv", row.names=1)
```


```{r}
library(ggplot2)
library(ggrepel)
library(ggokabeito) # Okabe-Ito
library(ggpubr)
```


## Assay names

```{r, fig.width = 11, fig.height = 4.5}
outcomes <- unique(MixedModel_complete$outcome)
```


Compare number (or proportion) of significant hits per omics, outcome variable, and model.
```{r, fig.width=11, fig.height=7}
require(dplyr)
require(tidyr)

names(MixedModel_complete)
MixedModel_complete$significant <- MixedModel_complete$p.adj <= 0.05
MixedModel_complete$method <- 'Nested_Mixed'

# calculate proportion of significant features per omics, outcome, data set/model
df <- MixedModel_complete %>%
  group_by(Assay, outcome, method) %>%
  #count(significant, name = "n_signif") %>%
  summarise(
      n_signif = sum(significant == TRUE),
      total = n()
  ) %>%
  #group_by(assay, outcome, method) %>%
  mutate(freq = n_signif / total * 100) %>%
  drop_na()
df<- df[df$outcome != "phenotypeRescue: repeat excised CRISPR DM1",] ##TODO, could be removed
df<- df[df$outcome != "phenotypeDiseased: DM1",]

df$`Dataset | model` <-  factor(df$method, levels = unique(MixedModel_complete$method)) 

new_labels <- c(
  "DiseaseDisease" = "All_Disease",
  "phenotypeDiseased: CHD2" = "CHD2",
  "phenotypeDiseased: NANS-CDG" = "NANS-CDG",
  "phenotype_DM1Diseased: DM1" = "DM1",
  "DM1_vs_rescue" = "DM1",
  "sexmale"= "Sex"
)

new_omics_labels <- config$experiment_names$shorter_names

ggp <- ggplot(df) + 
  geom_col(aes_string(
    x = "`Dataset | model`", y = "freq", fill = "`Dataset | model`"),
    position = "dodge") + #position_dodge2(0.9, preserve = "total")) +
  geom_text(aes_string(x = "`Dataset | model`", y = 0.3, label="n_signif"),
            position = position_dodge(width =0.9),
            size=5) +
  # facet_grid(rows = vars(Assay), cols = vars(outcome),
  #            labeller = labeller(outcome = new_labels)) +
  facet_grid(rows = vars(Assay), cols = vars(outcome),
             labeller = labeller(outcome = new_labels, Assay = new_omics_labels)) +
  theme_bw() +
  theme(axis.text.x = element_blank(), #element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text.x = element_text(angle = 0),
        strip.text.y = element_text(angle = 0),
        axis.title.x = element_blank(),
        panel.spacing = unit(0, "lines"),
        legend.position = "bottom",
        text = element_text(size = 17)) +
  scale_fill_okabe_ito(order = c(1, 3, 7, 5)) +
  scale_y_continuous(
    name = stringr::str_wrap(
      "Percentage of significantly associated omics features",
      width = 30),
    breaks = c(0, 0.5, 1, 1.5),
    labels = c("0", "", "1.0", ""))


#dir.create(outdir)

ggsave(file.path(outdir, "MixedModelproportion_signif_hits.pdf"), ggp,
      width = 26.04, height = 15.46, units = "cm", dpi = 300)

ggp

```




## Functions for plotting features per disease group
```{r}
source_paths <- c(
  "../support_functions/Feature_plots.R",      # when running in Rstudio
  "support_functions/Feature_plots.R"                    # when running in Nextflow
)

for (path in source_paths) {
  if (file.exists(path)) {
    source(path)
    break
  }
}
```


# Metabolomics TML
```{r}
assay_name <- assay_names[grepl('TML', assay_names)]
make_boxplots_signif_feature(assay_name)
```

## Proteomics
```{r}
assay_name <- assay_names[grepl('Prot', assay_names)]
make_boxplots_signif_feature(assay_name)
```

# Metabolomics LUMC
```{r}
assay_name <- assay_names[grepl('LUMC', assay_names)]
make_boxplots_signif_feature(assay_name)
```

# Lipidomics, targeted
```{r}
assay_name <- assay_names[grepl('Untargeted', assay_names)]
make_boxplots_signif_feature(assay_name)
```

# Lipidomics, untargeted
```{r}
assay_name <- assay_names[grepl('Targeted', assay_names)]
make_boxplots_signif_feature(assay_name)m
```

# RNAseq
```{r}
assay_name <- assay_names[grepl('RNAseq', assay_names)]
make_boxplots_signif_feature(assay_name)
```

# EMseq
```{r}
assay_name <- assay_names[grepl('EMseq', assay_names)]
make_boxplots_signif_feature(assay_name, ylab = 'beta-value')
```


## Volcano plots function
```{r}
library(ggrepel)
results <- MixedModel_DM1

volcano_plot <- function(results, disease_group, assay_name) {

  results$logp <- -log10(results$p.adj)

  logFC_threshold <- 0.5
  pval_threshold <- 0.05

  results$Sign <- "Not Significant"
  results$Sign[results$coefficient > logFC_threshold & results$p.adj < 0.05] <- 'Upregulated'
  results$Sign[results$coefficient < -logFC_threshold & results$p.adj < 0.05] <- 'Downregulated'
  
  ## Label for only significant features
  results$Label <- ifelse(results$p.adj < pval_threshold & abs(results$coefficient) > logFC_threshold,
                                    results$Feature_plot,
                                    NA)
  
  ggp <- ggplot(results, aes(x = coefficient, y = logp, color = Sign, label = Label)) +
    geom_point(alpha =0.7, size =2) +
    geom_text_repel(size =3, box.padding = 0.3,max.overlaps = 50) +
    scale_color_manual(values = c("Not Significant"= "gray",
                                  "Upregulated"= "red",
                                  "Downregulated"= "blue")) + 
    geom_vline(xintercept = c(-logFC_threshold, logFC_threshold), linetype = "dashed") +
    geom_hline(yintercept = -log10(pval_threshold), linetype = "dashed") +
    labs(title = paste0(assay_name)) +
    theme_minimal()
  
  ## Save image
  assay_name <- strsplit(assay_name, "\\|")[[1]][1]
  assay_name <- gsub(" ", "", assay_name)
  full_path <- file.path(outdir, 'Volcano', disease_group)
  if (!dir.exists(full_path)) {
    dir.create(full_path, recursive = T)
  }
  file_path <- file.path(full_path, paste0(assay_name, '.png'))
  ggsave(file_path, ggp,
      width = 25.04, height = 17.46, units = "cm", dpi = 300)
  
  return(ggp)
}
```


## CHD2
```{r}
Results_CHD2 <- MixedModel_complete[MixedModel_phenotype$outcome == 'phenotypeDiseased: CHD2',]
lapply(assay_names, function(x) {
  df <- Results_CHD2[Results_CHD2$Assay == x,]
  volcano_plot(df, 'CHD2', x)
})
```


## NANS-CDG
```{r}
Results_NANS <- MixedModel_complete[MixedModel_phenotype$outcome == 'phenotypeDiseased: NANS-CDG',]

lapply(assay_names, function(x) {
  df <- Results_NANS[Results_NANS$Assay == x,]
  volcano_plot(df, 'NANS-CDG', x)
})

```

## DM1
```{r}
lapply(assay_names, function(x) {
  df <- MixedModel_DM1[MixedModel_DM1$Assay == x,]
  volcano_plot(df, 'DM1', x)
})
```

## All disease
```{r}
lapply(assay_names, function(x) {
  df <- MixedModel_complete_Disease[MixedModel_complete_Disease$Assay == x,]
  volcano_plot(df, 'All_Disease', x)
})
```

## Sex-associations
```{r}
lapply(assay_names, function(x) {
  df <- MixedModel_complete_Sex[MixedModel_complete_Sex$Assay == x,]
  volcano_plot(df, 'Sex', x)
})
```





## Session information
```{r}
sessionInfo()
```
