---
title: "Molecular Markers"
output: html_document
date: "2024-11-15"
params:
  outdir: Z:/data/iPSC_data/results/MolMarkers
  mae_hdf5_path: Z:/data/iPSC_data/iPSC_MAE/h5MAE
  neuron_markers: Markers_files/Neuron_markers.csv
  TMEM_mappings: Markers_files/TMEM_mappings.csv
  config_file: 'config.yml'
---

# Visualize results of linear modelling

## Parameters
```{r}
print(params)
outdir <- params$outdir

# If outdir not exists set to current dir
if (!dir.exists(outdir)){
    outdir <- getwd()
}
```


## Read data and metadata: MultiAssayExperiment
```{r}
require(SummarizedExperiment)
suppressWarnings(library(MultiAssayExperiment)) 


mae <- loadHDF5MultiAssayExperiment(
  dir = params$mae_hdf5_path, 
  prefix = 'mae')
```

## Config file
```{r}
config <- yaml::read_yaml(params$config_file)
```

## Assay selection
```{r}
assay_names <- config$experiment_names$general_selection
assay_names
```


## Phenotypic data
```{r}
pheno_df <- as.data.frame(colData(mae))
```


```{r}
library(patchwork)
library(pheatmap)
library(ggplot2)
library(ggrepel)
library(ggokabeito) # Okabe-Ito
library(ggpubr)
library(ggcorrplot)
```


## Functions for plotting features per disease group
```{r}
source_paths <- c(
  "../support_functions/Feature_plots.R",      # when running in Rstudio
  "Feature_plots.R"          # when running in Nextflow
)

for (path in source_paths) {
  if (file.exists(path)) {
    source(path)
    break
  }
}
```


## Neuronal markers

```{r}
assay_name <- assay_names[grepl('RNAseq', assay_names)]

rna_plot_df <- prepare_boxplot_df(assay_name)

boxplot_per_group(rna_plot_df, 'ENSG00000178403', assay_name, 'NEUROG2', 'Log-scaled intensity', phenotype = 'sample.description', color_group = 'phenotype')
boxplot_per_group(rna_plot_df, 'ENSG00000078018', assay_name, 'MAP2', 'Log-scaled intensity', phenotype = 'sample.description', color_group = 'phenotype')
boxplot_per_group(rna_plot_df, 'ENSG00000258947', assay_name, 'Beta III tubulin', 'Log-scaled intensity', phenotype = 'sample.description', color_group = 'phenotype')
boxplot_per_group(rna_plot_df, 'ENSG00000119042', assay_name, 'SATB2', 'Log-scaled intensity', phenotype = 'sample.description', color_group = 'phenotype')
boxplot_per_group(rna_plot_df, 'ENSG00000136535', assay_name, 'TBR1', 'Log-scaled intensity', phenotype = 'sample.description', color_group = 'phenotype')
```

## Protein levels
```{r}
assay_name <- assay_names[grepl('Prote', assay_names)]

protein_plot_df <- prepare_boxplot_df(assay_name)

boxplot_per_group(protein_plot_df, 'Q9H2A3', assay_name, 'NEUROG2', 'Log-scaled intensity', phenotype = 'sample.description', color_group = 'phenotype')
boxplot_per_group(protein_plot_df, 'P11137', assay_name, 'MAP2', 'Log-scaled intensity', phenotype = 'sample.description', color_group = 'phenotype')
boxplot_per_group(protein_plot_df, 'Q13509', assay_name, 'Beta III tubulin', 'Log-scaled intensity', phenotype = 'sample.description', color_group = 'phenotype')

```

## Mature neuronal markers
```{r}
markers_df <- read.csv(params$neuron_markers)
```


## Overview figure of molecular markers
```{r}
assay_name <- assay_names[grepl('Prote', assay_names)]
df_input <- prepare_boxplot_df(assay_name)

df_mature <- get_long_table_markers(df_input, omics_id = 'Uniprot', stage = 'Mature')
df_early <- get_long_table_markers(df_input, omics_id = 'Uniprot', stage = 'Early')

ggp_mature <- boxplot_groups(df_mature, title='Mature markers')
ggp_early <- boxplot_groups(df_early, title='Early markers')

ggp_combined <- ggp_early / ggp_mature +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom", 
        legend.justification = "center",
        legend.direction = "horizontal",
        legend.box = 'horizontal')
    
#ggp_combined

#ggsave(file.path(outdir, 'Markers_prot.pdf'), plot = ggp_combined, width = 15, height = 10)
```


### Heatmap proteomics markers
```{r}
markers_df_prt <- markers_df[!is.na(markers_df$Uniprot),]

df_input_heatmap <- df_input[, unique(markers_df_prt$Uniprot)]

ggp <- make_heatmap(df_input_heatmap, 
                    pheno_df$sample.description,
                    unique(markers_df_prt$Gene_symbol),
                    markers_df_prt)

ggsave(file.path(outdir, 'Heatmap_Neuron_Prot.pdf'), ggp, width = 12, height = 14)
```

## Combine all neuronal markers, and correlate with NEUROG2
```{r}
markers_df_prot_neuron <- markers_df_prt[markers_df_prt$Stage != 'Ipsc',]

df_prot <- df_input[, unique(markers_df_prot_neuron$Uniprot)]
cor_matrix <- cor(df_prot)
colnames(cor_matrix) <- unique(markers_df_prot_neuron$Gene_symbol)
rownames(cor_matrix) <- unique(markers_df_prot_neuron$Gene_symbol)
corrplot::corrplot(cor_matrix, method = 'color', type = 'upper', tl.col='black', tl.srt=45, order = 'hclust')
```



```{r}
# RNA data
assay_name <- assay_names[grepl('RNAseq', assay_names)]
df_input <- prepare_boxplot_df(assay_name)

df_mature <- get_long_table_markers(df_input, omics_id = 'Ensembl', stage = 'Mature')
df_early <- get_long_table_markers(df_input, omics_id = 'Ensembl', stage = 'Early')
df_NPC <- get_long_table_markers(df_input, omics_id = 'Ensembl', stage = 'Ipsc')

ggp_mature <- boxplot_groups(df_mature, title='Mature markers')
ggp_early <- boxplot_groups(df_early, title='Early markers')
ggp_ipsc <- boxplot_groups(df_NPC, title= 'iPSC markers')

ggp_combined <- ggp_ipsc / ggp_early / ggp_mature +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom", 
        legend.justification = "center",
        legend.direction = "horizontal",
        legend.box = 'horizontal')
    

#ggp_combined
#ggsave(file.path(outdir, 'Markers_rna.pdf'), plot = ggp_combined, width = 15, height = 10)
```

#### Heatmap neuronal markers genes
```{r}
markers_df_rna <- markers_df[!is.na(markers_df$Ensembl),]

df_input_heatmap <- df_input[, unique(markers_df_rna$Ensembl)]

ggp <- make_heatmap(df_input_heatmap, 
                    pheno_df$sample.description,
                    unique(markers_df_rna$Gene_symbol),
                    markers_df_rna)

ggsave(file.path(outdir, 'Heatmap_Neuron_RNA.pdf'), ggp, width = 14, height = 14)
```

## Combine all neuronal markers, and correlate with NEUROG2
```{r}
markers_df_rna_neuron <- markers_df_rna[markers_df_rna$Stage != 'Ipsc',]

df_rna <- df_input[, unique(markers_df_rna_neuron$Ensembl)]
cor_matrix <- cor(df_rna)
colnames(cor_matrix) <- unique(markers_df_rna_neuron$Gene_symbol)
rownames(cor_matrix) <- unique(markers_df_rna_neuron$Gene_symbol)
corrplot::corrplot(cor_matrix, method = 'color', type = 'upper', tl.col='black', tl.srt=45, order = 'hclust')
```


## Correlate NEUROG2 RNA and protein levels
```{r}
## Select RNA markers that are also present in proteomics data
markers_df_rna_neuron_overlap <- markers_df_rna[markers_df_rna$Uniprot %in% markers_df_prot_neuron$Uniprot,]
df_rna_overlap_prot <- df_rna[, unique(markers_df_rna_neuron_overlap$Ensembl)]

## Use same set of set samples for proteomics and RNAseq
df_prot_overlap <- df_prot[rownames(df_rna_overlap_prot),]

## Use same gene names 
colnames(df_rna_overlap_prot) <- markers_df_rna_neuron_overlap$Gene_symbol
colnames(df_prot_overlap) <- markers_df_prot_neuron$Gene_symbol


## Make pairwise correlations per marker across omics data sets
omics_correlations <- sapply(names(df_rna_overlap_prot), function(col) cor(df_rna_overlap_prot[[col]], df_prot_overlap[[col]], use = 'complete.obs'))

markers_correlations <- data.frame(Neuron_marker = names(omics_correlations), Pearson_cor = omics_correlations)
markers_correlations[order(markers_correlations$Pearson_cor),]
```


## Disease markers

### NANS-CDG
```{r}
assay_name <- assay_names[grepl('TML', assay_names)]
metabolomics_plot_df <- prepare_boxplot_df(assay_name)

boxplot_per_group(metabolomics_plot_df, 'TML_neg3025', assay_name, 'TML_neg3025_ManNAc-6P', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')
boxplot_per_group(metabolomics_plot_df, 'TML_neg3622', assay_name, 'TML_neg3622_ManNAc-6P', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')
boxplot_per_group(metabolomics_plot_df, 'TML_neg3009', assay_name, 'TML_neg3009_ManNAc-6P', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')
boxplot_per_group(metabolomics_plot_df, 'TML_pos4085', assay_name, 'TML_pos4085_ManNAc-6P', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')
boxplot_per_group(metabolomics_plot_df, 'TML_pos4106', assay_name, 'TML_pos4106_ManNAc-6P', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')

boxplot_per_group(metabolomics_plot_df, 'TML_pos2344', assay_name, 'TML_pos2344_ManNAc', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')

boxplot_per_group(metabolomics_plot_df, 'TML_pos339', assay_name, 'Creatinine', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')
boxplot_per_group(metabolomics_plot_df, 'TML_pos666', assay_name, 'Creatinine', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')

```

## ManNAc6P
```{r}
ggp <- boxplot_per_group(metabolomics_plot_df, 'TML_neg3025', assay_name, 'TML_neg3025_ManNAc-6P', 'Log-scaled intensity', phenotype = 'phenotype', color_group = 'phenotype')
ggp
ggsave(file.path(outdir, 'ManNAC6p.pdf'), ggp)
```


### DM1
#### RNA levels
```{r}
assay_name <- assay_names[grepl('RNAseq', assay_names)]

rna_plot_df <- prepare_boxplot_df(assay_name)

boxplot_per_group(rna_plot_df, 'ENSG00000149187', assay_name, 'CELF1', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr')
boxplot_per_group(rna_plot_df, 'ENSG00000048740', assay_name, 'CELF2', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr')
boxplot_per_group(rna_plot_df, 'ENSG00000159409', assay_name, 'CELF3', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr')
boxplot_per_group(rna_plot_df, 'ENSG00000161082', assay_name, 'CELF5', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr')
```

#### Protein levels
```{r}
assay_name <- assay_names[grepl('Prot', assay_names)]

protein_plot_df <- prepare_boxplot_df(assay_name)

boxplot_per_group(protein_plot_df, 'Q92879', assay_name, 'CELF1', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr')
boxplot_per_group(protein_plot_df, 'O95319', assay_name, 'CELF2', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr')
boxplot_per_group(protein_plot_df, 'Q5SZQ8', assay_name, 'CELF3', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr')

```


### CHD2
```{r}
# RNA
assay_name <- assay_names[grepl('RNAseq', assay_names)]
rna_plot_df <- prepare_boxplot_df(assay_name)
boxplot_per_group(rna_plot_df, 'ENSG00000185896', assay_name, 'LAMP1', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr')

# PROTEIN
assay_name <- assay_names[grepl('Prot', assay_names)]
protein_plot_df <- prepare_boxplot_df(assay_name)
boxplot_per_group(protein_plot_df, 'P11279', assay_name, 'LAMP1', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr')

# CHD1 protein levels
boxplot_per_group(protein_plot_df, 'O14646', assay_name, 'CHD1', 'Log2-scaled intensity', phenotype = 'observation.unit.identifier.crispr')
```

## TMEM132 
```{r}
TMEM_mappings <- read.csv(params$TMEM_mappings) 
```

```{r}
# RNA
assay_name <- assay_names[grepl('RNAseq', assay_names)]
rna_plot_df <- prepare_boxplot_df(assay_name)
#rna_plot_df_DM_2 <- rna_plot_df[rna_plot_df$observation.unit.identifier.crispr == 'DM_2',]
boxplot_per_group(rna_plot_df, 'ENSG00000139364', assay_name, 'TMEM132B', 'Log-scaled intensity', phenotype = 'sample.description', width = 20, height = 12)
boxplot_per_group(rna_plot_df, 'ENSG00000181234', assay_name, 'TMEM132C', 'Log-scaled intensity', phenotype = 'sample.description', width = 20, height = 12)
boxplot_per_group(rna_plot_df, 'ENSG00000151952', assay_name, 'TMEM132D', 'Log-scaled intensity', phenotype = 'sample.description', width = 20, height = 12)
boxplot_per_group(rna_plot_df, 'ENSG00000181291', assay_name, 'TMEM132E', 'Log-scaled intensity', phenotype = 'sample.description', width = 20, height = 12)
```

## DPP6
```{r}
boxplot_per_group(rna_plot_df, 'ENSG00000130226', assay_name, 'DPP6', 'Log-scaled intensity', phenotype = 'sample.description')
```

## RIMBP2
```{r}
boxplot_per_group(rna_plot_df, 'ENSG00000060709', assay_name, 'RIMBP2', 'Log-scaled intensity', phenotype = 'sample.description')
```

## GLT1D1
```{r}
boxplot_per_group(rna_plot_df, 'ENSG00000151948', assay_name, 'GLT1D1', 'Log-scaled intensity', phenotype = 'sample.description')
```

## RNF32-DT
```{r}
boxplot_per_group(rna_plot_df, 'ENSG00000182648', assay_name, 'RNF32-DT', 'Log-scaled intensity', phenotype = 'sample.description')
```

## NKX2-5
```{r}
boxplot_per_group(rna_plot_df, 'ENSG00000183072', assay_name, 'NKX2-5', 'Log-scaled intensity', phenotype = 'sample.description')
```

## FZD10
```{r}
boxplot_per_group(rna_plot_df, 'ENSG00000111432', assay_name, 'FZD10', 'Log-scaled intensity', phenotype = 'sample.description')
```




## Lipidomics Classes
```{r}
library(viridis)

assay_name <- assay_names[grepl('Targeted', assay_names)]
se <- experiments(mae)[assay_name]
df <- as.data.frame(t(assays(se)[[1]]))



feature_metadata <- data.frame(colnames(df))
feature_metadata$Class <- sub("\\..*", "",  names(df))
#feature_metadata$colnames.df. <- NULL
rownames(feature_metadata) <- colnames(df)

## Remove TG's
feature_metadata <- data.frame(feature_metadata[feature_metadata$Class != "TG",])

df <- df[, rownames(feature_metadata)]
feature_metadata$colnames.df. <- NULL

anno_col <- pheno_df[rownames(df),]
anno_col <- anno_col[, "sample.description"]
anno_col <- data.frame(anno_col)
rownames(anno_col) <- rownames(df)

df_heatmap <- data.frame(t(df))

ann_colors <- list(
  anno_col = setNames(viridis(length(unique(anno_col$anno_col))),
                      levels(anno_col$anno_col)),
  Class = setNames(viridis(length(unique(feature_metadata$Class))),
                      levels(feature_metadata$Class))
)

## Function to make heatmap
make_heatmap <- function(df){
  ggp <- pheatmap(
    df,
    cluster_rows = T,
    cluster_cols = T,
    show_colnames = T,
    show_rownames = F,
    annotation_row = feature_metadata,
    annotation_col = anno_col,
    scale = "column",
    #annotation_colors = ann_colors
  
  )
  return(ggp)
}

ggp <- make_heatmap(df_heatmap)
ggsave(filename = file.path(outdir, "Lipidomics_Heatmap_noTG.pdf"), ggp, width = 14, height = 12)
```


## Metabolite hits on MOFA factor 1

### Glutamine
```{r}
assay_name <- assay_names[grepl('LUMC', assay_names)]
metabolomics_plot_df <- prepare_boxplot_df(assay_name)

boxplot_per_group(metabolomics_plot_df, 'pos_1799', assay_name, 'pos_1799', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')
boxplot_per_group(metabolomics_plot_df, 'pos_1799', assay_name, 'pos_1799', 'Log-scaled intensity', phenotype = 'sex')
boxplot_per_group(metabolomics_plot_df, 'pos_1799', assay_name, 'pos_1799', 'Log-scaled intensity', phenotype = 'culturing.strategy')
boxplot_per_group(metabolomics_plot_df, 'pos_1799', assay_name, 'pos_1799', 'Log-scaled intensity', phenotype = 'cell.line.background')
boxplot_per_group(metabolomics_plot_df, 'pos_1799', assay_name, 'pos_1799', 'Log-scaled intensity', phenotype = 'reprogramming.method')
boxplot_per_group(metabolomics_plot_df, 'pos_1799', assay_name, 'pos_1799', 'Log-scaled intensity', phenotype = 'tissues.from.which.cells.were.derived')
boxplot_per_group(metabolomics_plot_df, 'pos_1799', assay_name, 'pos_1799', 'Log-scaled intensity', color_group =  'reprogramming.method')
boxplot_per_group(metabolomics_plot_df, 'pos_1799', assay_name, 'pos_1799', 'Log-scaled intensity', color_group =  'tissues.from.which.cells.were.derived')
```

### Doxycycline
```{r}
assay_name <- assay_names[grepl('LUMC', assay_names)]
metabolomics_plot_df <- prepare_boxplot_df(assay_name)


boxplot_per_group(metabolomics_plot_df, 'pos_2595', assay_name, 'Doxycline', 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')
boxplot_per_group(metabolomics_plot_df, 'pos_2595', assay_name, 'Doxycline', 'Log-scaled intensity', phenotype = 'sex')
boxplot_per_group(metabolomics_plot_df, 'pos_2595', assay_name, 'Doxycline', 'Log-scaled intensity', phenotype = 'culturing.strategy')
boxplot_per_group(metabolomics_plot_df, 'pos_2595', assay_name, 'Doxycline', 'Log-scaled intensity', phenotype = 'cell.line.background')
boxplot_per_group(metabolomics_plot_df, 'pos_2595', assay_name, 'Doxycline', 'Log-scaled intensity', phenotype = 'reprogramming.method')
boxplot_per_group(metabolomics_plot_df, 'pos_2595', assay_name, 'Doxycline', 'Log-scaled intensity', phenotype = 'tissues.from.which.cells.were.derived')
boxplot_per_group(metabolomics_plot_df, 'pos_2595', assay_name, 'Doxycline', 'Log-scaled intensity', color_group =  'reprogramming.method')
boxplot_per_group(metabolomics_plot_df, 'pos_2595', assay_name, 'Doxycline', 'Log-scaled intensity', color_group =  'tissues.from.which.cells.were.derived')
```


```{r}
cor_plot(metabolomics_plot_df$pos_1799, 
         as.numeric(metabolomics_plot_df$PassageNumber),
         'Metabolite levels', 'Passage numbers')
```

```{r}
cor_plot(metabolomics_plot_df$pos_2595, 
         as.numeric(metabolomics_plot_df$PassageNumber), 
         'Metabolite levels', 'Passage numbers')
```


## Correlate Glutamine levels with transporter protein levels
```{r}
#Q15758 SLC1A5	Neutral amino acid transporter B(0)

proteomics_plot_df_mtblmcs <- protein_plot_df[rownames(metabolomics_plot_df),]

ggp_pos1799 <- cor_plot(metabolomics_plot_df$pos_1799, 
         proteomics_plot_df_mtblmcs$Q15758,
         'Glutamine metabolite (pos_1799) levels', 'SLC1A5 protein levels', title = ' ', method = 'spearman')


ggp_pos1901 <- cor_plot(metabolomics_plot_df$pos_1901, 
         proteomics_plot_df_mtblmcs$Q15758,
         'Glutamine metabolite (pos_1901) levels', 'SLC1A5 protein levels', title = '', method = 'spearman')


ggp_pos1708 <- cor_plot(metabolomics_plot_df$pos_1708, 
         proteomics_plot_df_mtblmcs$Q15758,
         'Glutamine metabolite (pos_1708) levels', 'SLC1A5 protein levels', title = 'Glutamine and SLCA1A5 correlation', method = 'spearman')


ggsave(file.path(outdir, 'Gln_SLC1A4_Cor.pdf'), ggarrange(ggp_pos1708, ggp_pos1799, ggp_pos1901), width = 10)
```

```{r}
#Q96QD8 SLC38A2	Sodium-coupled neutral amino acid symporter 2

ggp_pos1799 <- cor_plot(metabolomics_plot_df$pos_1799, 
         proteomics_plot_df_mtblmcs$Q96QD8,
         'Glutamine metabolite (pos_1799) levels', 'SLC38A2 protein levels', 
         title = ' ', 
         method = 'spearman')


ggp_pos1901 <- cor_plot(metabolomics_plot_df$pos_1901, 
         proteomics_plot_df_mtblmcs$Q96QD8,
         'Glutamine metabolite (pos_1901) levels', 'SLC38A2 protein levels', 
         title = ' ',
         method = 'spearman')


ggp_pos1708 <- cor_plot(metabolomics_plot_df$pos_1708, 
         proteomics_plot_df_mtblmcs$Q96QD8,
         'Glutamine metabolite (pos_1708) levels', 'SLC38A2 protein levels', 
         title = 'Glutamine and SLC38A2 correlation',
         method = 'spearman')


ggsave(file.path(outdir, 'Gln_SLC38A2_Cor.pdf'), ggarrange(ggp_pos1708, ggp_pos1799, ggp_pos1901), width = 10)
```



## Kinetin-9-riboside (putative)
```{r}
assay_name <- assay_names[grepl('LUMC', assay_names)]
metabolomics_plot_df <- prepare_boxplot_df(assay_name)

putative_annotation_id <- 'neg_5241'
putative_annotation <- 'Kinetin-9-riboside'

boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'sex')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'culturing.strategy')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'cell.line.background')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'reprogramming.method')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'tissues.from.which.cells.were.derived')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', color_group =  'reprogramming.method')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', color_group =  'tissues.from.which.cells.were.derived')
```  


## N6-Isopentenlyadenosine (putative)
```{r}
assay_name <- assay_names[grepl('LUMC', assay_names)]
metabolomics_plot_df <- prepare_boxplot_df(assay_name)

putative_annotation_id <- 'neg_1684'
putative_annotation <- 'N6-Isopentenlyadenosine'

boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'sex')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'culturing.strategy')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'cell.line.background')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'reprogramming.method')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'tissues.from.which.cells.were.derived')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', color_group =  'reprogramming.method')
boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', color_group =  'tissues.from.which.cells.were.derived')
```  
```{r}
cor_plot(metabolomics_plot_df$neg_5241, 
         as.numeric(metabolomics_plot_df$PassageNumber), 
         'Metabolite levels', 'Passage numbers')
```

```{r}
cor_plot(metabolomics_plot_df$neg_1684, 
         as.numeric(metabolomics_plot_df$PassageNumber),
         'Metabolite levels', 'Passage numbers')
```

## Uridine-derivatives
```{r}
assay_name <- assay_names[grepl('LUMC', assay_names)]
metabolomics_plot_df <- prepare_boxplot_df(assay_name)

putative_annotation_id <- 'neg_71'
putative_annotation <- 'Uridine monophosphate'

boxplot_per_group(metabolomics_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')
```


```{r}
assay_name <- assay_names[grepl('TML', assay_names)]
metabolomics_TML_plot_df <- prepare_boxplot_df(assay_name)

putative_annotation_id <- 'TML_neg6366'
putative_annotation <- 'Uridine diphosphate glucose/galactose'

boxplot_per_group(metabolomics_TML_plot_df, putative_annotation_id, assay_name, putative_annotation, 'Log-scaled intensity', phenotype = 'observation.unit.identifier.crispr', color_group = 'phenotype')
```

```{r}
metabolomics_TML_plot_df <- metabolomics_TML_plot_df[rownames(metabolomics_plot_df),]

cor_plot(metabolomics_plot_df$neg_71, 
         metabolomics_TML_plot_df$TML_neg6366, 
         'RUMC', 'LUMC', 
         title = 'Correlation between RUMC and LUMC measurements')
```



## Plot RNA levels of genes with large deletion in DM1 patient 2 clone 2
```{r}
# RNA
assay_name <- assay_names[grepl('RNAseq', assay_names)]
rna_plot_df <- prepare_boxplot_df(assay_name)

## CLEC4A is not present in the data

## FOXJ2
boxplot_per_group(rna_plot_df, 'ENSG00000065970', assay_name, 'FOXJ2', 'Log-scaled intensity', phenotype = 'sample.description')

## NECAP1
boxplot_per_group(rna_plot_df, 'ENSG00000089818', assay_name, 'NECAP1', 'Log-scaled intensity', phenotype = 'sample.description')

## C3AR1
boxplot_per_group(rna_plot_df, 'ENSG00000171860', assay_name, 'C3AR1', 'Log-scaled intensity', phenotype = 'sample.description')
```



## Session information
```{r}
sessionInfo()
```
